---
title: "Differential expression meta analysis"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  output_dir: "differential_expression_meta"
  metadata: "combined_rna_metadata.csv"
  analyze_specific_trial: ""
  run_meta_analysis: FALSE
  generate_heatmaps: TRUE
---

Author comment: 

File description comment:
The purpose of this script is to aggregate differential expression analysis results and integrate them to derive a new response signature score
The NKI pancancer signature score was defined as the difference between ssGSEA scores for the positive and negative component

# Loading Packages

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE, dev = "png"
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# Install if necessary install.packages("BiocManager")
# Install if necessary BiocManager::install("fgsea")
library(tidyverse)
library(ggrepel)
library(here)
library(DESeq2)
library(ggpubr)
library(rstatix)
library(corrplot)
source("helper_functions.R")
library(meta)
library(enrichR)
library(VennDiagram)
library(ggVennDiagram)
library(ggpointdensity)
library(RobustRankAggreg)

set.seed(42)
select <- dplyr::select
rename <- dplyr::rename
here <- here::here
```

# ----------

# Setup output folder

```{r Setup, echo = FALSE, include = FALSE}
export <- params$export
analyze_specific_trial <- params$analyze_specific_trial
run_meta_analysis <- params$run_meta_analysis   
generate_heatmaps <- params$generate_heatmaps

# generate output directory
output_dir <- here("data", "processed", params$output_dir)
print(output_dir)
output_dir_plots <- here(output_dir, "plots")

data_dir <- here("data", "processed")
combined_dir <- here(data_dir, "gene_signatures")
meta_rna_dir <- here("data", "processed", "gene_signatures_meta") 

# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)
    dir.create(output_dir_plots)
} else {
  print("Output folder already exists")
}

```

# ----------

# Load RNAseq data

```{r load rnaseq data, include = TRUE}
df_counts <- read_csv2(here("data", "processed", "rnaseq_data",
                             "combined_rnaseq_counts_p.csv")) %>%
  column_to_rownames("gene")

rna_metadata <- read_csv2(
  here("data", "processed", "metadata", params$metadata))

if (analyze_specific_trial != ""){
  rna_metadata <- rna_metadata %>% filter(Trial == analyze_specific_trial)
}

rna_metadata <- rna_metadata %>%
  mutate(
      Treatment = factor(Treatment, levels = c("Pre", "Post")),
      CF_sample = as.factor(CF_sample),
      Trial = as.factor(Trial),
      Response = as.factor(Response),
      Response_detail = as.factor(Response_detail),
      MSI_MSS = as.factor(MSI_MSS),
      BRAF = as.factor(BRAF))

rna_metadata %>% mutate_if(is.character, as.factor) %>% summary()

# Remove genes with all zero counts, might be detected in specific trials only
df_counts <- df_counts[, colnames(df_counts) %in%
                          rna_metadata$CF_sample]
df_counts <- df_counts[rowSums(df_counts) != 0, ]

unique_patient_id <- rna_metadata %>% select(Trial, Patient) %>% unique()
table(unique_patient_id$Trial)
 
paired_samples <- table(rna_metadata$Patient, rna_metadata$Treatment) %>%
  as.data.frame() %>%
  setNames(c("Patient", "Treatment", "Freq")) %>%
  pivot_wider(values_from = Freq, names_from = Treatment) %>%
  filter(Pre >  0 & Post > 0) %>%
  left_join(unique_patient_id)

histology_df <- data.frame(
  Trial = c("BELLINI", "IMCISION", "NABUCCO", "NICHE-MSS", "NICHE-MSI",
            "OpACIN_neo", "PRADO", "MATISSE"),
  Histology = c("TBNC", "HNSCC", "Urothelial", "Colon", "Colon",
                "LN_melanoma", "LN_melanoma", "CSCC"))

```

# Annotation color

```{r}

# Define colours for each group
Color_NR <- "#F8766D"
Color_R <- "#00BA38" # "#619CFF"
Color_NA <- "#CD9600"
Color_Pre <- "#619CFF"
Color_Post <- "#F564E3"

Color_HNSCC <- "#009292"
Color_CSCC <- "#CD9600"
Color_Urothelial <- "#B66DFF"
Color_LN_melanoma <- "#00BFC4"
Color_Colon <- "#F8766D"
Color_Colon_MSI <- "#B73A3A"
Color_Colon_MSS <- "#23C26F"
Color_TNBC <- "#FEB5DA"
Color_BELLINI <- "#FEB5DA"
Color_IMCISION <- "#009292"
Color_MATISSE <- "#CD9600"
Color_NABUCCO <- "#B66DFF"
Color_BELLINI <- "#FEB5DA"
Color_NICHE_MSS <- "#23C26F"
Color_NICHE_MSI <- "#B73A3A"
Color_OpACIN_neo <- "#00BFC4"
Color_PRADO <- "#619CFF"

annotation_color_list = list(
  Response = c(NR = Color_NR, R = Color_R),
  Treatment = c(Pre = Color_Pre, Post = Color_Post),
  Histology = c(LN_melanoma = Color_LN_melanoma, Urothelial = Color_Urothelial,
                Colon = Color_Colon, HNSCC = Color_HNSCC, TBNC = Color_TNBC,
                CSCC = Color_CSCC),
  Trial = c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO))

trial_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_colors_combined <- c(Combined = "#999999", BELLINI = Color_BELLINI,
                           IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_Colon_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

histology_msi_mss_colors <- c(`Colon-MSI`= Color_Colon_MSI,
                              `Colon-MSS`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              LN_melanoma = Color_LN_melanoma,
                              TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)

```

# ----------

# Load gene sets

```{r Load msigdb hallmarks, echo = FALSE}
gene_signatures <- read_csv2(
  here("data", "processed", "signatures", "gene_signatures.csv"))

gene_signatures <- gene_signatures %>% filter(!is.na(Gene))

signature_overview <- read_csv2(
  here("data", "external", "signatures",
       "_overview_of_signatures.csv")) 

gene_signatures_meta_analysis <- signature_overview %>%
  filter(Meta_analysis == "Yes")

gene_signatures_meta_analysis %>% knitr::kable()

signature_overview %>%
  filter(!Signature == "TMB") %>%
  select(-c(Include, Obs, Signature, Signature_abbreviation, DOI, Class)) %>%
  mutate_if(is.character, as.factor) %>%
  summary(maxsum = 20) # %>% knitr::kable()

gene_signatures_meta_analysis %>%
  select(-c(Include, Obs, Signature, Signature_abbreviation, DOI, Class)) %>%
  mutate_if(is.character, as.factor) %>%
  summary(maxsum = 20) # %>% knitr::kable()

gene_signatures_l <- split(
  gene_signatures$Gene, gene_signatures$Signature)
gene_signatures_l <- gene_signatures_l[gene_signatures$Signature %>% unique()]

print(knitr::kable(gene_signatures %>%
        select(Signature, Class, Geneset_size) %>%
        unique()), n = n_distinct(gene_signatures$Signature))

gene_signatures_ssgsea <- gene_signatures %>% filter(use_ssgsea)

gene_signatures_ssgsea_l <- split(
  gene_signatures_ssgsea$Gene, gene_signatures_ssgsea$Signature)

ssgsea_signature_names <- names(gene_signatures_ssgsea_l)

gene_features <- read.csv2(
  file = here("data", "external", "biomaRt_gene_features.csv"))

gene_descriptions <- gene_features %>%
              select(hgnc_symbol, description) %>%
              rename("gene" = "hgnc_symbol") %>%
              unique()

```

# ----------

# Tissue specific genes human protein atlas

```{r}
list.files(here("data", "external", "human_protein_atlas"))

breast_tissue_rna_genes <- read_delim(
  here("data", "external", "human_protein_atlas",
       "tissue_category_rna_breast_Tissue.tsv"))

intestine_tissue_rna_genes <- read_delim(
  here("data", "external", "human_protein_atlas",
       "tissue_category_rna_intestine_Tissue.tsv"))

lymphoid_tissue_rna_genes <- read_delim(
  here("data", "external", "human_protein_atlas",
       "tissue_category_rna_lymphoid.tsv"))

skin_tissue_rna_genes <- read_delim(
  here("data", "external", "human_protein_atlas",
       "tissue_category_rna_skin_Tissue.tsv"))

urinary_tissue_rna_genes <- read_delim(
  here("data", "external", "human_protein_atlas",
       "tissue_category_rna_urinary.tsv"))

# Tissue Enriched: The gene has at least a four-fold higher mRNA level in a particular tissue compared to any other individual tissue. The gene is highly localized or specific to that single organ.

# Tissue Enhanced: The gene has at least a four-fold higher mRNA level in a particular tissue compared to the average level across all other tissues. It shows elevated expression in one tissue but may also be expressed at moderate levels in multiple other organs.

table(breast_tissue_rna_genes$`RNA tissue specificity`)
table(intestine_tissue_rna_genes$`RNA tissue specificity`)
table(lymphoid_tissue_rna_genes$`RNA tissue specificity`)
table(skin_tissue_rna_genes$`RNA tissue specificity`)
table(urinary_tissue_rna_genes$`RNA tissue specificity`)

```

# ----------

# Aggregate Differential Expression Results

## Load DESEq data per trial Response

```{r}
# Pre R x NR
deseq_r1_bellini <- read_csv2(
     here("data", "processed",
         "differential_expression_BELLINI", "deseq.results_r1.csv")) %>%
     mutate(Trial = "BELLINI", contrast = "Pre R x NR")
 
 deseq_r1_imcision <- read_csv2(
     here("data", "processed",
         "differential_expression_IMCISION", "deseq.results_r1.csv")) %>%
     mutate(Trial = "IMCISION", contrast = "Pre R x NR")
 
 deseq_r1_nabucco <- read_csv2(
     here("data", "processed",
         "differential_expression_NABUCCO", "deseq.results_r1.csv")) %>%
     mutate(Trial = "NABUCCO", contrast = "Pre R x NR")
 
 deseq_r1_niche_mss <- read_csv2(
     here("data", "processed",
         "differential_expression_NICHE-MSS", "deseq.results_r1.csv")) %>%
     mutate(Trial = "NICHE-MSS", contrast = "Pre R x NR")
 
 deseq_r1_OpACIN_neo <- read_csv2(
     here("data", "processed",
         "differential_expression_OpACIN_neo", "deseq.results_r1.csv")) %>%
     mutate(Trial = "OpACIN_neo", contrast = "Pre R x NR")
 
  deseq_r1_prado <- read_csv2(
     here("data", "processed",
         "differential_expression_PRADO", "deseq.results_r1.csv")) %>%
     mutate(Trial = "PRADO", contrast = "Pre R x NR")
 
 deseq_r1_matisse <- read_csv2(
     here("data", "processed",
         "differential_expression_MATISSE", "deseq.results_r1.csv")) %>%
     mutate(Trial = "MATISSE", contrast = "Pre R x NR")
```

## Combine

```{r}
deseq_r1_c <- deseq_r1_bellini %>%
   rbind(deseq_r1_imcision) %>%
   rbind(deseq_r1_nabucco) %>%
   rbind(deseq_r1_niche_mss) %>%
   rbind(deseq_r1_OpACIN_neo) %>%
   rbind(deseq_r1_prado) %>%
   rbind(deseq_r1_matisse)
 
 deseq_r1_c <- deseq_r1_c %>% filter(!Trial %in% "Combined")
 deseq_r1_c %>% write_csv2(here(output_dir, "deseq_r1_c.csv"))

 deseq_r1 <- read_csv2(
     here("data", "processed",
         "differential_expression", "deseq.results_r1.csv")) %>%
     mutate(Trial = "Combined", contrast = "Pre R x NR")
 
```

```{r, fig.height = 20, fig.width = 20}
deseq_r1_c %>%
  filter(abs(log2FoldChange) > log2(1.25)) %>%
  select(lfcSE, stat, pvalue, padj, Trial) %>%
  factor_summary()

deseq_r1_c %>%
  filter(abs(log2FoldChange) > 1.25) %>%
  select(lfcSE, stat, pvalue, padj, Trial) %>%
  factor_summary()

deseq_r1_c %>%
  ggplot(aes(x = log2FoldChange, y = -log10(pvalue), col = Trial)) +
  geom_point() + theme_Publication() + facet_wrap(~Trial, scales = "free") +
  geom_hline(yintercept = -log10(0.05))

deseq_r1_c %>%
  filter(abs(log2FoldChange) > log2(1.25)) %>%
  ggplot(aes(x = log2FoldChange, y = -log10(pvalue), col = Trial)) +
  geom_point() + theme_Publication() + facet_wrap(~Trial, scales = "free") +
  geom_hline(yintercept = -log10(0.05))

```

## Check Intersections - consistent up and downregulation across trials

```{r}
deseq_r1_c <- read_csv2(here(output_dir, "deseq_r1_c.csv"))

deseq_r1_c$log2FoldChange %>% summary()

deseq_r1_c %>% filter(log2FoldChange >= 0) %>% summary()

n_distinct(deseq_r1_c$gene)

genes_with_na_log2fc <- deseq_r1_c %>% filter(is.na(log2FoldChange))
table(genes_with_na_log2fc$gene) %>% as.data.frame() %>% filter(Freq > 1)
# table(genes_with_na_log2fc$gene, genes_with_na_log2fc$Trial)

deseq_r1_c %>% filter(is.na(log2FoldChange)) %>% select(baseMean) %>% summary()

gene_n_trials <- table(deseq_r1_c$gene,
      !is.na(deseq_r1_c$log2FoldChange)) %>%
  as.data.frame()  %>%
  setNames(c("gene", "not_na_log2fc", "n_trials"))%>%
  filter(not_na_log2fc == TRUE) %>%
  select(-not_na_log2fc)

gene_n_trials %>%
  ggplot(aes(x = n_trials)) + geom_histogram() + theme_Publication() +
  labs(x = "n_trials (not NA log2FC)")
```

## Define deseq_r1_meta_df: log2FC across trials

```{r}
# log2Fold changes for each trial - consisten directions
# Add mean log2FC, number of trials with response x non response enrichment, 
deseq_r1_wide <- deseq_r1_c %>%
  select(gene, log2FoldChange, Trial) %>%
  pivot_wider(names_from = Trial, values_from = log2FoldChange) 
deseq_r1_meta_df <- deseq_r1_wide

dim(deseq_r1_wide)

gene_direction <- deseq_r1_c %>%
  group_by(gene) %>%
  summarise(
    n_responder_enrichment = sum(log2FoldChange > 0, na.rm = TRUE),
    n_nresponder_enrichment = sum(log2FoldChange < 0, na.rm = TRUE),
    direction_consistent = (n_responder_enrichment == 0 |
                              n_nresponder_enrichment == 0),
    dominant_direction = ifelse(
      n_responder_enrichment >= n_nresponder_enrichment, "R", "NR")
  )

gene_sig_counts <- deseq_r1_c %>%
  mutate(
    significant_pvalue = pvalue < 0.05,
    significant_padj = padj < 0.05) %>%
  group_by(gene) %>%
  summarise(
    n_sig_pvalue = sum(as.numeric(significant_pvalue), na.rm = TRUE),
    n_sig_padj = sum(as.numeric(significant_padj), na.rm = TRUE))
  
deseq_r1_meta_df <- deseq_r1_meta_df %>%
  left_join(gene_direction) %>%
  left_join(gene_sig_counts) %>%
  left_join(gene_n_trials)

deseq_r1_meta_df$mean_log2fc <- rowMeans(
  deseq_r1_wide %>% column_to_rownames("gene"), na.rm = TRUE)
deseq_r1_meta_df$median_log2fc <- rowMedians(
  deseq_r1_wide %>% column_to_rownames("gene") %>% as.matrix(), na.rm = TRUE)

deseq_r1_meta_df <- deseq_r1_meta_df %>% left_join(gene_descriptions)

deseq_r1_meta_df %>% write_csv2(here(output_dir, "deseq_r1_meta_df.csv"))
deseq_r1_meta_df <- read_csv2(here(output_dir, "deseq_r1_meta_df.csv"))

```

# ----------

```{r}
deseq_r1_meta_df %>%
  ggplot(aes(x = n_trials, y = mean_log2fc)) +
  geom_pointdensity() + theme_Publication()+
  guides(colour = guide_colourbar(barwidth = 10))

deseq_r1_meta_df %>%
  ggplot(aes(x = n_trials, y = median_log2fc)) +
  geom_pointdensity() + theme_Publication()+
  guides(colour = guide_colourbar(barwidth = 10))

```

# Explore direction consistency

```{r}
# You may relax the filter adding '- 1' to the number of trials
# deseq_r1_consistent_upreg <- deseq_r1_meta_df %>%
#   filter(n_responder_enrichment >= n_distinct(deseq_r1_c$Trial)) %>%
#   arrange(-mean_log2fc)

deseq_r1_consistent_upreg <- deseq_r1_meta_df %>%
  filter(n_trials == n_distinct(deseq_r1_c$Trial)) %>%
  filter(direction_consistent == TRUE) %>%
  filter(dominant_direction == "R") %>%
  arrange(-mean_log2fc)

nrow(deseq_r1_consistent_upreg)
deseq_r1_consistent_upreg  %>% select(-description) %>% factor_summary()
table(deseq_r1_consistent_upreg$n_sig_pvalue == 0)
table(deseq_r1_consistent_upreg$n_sig_padj == 0)

deseq_r1_consistent_downreg <-  deseq_r1_meta_df  %>%
  filter(n_trials == n_distinct(deseq_r1_c$Trial)) %>%
  filter(direction_consistent == TRUE) %>%
  filter(dominant_direction == "NR") %>%
  arrange(-mean_log2fc)

deseq_r1_consistent_downreg %>% select(-description)  %>% factor_summary()
table(deseq_r1_consistent_downreg$n_sig_pvalue == 0)
table(deseq_r1_consistent_downreg$n_sig_padj == 0)
```

```{r}

deseq_r1_consistent_upreg %>%
  ggplot(aes(x = mean_log2fc)) + geom_histogram() +
  theme_Publication() + geom_vline(xintercept = log2(1.25))

deseq_r1_consistent_downreg %>%
  ggplot(aes(x = mean_log2fc)) + geom_histogram() +
  theme_Publication() + geom_vline(xintercept = -log2(1.25))

deseq_r1_consistent_downreg %>%
  filter(mean_log2fc < -log2(1.25) & median_log2fc < -log2(1.25)) %>%
  arrange(-mean_log2fc) %>% pull(gene)

deseq_r1_consistent_upreg_f <- deseq_r1_consistent_upreg %>%
  filter(mean_log2fc > log2(1.25) & median_log2fc > log2(1.25))

dim(deseq_r1_consistent_upreg_f)

deseq_r1_consistent_downreg_f <- deseq_r1_consistent_downreg %>%
  filter(mean_log2fc < -log2(1.25) & median_log2fc < -log2(1.25))

dim(deseq_r1_consistent_downreg_f)

```

# Explore magnitude

# Check mean log2fc vs number of trials

```{r}
deseq_r1_meta_df %>%
  ggplot(aes(x = mean_log2fc, y = median_log2fc,
             col = n_responder_enrichment)) + geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed")

deseq_r1_meta_df %>%
  ggplot(aes(x = n_responder_enrichment, y = mean_log2fc)) +
  geom_point() + theme_Publication() +
  geom_hline(yintercept = 0, linetype = "dashed")

deseq_r1_meta_df %>%
  filter(abs(mean_log2fc) > log2(1.25)) %>%
  ggplot(aes(x = n_responder_enrichment, y = mean_log2fc)) +
  geom_point() + theme_Publication() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = -log2(1.25), linetype = "dashed", col = "red") +
  geom_hline(yintercept = log2(1.25), linetype = "dashed", col = "red") +
  geom_hline(yintercept = -log2(1.5), linetype = "dashed", col = "red") +
  geom_hline(yintercept = log2(1.5), linetype = "dashed", col = "red") +
  geom_hline(yintercept = -log2(1.75), linetype = "dashed", col = "red") +
  geom_hline(yintercept = log2(1.75), linetype = "dashed", col = "red")

```

```{r}
deseq_r1 %>%
  left_join(deseq_r1_meta_df %>% select(gene, mean_log2fc)) %>%
  ggplot(aes(x = log2FoldChange, y = mean_log2fc)) +
  geom_pointdensity() + theme_Publication() +
  guides(colour = guide_colourbar(barwidth = 10))

```

## Venn diagrams

### Upregulated

```{r venn diagram upreg R, fig.height = 20, fig.width = 20}
deseq_r1_c_upreg_any <- deseq_r1_c %>%
  filter(log2FoldChange > 0) %>%
  filter(!Trial %in% c("Combined"))

deseq_r1_c_upreg_split <- split(deseq_r1_c_upreg_any$gene,
                                deseq_r1_c_upreg_any$Trial)
venn_list <- lapply(split(deseq_r1_c_upreg_any$gene,
                          deseq_r1_c_upreg_any$Trial), unique)
ggVennDiagram(venn_list)

# Filter above log2(1.25) ~ 0.32
deseq_r1_c_upreg <- deseq_r1_c %>%
  filter(log2FoldChange > log2(1.25)) %>%
  filter(!Trial %in% c("Combined"))

deseq_r1_c_upreg_split <- split(deseq_r1_c_upreg$gene,
                                deseq_r1_c_upreg$Trial)
venn_list <- lapply(split(deseq_r1_c_upreg$gene,
                          deseq_r1_c_upreg$Trial), unique)
ggVennDiagram(venn_list)

```

```{r Venn upset plot upreg}
deseq_r1_c_upreg <- deseq_r1_c %>% filter(log2FoldChange > 0)

deseq_r1_c_upreg_split <- split(deseq_r1_c_upreg$gene,
                                deseq_r1_c_upreg$Trial)
venn_list <- lapply(split(deseq_r1_c_upreg$gene,
                          deseq_r1_c_upreg$Trial), unique)
ggVennDiagram(venn_list, force_upset = TRUE)

deseq_r1_c_upreg <- deseq_r1_c %>% filter(log2FoldChange > log2(1.25))

deseq_r1_c_upreg_split <- split(deseq_r1_c_upreg$gene,
                                deseq_r1_c_upreg$Trial)
venn_list <- lapply(split(deseq_r1_c_upreg$gene,
                          deseq_r1_c_upreg$Trial), unique)
ggVennDiagram(venn_list, force_upset = TRUE)

```

### Downregulated

```{r, fig.height = 20, fig.width = 20}
deseq_r1_c_downreg <- deseq_r1_c %>%
  filter(log2FoldChange < 0) %>%
  filter(!Trial %in% c("Combined"))

deseq_r1_c_downreg_split <- split(deseq_r1_c_downreg$gene,
                                  deseq_r1_c_downreg$Trial)
venn_list <- lapply(split(deseq_r1_c_downreg$gene,
                          deseq_r1_c_downreg$Trial), unique)
ggVennDiagram(venn_list)

deseq_r1_c_downreg <- deseq_r1_c %>%
  filter(log2FoldChange < -(log2(1.25))) %>%
  filter(!Trial %in% c("Combined"))

deseq_r1_c_downreg_split <- split(deseq_r1_c_downreg$gene,
                                  deseq_r1_c_downreg$Trial)
venn_list <- lapply(split(deseq_r1_c_downreg$gene,
                          deseq_r1_c_downreg$Trial), unique)
ggVennDiagram(venn_list)
```

```{r Venn upset plot downreg}
deseq_r1_c_downreg <- deseq_r1_c %>% filter(log2FoldChange < 0)
deseq_r1_c_downreg_split <- split(deseq_r1_c_downreg$gene,
                                  deseq_r1_c_downreg$Trial)
venn_list <- lapply(split(deseq_r1_c_downreg$gene,
                          deseq_r1_c_downreg$Trial), unique)
ggVennDiagram(venn_list, force_upset = TRUE)

deseq_r1_c_downreg <- deseq_r1_c %>% filter(log2FoldChange< -(log2(1.25)))
deseq_r1_c_downreg_split <- split(deseq_r1_c_downreg$gene,
                                  deseq_r1_c_downreg$Trial)
venn_list <- lapply(split(deseq_r1_c_downreg$gene,
                          deseq_r1_c_downreg$Trial), unique)
ggVennDiagram(venn_list, force_upset = TRUE)

```

# Check signature genes
```{r, fig.height = 12, fig.width = 8}

meta_combined_effects_random_effects <- read_csv2(
  here(meta_rna_dir, "meta_combined_effects_random_effects.csv")) %>%
  arrange(TE.random) %>%
  mutate(Variable = ifelse(Variable == "Gamma_delta_TCR_Yu",
                           "Gamma_delta_TCRs", Variable))

gene_signatures %>% 
  filter(Signature %in% gene_signatures_meta_analysis$Signature) %>%
  left_join(deseq_r1_meta_df %>%
              rename(Gene = gene)) %>%
  ggplot(aes(x = factor(Signature,
                        levels = meta_combined_effects_random_effects$Variable),
             y = mean_log2fc)) +
  geom_boxplot() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "", x = "", y = "Mean Log2FC across trials")

gene_signatures %>% 
  filter(Signature %in% gene_signatures_meta_analysis$Signature) %>%
  left_join(deseq_r1_meta_df %>%
              rename(Gene = gene)) %>%
  ggplot(aes(x = factor(Signature,
                        levels = meta_combined_effects_random_effects$Variable),
             y = median_log2fc)) +
  geom_boxplot() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "", x = "", y = "Median Log2FC across trials")

```

# ----------

# Check mean log2fc cutoffs x n genes
## Response

```{r}
cutoffs <- seq(0, 2, by = 0.05)

deseq_r1_meta_cutoff_R <- lapply(cutoffs, function(cut) {
  passed <- deseq_r1_meta_df %>% filter(mean_log2fc >= cut)

  data.frame(
    cutoff = cut,
    n_genes = nrow(passed),
    avg_mean_log2fc = mean(passed$mean_log2fc, na.rm = TRUE)
  )
}) %>%
  bind_rows()
  
ggplot(deseq_r1_meta_cutoff_R, aes(x = cutoff, y = n_genes)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_point(aes(color = avg_mean_log2fc), size = 2.5) +
  scale_color_viridis_c() +
  labs(
    x = "mean_log2fc cutoff",
    y = "Number of genes passing",
    color = "Avg mean_log2fc",
    title = "Filtering by mean_log2fc: cutoff vs number of genes"
  ) +
  theme_minimal(base_size = 14)
```

## Non response

```{r}
cutoffs <- seq(0, -2, by = -0.05)

deseq_r1_meta_cutoff_NR <- lapply(cutoffs, function(cut) {
  passed <- deseq_r1_meta_df %>% filter(mean_log2fc <= cut)

  data.frame(
    cutoff = cut,
    n_genes = nrow(passed),
    avg_mean_log2fc = mean(passed$mean_log2fc, na.rm = TRUE)
  )
}) %>%
  bind_rows()
  
ggplot(deseq_r1_meta_cutoff_NR, aes(x = cutoff, y = n_genes)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_point(aes(color = avg_mean_log2fc), size = 2.5) +
  scale_color_viridis_c() +
  labs(
    x = "mean_log2fc cutoff",
    y = "Number of genes passing",
    color = "Avg mean_log2fc",
    title = "Filtering by mean_log2fc: cutoff vs number of genes"
  ) +
  theme_minimal(base_size = 14)
```

# ----------

# Define Positive component - Response 1.25 signature

```{r}
# Not based on mean log2fc only
# Consistency and Log2FC of genes above cutoff across all (measured, min 4) trials
deseq_r1_c_upreg_1.25 <- deseq_r1_c %>%
  left_join(gene_n_trials) %>%
  filter(log2FoldChange > log2(1.25)) %>%
  group_by(gene) %>%
  filter(n() >= n_trials)  %>% # Direction + magnitude consistency must be equal
  unique() %>%
  filter(n_trials > 4) %>% # Exclude genes measured in 4 or less 4 trials (only 5,6,7)
  ungroup() %>%
  pull(gene) %>%
  unique()

deseq_r1_c_upreg_1.25
length(deseq_r1_c_upreg_1.25)

deseq_r1_meta_df %>%
  filter(gene %in% deseq_r1_c_upreg_1.25)

deseq_r1_meta_df %>%
  filter(gene %in% deseq_r1_c_upreg_1.25) %>%
  arrange(-mean_log2fc) %>%
  write_csv2(here(output_dir, "pancan_response_signature_1.25.csv"))

pancan_response_signature_1.25 <- read_csv2(
  here(output_dir, "pancan_response_signature_1.25.csv"))
```


```{r}
gene_descriptions %>%
  filter(gene %in% deseq_r1_c_upreg_1.25)

gene_descriptions %>%
  filter(gene %in% deseq_r1_c_upreg_1.25) %>%
  left_join(deseq_r1_meta_df)

pancan_response_signature_1.25 %>%
  left_join(gene_descriptions) %>%
  relocate(cols = "description", .after = "gene") %>%
  arrange(-mean_log2fc) 

pancan_response_signature_1.25 %>%
  left_join(gene_descriptions) %>%
  relocate(cols = "description", .after = "gene") %>%
  arrange(-mean_log2fc)  %>%
  view()

pancan_response_signature_1.25 %>%
  filter(gene %in% breast_tissue_rna_genes$Gene)
pancan_response_signature_1.25 %>%
  filter(gene %in% intestine_tissue_enriched$Gene)
pancan_response_signature_1.25 %>%
  filter(gene %in% skin_tissue_enriched$Gene)
pancan_response_signature_1.25 %>%
  filter(gene %in% urinary_tissue_enriched$Gene)
pancan_response_signature_1.25 %>%
  filter(gene %in% lymphoid_tissue_enriched$Gene)

```

# Define negative component - Non-Response 1.25 signature

```{r}
# Not based on mean log2fc only
# Consistency and Log2FC of genes above cutoff across all (measured, min 4) trials
deseq_r1_c_downreg_1.25 <- deseq_r1_c %>%
  left_join(gene_n_trials) %>%
  filter(log2FoldChange < -log2(1.25)) %>%
  group_by(gene) %>%
  filter(n() >= n_trials)  %>%
  unique() %>%
  filter(n_trials > 4) %>%
  ungroup() %>%
  pull(gene) %>%
  unique()

deseq_r1_c_downreg_1.25
length(deseq_r1_c_downreg_1.25)

deseq_r1_meta_df %>%
  filter(gene %in% deseq_r1_c_downreg_1.25)

deseq_r1_meta_df %>%
  filter(gene %in% deseq_r1_c_downreg_1.25) %>%
  write_csv2(here(output_dir, "pancan_nonresponse_signature_1.25.csv"))

pancan_nonresponse_signature_1.25 <- read_csv2(
  here(output_dir, "pancan_nonresponse_signature_1.25.csv"))
```


```{r}
gene_descriptions %>%
  filter(gene %in% deseq_r1_c_downreg_1.25)

gene_descriptions %>%
  filter(gene %in% deseq_r1_c_downreg_1.25) %>%
  left_join(deseq_r1_meta_df)

deseq_r1_meta_df %>%
  filter(gene %in% deseq_r1_c_downreg_1.25) %>%
  left_join(gene_descriptions) %>%
  relocate(cols = "description", .after = "gene")

deseq_r1_meta_df %>%
  filter(gene %in% deseq_r1_c_downreg_1.25) %>%
  left_join(gene_descriptions) %>%
  relocate(cols = "description", .after = "gene") %>% view()

pancan_nonresponse_signature_1.25 %>%
  filter(gene %in% breast_tissue_rna_genes$Gene)
pancan_nonresponse_signature_1.25 %>%
  filter(gene %in% intestine_tissue_enriched$Gene)
pancan_nonresponse_signature_1.25 %>%
  filter(gene %in% skin_tissue_enriched$Gene)
pancan_nonresponse_signature_1.25 %>%
  filter(gene %in% urinary_tissue_enriched$Gene)
pancan_nonresponse_signature_1.25 %>%
  filter(gene %in% lymphoid_tissue_enriched$Gene)

```

# Cross check signatures
```{r}
gene_signatures %>% factor_summary()
gene_signatures %>%
  filter(!Signature %in% c("pancancer_response_sig_1.25")) %>%
  filter(Gene %in% pancan_response_signature_1.25$gene)

gene_signatures %>%
  filter(!Signature %in% c("pancan_nonresponse_signature_1.25")) %>%
  filter(Gene %in% pancan_nonresponse_signature_1.25$gene )

```

## Enrichr

```{r}
dbs <- c("GO_Biological_Process_2021",
         "Reactome_2022", "GO_Biological_Process_2025",
         "KEGG_2026", "Reactome_Pathways_2024")
```

## Enrichr signatures

```{r}
deseq_r1_meta_df %>%
  filter(gene %in% pancan_response_signature_1.25$gene) %>%
  dim()

deseq_r1_meta_df %>%
  filter(gene %in% pancan_response_signature_1.25$gene) %>%
  filter(!str_detect(gene, "ENSG")) %>%
  dim()

run_enrichr(
  pancan_response_signature_1.25 %>%
    filter(!str_detect(gene, "ENSG")),
    dbs, "deseq_r1_c_upreg_1.25_symbol.rds")
```

```{r}
deseq_r1_meta_df %>%
  filter(gene %in% pancan_nonresponse_signature_1.25$gene) %>%
  dim()

deseq_r1_meta_df %>%
  filter(gene %in% pancan_nonresponse_signature_1.25$gene) %>%
  filter(!str_detect(gene, "ENSG")) %>%
  dim()

run_enrichr(
  pancan_nonresponse_signature_1.25 %>%
    filter(!str_detect(gene, "ENSG")),
    dbs, "deseq_r1_c_downreg_1.25_symbol.rds")
```

# ----------

# Plot log2FC over consistent directions

```{r, fig.height = 40, fig.width = 10}
deseq_r1_c %>%
  filter(gene %in% deseq_r1_consistent_upreg$gene) %>%
  left_join(deseq_r1_consistent_upreg %>%
              select(gene, n_responder_enrichment,
                     n_nresponder_enrichment, mean_log2fc)) %>%
  ggplot(aes(x = reorder(gene, mean_log2fc),
             y = log2FoldChange, col = Trial)) +
  geom_point() + geom_hline(yintercept = 0) +
  theme_Publication() + coord_flip() +
  scale_color_manual(values = trial_colors_combined)
```

```{r, fig.height = 30, fig.width = 10}
deseq_r1_c %>%
  filter(gene %in% deseq_r1_consistent_upreg_f$gene)  %>%
  left_join(deseq_r1_consistent_upreg_f %>%
              select(gene, n_responder_enrichment,
                     n_nresponder_enrichment, mean_log2fc))%>%
  ggplot(aes(x = reorder(gene, mean_log2fc),
             y = log2FoldChange, col = Trial)) +
  geom_point() + geom_hline(yintercept = 0) +
  theme_Publication() + coord_flip()+
  scale_color_manual(values = trial_colors_combined)
```

```{r, fig.height = 40, fig.width = 10}
deseq_r1_c %>%
  filter(gene %in% deseq_r1_consistent_downreg$gene) %>%
  left_join(deseq_r1_consistent_downreg %>%
              select(gene, n_responder_enrichment,
                     n_nresponder_enrichment, mean_log2fc)) %>%
  ggplot(aes(x = reorder(gene, mean_log2fc),
             y = log2FoldChange, col = Trial)) +
  geom_point() + geom_hline(yintercept = 0) +
  theme_Publication() + coord_flip()+
  scale_color_manual(values = trial_colors_combined)

```

```{r, fig.height = 40, fig.width = 10}
deseq_r1_c %>%
  filter(gene %in% deseq_r1_consistent_downreg_f$gene) %>%
  left_join(deseq_r1_consistent_downreg_f %>%
              select(gene, n_responder_enrichment,
                     n_nresponder_enrichment, mean_log2fc)) %>%
  ggplot(aes(x = reorder(gene, mean_log2fc), y = log2FoldChange, col = Trial)) +
  geom_point() + geom_hline(yintercept = 0) +
  theme_Publication() + coord_flip()+
  scale_color_manual(values = trial_colors_combined)

```

## Visualize signatures

```{r, fig.height = 8, fig.width = 8}

deseq_r1_c %>%
  filter(gene %in% pancan_response_signature_1.25$gene) %>%
  left_join(deseq_r1_meta_df %>%
              select(gene, n_responder_enrichment,
                     n_nresponder_enrichment, mean_log2fc)) %>%
  ggplot(aes(x = reorder(gene, mean_log2fc),
             y = log2FoldChange, col = Trial)) +
  geom_point() +
  geom_point(aes(x = gene, y = mean_log2fc, col = "black"),
             shape = 5,
             data = deseq_r1_meta_df %>%
               filter(gene %in% pancan_response_signature_1.25$gene)) +
  geom_hline(yintercept = 0) +
  theme_Publication() + coord_flip() +
  scale_color_manual(values = trial_colors_combined) +
  geom_hline(yintercept = log2(1.25), linetype = "dashed") +
  labs(x = "")
```

```{r, fig.height = 8, fig.width = 8}
pancan_nonresponse_signature_1.25
deseq_r1_c %>%
  filter(gene %in% pancan_nonresponse_signature_1.25$gene) %>%
  left_join(deseq_r1_meta_df %>%
              select(gene, n_responder_enrichment,
                     n_nresponder_enrichment, mean_log2fc)) %>%
  ggplot(aes(x = reorder(gene, mean_log2fc),
             y = log2FoldChange, col = Trial)) +
  geom_point() +
  geom_point(aes(x = gene, y = mean_log2fc, col = "black"),
             shape = 5,
             data = deseq_r1_meta_df %>%
               filter(gene %in% pancan_nonresponse_signature_1.25$gene)) +
  geom_hline(yintercept = 0) +
  theme_Publication() + coord_flip() +
  scale_color_manual(values = trial_colors_combined) +
  geom_hline(yintercept = -log2(1.25), linetype = "dashed") +
  labs(x = "")
```

# ----------

# Lint script

```{r}
# lint("~/NKI_pancancer_neoadj_io/src/differential_expression_meta.Rmd")
lint(
  filename = "~/NKI_pancancer_neoadj_io/src/differential_expression_meta.Rmd",
  linters = unused_import_linter()
)
```

# ----------

# sessionInfo - Package Versions

```{r sessionInfo}
sessionInfo()
```
