---
title: "Differential expression"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  export: TRUE
  output_dir: "differential_expression"
  run_deseq_analysis: TRUE
  analyze_specific_trial: ""
  exclude_clinical_response: FALSE
  response_definition: "Original" # MPR pCR

---

Author comment:
File description comment:
The purpose of this script is to perform differential expression analysis
for neoadjuvant IO samples
It receives metadata and RNAseq molecular counts as inputs
Outputting DESeq2 results and volcano plots

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)
```

# Loading Packages
```{r}
# Install if necessary install.packages("BiocManager")
# Install if necessary BiocManager::install("DESeq2")
source("helper_functions.R")
library(tidyverse)
library(ggrepel)
library(here)
library(DESeq2)
library(plotly)
library(paletteer)
library(enrichR)
library(lintr)

set.seed(42)

select <- dplyr::select
rename <- dplyr::rename
here <- here::here
```

# ----------

# Function definitions
```{r}
make_dist_heatmap <- function(vsd, annotation_file, annotation_colors,
                              output_file,
                             width = 8, height = 8, cutree_cols = NA) {

  sample_dists <- dist(t(assay(vsd)[, rownames(annotation_file)]),
                       method = "euclidean")
  sample_dists_m <- as.matrix(sample_dists)

  sample_dist_heatmap <- pheatmap(sample_dists_m,
           annotation_col = annotation_file,
           col = paletteer_d(palette = "RColorBrewer::RdYlBu", direction = -1),
           clustering_distance_rows = sample_dists,
           clustering_distance_cols = sample_dists,
           show_colnames = FALSE,
           fontsize_col = 7,
           show_rownames = TRUE,
           cutree_cols = cutree_cols,
           annotation_colors = annotation_colors
           )

  save_pheatmap_pdf(x = sample_dist_heatmap,
                    filename = here(
                      output_dir, output_file),
                    width = width, height = height)
}

# Volcano plot function with nominal p-values
make_volcano_plot <- function(
  dataset, output_filename, alpha_threshold = 0.01, log2fcthresh = 1,
  export = TRUE, width = 6, height = 4) {

   dataset <- dataset %>% mutate(
  within_threshold = (pvalue < alpha_threshold &
                        abs(log2FoldChange) > log2fcthresh))

  plot <- dataset %>% ggplot(
  aes(x = log2FoldChange, y = -log10(pvalue + 1e-300), shape = pvalue > 0.01)) +
  geom_point(alpha = 0.8) +
  geom_vline(xintercept = 0, col = "black") +
  geom_vline(xintercept = c(-log2fcthresh, log2fcthresh), col = "red") +
    geom_hline(yintercept = -log10(alpha_threshold), colour = "red") +
  geom_text_repel(
  aes(x = log2FoldChange, y = -log10(pvalue + 1e-300), label = gene),
  data = dataset %>% filter(within_threshold), colour = "grey20",
  min.segment.length = 0) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(
    title = "Volcano plot Deseq2 Wald",
    x = "Effect size: log2 (fold-change)",
    y = "-log10 (p-value)",
    shape = "p-value > 0.01"
  )  +
    scale_color_gradient2(low = "red", mid = "gray",
                          high = "blue", midpoint = 1) +
    theme_Publication()

  print(plot)

  if (export) {
  ggsave(plot = plot, file = here(output_dir, output_filename),
         width = width, height = height, dpi = 300)
  }
}

# Volcano plot function with adjusted p-values (FDRq)
make_volcano_plot_fdr <- function(
  dataset, output_filename, alpha_threshold = 0.01, log2fcthresh = 2,
  export = TRUE, width = 6, height = 4) {

   dataset <- dataset %>% mutate(
  within_threshold = (padj < alpha_threshold &
                        abs(log2FoldChange) > log2fcthresh))

  plot <- dataset %>% ggplot(
  aes(x = log2FoldChange, y = -log10(padj + 1e-300), shape = padj > 0.01)) +
  geom_point(alpha = 0.8) +
  geom_vline(xintercept = 0, col = "black") +
  geom_vline(xintercept = c(-log2fcthresh, log2fcthresh), col = "red") +
    geom_hline(yintercept = -log10(alpha_threshold), colour = "red") +
  geom_text_repel(
  aes(x = log2FoldChange, y = -log10(padj + 1e-300), label = gene),
  data = dataset %>% filter(within_threshold), colour = "grey20",
  min.segment.length = 0) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(
    title = "Volcano plot Deseq2 Wald",
    x = "Effect size: log2 (fold-change)",
    y = "-log10 (FDRq)",
    shape = "FDRq > 0.01"
  )  +
    scale_color_gradient2(low = "red", mid = "gray",
                          high = "blue", midpoint = 1) +
    theme_Publication()

  print(plot)

  if (export) {
  ggsave(plot = plot, file = here(output_dir, output_filename),
         width = width, height = height, dpi = 300)
  }
}

run_deseq <- function(count_data, design_df,
                      output_suffix, design_formula,
                      contrast_col, contrast_ref, contrast_exp,
                      output_dir = output_dir, export) {
  
  deseq_rnaseq_data <- count_data %>% select(all_of(design_df$CF_sample))

  df_diff_design <- design_df

  # Change factor levels to compare contrasts (default is alphabetical)
  df_diff_design[contrast_col] <- factor(
    df_diff_design[, contrast_col, drop = TRUE],
    levels = c(contrast_ref, contrast_exp))

  if (sum(df_diff_design$CF_sample == colnames(deseq_rnaseq_data)) !=
      length(df_diff_design$CF_sample)) {
    print("sample column names and design do not match")
    return()
  }

  deseq_dataset <- DESeqDataSetFromMatrix(
    countData = deseq_rnaseq_data,
    colData = df_diff_design,
    design = as.formula(design_formula)
  ) %>%
    estimateSizeFactors()

  # create DESeq object
  dds_wald <- DESeq(deseq_dataset, test = "Wald")
  saveRDS(dds_wald,
          file = here(output_dir, sprintf("deseq_%s.rds", output_suffix)))

  res_wald <- results(dds_wald, cooksCutoff = FALSE)
  print(summary(res_wald))

  df_deseq_wald <- res_wald %>%
    as.data.frame() %>%
    rownames_to_column("gene")

  # Export
  df_deseq_wald %>% write.csv2(
    here(output_dir,
              sprintf("deseq.results_%s.csv", output_suffix)),
    row.names = FALSE)

}


select_top_extremes <- function(all_scores, top_n = 100) {
  # function that returns TRUE/FALSE for rankings within the top_n values
  increasing <- sort(all_scores, decreasing = FALSE)[1:top_n]
  decreasing <- sort(all_scores, decreasing = TRUE)[1:top_n]
  return(all_scores %in% increasing | all_scores %in% decreasing)
}


select_pval <- function(all_pvals, threshold = 0.01) {
  # function that returns TRUE/FALSE for pvalues below threshold
  return(all_pvals < threshold)
}

make_deg_heatmap <- function(count_data, samples, deg_df, annotation_df,
                             filename, width, height, palette_length = 50) {

  deg_df <- deg_df %>% arrange(stat)
  count_data <- count_data[deg_df$gene, samples]

  z_norm <- t(apply(
    as.matrix(count_data), 1,
    function(x) scale(x, center = TRUE, scale = TRUE)))

  colnames(z_norm) <- samples

  my_color <- colorRampPalette(c("blue", "white", "red"))(palette_length)
  # length(breaks) == length(palette_length) + 1
  # use floor and ceiling to deal with even/odd length pallettelengths
  my_breaks <- c(seq(min(z_norm), 0,
                    length.out = ceiling(palette_length / 2) + 1),
                seq(max(z_norm) / palette_length, max(z_norm),
                    length.out = floor(palette_length / 2)))

  deg_heatmap <- pheatmap(z_norm,
           annotation_col = annotation_df,
           col = my_color,
           breaks = my_breaks,
           show_colnames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE
           )
  save_pheatmap_pdf(deg_heatmap,
                    filename,
                    width, height)
}

```

# ----------
# Setup output folder
```{r}
print(params)

export <- params$export

# generate output directory
output_dir <- here("data", "processed", params$output_dir)
print(output_dir)

# overwrite clinical exclusion in case of MPR/pCR response
exclude_clinical_response <- params$exclude_clinical_response
if (params$response_definition %in% c("MPR", "pCR")) {
  exclude_clinical_response <- TRUE
}

# de_dir <- here("data", "processed", "differential_expression")
de_dir <- output_dir
analyze_specific_trial <- params$analyze_specific_trial
response_definition <- params$response_definition

run_r_vs_nr <- !(
  analyze_specific_trial %in% c("NICHE-MSI") &
    response_definition == "Original")

print(analyze_specific_trial)
print(run_r_vs_nr)
print(exclude_clinical_response)
print(response_definition)

# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)
} else {
  print("Output folder already exists")
}

```

# ----------

# Load metadata
```{r}
metadata <- read_csv2(
  here("data", "processed", "metadata", "combined_rna_metadata.csv"))

patient_metadata_combined <- read_csv2(
  here("data", "processed", "metadata", "patient_metadata_combined.csv"))

# Filter for specific trial
if (analyze_specific_trial != ""){
  metadata <- metadata %>% filter(Trial == analyze_specific_trial)
  }

metadata <- metadata %>%
  mutate(
      Treatment = factor(Treatment, levels = c("Pre", "Post")),
      CF_sample = as.factor(CF_sample),
      MSI_MSS = as.factor(MSI_MSS),
      BRAF = as.factor(BRAF),
      Patient = as.factor(Patient),
      Cohort = as.factor(Cohort),
      Response = as.factor(Response),
      Response_detail = as.factor(Response),
      Trial = as.factor(Trial))

```

## Adapt Response Definition
```{r}
metadata <- metadata %>%
    mutate(Response_original = Response) %>%
  left_join(patient_metadata_combined %>%
              select(Patient, Response_type, Response_pCR, Response_MPR)) %>%
  mutate(Response_analysis = response_definition,
         exclude_clinical_response = exclude_clinical_response)

# Automatic in case response definition is MPR and pCR
if (exclude_clinical_response) {
  metadata <- metadata %>%
    filter(Response_type == "Pathological")
}

if (response_definition == "MPR") {
  metadata <- metadata %>%
    filter(Response_type == "Pathological") %>%
    mutate(Response = ifelse(Response_MPR == "MPR", "R", "NR"))
}

if (response_definition == "pCR") {
  metadata <- metadata %>%
    filter(Response_type == "Pathological") %>%
    mutate(Response = ifelse(Response_pCR == "pCR", "R", "NR"))
}

```

# ----------
# Load RNAseq data
```{r load RNA data}
df_counts <- read_csv2(
  here("data", "processed", "rnaseq_data",
       "combined_rnaseq_counts_p.csv")) %>%
  column_to_rownames("gene")

rpm_counts <- read_csv2(here("data", "processed", "rnaseq_data",
                             "combined_rnaseq_rpm.csv")) %>%
  column_to_rownames("gene")

# Remove genes with all zero counts, might be detected in specific trials only
df_counts <- df_counts[, colnames(df_counts) %in%
                          metadata$CF_sample]
df_counts <- df_counts[rowSums(df_counts) != 0, ]

rpm_counts <- rpm_counts[, colnames(rpm_counts) %in%
                          metadata$CF_sample]
rpm_counts <- rpm_counts[rowSums(rpm_counts) != 0, ]

gene_features <- read.csv2(
  file = here("data", "external", "biomaRt_gene_features.csv"))

gene_descriptions <- gene_features %>%
              select(hgnc_symbol, description) %>%
              rename("gene" = "hgnc_symbol") %>%
              unique()

```

# Filter for samples with RNA data
```{r}
metadata <- metadata %>% filter(CF_sample %in% colnames(df_counts))
```

```{r}
table(metadata$Response)
table(metadata$Response, metadata$Response_MPR)
table(metadata$Response, metadata$Response_pCR)
table(metadata$Response, metadata$Response_original)

metadata %>%
  select(Trial, Response, Response_type, Response_MPR, Response_pCR,
         Response_original, Response_analysis) %>%
  factor_summary()

table(metadata$Treatment, metadata$Response)

unique_patient_id <- metadata %>% select(Trial, Patient) %>% unique()
table(unique_patient_id$Trial)

```

# Data structure - sample distances
```{r sample distances}
z_norm <- t(apply(
  as.matrix(df_counts[, colnames(df_counts) %in% metadata$CF_sample]), 1,
                  function(x) scale(x, center = TRUE, scale = TRUE)))

colnames(z_norm) <- colnames(df_counts)[
  colnames(df_counts) %in% metadata$CF_sample]

annot_df <- metadata %>%
  tibble::column_to_rownames("CF_sample") %>%
  select(Treatment, Response, Trial)

sample_dists <- dist(t(z_norm), method = "euclidean")
sample_dists_m <- as.matrix(sample_dists)

# create colours for each group
response_col <- c("#F8766D", "#00BA38")
names(response_col) <- unique(annot_df$Response)
treatment_col <- c("#F564E3", "#619CFF")
names(treatment_col) <- unique(annot_df$Treatment)
annotation_col <- list(Response = response_col, Treatment = treatment_col)
annotation_col = list(
  Response = c(NR="#F8766D", R="#00BA38"),
  Treatment = c(Pre="#F564E3", Post="#619CFF"))

sample_dist_heatmap <- pheatmap(sample_dists_m,
           annotation_col = annot_df,
           col = paletteer_d(palette = "RColorBrewer::RdYlBu", direction = -1),
           clustering_distance_rows = sample_dists,
           clustering_distance_cols = sample_dists,
           show_colnames = FALSE,
           show_rownames = FALSE,
           annotation_colors = annotation_col
           )

dds <- DESeqDataSetFromMatrix(
  countData = df_counts[, colnames(df_counts) %in%
                          metadata$CF_sample],
  colData = metadata,
  design = ~1
)

vsd <- vst(dds, blind = FALSE)

colSums(assay(vsd)) %>% as.data.frame() %>% summary()
```

```{r, fig.width = 14, fig.height = 14}
make_dist_heatmap(vsd = vsd, annotation_file = annot_df,
                  annotation_colors = annotation_col,
                  output_file = "distance_pheatmap_batch.pdf",
                             width = 10, height = 8)
```

# ----------

# PCA
```{r PCA}
pca <- prcomp(t(z_norm))
percent_var <- pca$sdev^2 / sum(pca$sdev^2)

percent_var <- percent_var %>%
  as.data.frame() %>%
  setNames("Var_exp") %>%
  rownames_to_column("PC") %>%
  mutate(Cumvar = cumsum(Var_exp), PC = as.numeric(PC))

percent_var %>%
  filter(PC < 50) %>%
  ggplot(aes(x = PC, y = Var_exp)) +
  geom_col() +
  geom_point(aes(x = PC, y = Cumvar)) +
  labs(x = "Principal Component", y = "(Cumulative) Variance Explained") +
  theme_Publication()

n_pcs <- 20

pca_heatmap <- pheatmap(t(pca$x[, 1:n_pcs]),
         annotation_col = annot_df,
         col = paletteer_d(palette = "RColorBrewer::RdYlBu", direction = -1),
         show_colnames = FALSE,
         cluster_rows = FALSE,
         annotation_colors = annotation_col
         )
```

```{r PCA on VSD data}
# PCA on VSD
ntop <- 500
rv <- rowVars(assay(vsd))
pca_selection <- order(rv, decreasing = TRUE)[seq_len(min(ntop,
        length(rv)))]
pca <- prcomp(t(assay(vsd)[pca_selection, ]))
percent_var <- pca$sdev^2 / sum(pca$sdev^2)

percent_var <- percent_var %>%
  as.data.frame() %>%
  setNames("Var_exp") %>%
  rownames_to_column("PC") %>%
  mutate(Cumvar = cumsum(Var_exp), PC = as.numeric(PC))

percent_var %>%
  filter(PC < 50) %>%
  ggplot(aes(x = PC, y = Var_exp)) +
  geom_col() +
  geom_point(aes(x = PC, y = Cumvar)) +
  labs(x = "Principal Component", y = "(Cumulative) Variance Explained") +
  theme_Publication()

pc1_var <- percent_var[1, 2] %>% round(2) * 100
pc2_var <- percent_var[2, 2] %>% round(2) * 100

# 10 PCs seems enough
n_pcs <- 10

pca_heatmap <- pheatmap(t(pca$x[, 1:n_pcs]),
         annotation_col = annot_df,
         col = paletteer_d(palette = "RColorBrewer::RdYlBu", direction = -1),
         show_colnames = FALSE,
         cluster_rows = FALSE,
         cutree_cols = 5,
         annotation_colors = annotation_col
         )

save_pheatmap_pdf(x = pca_heatmap,
                  filename = here(
                    output_dir, "pca_pheatmap.pdf"),
                  width = 8, height = 7)

saveRDS(pca, here(output_dir, "pca.RDS"))
```

```{r PCA plots}
pca_data <- plotPCA(vsd,
                    intgroup = c("Treatment",
                                 "Response", "Patient", "Trial"),
                    returnData = TRUE)

plotPCA(vsd, intgroup = c("Response")) +
  theme_Publication() +
  geom_point(size = 0.5)
ggsave(file = file.path(output_dir, "pca_response.png"),
       width = 6, height = 6, dpi = 300)

plotPCA(vsd, intgroup = c("Treatment")) +
  theme_Publication()
ggsave(file = file.path(output_dir, "pca_treatment.png"),
       width = 6, height = 6, dpi = 300)

plotPCA(vsd, intgroup = c("Treatment", "Response")) +
  theme_Publication()
ggsave(file = file.path(output_dir, "pca_treatment_response.png"),
       width = 6, height = 6, dpi = 300)

if (analyze_specific_trial == ""){
  plotPCA(vsd, intgroup = c("Trial")) +
  theme_Publication() +
  geom_point(size = 0.5)
  ggsave(file = file.path(output_dir, "pca_trial.png"),
         width = 6, height = 6, dpi = 300)
  
  plotPCA(vsd, intgroup = c("Histology")) +
    theme_Publication() +
    geom_point(size = 0.5)
  ggsave(file = file.path(output_dir, "pca_histology.png"),
         width = 6, height = 6, dpi = 300)
  
  plotPCA(vsd, intgroup = c("Treatment", "Trial")) +
    theme_Publication()
  
  pca_data %>%
    ggplot(aes(x = PC1, y = PC2, col = Trial, shape = Treatment)) +
    geom_point() +
    theme_Publication() +
    guides(col = guide_legend(nrow = 3, byrow = TRUE),
           shape = guide_legend(nrow = 2, byrow = TRUE)) +
    labs(x = sprintf("PC1: %s%% variance", pc1_var),
         y = sprintf("PC2: %s%% variance", pc2_var))
  ggsave(file = file.path(output_dir, "pca_trial_treatment.png"),
         width = 6, height = 6, dpi = 300)
  
  pca_data %>%
    ggplot(aes(x = PC1, y = PC2, col = Trial, shape = Response)) +
    geom_point() +
    theme_Publication()  +
    guides(col = guide_legend(nrow = 3, byrow = TRUE),
           shape = guide_legend(nrow = 2, byrow = TRUE)) +
    labs(x = sprintf("PC1: %s%% variance", pc1_var),
         y = sprintf("PC2: %s%% variance", pc2_var))
  ggsave(file = file.path(output_dir, "pca_trial_response.png"),
         width = 6, height = 6, dpi = 300)
}

```

```{r PCA loadings, fig.height = 8, fig.width = 6}
# Check the loadings PC1
loadings <- as.data.frame(pca$rotation) %>% arrange(PC1)

loadings[c(1:20, (nrow(loadings) - 20):nrow(loadings)), ] %>%
  rownames_to_column("Gene") %>%
  ggplot(aes(x = reorder(Gene, PC1), y = PC1)) +
  geom_col() +
  coord_flip() +
  labs(x = "Top 40 PC1 Loadings") +
  theme_Publication()

# Check the loadings PC2

loadings <- as.data.frame(pca$rotation) %>% arrange(PC2)

loadings[c(1:20, c(nrow(loadings) - 20):nrow(loadings)), ] %>%
  rownames_to_column("Gene") %>%
  ggplot(aes(x = reorder(Gene, PC2), y = PC2)) +
  geom_col() +
  coord_flip() +
  labs(x = "Top 40 PC2 Loadings") +
  theme_Publication()

```

# ----------
# DESeq2
```{r Treatment_DE, echo = TRUE}
run_deseq_analysis <-  params$run_deseq_analysis
alpha_threshold <- 0.05
log2fcthresh <- 1

```

# ----------

## Responders x non-responders DE
```{r Responder_DE, echo = TRUE}
design_r1 <- metadata %>%
  select(Treatment, Patient, CF_sample, Response, Trial) %>%
  filter(Treatment %in% c("Pre"))

table(design_r1$Trial, design_r1$Response)

run_deseq_analysis <-  params$run_deseq_analysis

# metadata <- metadata %>% filter(Trial == analyze_specific_trial)

response_design_formula <- "~ Response"

if (analyze_specific_trial == ""){
  response_design_formula <- "~ Trial + Response"
}

if (run_deseq_analysis & run_r_vs_nr) {
  print(Sys.time())
  run_deseq(count_data = df_counts, design_df = design_r1,
            output_suffix = "r1",
            design_formula = response_design_formula,
            contrast_col = "Response",
            contrast_ref = "NR", contrast_exp = "R",
            output_dir = output_dir, export)
  print(Sys.time())

}

if (run_r_vs_nr) {
  deseq_r1 <- read_csv2(
  here(de_dir, "deseq.results_r1.csv")) %>%
  left_join(gene_descriptions) %>%
  mutate(ranking = -log10(pvalue) * log2FoldChange)
  
  make_volcano_plot_fdr(deseq_r1,
                        alpha_threshold = alpha_threshold,
                        log2fcthresh = log2fcthresh,
                 "volcano_fdr_r1.png", export)
  
}

# Run naive direct response formula - all samples pooled analysis
if (analyze_specific_trial == "") {
  response_design_formula <- "~ Response"
  
  run_deseq_analysis <- TRUE
  if (run_deseq_analysis & run_r_vs_nr) {
    run_deseq(count_data = df_counts, design_df = design_r1,
              output_suffix = "r1_not_modelling_trial",
              design_formula = response_design_formula,
              contrast_col = "Response",
              contrast_ref = "NR", contrast_exp = "R",
              output_dir = output_dir, export)
  }

  deseq_r1_not_modelling_trial <- read_csv2(
    here(de_dir, "deseq.results_r1_not_modelling_trial.csv")) %>%
    left_join(gene_descriptions) %>%
    mutate(ranking = -log10(pvalue) * log2FoldChange)
  
  make_volcano_plot_fdr(deseq_r1_not_modelling_trial,
                        alpha_threshold = alpha_threshold,
                        log2fcthresh = log2fcthresh,
                 "volcano_fdr_r1_not_modelling_trial.png", export)
  
}

```

# ----------

# Get Differentially expressed genes
```{r Get top DEG}
# Either top_n with padj < threshold, or all genes with padj < threshold
top_n <- 100

if (run_r_vs_nr){
  deseq_r1_fdrq <- deseq_r1[
  select_pval(deseq_r1$padj, 0.05), ] %>%
  filter(!is.na(gene)) %>%
  arrange(padj)
  
  deseq_r1_top <- deseq_r1[
    select_top_extremes(deseq_r1$stat, top_n), ] %>%
    arrange(stat) %>%
    filter(padj < 0.05)
  
  print(nrow(deseq_r1_fdrq))
  print(nrow(deseq_r1_top))
  
  deseq_r1_fdrq %>% write.csv2(
    here(output_dir, "deseq_r1_fdrq.csv"), row.names = FALSE)

  deseq_r1_top %>% write.csv2(
    here(output_dir, "deseq_r1_top.csv"), row.names = FALSE)
  
}
```

# ----------

# Plot DEG Z-score heatmaps
```{r DEG heatmaps, fig.width = 10, fig.height = 17}

annot_df <- metadata %>%
  tibble::column_to_rownames("CF_sample") %>%
  select(Treatment, Response, Trial) %>%
  mutate_if(is.numeric, as.factor)

if (run_r_vs_nr){
  make_deg_heatmap(assay(vsd),
                   samples = as.character(
                     (design_r1 %>% arrange(Response))$CF_sample),
                   deg_df = deseq_r1_top,
                   annotation_df = annot_df,
                   filename = here(output_dir, "DEG_r1.pdf"),
                   height = 23)
  
  # DEG heatmap filter less genes
  make_deg_heatmap(assay(vsd),
                   samples = as.character(
                     (design_r1 %>% arrange(Response))$CF_sample),
                   deg_df = deseq_r1_top[
                     select_top_extremes(deseq_r1_top$stat, 25), ],
                   annotation_df = annot_df,
                   filename = here(output_dir, "DEG_r1_f.pdf"),
                   height = 8)
}
```

# ----------
# EnrichR Pathway Enrichment on Top DE genes
```{r, results = "asis"}
setEnrichrSite("Enrichr") # Human genes
# Check all_dbs <- listEnrichrDbs()

dbs <- c("GO_Biological_Process_2021",
         "Reactome_2022", "GO_Biological_Process_2025",
         "KEGG_2026", "Reactome_Pathways_2024")
```


```{r enrichr Response, results = "asis"}
if (run_r_vs_nr){
  # print("enrichr r1  \n")
  # print(nrow(deseq_r1_fdrq))
  # run_enrichr(deseq_r1_fdrq, dbs, "enrichr_r1.rds")

  ## Directional enrichr
  print("enrichr r1 lfc > 0:  \n")
  print(nrow(deseq_r1_fdrq %>% filter(log2FoldChange > 0)))

  run_enrichr(deseq_r1_fdrq %>% filter(log2FoldChange > 0),
            dbs, "enrichr_r1_up_fdrq.rds")
  
  print("enrichr r1 lfc < 0:  \n")
  print(nrow(deseq_r1_fdrq %>% filter(log2FoldChange < 0)))

  run_enrichr(deseq_r1_fdrq %>% filter(log2FoldChange < 0),
            dbs, "enrichr_r1_down_fdrq.rds")
  
  enrichr_r1 <- readRDS(here(output_dir, "enrichr_r1.rds"))
}

```

# ----------

# Plot RPM boxplots
```{r RPM boxplots}
if (run_r_vs_nr){
  deseq_r1_top10 <- deseq_r1_top[
    select_top_extremes(deseq_r1_top$stat, 5), ]

  filter_top_genes <- deseq_r1_top10$gene
  
  
  rpm_counts %>% select(metadata$CF_sample)
  
  rpm_counts_f <- rpm_counts[filter_top_genes,
                             as.character(metadata$CF_sample)]
  
  pseudolog2_rpm_counts <- log2(rpm_counts_f + 1)
  pseudolog2_rpm_counts <- pseudolog2_rpm_counts %>%
    t() %>%
    as.data.frame() %>%
    rownames_to_column("CF_sample") %>%
    left_join(metadata)
  
  colnames(pseudolog2_rpm_counts) <- gsub(" |-", "_",
                                          colnames(pseudolog2_rpm_counts))
  
  for (current_gene in deseq_r1_top10$gene) {
    print(current_gene)
    gene_col <- gsub(" |-", "_", current_gene)
    
    if (analyze_specific_trial == "") {
        plot <- pseudolog2_rpm_counts %>%
          filter(Treatment == "Pre") %>%
          ggplot(aes_string(x = "Response", y = gene_col, fill = "Trial")) +
          geom_boxplot(outlier.shape = NA) +
          geom_point(position = position_jitterdodge(jitter.width = 0.15),
                     aes(group = Response,
                       col = Response), size = 2.5) +
            labs(col = "Response") +
            expand_limits(y = 0) +
            labs(x = "",
                 y = sprintf("%s log2 (RPM+1)", current_gene),
                 title = "Pre-treatment") +
            theme(axis.text.x = element_text(
              angle = 90, vjust = 0.5, hjust = 1)) +
            theme_Publication() +
            theme(legend.spacing.x = unit(0.3, "cm")) +
      theme(legend.key.size = unit(0.75, 'cm')) +
      scale_color_manual(values = c("#F8766D", "#00BA38")) +
      guides(fill = guide_legend(nrow = 3, byrow = TRUE),
             col = guide_legend(nrow = 2, byrow = TRUE))
        print(plot)
    }
    
    if (analyze_specific_trial != "") {
      plot <- pseudolog2_rpm_counts %>%
          filter(Treatment == "Pre") %>%
          ggplot(aes_string(x = "Response", y = gene_col, fill = "Response")) +
          geom_boxplot(outlier.shape = NA) +
          geom_point(position = position_jitterdodge(jitter.width = 0.15),
                     aes(group = Response), size = 1.5) +
            labs(col = "Response") +
            expand_limits(y = 0) +
            labs(x = "",
                 y = sprintf("%s log2 (RPM+1)", current_gene),
                 title = "Pre-treatment") +
            theme(axis.text.x = element_text(
              angle = 90, vjust = 0.5, hjust = 1)) +
            theme_Publication() +
            theme(legend.spacing.x = unit(0.3, "cm")) +
      theme(legend.key.size = unit(0.75, 'cm')) +
      scale_fill_manual(values = c("#F8766D", "#00BA38"))
      print(plot)
    }
  }
}

```

# ----------

# Lint script

```{r}
# lint("~/NKI_pancancer_neoadj_io/src/differential_expression.Rmd")
lint(filename = "~/NKI_pancancer_neoadj_io/src/differential_expression.Rmd",
  linters = unused_import_linter()
)
```

# ----------
# sessionInfo - Package Versions
```{r sessionInfo}
sessionInfo()
```
