---
title: "gene signature scores meta analysis"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  output_dir: "gene_signatures_meta_MPR"
  metadata: "combined_rna_metadata.csv"
  analyze_specific_trial: ""
  run_meta_analysis: TRUE
  generate_heatmaps: TRUE
  use_demo: FALSE
  response_definition: "MPR" # MPR pCR Original
  exclude_monotherapy: FALSE
  exclude_niche_msi: FALSE
---

Author comment:
File description comment:
The purpose of this script is to perform a meta-analysis of signature effect sizes, comparing responders and non-responders within trials

It receives the individual signature scores for different trials, as well as the aggregated group means and SD for SMD calculations
and outputs meta-analysis results

# Loading Packages
```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# Install if necessary install.packages("BiocManager")
# Install if necessary BiocManager::install("fgsea")
library(tidyverse)
library(ggrepel)
library(here)
library(fgsea)
library(pheatmap)
library(DESeq2)
library(paletteer)
library(ggpubr)
library(rstatix)
source("helper_functions.R")
library(meta)
library(fgsea)
library(export)
library(lintr)

set.seed(42)
select <- dplyr::select
rename <- dplyr::rename
here <- here::here
```

# ----------

# Setup output folder
```{r, echo = FALSE, include = FALSE}
print(params)

export <- params$export
analyze_specific_trial <- params$analyze_specific_trial
run_meta_analysis <- params$run_meta_analysis   
generate_heatmaps <- params$generate_heatmaps
response_definition <- params$response_definition

# generate output directory
output_dir <- here("data", "processed", params$output_dir)
print(output_dir)
output_dir_plots <- here(output_dir, "plots")

# overwrite clinical exclusion in case of MPR/pCR response
exclude_clinical_response <- params$exclude_clinical_response
if (params$response_definition %in% c("MPR", "pCR")) {
  exclude_clinical_response <- TRUE
}
exclude_monotherapy <- params$exclude_monotherapy


data_dir <- here("data", "processed")
input_dir <- "processed"
if (params$use_demo) {
  # data_dir <- here("data", "processed_demo")
  input_dir <- "processed_demo"
}

combined_dir <- here(data_dir, "gene_signatures_MPR")
combined_dir_excl_dmmr <- here(data_dir, "gene_signatures_excl_dmmr")

# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)
    dir.create(output_dir_plots)
} else {
  print("Output folder already exists")
}

```

# ----------
# Load RNAseq data
```{r load rnaseq data, include = TRUE}
df_counts <- read_csv2(here("data", input_dir, "rnaseq_data",
                             "combined_rnaseq_counts_p.csv")) %>%
  column_to_rownames("gene")

rna_metadata <- read_csv2(
  here("data", input_dir, "metadata", params$metadata))

if (analyze_specific_trial != ""){
  rna_metadata <- rna_metadata %>% filter(Trial == analyze_specific_trial)
}

rna_metadata <- rna_metadata %>%
  mutate(
      Treatment = factor(Treatment, levels = c("Pre", "Post")),
      CF_sample = as.factor(CF_sample),
      Trial = as.factor(Trial),
      Response = as.factor(Response),
      Response_detail = as.factor(Response_detail),
      MSI_MSS = as.factor(MSI_MSS),
      BRAF = as.factor(BRAF))

rna_metadata %>% mutate_if(is.character, as.factor) %>% summary()

# Remove genes with all zero counts, might be detected in specific trials only
df_counts <- df_counts[, colnames(df_counts) %in%
                          rna_metadata$CF_sample]
df_counts <- df_counts[rowSums(df_counts) != 0, ]


unique_patient_id <- rna_metadata %>% select(Trial, Patient) %>% unique()
table(unique_patient_id$Trial)
 
histology_df <- data.frame(
  Trial = c("BELLINI", "IMCISION", "NABUCCO", "NICHE-MSS", "NICHE-MSI",
            "OpACIN_neo", "PRADO", "MATISSE"),
  Histology = c("TBNC", "HNSCC", "Urothelial", "Colon", "Colon",
                "LN_melanoma", "LN_melanoma", "CSCC"))

# Triple Negative Breast cancer, Head and Neck Squamous Cell Carcinoma,
# Urothelial Carcinoma (UC), Colon cancer (CC), 
# cutaneous squamous cell carcinoma (CSCC) 

```

# Annotation color
```{r}
# Define colours for each group
Color_NR <- "#F8766D"
Color_R <- "#00BA38" # "#619CFF"
Color_NA <- "#CD9600"
Color_Pre <- "#619CFF"
Color_Post <- "#F564E3"

Color_HNSCC <- "#009292"
Color_CSCC <- "#CD9600"
Color_Urothelial <- "#B66DFF"
Color_LN_melanoma <- "#00BFC4"
Color_Colon <- "#F8766D"
Color_Colon_MSI <- "#B73A3A"
Color_Colon_MSS <- "#23C26F"
Color_TNBC <- "#FEB5DA"
Color_BELLINI <- "#FEB5DA"
Color_IMCISION <- "#009292"
Color_MATISSE <- "#CD9600"
Color_NABUCCO <- "#B66DFF"
Color_BELLINI <- "#FEB5DA"
Color_NICHE_MSS <- "#23C26F"
Color_NICHE_MSI <- "#B73A3A"
Color_OpACIN_neo <- "#00BFC4"
Color_PRADO <- "#619CFF"

annotation_color_list = list(
  Response = c(NR = Color_NR, R = Color_R),
  Treatment = c(Pre = Color_Pre, Post = Color_Post),
  Histology = c(LN_melanoma = Color_LN_melanoma, Urothelial = Color_Urothelial,
                Colon = Color_Colon, HNSCC = Color_HNSCC, TBNC = Color_TNBC,
                CSCC = Color_CSCC),
  Trial = c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO))

trial_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_colors_combined <- c(Combined = "#999999", BELLINI = Color_BELLINI,
                           IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_Colon_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

histology_msi_mss_colors <- c(`Colon-MSI`= Color_Colon_MSI,
                              `Colon-MSS`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              LN_melanoma = Color_LN_melanoma,
                              TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)

```

# ----------

# Load gene sets
```{r Load msigdb hallmarks}
gene_signatures <- read_csv2(
  here("data", input_dir, "signatures", "gene_signatures.csv"))

gene_signatures <- gene_signatures %>% filter(!is.na(Gene))

signature_overview <- read_csv2(
  here("data", input_dir, "gene_signatures",
       "_overview_of_signatures.csv")) 
  # mutate(ifelse(Signature == "B_catenin_signature_pancan_22_Luke_et_al", ))

signature_overview$Signature <- gsub(
  "B_catenin_signature_pancan_22_Luke_et_al",
  "B_catenin_signature_Luke_et_al", signature_overview$Signature)

# signature_overview$Signature <- gsub(
#   "Gamma_delta_TCR_Yu",
#   "Gamma_delta_TCRs", signature_overview$Signature)

gene_signatures_meta_analysis <- signature_overview %>%
  filter(Meta_analysis == "Yes")

# Export
# gene_signatures %>%
#   filter(Signature %in% gene_signatures_meta_analysis$Signature) %>%
#   write_csv2(here(figures_dir, "gene_signatures_meta_supp_table.csv"))

gene_signatures_meta_analysis %>% knitr::kable()

signature_overview %>%
  filter(!Signature == "TMB") %>%
  select(-c(Include, Signature, Signature_abbreviation, DOI, Class)) %>%
  mutate_if(is.character, as.factor) %>%
  summary(maxsum = 20)

gene_signatures_meta_analysis %>%
  select(-c(Include, Signature, Signature_abbreviation, DOI, Class)) %>%
  mutate_if(is.character, as.factor) %>%
  summary(maxsum = 20)

gene_signatures_l <- split(
  gene_signatures$Gene, gene_signatures$Signature)
gene_signatures_l <- gene_signatures_l[gene_signatures$Signature %>% unique()]

print(knitr::kable(gene_signatures %>%
        select(Signature, Class, Geneset_size) %>%
        unique()), n = n_distinct(gene_signatures$Signature))

gene_signatures_ssgsea <- gene_signatures %>% filter(use_ssgsea)

gene_signatures_ssgsea_l <- split(
  gene_signatures_ssgsea$Gene, gene_signatures_ssgsea$Signature)

ssgsea_signature_names <- names(gene_signatures_ssgsea_l)

```

# Load ssGSEA and RPM data
```{r}
# combined analysis
ssgsea_results <- read_csv2(here(combined_dir, "ssgsea_results.csv")) %>%
  column_to_rownames("Signature")

ssgsea_results_df <- ssgsea_results %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("CF_sample") %>%
  left_join(rna_metadata) %>%
  left_join(histology_df)

ssgsea_results_df %>% write_csv2(here(output_dir, "ssgsea_results_df.csv"))

signature_response_ssgsea_raw <- read_csv2(
  here(combined_dir, "signature_response_ssgsea.csv")) %>%
  mutate(Trial = "Combined")

# signature_response_ssgsea_excl_dmmr <- read_csv2(
#   here(combined_dir_excl_dmmr, "signature_response_ssgsea.csv")) %>%
#   mutate(Trial = "Combined_excl_dMMR")

signature_rpm_data <- read_csv2(
  here(combined_dir, "signature_rpm_data.csv"))

# signature_response_rpm_expression_excl_dmmr <- read_csv2(
#   here(combined_dir_excl_dmmr, "signature_response_rpm_expression.csv")) %>%
#   mutate(Trial = "Combined_excl_dMMR")
```



## Adapt Response Definition
```{r}
compare_r_nr_trials <- c("Combined", "Combined_excl_dMMR", "BELLINI",
                         "IMCISION", "MATISSE", "NABUCCO", "NICHE-MSS",
                         "OpACIN_neo", "PRADO")

if (params$response_definition %in% c("MPR")) {
  compare_r_nr_trials <- c("Combined", "Combined_excl_dMMR", "BELLINI",
                         "IMCISION", "MATISSE", "NABUCCO",
                         "NICHE-MSS", "OpACIN_neo", "PRADO")
}

if (params$response_definition %in% c("pCR")) {
  compare_r_nr_trials <- c("Combined", "Combined_excl_dMMR", "BELLINI",
                          "MATISSE", "NABUCCO", "NICHE-MSI", "NICHE-MSS", 
                         "OpACIN_neo", "PRADO")
}

patient_metadata_combined <- read_csv2(
  here("data", "processed", "metadata", "patient_metadata_combined.csv"))

ssgsea_results_df <- ssgsea_results_df %>%
    mutate(Response_original = Response) %>%
  left_join(patient_metadata_combined %>%
              select(Patient, Response_type, Response_pCR, Response_MPR)) %>%
  mutate(Response_analysis = response_definition,
         exclude_clinical_response = exclude_clinical_response)

signature_rpm_data_df <- signature_rpm_data %>%
    mutate(Response_original = Response) %>%
  left_join(patient_metadata_combined %>%
              select(Patient, Response_type, Response_pCR, Response_MPR)) %>%
  mutate(Response_analysis = response_definition,
         exclude_clinical_response = exclude_clinical_response)

# overwrite clinical exclusion in case of MPR/pCR response
if (params$response_definition %in% c("MPR", "pCR")) {
  exclude_clinical_response <- TRUE
}

# Automatic in case response definition is MPR and pCR
if (exclude_clinical_response) {
  ssgsea_results_df <- ssgsea_results_df %>%
    filter(Response_type == "Pathological")
  
}

if (response_definition == "MPR") {
  ssgsea_results_df <- ssgsea_results_df %>%
    filter(Response_type == "Pathological") %>%
    mutate(Response = ifelse(Response_MPR == "MPR", "R", "NR"))
  
  signature_rpm_data_df <- signature_rpm_data_df %>%
    filter(Response_type == "Pathological") %>%
    mutate(Response = ifelse(Response_MPR == "MPR", "R", "NR"))
}

if (response_definition == "pCR") {
  ssgsea_results_df <- ssgsea_results_df %>%
    filter(Response_type == "Pathological") %>%
    mutate(Response = ifelse(Response_pCR == "pCR", "R", "NR"))
  
  signature_rpm_data_df <- signature_rpm_data_df %>%
    filter(Response_type == "Pathological")%>%
    mutate(Response = ifelse(Response_pCR == "pCR", "R", "NR"))
}

if (exclude_monotherapy) {
  ssgsea_results_df <- ssgsea_results_df %>%
    filter(Cohort == "Nivo_ipi") 
  
  signature_rpm_data_df <- signature_rpm_data_df  %>%
    filter(Cohort == "Nivo_ipi") 
}

ssgsea_results_df$Cohort %>% table()

# Sensitivity analysis excluding NICHE2 dMMR MSI samples
if (params$exclude_niche_msi) {
  ssgsea_results_df <- ssgsea_results_df %>%
    filter(!Trial %in% c("NICHE-MSI"))
  
  signature_rpm_data_df <- signature_rpm_data_df %>%
    filter(!Trial %in% c("NICHE-MSI"))
}

ssgsea_results_df %>%
  select(colnames(patient_metadata_combined)[
    colnames(patient_metadata_combined) %in% colnames(ssgsea_results_df)]) %>%
  mutate_if(is.character, as.factor) %>% summary()

table(ssgsea_results_df$Trial, ssgsea_results_df$Response)

ssgsea_results_df$Trial %>% table()


ssgsea_results_df_pre <- ssgsea_results_df %>% filter(Treatment == "Pre")

table(ssgsea_results_df_pre$Trial, ssgsea_results_df_pre$Response)

ssgsea_results_df_pre$Trial %>% table()

```

# Load ssGSEA data expression contrasts

```{r}

## Individual trials Response
signature_response_ssgsea_bellini <- read_csv2(
  here(data_dir, "gene_signatures_MPR_BELLINI",
       "signature_response_ssgsea.csv")) %>%
  mutate(Trial = "BELLINI") 
 
signature_response_ssgsea_imcision <- read_csv2(
  here(data_dir, "gene_signatures_MPR_IMCISION",
       "signature_response_ssgsea.csv")) %>%
  mutate(Trial = "IMCISION") 

signature_response_ssgsea_nabucco <- read_csv2(
  here(data_dir, "gene_signatures_MPR_NABUCCO",
       "signature_response_ssgsea.csv")) %>%
  mutate(Trial = "NABUCCO") 

signature_response_ssgsea_niche_mss <- read_csv2(
  here(data_dir, "gene_signatures_MPR_NICHE-MSS",
       "signature_response_ssgsea.csv")) %>%
  mutate(Trial = "NICHE-MSS") 

signature_response_ssgsea_opacin_neo <- read_csv2(
  here(data_dir, "gene_signatures_MPR_OpACIN_neo",
       "signature_response_ssgsea.csv")) %>%
  mutate(Trial = "OpACIN_neo") 

signature_response_ssgsea_prado <- read_csv2(
  here(data_dir, "gene_signatures_MPR_PRADO",
       "signature_response_ssgsea.csv")) %>%
  mutate(Trial = "PRADO") 

signature_response_ssgsea_matisse <- read_csv2(
  here(data_dir, "gene_signatures_MPR_MATISSE",
       "signature_response_ssgsea.csv")) %>%
  mutate(Trial = "MATISSE") 

signature_response_ssgsea_c <- signature_response_ssgsea_bellini %>%
  rbind(signature_response_ssgsea_imcision) %>%
  rbind(signature_response_ssgsea_nabucco) %>%
  rbind(signature_response_ssgsea_niche_mss) %>%
  rbind(signature_response_ssgsea_opacin_neo) %>%
  rbind(signature_response_ssgsea_prado)  %>%
  rbind(signature_response_ssgsea_matisse) %>%
  rbind(signature_response_ssgsea_raw) # %>%
  # rbind(signature_response_ssgsea_excl_dmmr)

signature_response_ssgsea_c <- signature_response_ssgsea_c %>%
  mutate(mean_delta = mean_responders - mean_nresponders,
         pooled_sd = sqrt((sd_responders^2 + sd_nresponders^2) / 2),
         effect_size = mean_delta / pooled_sd)

signature_response_ssgsea_c <- signature_response_ssgsea_c %>%
  filter(signature %in% signature_overview$Signature)

signature_response_ssgsea_c %>%
  write_csv2(here(output_dir, "signature_response_ssgsea_c.csv"))

annot_df <- ssgsea_results_df %>%
    tibble::column_to_rownames("CF_sample") %>%
    select(Response, Treatment, Trial, Histology) %>%
    mutate_if(is.numeric, as.factor)

```

# Load RPM data expression contrasts
```{r}

signature_response_rpm_expression_raw <- read_csv2(
  here(combined_dir, "signature_response_rpm_expression.csv")) %>%
  mutate(Trial = "Combined")

# signature_response_rpm_expression_excl_dmmr <- read_csv2(
#   here(combined_dir_excl_dmmr, "signature_response_rpm_expression.csv")) %>%
#   mutate(Trial = "Combined_excl_dMMR")

## Individual trials Response
signature_response_rpm_expression_bellini <- read_csv2(
  here(data_dir, "gene_signatures_MPR_BELLINI",
       "signature_response_rpm_expression.csv")) %>%
  mutate(Trial = "BELLINI") 
 
signature_response_rpm_expression_imcision <- read_csv2(
  here(data_dir, "gene_signatures_MPR_IMCISION",
       "signature_response_rpm_expression.csv")) %>%
  mutate(Trial = "IMCISION") 

signature_response_rpm_expression_nabucco <- read_csv2(
  here(data_dir, "gene_signatures_MPR_NABUCCO",
       "signature_response_rpm_expression.csv")) %>%
  mutate(Trial = "NABUCCO") 

signature_response_rpm_expression_niche_mss <- read_csv2(
  here(data_dir, "gene_signatures_MPR_NICHE-MSS",
       "signature_response_rpm_expression.csv")) %>%
  mutate(Trial = "NICHE-MSS") 

signature_response_rpm_expression_opacin_neo <- read_csv2(
  here(data_dir, "gene_signatures_MPR_OpACIN_neo",
       "signature_response_rpm_expression.csv")) %>%
  mutate(Trial = "OpACIN_neo") 

signature_response_rpm_expression_prado <- read_csv2(
  here(data_dir, "gene_signatures_MPR_PRADO",
       "signature_response_rpm_expression.csv")) %>%
  mutate(Trial = "PRADO") 

signature_response_rpm_expression_matisse <- read_csv2(
  here(data_dir, "gene_signatures_MPR_MATISSE",
       "signature_response_rpm_expression.csv")) %>%
  mutate(Trial = "MATISSE") 

signature_response_rpm_expression_c <- signature_response_rpm_expression_bellini %>%
  rbind(signature_response_rpm_expression_imcision) %>%
  rbind(signature_response_rpm_expression_nabucco) %>%
  rbind(signature_response_rpm_expression_niche_mss) %>%
  rbind(signature_response_rpm_expression_opacin_neo) %>%
  rbind(signature_response_rpm_expression_prado)  %>%
  rbind(signature_response_rpm_expression_matisse) %>%
  rbind(signature_response_rpm_expression_raw) # %>%
  # rbind(signature_response_rpm_expression_excl_dmmr)

signature_response_rpm_expression_c <- signature_response_rpm_expression_c %>%
  mutate(mean_delta = mean_responders - mean_nresponders,
         pooled_sd = sqrt((sd_responders^2 + sd_nresponders^2) / 2),
         effect_size = mean_delta / pooled_sd)

signature_overview$Signature[
  !signature_overview$Signature %in%
    signature_response_rpm_expression_c$signature]

signature_response_rpm_expression_c$signature[
  !signature_response_rpm_expression_c$signature %in% signature_overview$Signature] %>%
  unique()
signature_response_rpm_expression_c$signature %>%
  n_distinct()

signature_response_rpm_expression_c <- signature_response_rpm_expression_c %>%
  filter(signature %in% signature_overview$Signature)

signature_response_rpm_expression_c %>%
  write_csv2(here(output_dir, "signature_response_rpm_expression_c.csv"))

signature_response_rpm_expression_c$signature[
  !signature_response_rpm_expression_c$signature %in% signature_overview$Signature] %>%
  unique()
signature_response_rpm_expression_c$signature %>%
  n_distinct()
```

# ----------
# Merge ssgsea and RPM
```{r}
signature_response_ssgsea_c <- read_csv2(
  here(output_dir, "signature_response_ssgsea_c.csv"))

signature_response_rpm_expression_c <- read_csv2(
  here(output_dir, "signature_response_rpm_expression_c.csv"))

signature_response_c <- signature_response_ssgsea_c %>%
  mutate(signature_score = "ssgsea") %>%
  rbind(signature_response_rpm_expression_c %>%
  filter(!signature %in% signature_response_ssgsea_c$signature) %>%
    mutate(signature_score = "log(RPM+1)"))

table(
  gene_signatures_meta_analysis$Signature %in% signature_response_c$signature)
gene_signatures_meta_analysis$Signature[
  !gene_signatures_meta_analysis$Signature %in% signature_response_c$signature]
```

```{r}
# Compute Hedge's g instead of Cohen's d
signature_response_c <- signature_response_c %>%
  mutate(effect_size_cohens_d = effect_size) %>%
  mutate(weighted_pooled_sd = sqrt(
    ((sd_responders^2 * (n_responders - 1)) + (sd_nresponders^2 * (n_nresponders - 1))) / (n_responders + n_nresponders - 2)
    )
    ) %>%
  mutate(effect_size_hedges_g = mean_delta / weighted_pooled_sd)

signature_response_c %>%
  ggplot(aes(x = pooled_sd, y = weighted_pooled_sd)) +
  geom_point() + theme_Publication() + geom_abline(intercept = 0, slope = 1)
signature_response_c %>%
  ggplot(aes(x = effect_size_cohens_d, y = effect_size_hedges_g)) +
  geom_point() + theme_Publication() + geom_abline(intercept = 0, slope = 1)

signature_response_c <- signature_response_c %>%
  mutate(effect_size_hedges_g_c = ifelse(
    n < 50, effect_size_hedges_g * ((n - 3) / (n - 2.25)) * sqrt((n - 2)/n),
    effect_size_hedges_g))

signature_response_c %>%
  filter(n < 50) %>%
  ggplot(aes(x = effect_size_hedges_g, y = effect_size_hedges_g_c)) +
  geom_point() + theme_Publication() + geom_abline(intercept = 0, slope = 1)

```

```{r}
signature_response_c %>%
  write_csv2(here(output_dir, "signature_response_c.csv"))

signature_response_c <- read_csv2(
  here(output_dir, "signature_response_c.csv"))

signature_response_c$signature[
  !signature_response_c$signature %in% signature_overview$Signature] %>%
  unique()
```

# PCA
```{r}
rna_metadata_f <- rna_metadata %>% filter(CF_sample %in% colnames(df_counts))
rna_metadata %>% filter(!CF_sample %in% rna_metadata_f$CF_sample) 

rna_metadata_f <- rna_metadata_f %>% filter(Treatment == "Pre")

dds <- DESeqDataSetFromMatrix(
  countData = df_counts[, colnames(df_counts) %in%
                          rna_metadata_f$CF_sample],
  colData = rna_metadata_f,
  design = ~1
)

vsd <- vst(dds, blind = FALSE)

# PCA on VSD
ntop <- 500
rv <- rowVars(assay(vsd))
pca_selection <- order(rv, decreasing = TRUE)[seq_len(min(ntop,
        length(rv)))]
pca <- prcomp(t(assay(vsd)[pca_selection, ]))
percent_var <- pca$sdev^2 / sum(pca$sdev^2)

percent_var <- percent_var %>%
  as.data.frame() %>%
  setNames("Var_exp") %>%
  rownames_to_column("PC") %>%
  mutate(Cumvar = cumsum(Var_exp), PC = as.numeric(PC))

percent_var %>%
  filter(PC < 50) %>%
  ggplot(aes(x = PC, y = Var_exp)) +
  geom_col() +
  geom_point(aes(x = PC, y = Cumvar)) +
  labs(x = "Principal Component", y = "(Cumulative) Variance Explained") +
  theme_Publication()

pc1_var <- percent_var[1, 2] %>% round(2) * 100
pc2_var <- percent_var[2, 2] %>% round(2) * 100

# 10 PCs seems enough
n_pcs <- 10

annot_df <- rna_metadata %>%
  tibble::column_to_rownames("CF_sample") %>%
  select(Treatment, Response, Trial)

pca_heatmap <- pheatmap(t(pca$x[, 1:n_pcs]),
         annotation_col = annot_df,
         col = paletteer_d(palette = "RColorBrewer::RdYlBu", direction = -1),
         show_colnames = FALSE,
         cluster_rows = FALSE,
         cutree_cols = 5,
         annotation_colors = annotation_color_list
         )

save_pheatmap_pdf(x = pca_heatmap,
                  filename = here(
                    output_dir, "pca_pheatmap.pdf"),
                  width = 8, height = 7)

saveRDS(pca, here(output_dir, "pca.RDS"))

pca_data <- plotPCA(vsd,
                    intgroup = c("Treatment", "Response", "Patient", "Trial"),
                    returnData = TRUE)

plotPCA(vsd, intgroup = c("Trial")) +
  theme_Publication() +
  geom_point(size = 0.5)
ggsave(file = file.path(output_dir, "pca_trial.png"),
         width = 6, height = 6, dpi = 300)
  
plotPCA(vsd, intgroup = c("Histology")) +
  theme_Publication() +
  geom_point(size = 0.5) 
ggsave(file = file.path(output_dir, "pca_histology.png"),
       width = 6, height = 6, dpi = 300)

plotPCA(vsd, intgroup = c("Treatment", "Trial")) +
  theme_Publication()

pca_data %>%
  ggplot(aes(x = PC1, y = PC2, col = Trial, shape = Treatment)) +
  geom_point() +
  theme_Publication() +
  guides(col = guide_legend(nrow = 3, byrow = TRUE),
         shape = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = sprintf("PC1: %s%% variance", pc1_var),
       y = sprintf("PC2: %s%% variance", pc2_var)) +
  scale_color_manual(values = trial_colors)
ggsave(file = file.path(output_dir, "pca_trial_treatment.png"),
       width = 6, height = 6, dpi = 300)

pca_data %>%
  ggplot(aes(x = PC1, y = PC2, col = Trial, shape = Response)) +
  geom_point() +
  theme_Publication()  +
  guides(col = guide_legend(nrow = 3, byrow = TRUE),
         shape = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = sprintf("PC1: %s%% variance", pc1_var),
       y = sprintf("PC2: %s%% variance", pc2_var)) +
  scale_color_manual(values = trial_colors)
ggsave(file = file.path(output_dir, "pca_trial_response.png"),
       width = 6, height = 6, dpi = 300)

pca_trial_plot <- pca_data %>%
  ggplot(aes(x = PC1, y = PC2, col = Trial, shape = Response)) +
  geom_point() +
  theme_Publication()  +
  guides(col = guide_legend(nrow = 3, byrow = TRUE),
         shape = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = sprintf("PC1: %s%% variance", pc1_var),
       y = sprintf("PC2: %s%% variance", pc2_var)) +
  scale_color_manual(values = trial_colors) +
  labs(col = "") +
  theme(legend.spacing.x = unit(0.3, "cm"))

graph2pdf(x = pca_trial_plot,
          file = here(output_dir, "_pca_trial_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 6,
          width = 6)

pca_data %>% write_csv2(here(output_dir, "pca_data.csv"))
```

# ----------
# Compare signature scores between histologies
```{r}
signatures_to_plot <- c("Ayers_IFNg_expanded_signature", "TIL_score",
                        "Tumor_proliferation_rate", "EMT_signature", "B_cells")

for (current_signature in signatures_to_plot) {
  print(current_signature)
  plot <- ssgsea_results_df %>% ggplot(aes_string(x = "Trial",
                                   y = current_signature,
                                   fill = "Response")) +
    geom_boxplot() +
    theme_Publication()  +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    scale_fill_manual(values = annotation_color_list$Response) +
  facet_wrap(~Treatment, scales = "free_x") +
    labs(y = sprintf("ssGSEA\n%s", current_signature))
  print(plot)
}

```

## Heatmaps
```{r, fig.height = 14, fig.width = 14}

if (generate_heatmaps){
  
  ssgsea_results_m <- ssgsea_results_df %>%
    select(all_of(c("CF_sample", names(gene_signatures_ssgsea_l))))
  
  ssgsea_results_m <- ssgsea_results_m %>%
    column_to_rownames("CF_sample") %>%
    mutate_if(is.character, as.numeric) %>%
    as.matrix() %>%
    t()
  
  matrix_colnames <- colnames(ssgsea_results_m)
  ssgsea_results_m <- t(apply(
    ssgsea_results_m, 1, function(x) scale(x, center = TRUE, scale = TRUE)))
  
  colnames(ssgsea_results_m) <- matrix_colnames
  
  annot_df <- ssgsea_results_df %>%
    tibble::column_to_rownames("CF_sample") %>%
    select(Response, Treatment, Trial, Histology) %>%
    mutate_if(is.numeric, as.factor)
  
  mat_breaks <- quantile_breaks(ssgsea_results_m, n = 11)
  
  
  ssgsea_heatmap <- pheatmap::pheatmap(ssgsea_results_m,
    annotation_col = annot_df,
    annotation_colors = annotation_color_list,
    col = paletteer_d(palette = "RColorBrewer::RdYlBu",
                      n = length(mat_breaks) - 1, direction = -1),
    show_colnames = FALSE,
    fontsize_row = 6,
    breaks = mat_breaks,
  )
  
  save_pheatmap_pdf(x = ssgsea_heatmap,
                    filename = here(output_dir_plots, "ssgsea_pheatmap.pdf"),
                    width = 11, height = 11)
  
  annot_df <- ssgsea_results_df %>%
    tibble::column_to_rownames("CF_sample") %>%
    select(Response, Treatment, Trial, Histology) %>%
    mutate_if(is.numeric, as.factor) %>%
    filter(Treatment == "Pre") %>%
    select(-Treatment)
  
  ssgsea_heatmap_pre <- pheatmap::pheatmap(
    ssgsea_results_m[, rownames(annot_df)],
    annotation_col = annot_df,
    annotation_colors = annotation_color_list,
    col = paletteer_d(palette = "RColorBrewer::RdYlBu",
                      n = length(mat_breaks) - 1, direction = -1),
    show_colnames = FALSE,
    fontsize_row = 6,
    breaks = mat_breaks,
  )
  
  save_pheatmap_pdf(x = ssgsea_heatmap_pre,
                    filename = here(output_dir_plots,
                                    "ssgsea_pheatmap_pre.pdf"),
                    width = 11, height = 11)
  
  annot_df <- ssgsea_results_df %>%
    tibble::column_to_rownames("CF_sample") %>%
    select(Response, Treatment, Trial, Histology) %>%
    mutate_if(is.numeric, as.factor) %>%
    filter(Treatment == "Pre") %>%
    select(-Treatment)  %>%
    arrange(Trial) %>%
    arrange(Response)
  
  # Pre Response
  ssgsea_heatmap_pre <- pheatmap::pheatmap(
    ssgsea_results_m[, rownames(annot_df)],
    annotation_col = annot_df,
    annotation_colors = annotation_color_list,
    col = paletteer_d(palette = "RColorBrewer::RdYlBu",
                      n = length(mat_breaks) - 1, direction = -1),
    show_colnames = FALSE,
    fontsize_row = 6,
    breaks = mat_breaks,
    cluster_cols = FALSE
  )
  
  save_pheatmap_pdf(x = ssgsea_heatmap_pre,
                    filename = here(output_dir_plots,
                                    "ssgsea_pheatmap_pre_response.pdf"),
                    width = 12, height = 12)
  
  # Pre Trial
  annot_df <- ssgsea_results_df %>%
    tibble::column_to_rownames("CF_sample") %>%
    select(Response, Treatment, Trial, Histology) %>%
    mutate_if(is.numeric, as.factor) %>%
    filter(Treatment == "Pre") %>%
    select(-Treatment)  %>%
    arrange(Response) %>%
    arrange(Trial)
  
  ssgsea_heatmap_pre <- pheatmap::pheatmap(
    ssgsea_results_m[, rownames(annot_df)],
    annotation_col = annot_df,
    annotation_colors = annotation_color_list,
    col = paletteer_d(palette = "RColorBrewer::RdYlBu",
                      n = length(mat_breaks) - 1, direction = -1),
    show_colnames = FALSE,
    fontsize_row = 6,
    breaks = mat_breaks,
    cluster_cols = FALSE
  )
  
  save_pheatmap_pdf(x = ssgsea_heatmap_pre,
                    filename = here(output_dir_plots,
                                    "ssgsea_pheatmap_pre_trial.pdf"),
                    width = 12, height = 12)
}
```

```{r, fig.width = 12, fig.height = 16}
signature_response_ssgsea_c %>%
  filter(contrast == "Pre R x NR") %>%
  ggplot(aes(x = Trial, y = signature, fill = effect_size)) +
  geom_tile() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  labs(y = "") +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggsave(here(output_dir_plots, "response_ssgsea_pre.png"),
       height = 19, width = 8)

```

# ----------
# Meta analysis
```{r}
safe_extract <- function(x) ifelse(is.null(x), NA, x)

run_smd_meta_analysis <- function(meta_analysis_df, variable_list,
                                  output_suffix, print_plot = FALSE,
                                  save_plot = FALSE) {
  
  meta_studies_results <- data.frame(matrix(nrow = 0, ncol = 0))
  meta_combined_effects <- data.frame(matrix(nrow = 0, ncol = 0))
  current_variable <- "CXCL9"
  for (current_variable in unique(variable_list)) {
    print(current_variable)
    cat("  \n")
    
    current_contrast = "R x NR"
    
    meta_analysis_df_var <- meta_analysis_df %>%
      filter(variable == current_variable)
        
    meta_analysis_res <- metacont(n.e = meta_analysis_df_var$n_responders,
                              mean.e = meta_analysis_df_var$mean_responders,
                              sd.e = meta_analysis_df_var$sd_responders,
                              n.c = meta_analysis_df_var$n_nresponders,
                              mean.c = meta_analysis_df_var$mean_nresponders,
                              sd.c = meta_analysis_df_var$sd_nresponders,
                              studlab = meta_analysis_df_var$Trial,
                              sm = "SMD",
                              method.smd = "Hedges",
                              prediction = TRUE,
                              common = FALSE,
                              random = TRUE,
                              title = sprintf(
                                "%s %s", current_variable, current_contrast))
    
    meta_random_eff <- data.frame(
      Variable = current_variable,
      Contrast = current_contrast, 
      sm = meta_analysis_res$sm,
      level.ma = meta_analysis_res$level.ma,
      null.effect = meta_analysis_res$null.effect, 
      k = meta_analysis_res$k,
      method = meta_analysis_res$method,
      TE.random = meta_analysis_res$TE.random, # Random-effects model
      seTE.random = meta_analysis_res$seTE.random,
      statistic.random = meta_analysis_res$statistic.random,
      pval.random = meta_analysis_res$pval.random,
      method.random.ci = meta_analysis_res$method.random.ci,
      df.random = meta_analysis_res$df.random,
      lower.random = meta_analysis_res$lower.random,
      upper.random = meta_analysis_res$upper.random,
      level.predict = meta_analysis_res$level.predict,
      lower.predict = meta_analysis_res$lower.predict,
      upper.predict = meta_analysis_res$upper.predict,
      method.tau = meta_analysis_res$method.tau,
      hakn = meta_analysis_res$hakn,
      TE.fixed = meta_analysis_res$TE.fixed, # Fixed-effect model (optional)
      seTE.fixed = meta_analysis_res$seTE.fixed,
      lower.fixed = meta_analysis_res$lower.fixed,
      upper.fixed = meta_analysis_res$upper.fixed,
      pval.fixed = meta_analysis_res$pval.fixed,
      Q = meta_analysis_res$Q, # Heterogeneity
      df.Q = meta_analysis_res$df.Q,
      pval.Q = meta_analysis_res$pval.Q,
      tau = meta_analysis_res$tau,
      tau2 = meta_analysis_res$tau^2,
      H = meta_analysis_res$H,
      H.lower = safe_extract(meta_analysis_res$H.lower),
      H.upper = safe_extract(meta_analysis_res$H.upper),
      I2 = meta_analysis_res$I2,
      I2.lower = safe_extract(meta_analysis_res$I2.lower),
      I2.upper = safe_extract(meta_analysis_res$I2.upper)
    )
    
    meta_analysis_signature_results <- as.data.frame(meta_analysis_res) %>%
      mutate(Variable = current_variable,
             Contrast = current_contrast)
    
    meta_studies_results <- meta_studies_results %>%
      rbind(meta_analysis_signature_results)
    
    meta_combined_effects <- meta_combined_effects %>%
      rbind(meta_random_eff)
    
    if (print_plot) {
      print(meta_analysis_res)
      
      forest(meta_analysis_res, studlab = TRUE, common = FALSE, random = TRUE,
       digits.mean = 2, digits.sd = 2,
       label.e = "R", label.c = "NR",
       leftlabs = c("Study", "N", "Mean", "SD", "N", "Mean", "SD"),
       allstudies = TRUE)
      
    }
    
    if (save_plot) {
      save_forestplot_pdf(forest_plot_data = meta_analysis_res,
       digits.mean = 2, digits.sd = 2,
       label.e = "R", label.c = "NR",
       leftlabs = c("Study", "N", "Mean", "SD", "N", "Mean", "SD"),
       allstudies = TRUE,
       title_text = sprintf("%s R vs. NR %s", current_variable, output_suffix),
                      filename = here(
                        output_dir,
                        sprintf("meta_analysis_wes_%s_%s.pdf",
                                current_variable, output_suffix)),
                      width = 12, height = 7)
    }
    
    cat("  \n")
  }
  
  meta_studies_results %>%
    write_csv2(here(output_dir,
                    sprintf("meta_studies_results_%s.csv", output_suffix)))
  meta_combined_effects %>%
    write_csv2(here(output_dir,
                    sprintf("meta_combined_effects_%s.csv", output_suffix)))
}

```


```{r}
run_smd_meta_analysis(meta_analysis_df = signature_response_c %>%
      filter(Trial %in% c("Combined")) %>%
      filter(contrast == "Pre R x NR") %>%
        mutate(variable = signature),
      variable_list = unique(gene_signatures_meta_analysis$Signature),
      output_suffix = "combined"
      )

run_smd_meta_analysis(meta_analysis_df = signature_response_c %>%
      filter(Trial %in% c("Combined_excl_dMMR")) %>%
      filter(contrast == "Pre R x NR") %>%
        mutate(variable = signature),
      variable_list = unique(gene_signatures_meta_analysis$Signature),
      output_suffix = "combined_excl_dMMR"
      )

run_smd_meta_analysis(meta_analysis_df = signature_response_c %>%
      filter(!Trial %in% c("Combined", "Combined_excl_dMMR")) %>%
      filter(contrast == "Pre R x NR") %>%
        mutate(variable = signature),
      variable_list = unique(gene_signatures_meta_analysis$Signature),
      output_suffix = "random_effects"
      )

```

# Combine
```{r}
meta_studies_results_combined <- read_csv2(
  here(output_dir, "meta_studies_results_combined.csv"))

meta_combined_effects_combined <- read_csv2(
  here(output_dir, "meta_combined_effects_combined.csv"))

# meta_studies_results_combined_excl_dMMR <- read_csv2(
#     here(output_dir, "meta_studies_results_combined_excl_dMMR.csv"))
# 
# meta_combined_effects_combined_excl_dMMR <- read_csv2(
#     here(output_dir, "meta_combined_effects_combined_excl_dMMR.csv"))

meta_studies_results_random_effects <- read_csv2(
  here(output_dir, "meta_studies_results_random_effects.csv"))

meta_combined_effects_random_effects <- read_csv2(
  here(output_dir, "meta_combined_effects_random_effects.csv"))

meta_studies_results_rna_all <- meta_studies_results_random_effects %>%
  full_join(meta_studies_results_combined) # %>%
  # full_join(meta_studies_results_combined_excl_dMMR)

meta_studies_results_rna_all %>%
  write_csv2(here(output_dir, "meta_studies_results_smd_all.csv"))

meta_studies_results_rna_all <- meta_studies_results_rna_all %>%
  left_join(gene_signatures_meta_analysis %>%
              select(Signature, Signature_abbreviation) %>%
              setNames(c("Variable", "label")))

```

```{r}
meta_combined_effects_random_effects_f <- meta_combined_effects_random_effects %>%
  mutate(meta_FDR = p.adjust(pval.random, method = "BH")) %>%
  arrange(TE.random)

meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = pval.random, y = meta_FDR)) +
  geom_point() + theme_Publication()

meta_combined_effects_random_effects_f %>%
  filter(meta_FDR < 0.05) %>%
  ggplot(aes(x = pval.random, y = meta_FDR)) +
  geom_point() + theme_Publication()

meta_combined_effects_random_effects_f %>%
  arrange(-TE.random) %>% 
  select(Variable, pval.random, meta_FDR) %>%
  mutate(pval.random = round(pval.random, 2)) %>%
  mutate(meta_FDR = round(meta_FDR, 2))

meta_combined_effects_random_effects_f %>%
  arrange(-TE.random) %>% 
  select(Variable, pval.random, meta_FDR) %>%
  mutate(pval.random = round(pval.random, 5)) %>%
  mutate(meta_FDR = round(meta_FDR, 5)) %>% view()

```

## Heterogeneity statistics
```{r}
meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = TE.random, y = TE.fixed)) + geom_point() +
  theme_Publication() + geom_abline(slope = 1, intercept = 0)

meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = Q, y = -log10(pval.Q))) + geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = Q, y = -log10(pval.Q), label = Variable),
                  data = meta_combined_effects_random_effects_f)

meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = tau2, y = -log10(pval.Q))) + geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = tau2, y = -log10(pval.Q), label = Variable),
                  data = meta_combined_effects_random_effects_f)

meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = I2, y = -log10(pval.Q))) + geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = I2, y = -log10(pval.Q), label = Variable),
                  data = meta_combined_effects_random_effects_f)

meta_combined_effects_random_effects_f %>%
  filter(pval.random < 0.05) %>%
  ggplot(aes(x = I2, y = -log10(pval.Q))) + geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = I2, y = -log10(pval.Q), label = Variable),
                  data = meta_combined_effects_random_effects_f %>%
  filter(pval.random < 0.05))

meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = Q, y = I2)) + geom_point() + theme_Publication() +
  geom_text_repel(aes(x = Q, y = I2, label = Variable),
                  data = meta_combined_effects_random_effects_f)

meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = tau2, y = I2)) + geom_point() + theme_Publication() +
  geom_text_repel(aes(x = tau2, y = I2, label = Variable),
                  data = meta_combined_effects_random_effects_f)

```

```{r, fig.height = 11, fig.width = 9}
meta_combined_effects_random_effects %>%
  arrange(pval.random) %>%
  select(c(Variable, Contrast, TE.random, pval.random,
           lower.random, upper.random)) %>%
  mutate(pval.random = round(pval.random, 5)) %>%
  mutate_at(.vars = c("TE.random", "lower.random", "upper.random"),
            .funs = function(x){round(x, digits = 3)}) %>% view()

# reorder((Variable, TE.random))
meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = factor(
    Variable, levels = meta_combined_effects_random_effects_f$Variable),
    y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip() + geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis SMD (95% CI)") +
  guides(size = guide_legend(nrow = 3))

ggsave(here(output_dir, "meta_combined_effects_random_effects_f.png"),
    height = 4, width = 5)


meta_studies_results_rna_all %>%
  mutate(label = cut(
    pval,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  )) %>%
  ggplot(aes(
    x = factor(studlab, levels = c("Combined", "Combined_excl_dMMR", "BELLINI",
                                   "IMCISION", "MATISSE", "NABUCCO",
                                   "NICHE-MSS", "OpACIN_neo", "PRADO")), 
    y = factor(
      Variable, levels = meta_combined_effects_random_effects_f$Variable),
    fill = TE,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

ggsave(here(output_dir, "meta_combined_effects_random_effects_pval.png"),
    height = 9, width = 8)

meta_combined_effects_random_effects_f %>%
  mutate(label = cut(
    pval.Q,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor("I2"), 
    y = factor(
      Variable, levels = meta_combined_effects_random_effects_f$Variable),
    fill = I2,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

ggsave(here(output_dir, "meta_combined_effects_random_effects_het.png"),
    height = 9, width = 4.5)

heterogeneity_plot <- meta_combined_effects_random_effects_f %>%
  mutate(label = cut(
    pval.Q,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor("I2"), 
    y = factor(
      Variable, levels = meta_combined_effects_random_effects_f$Variable),
    fill = I2,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

graph2pdf(heterogeneity_plot,
          here(output_dir, "meta_combined_effects_random_effects_het.pdf"),
    height = 9, width = 4.5)
```

```{r meta analysis, fig.height = 8.5, fig.width = 12, results = "asis"}
meta_studies_results <- data.frame(matrix(nrow = 0, ncol = 19))

meta_combined_effects <- data.frame(matrix(nrow = 0, ncol = 18))

if (run_meta_analysis) {
  current_gene_signature <- gene_signatures_meta_analysis$Signature[1]
  for (current_gene_signature in gene_signatures_meta_analysis$Signature) {
    print(current_gene_signature)
    cat("  \n")
    
    current_contrast = "Pre R x NR"
    meta_analysis_df <- signature_response_c %>%
      filter(Trial != "Combined") %>%
      filter(signature == current_gene_signature &
             contrast == current_contrast)
      
    meta_analysis_res <- metacont(n.e = meta_analysis_df$n_responders,
                              mean.e = meta_analysis_df$mean_responders,
                              sd.e = meta_analysis_df$sd_responders,
                              n.c = meta_analysis_df$n_nresponders,
                              mean.c = meta_analysis_df$mean_nresponders,
                              sd.c = meta_analysis_df$sd_nresponders,
                              studlab = meta_analysis_df$Trial,
                              sm = "SMD",
                              method.smd = "Hedges",
                              prediction = TRUE,
                              common = FALSE,
                              random = TRUE,
                              title = sprintf(
                                "%s %s", current_gene_signature,
                                current_contrast))
    
    meta_random_eff <- data.frame(
      Signature = current_gene_signature,
      Contrast = current_contrast, 
      sm = meta_analysis_res$sm,
      level.ma = meta_analysis_res$level.ma,
      null.effect = meta_analysis_res$null.effect, 
      k = meta_analysis_res$k,
      method = meta_analysis_res$method,
      TE.random = meta_analysis_res$TE.random,
      seTE.random = meta_analysis_res$seTE.random,
      statistic.random = meta_analysis_res$statistic.random,
      pval.random = meta_analysis_res$pval.random,
      method.random.ci = meta_analysis_res$method.random.ci,
      df.random = meta_analysis_res$df.random,
      lower.random = meta_analysis_res$lower.random,
      upper.random = meta_analysis_res$upper.random,
      level.predict = meta_analysis_res$level.predict,
      lower.predict = meta_analysis_res$lower.predict,
      upper.predict = meta_analysis_res$upper.predict
    )
    
    meta_analysis_signature_results <- as.data.frame(meta_analysis_res) %>%
      mutate(Signature = current_gene_signature,
             Contrast = current_contrast)
    
    meta_studies_results <- meta_studies_results %>%
      rbind(meta_analysis_signature_results)
    
    meta_combined_effects <- meta_combined_effects %>%
      rbind(meta_random_eff)
    
    forest(meta_analysis_res, studlab = TRUE, common = FALSE, random = TRUE,
       digits.mean = 2, digits.sd = 2,
       label.e = "R", label.c = "NR",
       leftlabs = c("Study", "N", "Mean", "SD", "N", "Mean", "SD"),
       allstudies = TRUE)

    save_forestplot_pdf(forest_plot_data = meta_analysis_res,
                     digits.mean = 2, digits.sd = 2,
                     label.e = "R", label.c = "NR",
                     leftlabs = c("Study", "N", "Mean", "SD",
                                  "N", "Mean", "SD"),
                     allstudies = TRUE,
                      filename = here(
                        output_dir_plots,
                        sprintf("meta_analysis_pre_%s.pdf",
                                current_gene_signature)),
                      width = 12, height = 7)
    cat("  \n")
  }
    
  meta_studies_results %>%
    write_csv2(here(output_dir, "meta_studies_results.csv"))
  meta_combined_effects %>%
    write_csv2(here(output_dir, "meta_combined_effects.csv"))
}
```

```{r, fig.height = 8.5, fig.width = 12, results = "asis"}
meta_studies_results <- read_csv2(
  here(output_dir, "meta_studies_results.csv"))
meta_combined_effects <- read_csv2(
  here(output_dir, "meta_combined_effects.csv"))

meta_combined_effects <- meta_combined_effects %>%
  arrange(TE.random)
meta_combined_effects_f <- meta_combined_effects %>%
  filter(pval.random < 0.05) %>%
  arrange(TE.random)

meta_combined_effects_f %>%
  arrange(pval.random) %>%
  select(c(Signature, Contrast, TE.random, pval.random,
           lower.random, upper.random)) %>%
  mutate(pval.random = round(pval.random, 5)) %>%
  mutate_at(.vars = c("TE.random", "lower.random", "upper.random"),
            .funs = function(x){round(x, digits = 3)}) %>% view()

meta_combined_effects %>%
  ggplot(aes(x = reorder(Signature, TE.random), y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip() + geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis SMD (95% CI)")
ggsave(here(output_dir_plots, "meta_combined_effects.png"),
    height = 8, width = 8)

meta_combined_effects_f %>%
  ggplot(aes(x = reorder(Signature, TE.random), y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip() + geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis SMD (95% CI)")
ggsave(here(output_dir_plots, "meta_combined_effects_f.png"),
    height = 8, width = 8)

signature_response_c %>%
  filter(contrast == "Pre R x NR") %>%
  filter(signature %in% meta_combined_effects$Signature) %>% 
  ggplot(aes(x = factor(Trial,
                        c("Combined", "BELLINI", "IMCISION", "MATISSE",
                          "NABUCCO", "NICHE-MSS", "OpACIN_neo", "PRADO")),
             y = factor(signature,
                                   meta_combined_effects$Signature),
             fill = effect_size)) +
  geom_tile() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  labs(x = "", y = "") +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggsave(here(output_dir_plots, "response_pre.png"),
       height = 10, width = 8)

signature_response_c %>%
  filter(contrast == "Pre R x NR") %>%
  filter(signature %in% meta_combined_effects_f$Signature) %>% 
  ggplot(aes(x = factor(Trial,
                        c("Combined", "BELLINI", "IMCISION", "MATISSE",
                          "NABUCCO", "NICHE-MSS", "OpACIN_neo", "PRADO")),
             y = factor(signature,
                                   meta_combined_effects_f$Signature),
             fill = effect_size)) +
  geom_tile() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  labs(x = "", y = "") +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggsave(here(output_dir_plots, "response_pre_f.png"),
       height = 10, width = 8)

```

## Plot signature scores significant in meta_analysis
```{r}
signatures_to_plot <- meta_combined_effects_f$Signature

for (current_signature in signatures_to_plot) {
  print(current_signature)
  
  if (current_signature %in% colnames(ssgsea_results_df)) {
    plot <- ssgsea_results_df %>% ggplot(aes_string(x = "Trial",
                                     y = current_signature,
                                     fill = "Response")) +
      geom_boxplot() +
      theme_Publication()  +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      scale_fill_manual(values = c("#F8766D", "#00BA38")) +
    facet_wrap(~Treatment, scales = "free_x") +
    labs(y = sprintf("%s\nssGSEA", current_signature))
    print(plot)
  }
  
  plot <- signature_rpm_data %>% ggplot(aes_string(x = "Trial",
                                   y = current_signature,
                                   fill = "Response")) +
    geom_boxplot() +
    theme_Publication()  +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    scale_fill_manual(values = c("#F8766D", "#00BA38")) +
  facet_wrap(~Treatment, scales = "free_x") +
    labs(y = sprintf("%s\nmean log2(RPM+1)", current_signature))
  print(plot)
}

```

```{r}
# Proliferation
current_signature = "Tumor_proliferation_rate"
plot <- ssgsea_results_df_pre %>%
  filter(!Trial %in% "NICHE-MSI") %>%
  ggplot(aes_string(x = "Trial",
                    y = current_signature,
                    fill = "Response")) +
      geom_boxplot() +
      theme_Publication()  +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      scale_fill_manual(values = c("#F8766D", "#00BA38")) +
    labs(y = sprintf("Proliferation ssGSEA", current_signature), x = "") +
  theme(axis.title.x = element_blank())
print(plot)
  
graph2pdf(plot, here(output_dir, "proliferation_ssgsea.pdf"),
          width = 5, height = 4.5)

current_signature = "H_E2F_Targets"
plot <- ssgsea_results_df_pre %>%
  filter(!Trial %in% "NICHE-MSI") %>%
  ggplot(aes_string(x = "Trial",
                    y = current_signature,
                    fill = "Response")) +
      geom_boxplot() +
      theme_Publication()  +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      scale_fill_manual(values = c("#F8766D", "#00BA38")) +
    labs(y = sprintf("E2F TargetsH ssGSEA", current_signature), x = "") +
  theme(axis.title.x = element_blank())
print(plot)
  
graph2pdf(plot, here(output_dir, "H_E2F_Targets_ssgsea.pdf"),
          width = 5, height = 4.5)

# Wnt B catenin
current_signature = "B_catenin_signature_Luke_et_al"
plot <- ssgsea_results_df_pre %>%
  filter(!Trial %in% "NICHE-MSI") %>%
  ggplot(aes_string(x = "Trial",
                    y = current_signature,
                    fill = "Response")) +
      geom_boxplot() +
      theme_Publication()  +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      scale_fill_manual(values = c("#F8766D", "#00BA38"))+
    labs(y = sprintf("B catenin signature ssGSEA", current_signature), x = "") +
  theme(axis.title.x = element_blank())
print(plot)

# CXCL9

graph2pdf(plot, here(output_dir, "B_catenin_ssgsea.pdf"),
          width = 5, height = 4.5)
```

# ----------
# Lint script
```{r}
# lint("~/NKI_pancancer_neoadj_io/src/gene_signatures_meta_MPR.Rmd")
lint(filename = "~/NKI_pancancer_neoadj_io/src/gene_signatures_meta_MPR.Rmd",
  linters = unused_import_linter()
)
```
# ----------
# sessionInfo - Package Versions
```{r sessionInfo}
sessionInfo()
```
