---
title: "NICHE response prediction"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  export: TRUE
  output_dir: "response_prediction"
  metadata: "combined_rna_metadata.csv"
  analyze_specific_trial: ""
---

Author comment:
File description comment:
The purpose of this script is to generate random forest models and linear signature models to check predictions for nICB response

# Loading Packages
```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# Install if necessary install.packages("BiocManager")
# Install if necessary BiocManager::install("DESeq2")
library(tidyverse)
library(readxl)
library(ggrepel)
library(here)
library(GSVA)
library(DESeq2)
library(paletteer)
library(ggpubr)
library(rstatix)
source("helper_functions.R")
library(ROCit)
library(corrplot)
library(caret)
library(glmnet)
library(ggpubr)
library(pROC)
library(plotly)
library(vip)
library(lme4)

set.seed(42)
select <- dplyr::select
rename <- dplyr::rename
here <- here::here
```

# ---------------

# Function definitions

# Annotation color
```{r}

# Define colours for each group
Color_NR <- "#F8766D"
Color_R <- "#00BA38" # "#619CFF"
Color_NA <- "#CD9600"
Color_Pre <- "#619CFF"
Color_Post <- "#F564E3"

Color_HNSCC <- "#009292"
Color_CSCC <- "#CD9600"
Color_Urothelial <- "#B66DFF"
Color_LN_melanoma <- "#00BFC4"
Color_Colon <- "#F8766D"
Color_Colon_MSI <- "#B73A3A"
Color_Colon_MSS <- "#23C26F"
Color_TNBC <- "#FEB5DA"
Color_BELLINI <- "#FEB5DA"
Color_IMCISION <- "#009292"
Color_MATISSE <- "#CD9600"
Color_NABUCCO <- "#B66DFF"
Color_BELLINI <- "#FEB5DA"
Color_NICHE_MSS <- "#23C26F"
Color_NICHE_MSI <- "#B73A3A"
Color_OpACIN_neo <- "#00BFC4"
Color_PRADO <- "#619CFF"


cosmic_colors <- c(Aflatoxin = "#d0e0e3", Aging = "#3d85c6", APOBEC = "#674ea7",
                   Aristolochic_acid = "#d9d2e9", dMMR = "#B73A3A",
                   HRD = "#e69138", Other = "#cccccc", POL1 = "#cc0000",
                   POLE_POLD1 = "#5b0f00", POLH = "#FEB5DA", ROS = "#6DB6FF",
                   SBS5 = "#ea9999", Smoking = "#38761d", TMZ = "#b5a7d6",
                   Tobacco_chewing = "#6aa84f", UV = "#f1c232")

annotation_color_list = list(
  Response = c(NR = Color_NR, R = Color_R), #R_NR = "#CD9600"
  Treatment = c(Pre = Color_Pre, Post = Color_Post),
  Histology = c(LN_melanoma = Color_LN_melanoma, Urothelial = Color_Urothelial,
                Colon = Color_Colon, HNSCC = Color_HNSCC, TBNC = Color_TNBC,
                cSCC = Color_CSCC),
  Trial = c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO))

trial_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_colors_combined <- c(Combined = "#999999", BELLINI = Color_BELLINI,
                           IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_Colon_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_label_dict <- data.frame(
  Trial = c("Combined", "BELLINI", "IMCISION", "MATISSE", "NABUCCO",
            "NICHE-MSS", "NICHE-MSI", "OpACIN_neo", "PRADO"),
  Trial_label = c("Pooled", "BELLINI", "IMCISION", "MATISSE", "NABUCCO",
                  "NICHE pMMR", "NICHE dMMR", "OpACIN neo", "PRADO")) %>%
  unique()

trial_label_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE pMMR` = Color_NICHE_MSS, `NICHE dMMR` = Color_NICHE_MSI,
            `OpACIN neo` = Color_OpACIN_neo, PRADO = Color_PRADO)

histology_msi_mss_colors <- c(`Colon-MSI`= Color_Colon_MSI,
                              `Colon-MSS`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              LN_melanoma = Color_LN_melanoma,
                              TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)

histology_msi_mss_label_dict <- data.frame(
  Histology_MSI_MSS = c("TBNC", "Colon-MSS", "HNSCC", "Urothelial",
                        "LN_melanoma", "LN_melanoma", "CSCC", "Colon-MSI"),
  histology_msi_mss_label = c("TBNC", "Colon pMMR", "HNSCC", "Urothelial",
                              "Melanoma", "Melanoma", "CSCC", "Colon dMMR")) %>%
  unique()

histology_msi_mss_label_colors <- c(`Colon dMMR`= Color_Colon_MSI,
                                    `Colon pMMR`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              Melanoma = Color_LN_melanoma, TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)
  
```


# ---------------
# Setup output folder
```{r, echo = FALSE, include = FALSE}
export <- params$export

# generate output directory
output_dir <- here::here("data", "processed", params$output_dir)
print(output_dir)
output_dir_plots <- here(output_dir, "plots")
prediction_dir <- output_dir

# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)
    dir.create(output_dir_plots)
} else {
  print("Output folder already exists")
}

```

```{r}
wes_meta_dir <- here("data", "processed", "WES_analysis_meta")
wes_dir <- here("data", "processed", "WES_analysis")
wes_rna_dir <- here("data", "processed", "WES_RNA_integration")
signature_dir <- here("data", "processed", "gene_signatures")
meta_rna_dir <- here("data", "processed", "gene_signatures_meta") 

```

# ---------------

# WES RNA
```{r}

wes_features_df <- read_csv2(here(wes_dir, "wes_features_df.csv"))
wes_features_df <- wes_features_df %>%
  filter(!(Trial %in% "NICHE-MSI" & Tumor_nr == "Tumor_2"))

wes_rna_features_df <- read_csv2(here(wes_rna_dir, "wes_rna_features_df.csv"))

wes_rna_features_df <- wes_rna_features_df %>%
  mutate(Response_n = ifelse(
    Response == "R",
    1, ifelse(
      Response == "NR", 0, NA))) %>%
  mutate(TMB_scaled = scale(TMB),
         UV_scaled = scale(UV_abs),
         APOBEC_scaled = scale(APOBEC_abs),
         TIL_score_scaled = scale(TIL_score),
         Proliferation_scaled = scale(Tumor_proliferation_rate),
         B_cat_scaled = scale(B_catenin_signature_Luke_et_al))

# Add patient N stage and sex information
wes_rna_features_df <- wes_rna_features_df %>%
  left_join(patient_metadata_combined %>%
              select(Patient, Sex, Baseline_N_h)) %>%
  mutate(Baseline_N_h = ifelse(Baseline_N_h == "Nx", "N+", Baseline_N_h)) %>%
  mutate(Cohort = ifelse(Cohort %in% c("Nivo_ipi_A", "Nivo_ipi_B", "Nivo_ipi_C"),
                         "Nivo_ipi", Cohort))

table(wes_rna_features_df$Cohort)

wes_rna_features_df_pre <- wes_rna_features_df %>%
  filter(Treatment == "Pre")

wes_rna_features_df_pre_f <- wes_rna_features_df_pre %>% filter(!is.na(TMB)) 
wes_rna_features_df_pre_f %>% dim()

table(gene_signatures_meta_analysis$Signature %in% colnames(wes_rna_features_df))

# Select Meta analysis predictor features
wes_rna_meta_features_df <- wes_rna_features_df %>%
  filter(Treatment == "Pre") %>%
  mutate(BRCA1_2 = BRCA1 | BRCA2) %>%
  mutate(HLA_alteration = ifelse(
    HLA_grouped %in% c("HLA-LOH", "HLA-DEL"), "HLA-Altered", HLA_grouped)) %>%
  select(Patient, Trial, Response, Sex, Baseline_N_h,
         gene_signatures_meta_analysis$Signature,
         TMB, APOBEC_abs, UV_abs, dMMR_abs, HRD_abs, Ploidy, Aneuploidy_score,
         scna_loss, "GIE_AP", "TP53", "BRCA1_2", HLA_alteration)  %>%
  left_join(
    wes_features_df %>%
      mutate(Purity = ifelse(is.na(Purity), Imputed_purity, Purity)) %>%
      select(Patient, Purity, TMB_frameshift, wgII_0.6_bases, "SIGPW_Wnt_B_Cat",
  "SIGPW_PI3K_signaling", "SIGPW_Cell_cycle",
  "SIGPW_TGFB_signaling", "GIE_IFNy", "RAS_RAF",
  pvactools_filtered_neoantigen_load, pyclone_clonal_TMB,
  pyclone_subclonal_TMB)
  ) %>%
  filter(!is.na(TMB))
```


# Load RNAseq data
```{r load rnaseq data, include = TRUE}
df_counts <- read_csv2(here("data", "processed", "rnaseq_data",
                             "combined_rnaseq_counts_p.csv"))
df_counts <- df_counts %>% column_to_rownames("gene")

rpm_counts <- read_csv2(here("data", "processed", "rnaseq_data",
                             "combined_rnaseq_rpm.csv"))
rpm_counts <- rpm_counts %>% column_to_rownames("gene")
 
metadata <- read_csv2(
  here("data", "processed", "metadata", params$metadata))

if (analyze_specific_trial != ""){
  metadata <- metadata %>% filter(Trial == analyze_specific_trial)
}

metadata <- metadata %>%
  mutate(
      Treatment = factor(Treatment, levels = c("Pre", "Post")),
      CF_sample = as.factor(CF_sample),
      Trial = as.factor(Trial),
      Response = as.factor(Response),
      Response_detail = as.factor(Response_detail),
      MSI_MSS = as.factor(MSI_MSS),
      BRAF = as.factor(BRAF))

metadata %>% mutate_if(is.character, as.factor) %>% summary()
metadata_pre <- metadata %>% filter(Treatment == "Pre")

df_counts <- df_counts[, colnames(df_counts) %in%
                          metadata_pre$CF_sample]
df_counts <- df_counts[rowSums(df_counts) != 0, ]


rpm_counts <- rpm_counts[, colnames(rpm_counts) %in%
                          metadata_pre$CF_sample]
rpm_counts <- rpm_counts[rowSums(rpm_counts) != 0, ]

ncol(df_counts)
  
unique_patient_id <- metadata_pre %>% select(Trial, Patient) %>% unique()
table(unique_patient_id$Trial)

metadata_pre$CF_sample[!metadata_pre$CF_sample %in% colnames(df_counts)]

ssgsea_results  <- read_csv2(here(signature_dir, "ssgsea_results.csv"))
ssgsea_results_df <- ssgsea_results %>%
  column_to_rownames("Signature") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("CF_sample") %>%
  left_join(metadata)
```

```{r}
patient_metadata_combined <- read_csv2(
  here("data", "processed", "metadata", "patient_metadata_combined.csv"))

patient_metadata_combined_model <- patient_metadata_combined %>%
  mutate(Response_n = ifelse(
    Response == "R",
    1, ifelse(
      Response == "NR", 0, NA)))%>%
  mutate(Baseline_N_h = ifelse(Baseline_N_h == "Nx", "N+", Baseline_N_h)) %>%
  mutate(Cohort = ifelse(
    Cohort %in% c("Nivo_ipi_A", "Nivo_ipi_B", "Nivo_ipi_C"),
    "Nivo_ipi", Cohort))

table(patient_metadata_combined_model$Baseline_N_h)
table(patient_metadata_combined_model$Cohort)

```

# Additional ssGSEA sig - pancancer signature
```{r}
differential_expression_meta_dir <- here("data", "processed",
                                         "differential_expression_meta")

pancan_response_signature_1.25 <- read_csv2(
  here(differential_expression_meta_dir, "pancan_response_signature_1.25.csv"))

pancan_nonresponse_signature_1.25 <- read_csv2(
  here(differential_expression_meta_dir,
       "pancan_nonresponse_signature_1.25.csv"))

pancan_response_signatures <- pancan_response_signature_1.25 %>%
              select(gene) %>%
              setNames("Gene") %>%
              mutate(Signature = "pancancer_response_sig_1.25") %>%
  full_join(pancan_nonresponse_signature_1.25 %>%
              select(gene) %>%
              setNames("Gene") %>%
              mutate(Signature = "pancancer_nonresponse_sig_1.25")
            ) %>%
  group_by(Signature) %>%
  add_count(name = "Geneset_size") %>%
  ungroup()
```

# Load gene sets
```{r}
signature_overview <- read_csv2(
  here("data", "external", "signatures",
       "_overview_of_signatures_20260624.csv")) 

signature_overview$Signature <- gsub(
  "B_catenin_signature_pancan_22_Luke_et_al",
  "B_catenin_signature_Luke_et_al", signature_overview$Signature)

gene_signatures_meta_analysis <- signature_overview %>%
  filter(Meta_analysis == "Yes")

gene_signatures_meta_analysis <- gene_signatures_meta_analysis %>%
  full_join(pancan_response_signatures)

gene_signatures_l <- split(
  gene_signatures_meta_analysis$Gene, gene_signatures_meta_analysis$Signature)
gene_signatures_l <- gene_signatures_l[
  gene_signatures_meta_analysis$Signature %>% unique()]

gene_features <- read.csv2(
  file = here("data", "external", "biomaRt_gene_features.csv"))

gene_descriptions <- gene_features %>%
              select(hgnc_symbol, description) %>%
              rename("gene" = "hgnc_symbol") %>%
              unique()

# RPM expression
signature_rpm_data <- read_csv2(
  here("data", "processed", "gene_signatures", "signature_rpm_data.csv"))
colnames(signature_rpm_data)
```

# ---------------
# Signature ssGSEA
```{r}
df_counts <- df_counts[, colnames(df_counts) %in%
                          metadata_pre$CF_sample]

dds <- DESeqDataSetFromMatrix(
  countData = df_counts,
  colData = metadata %>% filter(CF_sample %in% colnames(df_counts)),
  design = ~1
)

# Normalize and transform the data in the `DESeqDataSet` object using `vst()`
dds_norm <- vst(dds)

vst_data <- assay(dds_norm)

pancan_response_ssgsea_results <- gsva(
  vst_data,
  gene_signatures_l,
  method = "ssgsea",
  # Appropriate for vst transformed data
  kcdf = "Gaussian",
  # Minimum gene set size
  min.sz = 1,
  # Compute Gaussian-distributed scores
  mx.diff = TRUE,
  # Don't print out the progress bar
  verbose = FALSE
)
```

```{r, fig.width = 4, fig.height = 4}
pancan_response_ssgsea_results_df <- pancan_response_ssgsea_results %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("CF_sample") %>%
  left_join(metadata, by = "CF_sample") %>%
  mutate(pancancer_response_score =
           pancancer_response_sig_1.25 - pancancer_nonresponse_sig_1.25)

pancan_response_ssgsea_results_df %>%
  write_csv2(here(output_dir, "pancan_response_ssgsea_results_df.csv"))

```

```{r, fig.width = 4, fig.height = 4}

pancan_response_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_sig_1.25, fill = Response)) +
  geom_boxplot() + stat_compare_means()+ theme_Publication()+
  scale_fill_manual(values = c(Color_NR, Color_R))

pancan_response_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_nonresponse_sig_1.25,
             fill = Response)) +
  geom_boxplot() + stat_compare_means() + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R))

pancan_response_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + stat_compare_means() + theme_Publication()+
  scale_fill_manual(values = c(Color_NR, Color_R))

```

```{r, fig.height = 12, fig.width = 12}

pancan_response_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_sig_1.25, fill = Response)) +
  geom_boxplot() +
  stat_compare_means() + facet_wrap(~Trial) + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R))

pancan_response_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_nonresponse_sig_1.25,
             fill = Response)) +
  geom_boxplot() +
  stat_compare_means() + facet_wrap(~Trial) + theme_Publication()+
  scale_fill_manual(values = c(Color_NR, Color_R))

pancan_response_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + facet_wrap(~Trial) + stat_compare_means() +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R))
```

```{r, fig.height = 24, fig.width = 24}
ssgsea_results_df %>%
  filter(Treatment == "Pre") %>%
  left_join(pancan_response_ssgsea_results_df) %>%
  select(all_of(c(ssgsea_results$Signature,
                  "pancancer_response_sig_1.25",
                  "pancancer_nonresponse_sig_1.25",
                  "pancancer_response_score"))) %>%
  as.matrix()%>%
  cor() %>%
  corrplot(order = "AOE")

```

# ---------------
# Prediction Models single signatures

```{r}

include_prediction <- c(gene_signatures_meta_analysis$Signature,
         "pancancer_response_sig_1.25", "pancancer_nonresponse_sig_1.25", 
         "pancancer_response_score")
include_prediction
length(include_prediction)

sample_names <- rownames(prediction_data_m)

prediction_data <- ssgsea_results_df %>%
  filter(Treatment == "Pre") %>%
  left_join(pancan_response_ssgsea_results_df)%>%
  left_join(signature_rpm_data %>%
              select("CF_sample", "TCF1", "PD_1",
                     "PD_L1", "CXCL9", "CXCL13"))

prediction_data <- prediction_data %>%
  select(all_of(
    c(gene_signatures_meta_analysis$Signature[
      gene_signatures_meta_analysis$Signature %in%
        colnames(prediction_data)],
      "pancancer_response_sig_1.25", "pancancer_nonresponse_sig_1.25",
      "pancancer_response_score", "CF_sample")))

dim(prediction_data)

z_norm <- t(apply(
  prediction_data %>%
  column_to_rownames("CF_sample") %>%
  # select(include_prediction) %>%
  as.matrix(), 2,
  function(x) scale(x, center = TRUE, scale = TRUE)))

colnames(z_norm) <- sample_names

signature_z_norm_df <- z_norm %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("CF_sample") %>%
  left_join(metadata) %>%
  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
  select(-Cohort) %>%
  left_join(patient_metadata_combined_model %>%
              select(Patient, Baseline_N_h, Sex, Cohort))

table(signature_z_norm_df$Response, signature_z_norm_df$Response_n)

mat_breaks <- quantile_breaks(as.matrix(z_norm), n = 11)

pathway_heatmap <- pheatmap::pheatmap(z_norm,
  annotation_col = annot_df,
  show_colnames = FALSE, # Don't show sample labels
  fontsize_row = 6,
  breaks = mat_breaks,
  col = paletteer_d(palette = "RColorBrewer::RdYlBu",
                    n = length(mat_breaks) - 1, direction = -1)
)
```

```{r, fig.height = 8, fig.width = 8}

for (current_gene_signature in rownames(z_norm)) {

  cat(current_gene_signature, "  \n")
  cat(gene_signatures_l[current_gene_signature][[1]], "  \n")

  
  plot <- signature_z_norm_df %>% ggplot(
    aes_string(x = "Response", y = current_gene_signature, col = "Response")
    ) + geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15),
             aes(group = Response), size = 2.5) +
    labs(col = "Response") +
    expand_limits(y = 0) +
    labs(x = "", y = sprintf("%s\n Z-score ssGSEA, GSVA or Weighted Mean",
                             current_gene_signature),
                             title = signature_direction) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    theme_Publication() +
    theme(legend.spacing.x = unit(0.3, "cm")) +
    facet_wrap(~Trial, scales = "free")

  print(plot)
}

```

# ---------------
# Signature AUC
```{r}

get_signature_auc_simple_covariate_models <- FALSE

if (get_signature_auc_simple_covariate_models) {
    
  signature_prediction_data <- data.frame(matrix(nrow = 0, ncol = 7)) %>%
      setNames(c("Signature", "Intercept", "Estimate", "Intercept_pval",
                 "Estimate_pval", "AUC", "Predicted_direction"))
  
  signature_mixed_effects_covariate_prediction_data <- data.frame(
    matrix(nrow = 0, ncol = 7)) %>%
      setNames(c("Signature", "Intercept", "Estimate", "Intercept_pval",
                 "Estimate_pval", "AUC", "Predicted_direction"))
  
  dim(signature_z_norm_df)
  
  for (current_gene_signature in rownames(z_norm)) {
  
    cat(current_gene_signature, "  \n")
    cat(gene_signatures_l[current_gene_signature][[1]], "  \n")
  
    plot <- signature_z_norm_df %>% ggplot(
      aes_string(x = "Response", y = current_gene_signature, col = "Response")
      ) + geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.15),
               aes(group = Response, shape = MSI_MSS), size = 2.5) +
      labs(col = "Response") +
      expand_limits(y = 0) +
      labs(x = "", y = sprintf("%s\n Z-score ssGSEA",
                               current_gene_signature)) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      theme_Publication() +
      theme(legend.spacing.x = unit(0.3, "cm"))
  
    print(plot)
  
    current_formula <- sprintf("Response_n ~ %s", current_gene_signature)
   
    model <- glm(as.formula(current_formula), family = "binomial",
               data = signature_z_norm_df)
  
    print(summary(model))
  
    measure <- measureit(score =  model$fitted.values, class = model$y,
                         measure = c("ACC", "SENS", "SPEC", "FSCR"))
  
    predicted <- predict(model, signature_z_norm_df, type = "response")
  
    roc_empirical <- rocit(score = model$fitted.values, class = model$y)
    print(summary(roc_empirical))
  
    AUC <- summary(roc_empirical)[4, 1, drop = TRUE]
    AUC <- str_split(AUC, ":")[[1]][2]
    print(plot(roc_empirical, values = FALSE))
  
    signature_prediction_model <- broom::tidy(model) %>%
    select(term, estimate, p.value) %>%
    mutate(Signature = current_gene_signature) %>%
    pivot_wider(values_from = c(estimate, p.value), names_from = term) %>%
    setNames(c("Signature", "Intercept", "Estimate",
               "Intercept_pval", "Estimate_pval")) %>%
    mutate(AUC = AUC, Predicted_direction = ifelse(
      Estimate > 0, "ICB_sensitivity", "ICB_resistance"))
  
    signature_prediction_data <- signature_prediction_data %>%
      rbind(signature_prediction_model)
    
    print("Covariate model")
    
    # Glmer
    current_formula_covariates <- sprintf(
      "Response_n ~ %s + Sex + Baseline_N_h + Cohort + (1 | Trial)",
      current_gene_signature)
    
    mixed_model <- glmer(as.formula(current_formula_covariates),
                         family = "binomial", data = signature_z_norm_df)
  
    print(summary(mixed_model))
    coefs <- fixef(mixed_model)
    coef_table <-summary(mixed_model)$coefficients
    
    intercept_est  <- coef_table["(Intercept)", "Estimate"]
    intercept_pval <- coef_table["(Intercept)", "Pr(>|z|)"]
  
    signature_est  <- coef_table[current_gene_signature, "Estimate"]
    signature_pval <- coef_table[current_gene_signature, "Pr(>|z|)"]
    
    predicted <- predict(mixed_model, signature_z_norm_df, type = "response")
  
    plot <- data.frame(true_class, fitted_vals) %>%
      ggplot(aes(x = as.factor(true_class), y = fitted_vals)) + geom_boxplot() +
      labs(title = sprintf("Covariate model %s", current_gene_signature)) +
      theme_Publication()
    
    print(plot)
    fitted_vals <- fitted(mixed_model)
    true_class <- mixed_model@resp$y
    
    roc_empirical <- rocit(score = fitted_vals, class = true_class)
    print(summary(roc_empirical))
    
    AUC <- summary(roc_empirical)[4, 1, drop = TRUE]
    AUC <- str_split(AUC, ":")[[1]][2]
    print(plot(roc_empirical, values = FALSE))
  
    signature_prediction_mixed_model_covariates <- tibble(
      Signature = current_gene_signature,
      Intercept = intercept_est,
      Estimate = signature_est,
      Intercept_pval = intercept_pval,
      Estimate_pval = signature_pval,
      AUC = AUC,
      Predicted_direction = ifelse(
        signature_est > 0,
        "ICB_sensitivity",
        "ICB_resistance"))
    
    signature_mixed_effects_covariate_prediction_data <- signature_mixed_effects_covariate_prediction_data %>%
      rbind(signature_prediction_mixed_model_covariates)
  
  }
  
  signature_prediction_data <- signature_prediction_data %>%
    select(-Predicted_direction) %>%
    unique()
  signature_prediction_data
  
  signature_prediction_data %>%
    write_csv2(here(prediction_dir, "signature_prediction_data_with_msi.csv"))
  
  signature_mixed_effects_covariate_prediction_data %>%
    write_csv2(here(prediction_dir,
                    "signature_mixed_effects_covariate_prediction_data.csv"))
  
  signature_prediction_data <- read_csv2(
    here(output_dir, "signature_prediction_data_with_msi.csv"))

}

signature_prediction_data <- read_csv2(
  here(prediction_dir, "signature_prediction_data_with_msi.csv"))

signature_mixed_effects_covariate_prediction_data <- read_csv2(
  here(prediction_dir, "signature_mixed_effects_covariate_prediction_data.csv"))

```


# Plot AUC
```{r, fig.height = 8, fig.width = 8}
signature_prediction_data %>%
  ggplot(aes(x = reorder(Signature, Estimate), y = Estimate)) +
  geom_col() +
  expand_limits(y = 0) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(x = "") +
  coord_flip()

signature_prediction_data %>%
  mutate(AUC = as.numeric(AUC))  %>%
  ggplot(aes(x = reorder(Signature, AUC), y = AUC)) +
  geom_col() +
  expand_limits(y = 0) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  geom_hline(yintercept = 0.5, col = "red", linetype = "dashed") +
  labs(x = "") +
  coord_flip() +
  scale_y_continuous(breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7))

signature_prediction_data %>%
  filter(!Signature %in% c("pancancer_response_sig_1.25",
                           "pancancer_nonresponse_sig_1.25")) %>%
  mutate(AUC = as.numeric(AUC))  %>%
  ggplot(aes(x = reorder(Signature, AUC), y = AUC)) +
  geom_col() +
  expand_limits(y = 0) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  geom_hline(yintercept = 0.5, col = "red", linetype = "dashed") +
  labs(x = "") +
  coord_flip() +
  scale_y_continuous(breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7))

signature_mixed_effects_covariate_prediction_data %>%
  ggplot(aes(x = reorder(Signature, Estimate), y = Estimate)) +
  geom_col() +
  expand_limits(y = 0) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(x = "") +
  coord_flip()

signature_mixed_effects_covariate_prediction_data %>%
  mutate(AUC = as.numeric(AUC))  %>%
  ggplot(aes(x = reorder(Signature, AUC), y = AUC)) +
  geom_col() +
  expand_limits(y = 0) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  geom_hline(yintercept = 0.5, col = "red", linetype = "dashed") +
  labs(x = "") +
  coord_flip() +
  scale_y_continuous(breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9))

signature_mixed_effects_covariate_prediction_data %>%
  filter(!Signature %in% c("pancancer_response_sig_1.25",
                           "pancancer_nonresponse_sig_1.25")) %>%
  mutate(AUC = as.numeric(AUC))  %>%
  ggplot(aes(x = reorder(Signature, AUC), y = AUC)) +
  geom_col() +
  expand_limits(y = 0) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  geom_hline(yintercept = 0.5, col = "red", linetype = "dashed") +
  labs(x = "") +
  coord_flip() +
  scale_y_continuous(breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9))
```

# ---------------
# glmer model pancan signature + multivariate combinations
```{r}
prediction_data_df <- prediction_data %>%
  left_join(metadata) %>%
  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
  select(-Cohort) %>%
  left_join(patient_metadata_combined_model %>%
              select(Patient, Baseline_N_h, Sex, Cohort))  %>%
  left_join(wes_features_df %>% select(Patient, TMB)) %>%
  mutate(TMB_scaled = scale(TMB),
         TIL_score_scaled = scale(TIL_score),
         Proliferation_scaled = scale(Tumor_proliferation_rate),
         B_cat_scaled = scale(B_catenin_signature_Luke_et_al),
         pancancer_response_score_scaled = scale(pancancer_response_score))

# TMB * TIL score model
model14 <- glmer(Response_n ~ TMB_scaled * TIL_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = prediction_data_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
                  filter(!is.na(TMB)),
                family = binomial)

model15 <- glmer(Response_n ~ TMB_scaled * TIL_score_scaled + pancancer_response_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = prediction_data_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
                  filter(!is.na(TMB)),
                family = binomial)

summary(model14)
summary(model15)
anova(model14, model15)

# Proliferation * TIL score model
model17 <- glmer(Response_n ~ Proliferation_scaled * TIL_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = prediction_data_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
                  filter(!is.na(TMB)),
                family = binomial)

model18 <- glmer(Response_n ~ Proliferation_scaled * TIL_score_scaled + pancancer_response_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = prediction_data_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
                  filter(!is.na(TMB)),
                family = binomial)

summary(model17)
summary(model18)
anova(model17, model18)

model19 <- glmer(Response_n ~ TIL_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = prediction_data_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
                  filter(!is.na(TMB)),
                family = binomial)

summary(model19)
model20 <- glmer(Response_n ~ pancancer_response_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = prediction_data_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
                  filter(!is.na(TMB)),
                family = binomial)
summary(model20)
anova(model19, model20)

prediction_data_df %>%
  ggplot(aes(x = TIL_score, y = pancancer_response_score, color = Trial)) +
  geom_point() + stat_cor()

prediction_data_df %>%
  ggplot(aes(x = TIL_score, y = pancancer_response_score, color = Trial)) +
  geom_point() + stat_cor(method = "spearman") +
  scale_color_manual(values = trial_colors)
```


```{r}
get_multivariate_auc_models <- FALSE

if (get_multivariate_auc_models) {
  model_auc_data <- data.frame(
    matrix(nrow = 0, ncol = 2))
  mixed_effects_covariate_auc_data <- data.frame(
    matrix(nrow = 0, ncol = 7))
  # no_mixed_effects_auc_data <- data.frame(
  #   matrix(nrow = 0, ncol = 7))
  model_predictions_combined <- data.frame(
    matrix(nrow = 0, ncol = 2))
  
  model_formulas <- c(
    "Response_n ~ TMB_scaled",
    "Response_n ~ TIL_score_scaled",
    "Response_n ~ Proliferation_scaled",
    "Response_n ~ TMB_scaled * TIL_score_scaled",
    "Response_n ~ TMB_scaled * TIL_score_scaled + pancancer_response_score_scaled",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled + pancancer_response_score_scaled",
    "Response_n ~ pancancer_response_score_scaled")
    
  for (current_formula in model_formulas) {
    
    print(current_formula)
    model <- glm(as.formula(current_formula), family = "binomial",
               data = prediction_data_df %>% column_to_rownames("CF_sample"))
  
    print(summary(model))
  
    measure <- measureit(score =  model$fitted.values, class = model$y,
                         measure = c("ACC", "SENS", "SPEC", "FSCR"))

    predicted <- predict(model, prediction_data_df %>% column_to_rownames("CF_sample"), type = "response")
  
    roc_empirical <- rocit(score = model$fitted.values, class = model$y)
    print(summary(roc_empirical))
  
    AUC <- summary(roc_empirical)[4, 1, drop = TRUE]
    AUC <- str_split(AUC, ":")[[1]][2]
    print(plot(roc_empirical, values = FALSE))
  
    model_auc <- tibble(
      formula = current_formula,
      AUC = AUC,
      sensitivity = sum(model$fitted.values >= 0.5 & model$y == 1) / sum(model$y == 1), # Evaluated at 0.5, not youden optimal point
      specificity = sum(model$fitted.values < 0.5 & model$y == 0) / sum(model$y == 0)) # Evaluated at 0.5, not youden optimal point
    
    model_predictions <- tibble(
      formula = current_formula,
      CF_sample = prediction_data_df$CF_sample,
      Predictions = predicted)
    
    model_predictions <- as.data.frame(predicted) %>%
      setNames("Predictions") %>%
      rownames_to_column("CF_sample") %>%
      mutate(formula = current_formula)
    
    model_auc_data <- model_auc_data %>%
      rbind(model_auc)
    
    model_predictions_combined <- model_predictions_combined %>%
      rbind(model_predictions)
    }

  ranef_model_formulas <- c(
    "Response_n ~ Sex + Baseline_N_h + Cohort + (1 | Trial)",
    "Response_n ~ TMB_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial)",
    "Response_n ~ TIL_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial)",
    "Response_n ~ Proliferation_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial)",
    "Response_n ~ TMB_scaled * TIL_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial)",
    "Response_n ~ TMB_scaled * TIL_score_scaled + pancancer_response_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial)",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial)",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled + pancancer_response_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial)",
    "Response_n ~ pancancer_response_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial)")
    
  for (current_formula_covariates in ranef_model_formulas) {
    
    print(current_formula_covariates)

    # Glmer
    mixed_model <- glmer(as.formula(current_formula_covariates),
                         family = "binomial",
               data = prediction_data_df %>% column_to_rownames("CF_sample"))
  
    print(summary(mixed_model))
    coefs <- fixef(mixed_model)
    coef_table <-summary(mixed_model)$coefficients
    print(coef_table)
    
    if(str_detect(current_formula_covariates, "\\|")) {
      print(ranef(mixed_model)) 
    }
    
    predicted <- predict(mixed_model, prediction_data_df %>%
                           column_to_rownames("CF_sample"), type = "response")
  
    plot <- data.frame(true_class, fitted_vals) %>%
      ggplot(aes(x = as.factor(true_class), y = fitted_vals)) + geom_boxplot() +
      labs(title = sprintf("Covariate model %s", current_gene_signature)) +
      theme_Publication()
    
    print(plot)
    fitted_vals <- fitted(mixed_model)
    pred_class <- as.integer(fitted_vals >= 0.5)
    true_class <- mixed_model@resp$y
    
    roc_empirical <- rocit(score = fitted_vals, class = true_class)
    print(summary(roc_empirical))
    
    AUC <- summary(roc_empirical)[4, 1, drop = TRUE]
    AUC <- str_split(AUC, ":")[[1]][2]
    print(plot(roc_empirical, values = FALSE))
  
    as.integer(fitted_vals >= 0.5)

    mixed_model_auc <- tibble(
      formula = current_formula_covariates,
      AUC = AUC,
      sensitivity = sum(pred_class == 1 & true_class == 1) / sum(true_class == 1), # Evaluated at 0.5, not youden optimal point
      specificity = sum(pred_class == 0 & true_class == 0) / sum(true_class == 0) # Evaluated at 0.5, not youden optimal point)
      ) 

    model_predictions <- as.data.frame(fitted_vals) %>%
      setNames("Predictions") %>%
      rownames_to_column("CF_sample") %>%
      mutate(formula = current_formula_covariates)
    
    mixed_effects_covariate_auc_data <- mixed_effects_covariate_auc_data %>%
      rbind(mixed_model_auc)
  
    model_auc_data <- model_auc_data %>%
      rbind(mixed_model_auc)
    
    model_predictions_combined <- model_predictions_combined %>%
      rbind(model_predictions)
  }
  
  model_auc_data %>%
    write_csv2(here(prediction_dir, "model_auc_data.csv"))
  
  model_predictions_combined %>%
    write_csv2(here(prediction_dir, "model_predictions_combined.csv"))

}

model_auc_data <- read_csv2(
  here(prediction_dir, "model_auc_data.csv"))


model_predictions_combined <- read_csv2(
  here(prediction_dir, "model_predictions_combined.csv"))
  
model_auc_data %>% arrange(AUC)
model_auc_data %>%
  mutate(AUC = as.numeric(AUC)) %>%
  arrange(AUC) %>%
  mutate(AUC = round(AUC, 2),
         sensitivity = round(sensitivity, 2),
         specificity = round(specificity, 2))
```

```{r, fig.height = 14, fig.width = 16}
model_predictions_combined$formula %>% unique()

model_predictions_combined %>%
  left_join(prediction_data_df) %>%
  ggplot(aes(x = Response, y = Predictions)) +
  geom_hline(yintercept = 0.5, linetype = "dashed") + facet_wrap(~formula) +
  geom_boxplot() + stat_compare_means(method = "wilcox")
```


```{r, fig.height = 14, fig.width = 16}

model_predictions_combined %>%
  filter(formula == "Response_n ~ pancancer_response_score_scaled") %>%
  left_join(prediction_data_df) %>%
  ggplot(aes(x = Response, y = Predictions)) +
  geom_hline(yintercept = 0.5, linetype = "dashed") + facet_wrap(~Trial) +
  geom_boxplot() + stat_compare_means(method = "wilcox") +
  labs(title = "Response_n ~ pancancer_response_score")

model_predictions_combined %>%
  filter(formula == "Response_n ~ pancancer_response_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial)") %>%
  left_join(prediction_data_df) %>%
  ggplot(aes(x = Response, y = Predictions)) +
  geom_hline(yintercept = 0.5, linetype = "dashed") + facet_wrap(~Trial) +
  geom_boxplot() + stat_compare_means(method = "wilcox") +
  labs(title = "Response_n ~ pancancer_response_score + covariates")
```

# ---------------

# Volcano Plot DEG projection
```{r}
# project signatures in DE results
de_dir <- here("data", "processed", "differential_expression")

deseq_r1 <- read_csv2(
  here(de_dir, "deseq_r1_fdrq.csv")) %>%
  left_join(gene_descriptions) %>%
  mutate(ranking = -log10(pvalue) * log2FoldChange)

alpha_threshold <- 0.01
log2fcthresh <- 1

deseq_r1_volcano_df <- deseq_r1 %>%
  mutate(within_threshold = (padj < alpha_threshold &
                        abs(log2FoldChange) > log2fcthresh)) %>%
  left_join(
    pancan_response_signatures %>%
      filter(Signature %in% c("pancancer_response_sig_1.25",
                              "pancancer_nonresponse_sig_1.25")) %>%
      rename(gene = Gene)) %>%
  left_join(data.frame(
    Signature = c("pancancer_response_sig_1.25",
                  "pancancer_nonresponse_sig_1.25"),
    Class = c("ICB_sensitivity", "ICB_resistance"),
    Ref = c("Pancan", "Pancan"))
    ) %>%
  unique()


deseq_r1_volcano_df %>%
  ggplot(
  aes(x = log2FoldChange, y = -log10(padj + 1e-300),
      col = Class)) +
  geom_point(alpha = 0.5) +
  geom_point(alpha = 0.5, data = deseq_r1_volcano_df %>%
               filter(!is.na(Class))) +
  geom_vline(xintercept = 0, col = "black") +
  geom_vline(xintercept = c(-log2fcthresh, log2fcthresh), col = "red") +
    geom_hline(yintercept = -log10(alpha_threshold), colour = "red") +
  geom_text_repel(
    aes(x = log2FoldChange, y = -log10(padj + 1e-300), label = gene),
    data = deseq_r1_volcano_df %>%
  filter(Ref == "Pancan"),
    colour = "grey20",
    min.segment.length = 0) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(
    title = "Volcano plot Deseq2 Wald",
    x = "Effect size: log2 (fold-change)",
    y = "-log10 (FDRq)")  +
    theme_Publication()

deseq_r1_volcano_df %>%
  filter(Ref == "Pancan") %>%
  ggplot(
  aes(x = log2FoldChange, y = -log10(padj + 1e-300),
      col = Class)) +
  geom_point(alpha = 0.5) +
  geom_point(alpha = 0.5, data = deseq_r1_volcano_df %>%
               filter(!is.na(Class))) +
  geom_vline(xintercept = 0, col = "black") +
  geom_vline(xintercept = c(-log2fcthresh, log2fcthresh), col = "red") +
    geom_hline(yintercept = -log10(alpha_threshold), colour = "red") +
  geom_text_repel(
    aes(x = log2FoldChange, y = -log10(padj + 1e-300), label = gene),
    data = deseq_r1_volcano_df %>%
  filter(Ref == "Pancan"),
    colour = "grey20",
    min.segment.length = 0) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(
    title = "Volcano plot Deseq2 Wald",
    x = "Effect size: log2 (fold-change)",
    y = "-log10 (FDRq)")  +
    theme_Publication()

```

# ---------------

# Load data

```{r}
n_variants_tmb <- read_csv2(here(wes_dir, "n_variants_dp20_vaf_tmb.csv"))

wes_features_df <- read_csv2(here(wes_dir, "wes_features_df.csv"))
wes_features_df <- wes_features_df %>%
  filter(!(Trial %in% "NICHE-MSI" & Tumor_nr == "Tumor_2"))

pathway_level_mutation_df <- read_csv2(here(wes_dir,
                                            "pathway_level_mutation_df.csv"))
pathway_level_mutation_df <- pathway_level_mutation_df %>%
  filter(!(Trial %in% "NICHE-MSI" & Tumor_nr == "Tumor_2"))

facets_data_metrics <- read_csv2(here(wes_dir, "facets_data_metrics.csv"))
facets_data_metrics <- facets_data_metrics %>%
  filter(Tumor_Sample_Barcode %in% wes_features_df$Tumor_Sample_Barcode)

facets_aneuploidy_score_df <- read_csv2(
  here(wes_dir, "facets_aneuploidy_score_df.csv"))
facets_aneuploidy_score_df <- facets_aneuploidy_score_df %>%
  filter(CF_sample_tumor %in% wes_features_df$CF_sample_tumor)

combined_wes_smd <- read_csv2(here(wes_meta_dir, "combined_wes_smd.csv"))
trial_odds_ratio_df <- read_csv2(here(wes_meta_dir, "trial_odds_ratio_df.csv"))

meta_studies_results_smd_all <- read_csv2(
  here(wes_meta_dir, "meta_studies_results_smd_all.csv"))

meta_studies_results_or_all <- read_csv2(
  here(wes_meta_dir, "meta_studies_results_or_all.csv"))

# Add patient N stage and sex information
wes_features_df <- wes_features_df %>%
  left_join(patient_metadata_combined %>%
              select(Patient, Sex, Baseline_N_h)) %>%
  mutate(Baseline_N_h = ifelse(Baseline_N_h == "Nx", "N+", Baseline_N_h))

wes_features_df <- wes_features_df %>%
  mutate(Response_n = ifelse(
    Response == "R",
    1, ifelse(
      Response == "NR", 0, NA))) %>%
  mutate(TMB_scaled = scale(TMB),
         UV_scaled = scale(UV_abs),
         APOBEC_scaled = scale(APOBEC_abs)) 

wes_rna_meta_features_df$Response <- factor(
  wes_rna_meta_features_df$Response, levels = c("R", "NR"))

wes_rna_meta_features_df_excl_dmmr <- wes_rna_meta_features_df %>%
  filter(!Trial %in% c("NICHE-MSI"))

wes_rna_meta_features_df %>% factor_summary()

```

# ---------------
# Nested cross validation Random Forest 

## Data partitions 
### Balanced Outer Fold structure
```{r}

data_partition <- wes_rna_meta_features_df %>%
  mutate(row_index = rownames(wes_rna_meta_features_df)) %>%
  mutate(Patient_col = Patient) %>%
  column_to_rownames("Patient") %>%
  rename(Patient = Patient_col)

# stratify by Trial × Response for a balanced distribution
data_partition$Trial_response <- interaction(
  data_partition$Response, data_partition$Trial, drop = TRUE)
table(data_partition$Trial_response)

outer_folds <- createFolds(data_partition$Trial_response, k = 5,
                           returnTrain = FALSE)
length(outer_folds)
sapply(outer_folds, length)
cat_col = "Trial_response"

data_partition %>% write_csv2(here(output_dir, "data_partition.csv"))
```

### Check outer fold proportions
```{r}
factor_summary_prop <- function(object, maxsum = 20) {
  object %>%
    dplyr::mutate_if(is.character, as.factor) %>%
    purrr::map_df(~{
      x <- summary(.x, maxsum = maxsum)
      tibble::tibble(
        level = names(x),
        count = as.numeric(x),
        prop = count / sum(count)
      )
    }, .id = "variable")
}

data_partition %>%
          select(Trial_response, Trial, Response) %>%
          factor_summary_prop()

data_partition_folds_df <- data_partition %>%
          select(Trial_response, Trial, Response) %>%
          factor_summary_prop() %>%
  mutate(Partition = "All")

for (current_fold in names(outer_folds)){
  print(outer_folds[current_fold])
  print(length(outer_folds[[current_fold]]))

  print(data_partition[outer_folds[[current_fold]], ] %>%
          select(Trial_response, Trial, Response) %>%
          factor_summary_prop())
  
  current_data_partition_folds_df <- data_partition[
    outer_folds[[current_fold]], ] %>%
          select(Trial_response, Trial, Response) %>%
          factor_summary_prop() %>%
    mutate(Partition = paste0("Outer_", current_fold))

  data_partition_folds_df <- data_partition_folds_df %>%
    rbind(current_data_partition_folds_df)
}

data_partition_folds_df %>%
  filter(variable == "Trial") %>%
  ggplot(aes(x = level, y = prop)) + facet_wrap(~Partition) +
  geom_point() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

data_partition_folds_df %>%
  filter(variable == "Trial") %>%
  ggplot(aes(x = Partition, y = prop)) + facet_wrap(~level) +
  geom_point() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

outer_folds
for (current_fold in outer_folds) {
  print(current_fold)
  data_partition[current_fold, c("Trial", "Response", "Trial_response")] %>%
    factor_summary() %>%
    print()

}

```

## Define model data

```{r}
model_data <-  wes_rna_meta_features_df %>%
  column_to_rownames("Patient") %>%
  select(-Trial) %>%
  mutate(Sex = ifelse(Sex == "Female", 1, 0)) %>%
  mutate(Baseline_N_h = ifelse(Baseline_N_h == "N+", 1, 0)) %>%
  mutate(GIE_AP = ifelse(GIE_AP == TRUE, 1, 0)) %>%
  mutate(TP53 = ifelse(TP53 == TRUE, 1, 0)) %>%
  mutate(BRCA1_2 = ifelse(BRCA1_2 == TRUE, 1, 0))%>%
  mutate(HLA_alteration = ifelse(HLA_alteration == "HLA-Altered", 1, 0)) %>%
  mutate(SIGPW_Wnt_B_Cat = ifelse(SIGPW_Wnt_B_Cat == TRUE, 1, 0)) %>%
  mutate(SIGPW_PI3K_signaling = ifelse(SIGPW_PI3K_signaling == TRUE, 1, 0)) %>%
  mutate(SIGPW_Cell_cycle = ifelse(SIGPW_Cell_cycle == TRUE, 1, 0)) %>%
  mutate(SIGPW_TGFB_signaling = ifelse(SIGPW_TGFB_signaling == TRUE, 1, 0)) %>%
  mutate(GIE_IFNy = ifelse(GIE_IFNy == TRUE, 1, 0))  %>%
  mutate(RAS_RAF = ifelse(RAS_RAF == TRUE, 1, 0)) 
  
model_data$Response <- relevel(model_data$Response, ref = "NR")
levels(model_data$Response)
table(model_data$Response)
str(model_data)
factor_summary(model_data)
```

### Train outer 5-fold CV loop

```{r}
train_nested_cv_model <- FALSE

if (train_nested_cv_model) {
  # Store and report mean AUC and SD across 5 outer folds, also for sensitivity, specificity etc
  outer_fold_models <- list()
  outer_fold_results <- list()
  removed_predictors <- list()
  varimps <- list()

  # Iterate over outer folds
  for (i in seq_along(outer_folds)) {
    
    message("Outer fold: ", i)
    
    # Outer training and test sets indexes
    test_idx  <- outer_folds[[i]]
    train_idx <- setdiff(seq_len(nrow(data_partition)), test_idx)
  
    # Still has the partition information to balance the nested 10-CV inner loops
    train_outer <- data_partition[train_idx, ]
    test_outer  <- data_partition[test_idx, ]
    
    train_outer %>% select(Trial_response) %>% factor_summary() %>% print()
    test_outer %>% select(Trial_response) %>% factor_summary() %>% print()
    
    # Only contains the model data for prediction + Response
    train_data <- model_data[train_idx, ]
    test_data  <- model_data[test_idx, ]
    
    # Remove constant predictors inside this outer fold
    # unlikely given the distribution of these features
    constant_cols <- names(train_data)[sapply(train_data, function(x) length(unique(x)) == 1)]
    removed_predictors[[i]] <- constant_cols
    
     if (length(constant_cols) > 0) {
      train_data <- train_data[, !(names(train_data) %in% constant_cols), drop = FALSE]
      test_data  <- test_data[, !(names(test_data) %in% constant_cols), drop = FALSE]
    }
    
    # Create inner folds on the SAME data used for training 
    # 10 fold CV stratified by Trial_response
    inner_folds <- createFolds(
      train_outer$Trial_response, k = 10, returnTrain = TRUE)
    inner_indexOut <- lapply(
      inner_folds, function(idx) setdiff(seq_len(nrow(train_outer)), idx))
  
    # TrainControl using inner folds
    ctrl <- trainControl(
      method = "cv",
      number = 10,
      classProbs = TRUE,
      summaryFunction = twoClassSummary,
      savePredictions = TRUE,
      index = inner_folds,
      indexOut = inner_indexOut
    )
  
    # Train model on outer training set
    model <- train(
      Response ~ .,
      data = train_data,
      method = "rf",
      trControl = ctrl,
      metric = "ROC"
    )
  
    # Evaluate on outer test set
    pred <- predict(model, newdata = test_data, type = "prob")[, 2]
  
    outer_fold_results[[i]] <- data.frame(
      outer_fold = i,
      row_index = test_idx,
      truth = test_data$Response,
      pred = pred
    )
    
    outer_fold_models[[i]] <- model
    
    # Variable importance for this outer model
    varimps[[i]] <- varImp(model)$importance
    
  }
  
  outer_fold_models %>% saveRDS(here(output_dir, "outer_fold_models.rds"))
  outer_fold_results %>% saveRDS(here(output_dir, "outer_fold_results.rds"))
  removed_predictors %>% saveRDS(here(output_dir,
                                      "outer_fold_removed_predictors.rds"))
  varimps %>% saveRDS(here(output_dir, "outer_fold_varimps.rds"))
}

outer_fold_models <- readRDS(here(output_dir, "outer_fold_models.rds"))
outer_fold_results <- readRDS(here(output_dir, "outer_fold_results.rds"))
removed_predictors <- readRDS(here(output_dir,
                                   "outer_fold_removed_predictors.rds"))
varimps <- readRDS(here(output_dir, "outer_fold_varimps.rds"))

```

```{r}
all_outer <- bind_rows(outer_fold_results) %>%
  mutate(ground_truth = truth) %>%
  mutate(truth = ifelse(truth %in% c(1, "R"), 1, 0)) %>%
  mutate(outer_fold = paste0("fold_", outer_fold)) %>%
  mutate(row_index = as.character(row_index)) %>%
  left_join(data_partition %>%
              mutate(row_index  = as.character(1:nrow(data_partition))))

table(all_outer$ground_truth, all_outer$truth)
table(all_outer$Trial)
table(all_outer$Trial_response)
table(all_outer$outer_fold, all_outer$Trial)
table(all_outer$outer_fold, all_outer$Response)
table(all_outer$outer_fold, all_outer$Trial_response)

```

### ROC per outer folds
```{r}
plot(NULL, xlim=c(0,1), ylim=c(0,1),
     xlab="1 - Specificity", ylab="Sensitivity")

for (f in unique(all_outer$outer_fold)) {
  df_fold <- all_outer[all_outer$outer_fold == f, ]
  
  truth <- factor(df_fold$ground_truth, levels = c("NR", "R"))
  pred   <- df_fold$pred
  
  if (length(unique(truth)) < 2) next
  
  roc_obj <- roc(truth, pred)
  lines(1 - roc_obj$specificities, roc_obj$sensitivities, col = "grey70")
}

abline(0,1,lty=2)
```

```{r}
folds <- unique(all_outer$outer_fold)
roc_list <- list()

for (f in folds) {
  df_fold <- all_outer[all_outer$outer_fold == f, ]
  
  truth <- factor(df_fold$ground_truth, levels = c("NR", "R"))
  pred   <- df_fold$pred
  
  # Skip folds with only one class
  if (length(unique(truth)) < 2) {
    cat("Skipping", f, "→ only one class present\n")
    next
  }
  
  roc_list[[f]] <- roc(truth, pred, quiet = TRUE)
}

# Common grid of FPR values
fpr_grid <- seq(0, 1, length.out = 200)

# Interpolate TPR for each fold
tpr_mat <- sapply(roc_list, function(r) {
  approx(
    x = 1 - r$specificities,
    y = r$sensitivities,
    xout = fpr_grid
  )$y
})

mean_tpr <- rowMeans(tpr_mat, na.rm = TRUE)
sd_tpr   <- apply(tpr_mat, 1, sd, na.rm = TRUE)

plot(NULL, xlim=c(0,1), ylim=c(0,1),
     xlab="False Positive Rate", ylab="True Positive Rate")

# Per-fold ROC curves
for (r in roc_list) {
  lines(1 - r$specificities, r$sensitivities, col = "grey80")
}

# Shaded SD band
upper <- pmin(mean_tpr + sd_tpr, 1)
lower <- pmax(mean_tpr - sd_tpr, 0)
polygon(
  c(fpr_grid, rev(fpr_grid)),
  c(upper, rev(lower)),
  col = adjustcolor("skyblue", alpha.f = 0.3),
  border = NA
)

# Mean ROC curve
lines(fpr_grid, mean_tpr, col="blue", lwd = 2)

abline(0, 1, lty = 2)

```

```{r}
all_outer %>% ggplot(aes(x = ground_truth, y = pred, fill = Response)) +
  geom_boxplot() + theme_Publication()+
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  stat_compare_means(method = "wilcox")
```

```{r, fig.height = 8, fig.width = 8}
all_outer %>% ggplot(aes(x = ground_truth, y = pred, fill = Response)) +
  geom_boxplot() + theme_Publication() + facet_wrap(~outer_fold) +
  scale_fill_manual(values = c(Color_NR, Color_R))+
  stat_compare_means(method = "wilcox")
```

```{r, fig.height = 10, fig.width = 8}
all_outer %>% 
  mutate(outer_fold_ground_truth = paste0(ground_truth, "_", outer_fold)) %>%
  ggplot(aes(x = outer_fold_ground_truth, y = pred, fill = Response)) +
  geom_boxplot() + theme_Publication() + facet_wrap(~Trial) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = c(Color_NR, Color_R))
```

```{r, fig.height = 5, fig.width = 8}
all_outer %>% 
  ggplot(aes(x = Trial_response, y = pred, fill = Response)) +
  geom_boxplot() + theme_Publication() + facet_wrap(~outer_fold) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = c(Color_NR, Color_R))
```

### Summary of metrics across outer folds
```{r}
table(all_outer$ground_truth)
roc(all_outer$truth, all_outer$pred)

auc_per_fold <- all_outer %>%
  group_by(outer_fold) %>%
  summarise(
    auc = as.numeric(pROC::auc(pROC::roc(truth, pred)))
  )

auc_per_fold 

summary_auc <- auc_per_fold %>%
  summarise(
    mean_auc = mean(auc),
    sd_auc = sd(auc)
  )

summary_auc %>% round(2)

safe_sens <- function(truth, pred) {
  if (length(unique(truth)) < 2 || length(unique(pred)) < 2) return(NA_real_)
  caret::sensitivity(truth, pred, positive = "R")
}

safe_spec <- function(truth, pred) {
  if (length(unique(truth)) < 2 || length(unique(pred)) < 2) return(NA_real_)
  caret::specificity(truth, pred, negative = "NR")
}

results <- list()

for (folds in unique(all_outer$outer_fold)) {
  
  cat("Processing:", folds, "\n")
  
  df_fold <- all_outer[all_outer$outer_fold == folds, ]
  
  truth_f <- factor(df_fold$truth == 1, levels = c(FALSE, TRUE))
  pred_f  <- factor(df_fold$pred > 0.5, levels = c(FALSE, TRUE))
  
  # Check class presence
  if (length(unique(truth_f)) < 2) {
    cat("  Skipped:", folds, "→ only one class present in ground truth\n")
    sens <- NA_real_
    spec <- NA_real_
  } else if (length(unique(pred_f)) < 2) {
    cat("  Warning:", folds, "→ model predicted only one class\n")
    sens <- NA_real_
    spec <- NA_real_
  } else {
    sens <- caret::sensitivity(truth_f, pred_f, positive = TRUE)
    spec <- caret::specificity(truth_f, pred_f, negative = FALSE)
  }
  
  results[[folds]] <- data.frame(
    outer_fold = folds,
    sens = sens,
    spec = spec
  )
}

sens_spec_per_fold <- do.call(rbind, results)

sens_spec_per_fold %>%
  mutate(sens = round(sens, 2))%>%
  mutate(spec = round(spec, 2))

summary_sens_spec <- sens_spec_per_fold %>%
  summarise(
    mean_sens = mean(sens),
    sd_sens = sd(sens),
    mean_spec = mean(spec),
    sd_spec = sd(spec)
  )

summary_sens_spec %>% round(2)
```

### Varimps over outer folds
```{r, fig.height = 16, fig.width = 8}
varimp_df <- map2_df(
  varimps,
  seq_along(varimps),
  ~ mutate(.x, outer_fold = .y, Variable = rownames(.x))
)

varimp_summary <- varimp_df %>%
  group_by(Variable) %>%
  summarise(
    mean_imp = mean(Overall),
    sd_imp = sd(Overall)
  ) %>%
  arrange(desc(mean_imp))

varimp_df %>%
  ggplot(aes(x = fct_rev(factor(Variable, levels = varimp_summary$Variable)),
             y = Overall)) + geom_boxplot(outlier.shape = NA) +
  geom_point(aes(col = as.factor(outer_fold))) +
  coord_flip() + theme_Publication()

varimp_summary %>%
  ggplot(aes(x = reorder(Variable, mean_imp), y = mean_imp)) +
  geom_col() + coord_flip() + theme_Publication()

varimp_summary %>% write_csv2(here(output_dir, "varimp_summary.csv"))

varimp_df %>% filter(Overall == 100)
```

# ---------------
# LOSO Leave trial out outer folds
## Use each cohort as the outer test folds, nested CV on remaining pancan data
```{r}
train_nested_cv_model <- FALSE
if (train_nested_cv_model) {
  trial_outer_fold_models <- list()
  trial_outer_fold_results <- list()
  trial_outer_fold_removed_predictors <- list()
  trial_outer_fold_varimps <- list()
  
  # Iterate over trials as outer folds (Leave one study out - LOSO)
  for (current_trial in c("BELLINI", "IMCISION", "NABUCCO", "NICHE-MSS",
                          "OpACIN_neo", "PRADO", "MATISSE")) {
    
    message("Outer fold: ", current_trial)
    
    train_ids <- rownames(data_partition)[data_partition$Trial != current_trial]
    test_ids  <- rownames(data_partition)[data_partition$Trial == current_trial]
    
    # Still has the partition information to balance the nested 10-CV inner loops
    train_outer <- data_partition %>% filter(Patient %in% train_ids)
    test_outer  <- data_partition %>% filter(Patient %in% test_ids)
    
    train_outer %>% select(Trial, Trial_response) %>% factor_summary() %>% print()
    test_outer %>% select(Trial, Trial_response) %>% factor_summary() %>% print()
    
    # Only contains the model data for prediction + Response
    train_data <- model_data[data_partition$Trial != current_trial, ]
    test_data <- model_data[data_partition$Trial == current_trial, ]
    
    # Remove constant predictors inside this outer fold
    constant_cols <- names(train_data)[sapply(train_data, function(x) length(unique(x)) == 1)]
    trial_outer_fold_removed_predictors[[current_trial]] <- constant_cols
    
     if (length(constant_cols) > 0) {
      train_data <- train_data[, !(names(train_data) %in% constant_cols), drop = FALSE]
      test_data  <- test_data[, !(names(test_data) %in% constant_cols), drop = FALSE]
    }
    
    # Create inner folds on the SAME data used for training 
    # 10 fold CV stratified by Trial_response
    inner_folds_ids <- createFolds(train_outer$Trial_response, k = 10, returnTrain = TRUE)
    
    # Convert fold indices → sample_IDs
    inner_folds_ids <- lapply(inner_folds_ids, function(idx) rownames(train_outer)[idx])
    # Map sample_IDs to integer row positions in train_data
    inner_folds <- lapply(inner_folds_ids, function(ids) match(ids, rownames(train_data)))
    inner_indexOut <- lapply(inner_folds, function(idx) setdiff(seq_len(nrow(train_data)), idx))

    # TrainControl using inner folds
    ctrl <- trainControl(
      method = "cv",
      number = 10,
      classProbs = TRUE,
      summaryFunction = twoClassSummary,
      savePredictions = TRUE,
      index = inner_folds,
      indexOut = inner_indexOut
    )
  
    # Train model on outer training set
    model <- train(
      Response ~ .,
      data = train_data,
      method = "rf",
      trControl = ctrl,
      metric = "ROC"
    )
  
    # Evaluate on outer test set
    pred <- predict(model, newdata = test_data, type = "prob")[,2]
  
    trial_outer_fold_results[[current_trial]] <- data.frame(
      outer_fold = current_trial,
      sample_ID = rownames(test_data),
      Left_out_trial = current_trial,
      truth = test_data$Response,
      pred = pred,
      strategy = "LOSO"
    )
    
    trial_outer_fold_models[[current_trial]] <- model
    
    # Variable importance for this outer model
    trial_outer_fold_varimps[[current_trial]] <- varImp(model)$importance
    
  }
  
  trial_outer_fold_models %>%
    saveRDS(here(output_dir, "trial_outer_fold_models.rds"))
  trial_outer_fold_results %>%
    saveRDS(here(output_dir, "trial_outer_fold_results.rds"))
  trial_outer_fold_removed_predictors %>%
    saveRDS(here(output_dir, "trial_outer_fold_removed_predictors.rds"))
  trial_outer_fold_varimps %>%
    saveRDS(here(output_dir, "trial_outer_fold_varimps.rds"))
}

trial_outer_fold_models <- readRDS(
  here(output_dir, "trial_outer_fold_models.rds"))
trial_outer_fold_results <- readRDS(
  here(output_dir, "trial_outer_fold_results.rds"))
trial_outer_fold_removed_predictors <- readRDS(
  here(output_dir, "trial_outer_fold_removed_predictors.rds"))
trial_outer_fold_varimps <- readRDS(
  here(output_dir, "trial_outer_fold_varimps.rds"))

```

```{r}
LOSO_trial_outer_folds <- bind_rows(trial_outer_fold_results) %>%
  mutate(ground_truth = truth) %>%
  mutate(truth = ifelse(truth %in% c(1, "R"), 1, 0)) %>%
  mutate(outer_fold = paste0("fold_", outer_fold)) %>%
  mutate(Patient = sample_ID) %>%
  left_join(data_partition)

LOSO_trial_outer_folds %>% dim()

# Check encoding direction
table(LOSO_trial_outer_folds$ground_truth, LOSO_trial_outer_folds$truth)
table(LOSO_trial_outer_folds$outer_fold, LOSO_trial_outer_folds$Trial_response)
```

## Plot LOSO ROC
```{r}
plot(NULL, xlim=c(0,1), ylim=c(0,1),
     xlab="1 - Specificity", ylab="Sensitivity")

for (f in unique(LOSO_trial_outer_folds$outer_fold)) {
  print(f)
  df_fold <- LOSO_trial_outer_folds[LOSO_trial_outer_folds$outer_fold == f, ]
  
  truth <- factor(df_fold$ground_truth, levels = c("NR", "R"))
  pred   <- df_fold$pred
  
  if (length(unique(truth)) < 2) next
  
  roc_obj <- roc(truth, pred)
  print(roc_obj)
  lines(1 - roc_obj$specificities, roc_obj$sensitivities,
        col = trial_colors[unique(df_fold$Left_out_trial)])
}

abline(0,1,lty=2)

```

## Plot LOSO ROC + Nested CV

```{r}
plot(NULL, xlim=c(0,1), ylim=c(0,1),
     xlab="1 - Specificity", ylab="Sensitivity")

folds <- unique(all_outer$outer_fold)
roc_list <- list()

for (f in folds) {
  df_fold <- all_outer[all_outer$outer_fold == f, ]
  
  truth <- factor(df_fold$ground_truth, levels = c("NR", "R"))
  pred   <- df_fold$pred
  
  # Skip folds with only one class
  if (length(unique(truth)) < 2) {
    cat("Skipping", f, "→ only one class present\n")
    next
  }
  
  roc_list[[f]] <- roc(truth, pred, quiet = TRUE)
}

# Common grid of FPR values
fpr_grid <- seq(0, 1, length.out = 200)

# Interpolate TPR for each fold
tpr_mat <- sapply(roc_list, function(r) {
  approx(
    x = 1 - r$specificities,
    y = r$sensitivities,
    xout = fpr_grid
  )$y
})

mean_tpr <- rowMeans(tpr_mat, na.rm = TRUE)
sd_tpr   <- apply(tpr_mat, 1, sd, na.rm = TRUE)

# Per-fold ROC curves
for (r in roc_list) {
  lines(1 - r$specificities, r$sensitivities, col="grey80")
}

# Shaded SD band
upper <- pmin(mean_tpr + sd_tpr, 1)
lower <- pmax(mean_tpr - sd_tpr, 0)
polygon(
  c(fpr_grid, rev(fpr_grid)),
  c(upper, rev(lower)),
  col = adjustcolor("skyblue", alpha.f = 0.3),
  border = NA
)

# Mean ROC curve
lines(fpr_grid, mean_tpr, col="blue", lwd=2)

for (f in unique(LOSO_trial_outer_folds$outer_fold)) {
  print(f)
  df_fold <- LOSO_trial_outer_folds[LOSO_trial_outer_folds$outer_fold == f, ]
  
  truth <- factor(df_fold$ground_truth, levels = c("NR", "R"))
  pred   <- df_fold$pred
  
  if (length(unique(truth)) < 2) next
  
  roc_obj <- roc(truth, pred)
  print(roc_obj)
  lines(1 - roc_obj$specificities, roc_obj$sensitivities,
        col = trial_colors[unique(df_fold$Left_out_trial)])
}

abline(0, 1, lty = 2)
```

```{r}
LOSO_trial_outer_folds %>%
  ggplot(aes(x = ground_truth, y = pred, fill = Response)) +
  geom_boxplot() + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE) +
  labs(title = "Leave one study out set")
```

```{r, fig.height = 10, fig.width = 8}
LOSO_trial_outer_folds %>%
  ggplot(aes(x = ground_truth, y = pred, fill = Response)) + geom_boxplot() +
  facet_wrap(~Trial) + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE) + stat_compare_means(method = "wilcox") +
  labs(title = "Leave one study out set")

LOSO_trial_outer_folds %>%
  ggplot(aes(x = ground_truth, y = pred, fill = Left_out_trial)) +
  geom_boxplot() +
  facet_wrap(~Trial) + theme_Publication() +
  scale_fill_manual(values = c(trial_colors)) +
  labs(title = "Leave one study out set")
```

## Summary of metrics across LOSO
```{r}
roc(LOSO_trial_outer_folds$truth, LOSO_trial_outer_folds$pred)

LOSO_auc_per_fold <- LOSO_trial_outer_folds %>%
  group_by(outer_fold, Left_out_trial) %>%
  summarise(
    auc = as.numeric(pROC::auc(pROC::roc(truth, pred)))
  )

LOSO_auc_per_fold

LOSO_auc_per_fold %>%
  write_csv2(here(prediction_dir, "LOSO_auc_per_fold.csv"))
```


```{r}
LOSO_summary_auc <- LOSO_auc_per_fold %>%
  ungroup() %>%
  summarise(
    mean_auc = mean(auc),
    sd_auc = sd(auc)
  )

LOSO_summary_auc 
LOSO_summary_auc %>% round(2)

LOSO_results <- list()

for (folds in unique(LOSO_trial_outer_folds$outer_fold)) {
  
  cat("Processing:", folds, "\n")
  
  df_fold <- LOSO_trial_outer_folds[
    LOSO_trial_outer_folds$outer_fold == folds, ]
  
  truth_f <- factor(df_fold$truth == 1, levels = c(FALSE, TRUE))
  pred_f  <- factor(df_fold$pred > 0.5, levels = c(FALSE, TRUE))
  
  # Check class presence
  if (length(unique(truth_f)) < 2) {
    cat("  Skipped:", folds, "→ only one class present in ground truth\n")
    sens <- NA_real_
    spec <- NA_real_
  } else if (length(unique(pred_f)) < 2) {
    cat("  Warning:", folds, "→ model predicted only one class\n")
    sens <- NA_real_
    spec <- NA_real_
  } else {
    sens <- caret::sensitivity(truth_f, pred_f, positive = TRUE)
    spec <- caret::specificity(truth_f, pred_f, negative = FALSE)
  }
  
  LOSO_results[[folds]] <- data.frame(
    outer_fold = folds,
    sens = sens,
    spec = spec
  )
}

LOSO_results
LOSO_sens_spec_per_fold <- do.call(rbind, LOSO_results)

LOSO_summary_sens_spec <- LOSO_sens_spec_per_fold %>%
  summarise(
    mean_sens = mean(sens),
    sd_sens = sd(sens),
    mean_spec = mean(spec),
    sd_spec = sd(spec)
  )

LOSO_summary_sens_spec

LOSO_varimp_df <- map2_df(
  trial_outer_fold_varimps,
  names(trial_outer_fold_varimps),
  ~ mutate(.x, outer_fold = .y, Variable = rownames(.x))
) %>%
  mutate(Left_out_trial = outer_fold)

LOSO_varimp_summary <- LOSO_varimp_df %>%
  group_by(Variable) %>%
  summarise(
    mean_imp = mean(Overall),
    sd_imp = sd(Overall)
  ) %>%
  arrange(desc(mean_imp))

```

## Varimps over outer folds

```{r, fig.height = 16, fig.width = 8}
LOSO_varimp_df %>%
  ggplot(aes(x = fct_rev(factor(Variable, levels = varimp_summary$Variable)),
             y = Overall)) + geom_boxplot(outlier.shape = NA) +
  geom_point(aes(col = as.factor(Left_out_trial))) +
  coord_flip() + theme_Publication() +
  scale_color_manual(values = trial_colors) +
  labs(title = "Leave one study out sets")
```

# ---------------

# Per cohort RF LOOCV
```{r}

# Convert features to numeric, logistic becomes 1, categories get a reference 0/1
model_data_excl_dmmr <-  wes_rna_meta_features_df %>%
  filter(!Trial %in% c("NICHE-MSI")) %>%
  mutate(Patient = paste0(Patient, "_")) %>% # Prepare for adding numeric idx
    column_to_rownames("Patient") %>%
  mutate(Sex = ifelse(Sex == "Female", 1, 0)) %>%
  mutate(Baseline_N_h = ifelse(Baseline_N_h == "N+", 1, 0)) %>%
  mutate(GIE_AP = ifelse(GIE_AP == TRUE, 1, 0)) %>%
  mutate(TP53 = ifelse(TP53 == TRUE, 1, 0)) %>%
  mutate(BRCA1_2 = ifelse(BRCA1_2 == TRUE, 1, 0))%>%
  mutate(HLA_alteration = ifelse(HLA_alteration == "HLA-Altered", 1, 0)) %>%
  mutate(SIGPW_Wnt_B_Cat = ifelse(SIGPW_Wnt_B_Cat == TRUE, 1, 0)) %>%
  mutate(SIGPW_PI3K_signaling = ifelse(SIGPW_PI3K_signaling == TRUE, 1, 0)) %>%
  mutate(SIGPW_Cell_cycle = ifelse(SIGPW_Cell_cycle == TRUE, 1, 0)) %>%
  mutate(SIGPW_TGFB_signaling = ifelse(SIGPW_TGFB_signaling == TRUE, 1, 0)) %>%
  mutate(GIE_IFNy = ifelse(GIE_IFNy == TRUE, 1, 0))  %>%
  mutate(RAS_RAF = ifelse(RAS_RAF == TRUE, 1, 0)) 
  
model_data_excl_dmmr$Response <- relevel(model_data_excl_dmmr$Response,
                                         ref = "NR")
levels(model_data_excl_dmmr$Response)
table(model_data_excl_dmmr$Response)
str(model_data_excl_dmmr)
factor_summary(model_data_excl_dmmr)
head(rownames(model_data_excl_dmmr))

train_loocv_rf_models <- FALSE

if (train_loocv_rf_models) {
    rf_models <- list()
    removed_features <- list()
    
    for(current_trial in unique(model_data_excl_dmmr$Trial)) {
      print(sprintf("Processing trial: %s", current_trial))
      
      df_current <- subset(model_data_excl_dmmr,
                           Trial == current_trial) %>%
        select(-Trial) 
      factor_summary(df_current)
    
      # Detect and remove constant features
      constant_cols <- names(df_current)[sapply(
        df_current, function(x) length(unique(x)) == 1)]
      removed_features[[current_trial]] <- constant_cols
      
      df_current <- df_current[
        , !(names(df_current) %in% constant_cols), drop = FALSE]
      
      ctrl  <- trainControl(method  = "LOOCV",
                    classProbs = TRUE,
                    summaryFunction = twoClassSummary,
                    savePredictions = TRUE) 
      
      rf_models[[current_trial]] <- train(
        Response ~ ., 
        data = df_current,
        method = "rf",
        trControl = ctrl
      )
    }
    
    rf_models %>%
      saveRDS(here(output_dir, "random_forest_loocv_trial_fit.rds"))
    
    removed_features %>%
      saveRDS(here(output_dir, "random_forest_removed_features_trial_fit.rds"))
  }

rf_models <- readRDS(
  here(output_dir, "random_forest_loocv_trial_fit.rds"))

rf_removed_features <- readRDS(
  here(output_dir, "random_forest_removed_features_trial_fit.rds"))

rf_models
rf_removed_features

loocv_feature_imp <- bind_rows(
  lapply(names(rf_models), function(current_trial) {
    varImp(rf_models[[current_trial]])$importance %>%
      mutate(model = current_trial)
  }), .id = "Trial")

loocv_feature_imp <- bind_rows(
  lapply(names(rf_models), function(current_trial) {
    varImp(rf_models[[current_trial]])$importance %>%
      mutate(model = current_trial)
  })) %>% rownames_to_column("Variable")

loocv_feature_imp$Variable <- gsub("\\.{3}(\\d+)", "",
                                   loocv_feature_imp$Variable)

rf_get_pred_scores_auc_values <- function(model, trial){
  
  print(trial)
  print(sprintf("Best model mtry %s", model$bestTune$mtry))
  
  df <- model$pred
  # Subset to best tuning parameter
  best_mtry <- model$bestTune$mtry
  df_best <- df %>% filter(mtry == best_mtry)
  
  plot <- df_best %>%
    ggplot(aes(x = obs, y = NR)) + geom_boxplot() + geom_point() +
    labs(title = trial, y = "NR pred class score") +
    theme_Publication()
  print(plot)

  plot <- df_best %>%
    ggplot(aes(x = obs, y = R)) + geom_boxplot() + geom_point() +
    labs(title = trial, y = "R pred class score")+
    theme_Publication()
  print(plot)
  
    # Ensure correct factor levels
  truth <- factor(df_best$obs, levels = c("NR", "R"))
  pred  <- df_best$R
  
  # Skip if only one class present
  if (length(unique(truth)) < 2) {
    auc_val <- NA_real_
  } else {
    auc_val <- as.numeric(auc(truth, pred))
  }
  
  data.frame(
    Trial = trial,
    AUC = auc_val
  )
}

rf_loocv_auc <- lapply(
  names(rf_models),
  function(Trial) rf_get_pred_scores_auc_values(rf_models[[Trial]], Trial)
)

rf_loocv_auc <- do.call(rbind, rf_loocv_auc)
rf_loocv_auc %>% arrange(-AUC)

rf_loocv_auc %>% write_csv2(here(output_dir, "rf_loocv_auc.csv"))
```

# ---------------
# AUC comparison
```{r}
auc_comparison <- rf_loocv_auc %>%
  add_row(Trial = "Combined_outer_folds", AUC = summary_auc$mean_auc)

auc_comparison

auc_comparison %>%
  ggplot(aes(x = reorder(Trial, AUC), y = AUC, fill = Trial)) +
  geom_col() + theme_Publication() + coord_flip() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_fill_manual(values = trial_colors)

auc_comparison %>%
  mutate(model = "RF") %>%
  full_join(glmnet_loocv_auc %>%
  mutate(model = "glmnet")) %>%
  ggplot(aes(x = as.factor(model), y = AUC, col = Trial, group = Trial)) +
  geom_point() + geom_line(aes(group = Trial)) + theme_Publication() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_color_manual(values = trial_colors)

```

# Varimp comparison
```{r, fig.height = 16, fig.width = 8}

importance_df <- varimp_summary %>%
    mutate(model = "Combined") %>%
    rename(Overall = mean_imp) %>%
  full_join(loocv_feature_imp) %>%
  rename(Overall_var_imp = Overall)
  
importance_df_summary <- importance_df %>%
  group_by(Variable) %>%
  summarize(mean = mean(Overall_var_imp)) %>%
  arrange(-mean)

importance_df_summary

importance_df %>%
  ggplot(aes(x = factor(Variable, levels = rev(varimp_summary$Variable)),
             y = Overall_var_imp, col = model)) +
  geom_point() +
  coord_flip() + theme_Publication() +
  labs(x = "", title = "Variable importance") +
  scale_color_manual(values = trial_colors)

importance_df %>%
  ggplot(aes(x = factor(Variable, levels = rev(importance_df_summary$Variable)),
             y = Overall_var_imp, col = model),
         alpha = ifelse(model == "Combined", 1, 0.5)) +
  geom_point() +
  coord_flip() + theme_Publication() +
  labs(x = "", title = "Variable importance") +
  scale_color_manual(values = trial_colors)

# Reorder by top importance, should be top ranked in at least one model
importance_df %>%
  filter(model == "Combined") %>%
  ggplot(aes(x = reorder(Variable, Overall_var_imp),
             y = Overall_var_imp, col = model)) +
  geom_point() + coord_flip() + theme_Publication() +
  labs(x = "", title = "Variable importance (Reorder top 100)")

auc_comparison %>% write_csv2(here(prediction_dir, "auc_comparison.csv"))
importance_df %>% write_csv2(here(prediction_dir, "importance_df.csv"))

importance_df %>% filter(Overall_var_imp == 100)
```

# Aggregate all
```{r}
auc_comparison <- rf_loocv_auc %>%
  mutate(Training_set = "Trial_specific",
         outer_fold = NA) %>%
  add_row(Trial = "Combined_outer_folds",
          AUC = summary_auc$mean_auc,
          Training_set = "Pancancer Nested CV",
          outer_fold = "mean_outer_folds") %>%
  full_join(LOSO_auc_per_fold %>%
              rename(Trial = Left_out_trial, AUC = auc) %>%
              mutate(Training_set = "Pancancer LOSO")) %>%
  arrange(Trial)

auc_comparison %>%
  ggplot(aes(x = Training_set, y = AUC, col = Trial, group = Trial)) +
  geom_point() + geom_line(aes(group = Trial)) + theme_Publication() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_color_manual(values = trial_colors)

auc_comparison %>%
  filter(!Training_set %in% c("Pancancer Nested CV")) %>%
  ggplot(aes(x = Training_set, y = AUC, col = Trial, group = Trial)) +
  geom_point() + geom_line(aes(group = Trial)) + theme_Publication() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_color_manual(values = trial_colors)

```

```{r}
auc_comparison_aggregate_all <- rf_loocv_auc %>%
  mutate(Training_set = "Trial_specific",
         outer_fold = NA) %>%
  add_row(Trial = "Combined_outer_folds",
          AUC = summary_auc$mean_auc,
          Training_set = "Pancancer Nested CV",
          outer_fold = "mean_outer_folds") %>%
  full_join(LOSO_auc_per_fold %>%
              rename(Trial = Left_out_trial, AUC = auc) %>%
              mutate(Training_set = "Pancancer LOSO")) %>%
  arrange(Trial) %>%
  left_join(summary_auc %>%
              mutate(Trial = "Combined_outer_folds")) %>%
  full_join(auc_per_fold %>%
              rename(AUC = auc) %>%
              mutate(Trial = "Combined_outer_folds",
                     Training_set = "Pancancer Nested CV"))
  
print(auc_comparison_aggregate_all)

auc_comparison %>%
  left_join(summary_auc %>%
              mutate(Trial = "Combined_outer_folds")) %>%
  filter(!Training_set %in% c("Pancancer LOSO")) %>%
  ggplot(aes(x = reorder(Trial, AUC), y = AUC, fill = Trial,
             shape = Training_set)) +
  geom_col() + theme_Publication() + coord_flip() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_fill_manual(values = trial_colors)  +
  scale_color_manual(values = trial_colors)+
  geom_errorbar(
    aes(ymin = AUC - sd_auc, ymax = AUC + sd_auc),
    width = 0.2,
    linewidth = 0.5
  ) + geom_point(aes(x = Trial, y = AUC), data = auc_per_fold %>%
              mutate(AUC = auc, Trial = "Combined_outer_folds",
                     Training_set = "Pancancer Nested CV")) +
  geom_point(aes(x = Trial, y = AUC, col = "black"),
             data = auc_comparison %>%
               filter(Training_set == "Pancancer LOSO")) +
  labs(x = "") +
  guides(shape = guide_legend(nrow = 3), size = NULL)


auc_comparison_aggregate_all_plot <- auc_comparison_aggregate_all %>%
  filter(!(Training_set %in% c("Pancancer Nested CV") & outer_fold
           %in% c("fold_1", "fold_2", "fold_3", "fold_4", "fold_5"))) %>%
  filter(!Training_set %in% c("Pancancer LOSO")) %>%
  ggplot(aes(x = reorder(Trial, AUC), y = AUC, fill = Trial,
             shape = Training_set)) +
  geom_col() + theme_Publication() + coord_flip() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_fill_manual(values = trial_colors)  +
  scale_color_manual(values = trial_colors)+
  geom_errorbar(
    aes(ymin = AUC - sd_auc, ymax = AUC + sd_auc),
    width = 0.2,
    linewidth = 0.5
  ) + geom_point(aes(x = Trial, y = AUC), data = auc_comparison_aggregate_all) +
  geom_point(aes(x = Trial, y = AUC, col = "black"),
             data = auc_comparison_aggregate_all %>%
               filter(Training_set == "Pancancer LOSO")) +
  labs(x = "") +
  guides(shape = guide_legend(nrow = 3), size = NULL) +
  # guides(fill = guide_legend(nrow = 3), size = NULL) +
  guides(fill = NULL, col = NULL)

auc_comparison_aggregate_all_plot

graph2pdf(x = auc_comparison_aggregate_all_plot,
          file = here(output_dir, "_auc_comparison_aggregate_all_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4.5,
          width = 6.5)

auc_comparison_aggregate_all %>%
  write_csv2(here(output_dir, "auc_comparison_aggregate_all.csv"))
```


```{r, fig.height = 6, fig.width = 8}
auc_comparison_aggregate_all %>%
  filter(!outer_fold %in%
           c("fold_1", "fold_2", "fold_3", "fold_4", "fold_5")) %>%
  mutate(model = paste0("random_forest_", Trial, "_", Training_set)) %>%
  full_join(model_auc_data %>%
              mutate(AUC = as.numeric(AUC),
                     model = gsub("Response_n ~ ", "", formula))) %>%
  ggplot(aes(x = reorder(model, AUC), y = AUC)) + geom_col() + coord_flip() 

```


```{r}
auc_comparison %>%
  left_join(summary_auc %>%
              mutate(Trial = "Combined_outer_folds"))%>%
  ggplot(aes(x = reorder(Trial, AUC), y = AUC,
             col = Trial, shape = Training_set)) +
  geom_point() + theme_Publication() + coord_flip() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_color_manual(values = trial_colors)+
  geom_errorbar(
    aes(ymin = AUC - sd_auc, ymax = AUC + sd_auc),
    width = 0.2,
    linewidth = 0.5
  ) 

auc_comparison %>%
  left_join(summary_auc %>%
              mutate(Trial = "Combined_outer_folds"))%>%
  ggplot(aes(x = reorder(Trial, AUC), y = AUC,
             col = Training_set, shape = Training_set)) +
  geom_point() + theme_Publication() + coord_flip() +
  geom_hline(yintercept = 0.5, linetype = "dashed")  +
  geom_errorbar(
    aes(ymin = AUC - sd_auc, ymax = AUC + sd_auc),
    width = 0.2,
    linewidth = 0.5
  ) 

```

```{r, fig.height = 16, fig.width = 12}

importance_df %>%
  mutate(Trial = model) %>%
  full_join(LOSO_varimp_df %>%
              mutate(Trial = Left_out_trial,
                     model = paste("Leaving out:", Left_out_trial)) %>%
              rename(Overall_var_imp = Overall)) %>%
  mutate(Leave_one_study_out_set = ifelse(
    !is.na(Left_out_trial), "LOSO", "Other")) %>%
  ggplot(aes(x = factor(Variable, levels = rev(varimp_summary$Variable)),
             y = Overall_var_imp, col = Trial,
             shape = Leave_one_study_out_set),
         alpha = ifelse(model == "Combined", 1, 0.5)) +
  geom_point() +
  coord_flip() + theme_Publication() +
  labs(x = "", title = "Variable importance") +
  facet_wrap(~fct_rev(Leave_one_study_out_set)) +
  scale_color_manual(values = trial_colors)
```

```{r, fig.height = 16, fig.width = 8}

importance_df %>%
  full_join(LOSO_varimp_df %>%
              mutate(model = paste("Leaving out:", Left_out_trial)) %>%
              rename(Overall_var_imp = Overall)) %>%
  mutate(Leave_one_study_out_set = ifelse(!is.na(Left_out_trial),
                                          "LOSO", "Other")) %>%
  ggplot(aes(x = factor(Variable, levels = rev(varimp_summary$Variable)),
             y = Overall_var_imp, col = model,
             shape = Leave_one_study_out_set),
         alpha = ifelse(model == "Combined", 1, 0.5)) +
  geom_point() +
  coord_flip() + theme_Publication() +
  labs(x = "", title = "Variable importance") # +
  # scale_color_manual(values = trial_colors)
```


```{r, fig.height = 4, fig.width = 6}
auc_comparison_aggregate_all %>%
  filter(Training_set == "Trial_specific") %>%
  mutate(model = "Trial_specific_rf") %>%
  full_join(glmnet_loocv_auc %>%
  mutate(model = "glmnet")) %>%
  ggplot(aes(x = as.factor(model), y = AUC, col = Trial, group = Trial)) +
  geom_point() + geom_line(aes(group = Trial)) + theme_Publication() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_color_manual(values = trial_colors)

```

# ----------
# Lint script
```{r}
# lint("~/NKI_pancancer_neoadj_io/src/pancan_response_prediction.Rmd")
lint(filename = "~/NKI_pancancer_neoadj_io/src/pancan_response_prediction.Rmd",
  linters = unused_import_linter()
)
```

# ---------------
# sessionInfo - Package Versions
```{r sessionInfo}
sessionInfo()
```
