---
title: "neoadjuvant pancancer WES meta-analysis"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  export: TRUE
  output_dir: "WES_analysis_meta"
  metadata: ""
  filter_and_parse_vcfs: FALSE
  exclude_niche_msi: TRUE
  run_meta_analysis: TRUE
  response_definition: "Original" # MPR pCR
  exclude_monotherapy: FALSE
---

Author comment:
File description comment:
The purpose of this script is to parse results from WES analysis
vcf files, quality logs, comparing with RNAseq data
filter_and_parse_vcfs

It receives TRUST4 results as inputs
and outputs metrics calculated from TCR and BCR repertoires

# Loading Packages
```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# Install if necessary install.packages("tidyverse")
# Install if necessary BiocManager::install("DESeq2")

library(tidyverse)
source("helper_functions.R")
source("oncomatrix.R")
library(VariantAnnotation)
library(ggpubr)
library(paletteer)
library(meta)
library(ggrepel)
library(export)
library(lintr)
library(forestplot)
MakeDataFrameFromVCF <- getFromNamespace("MakeDataFrameFromVCF", "ICAMS")

set.seed(42)
select <- dplyr::select
rename <- dplyr::rename
here <- here::here
```

# Setup output folder

```{r Setup output, echo = FALSE, include = FALSE}
print(params)

export <- params$export
response_definition <- params$response_definition

# overwrite clinical exclusion in case of MPR/pCR response
exclude_clinical_response <- params$exclude_clinical_response
if (params$response_definition %in% c("MPR", "pCR")) {
  exclude_clinical_response <- TRUE
}
exclude_monotherapy <- params$exclude_monotherapy

# generate output directory
wes_dir <- here("data", "processed", "WES_analysis")
metadata_dir <- here("data", "processed", "metadata")
output_dir <- here("data", "processed", params$output_dir)
wes_meta_dir <- output_dir

print(output_dir)
print(exclude_clinical_response)
print(response_definition)

# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)
    } else {
  print("Output folder already exists")
}

```

# ----------
# Load metadata

```{r Load metadata}
dna_metadata <- read_csv2(here(metadata_dir, "combined_dna_metadata.csv"))

# WES target sizes
WES_target_sizes <- data.frame(
  Trial = c("IMCISION", "PRADO", "OpACIN_neo", "MATISSE",
            "NICHE-MSI", "NICHE-MSS", "NABUCCO", "BELLINI"),
  Kit = c(rep("Twist exome Refseq", 4), rep("Exome-IDT V1", 4)),
  WES_target_size = c(rep(36.72, 4), rep(39.02, 4)))

WES_target_sizes

histology_df <- data.frame(
  Trial = c("BELLINI", "IMCISION", "NABUCCO", "NICHE-MSS", "NICHE-MSI",
            "OpACIN_neo", "PRADO", "MATISSE"),
  Histology = c("TBNC", "HNSCC", "Urothelial", "Colon", "Colon",
                "LN_melanoma", "LN_melanoma", "CSCC"))

```

# Annotation color
```{r}

# Define colours for each group
Color_NR <- "#F8766D"
Color_R <- "#00BA38" # "#619CFF"
Color_NA <- "#CD9600"
Color_Pre <- "#619CFF"
Color_Post <- "#F564E3"
Color_HNSCC <- "#009292"
Color_CSCC <- "#CD9600"
Color_Urothelial <- "#B66DFF"
Color_LN_melanoma <- "#00BFC4"
Color_Colon <- "#F8766D"
Color_Colon_MSI <- "#B73A3A"
Color_Colon_MSS <- "#23C26F"
Color_TNBC <- "#FEB5DA"
Color_BELLINI <- "#FEB5DA"
Color_IMCISION <- "#009292"
Color_MATISSE <- "#CD9600"
Color_NABUCCO <- "#B66DFF"
Color_BELLINI <- "#FEB5DA"
Color_NICHE_MSS <- "#23C26F"
Color_NICHE_MSI <- "#B73A3A"
Color_OpACIN_neo <- "#00BFC4"
Color_PRADO <- "#619CFF"

annotation_color_list = list(
  Response = c(NR = Color_NR, R = Color_R),
  Treatment = c(Pre = Color_Pre, Post = Color_Post),
  Histology = c(LN_melanoma = Color_LN_melanoma, Urothelial = Color_Urothelial,
                Colon = Color_Colon, HNSCC = Color_HNSCC, TBNC = Color_TNBC,
                CSCC = Color_CSCC),
  Trial = c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO))

trial_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_colors_combined <- c(Combined = "#999999", BELLINI = Color_BELLINI,
                           IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_Colon_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

histology_msi_mss_colors <- c(`Colon-MSI`= Color_Colon_MSI,
                              `Colon-MSS`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              LN_melanoma = Color_LN_melanoma,
                              TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)

```

# ----------
# Load data

```{r}
n_variants_tmb <- read_csv2(here(wes_dir, "n_variants_dp20_vaf_tmb.csv"))

wes_features_df <- read_csv2(here(wes_dir, "wes_features_df.csv"))
wes_features_df <- wes_features_df %>%
  filter(!(Trial %in% "NICHE-MSI" & Tumor_nr == "Tumor_2"))

pathway_level_mutation_df <- read_csv2(here(wes_dir,
                                            "pathway_level_mutation_df.csv"))
pathway_level_mutation_df <- pathway_level_mutation_df %>%
  filter(!(Trial %in% "NICHE-MSI" & Tumor_nr == "Tumor_2"))

facets_data_metrics <- read_csv2(here(wes_dir, "facets_data_metrics.csv"))
facets_data_metrics <- facets_data_metrics %>%
  filter(Tumor_Sample_Barcode %in% wes_features_df$Tumor_Sample_Barcode)

facets_aneuploidy_score_df <- read_csv2(here(wes_dir,
                                             "facets_aneuploidy_score_df.csv"))
facets_aneuploidy_score_df <- facets_aneuploidy_score_df %>%
  filter(CF_sample_tumor %in% wes_features_df$CF_sample_tumor)

vaf_summaries_c <- read_csv2(here(wes_dir, "vaf_summaries_c.csv"))

```


## Adapt Response Definition
```{r}
compare_r_nr_trials <- c("Combined", "Combined_excl_dMMR", "BELLINI",
                         "IMCISION", "MATISSE", "NABUCCO", "NICHE-MSS",
                         "OpACIN_neo", "PRADO")

if (params$response_definition %in% c("MPR")) {
  compare_r_nr_trials <- c("Combined", "Combined_excl_dMMR", "BELLINI",
                         "IMCISION", "MATISSE", "NABUCCO", "NICHE-MSI",
                         "NICHE-MSS", "OpACIN_neo", "PRADO")
}

if (params$response_definition %in% c("pCR")) {
  compare_r_nr_trials <- c("Combined", "Combined_excl_dMMR", "BELLINI",
                          "MATISSE", "NABUCCO", "NICHE-MSI", "NICHE-MSS", 
                         "OpACIN_neo", "PRADO")
}

patient_metadata_combined <- read_csv2(
  here("data", "processed", "metadata", "patient_metadata_combined.csv"))

wes_features_df <- wes_features_df %>%
    mutate(Response_original = Response) %>%
  left_join(patient_metadata_combined %>%
              select(Patient, Response_type, Response_pCR, Response_MPR)) %>%
  mutate(Response_analysis = response_definition,
         exclude_clinical_response = exclude_clinical_response)

# overwrite clinical exclusion in case of MPR/pCR response
if (params$response_definition %in% c("MPR", "pCR")) {
  exclude_clinical_response <- TRUE
}

# Automatic in case response definition is MPR and pCR
if (exclude_clinical_response) {
  wes_features_df <- wes_features_df %>%
    filter(Response_type == "Pathological")
}

if (response_definition == "MPR") {
  wes_features_df <- wes_features_df %>%
    filter(Response_type == "Pathological") %>%
    mutate(Response = ifelse(Response_MPR == "MPR", "R", "NR"))
}

if (response_definition == "pCR") {
  wes_features_df <- wes_features_df %>%
    filter(Response_type == "Pathological") %>%
    mutate(Response = ifelse(Response_pCR == "pCR", "R", "NR"))
}

if (exclude_monotherapy) {
  wes_features_df <- wes_features_df %>%
    filter(Cohort == "Nivo_ipi") 
}

wes_features_df$Cohort %>% table()

# Sensitivity analysis excluding NICHE2 dMMR MSI samples
if (params$exclude_niche_msi) {
  wes_features_df <- wes_features_df %>% filter(!Trial %in% c("NICHE-MSI"))
}

wes_features_df %>%
  select(colnames(patient_metadata_combined)[
    colnames(patient_metadata_combined) %in% colnames(wes_features_df)]) %>%
  mutate_if(is.character, as.factor) %>% summary()

table(wes_features_df$Trial, wes_features_df$Response)

wes_features_df$Trial %>% table()
```

# ----------
# Get response SMDs
```{r}
calc_stat_r_vs_nr_general <- function(data, contrast,
                                      variable) {

  data_r <- data[data$Response == "R", variable, drop = TRUE]

  mean_r <- mean(data_r, na.rm = TRUE)
  sd_r <- sd(data_r, na.rm = TRUE)
  median_r <- median(data_r, na.rm = TRUE)
  n_r <- length(data_r)

  data_nr <- data[data$Response == "NR", variable, drop = TRUE]

  mean_nr <- mean(data_nr, na.rm = TRUE)
  sd_nr <- sd(data_nr, na.rm = TRUE)
  median_nr <- median(data_nr, na.rm = TRUE)
  n_nr <- length(data_nr)

  wilcoxon_test <- wilcox.test(data_r, data_nr,
                               alternative = "two.sided")
  n_total <- n_r + n_nr

  stat_df <- wilcoxon_test %>%
    broom::tidy() %>%
    mutate(
    "contrast" = contrast,
    "variable" = variable,
    "n" = n_total,
    "mean_responders" = mean_r,
    "sd_responders" = sd_r,
    "median_responders" = median_r,
    "n_responders" = n_r,
    "mean_nresponders" = mean_nr,
    "sd_nresponders" = sd_nr,
    "median_nresponders" = median_nr,
    "n_nresponders" = n_nr)

  return(stat_df)
}

get_trial_smd_estimates <- function(data, trial_list, variable) {
  
  general_smd_df <- data.frame(matrix(nrow = 0, ncol = 0))
  for (current_trial in trial_list) {
    current_data = data
    print(current_trial)

    if (! current_trial %in% c("Combined", "Combined_excl_dMMR")) {
      current_data <- data %>% filter(Trial == current_trial)
    }
    
    if (current_trial == "Combined_excl_dMMR") {
      current_data <- data %>% filter(Trial != c("NICHE-MSI"))
    }
    
    current_variable_smd_estimates = calc_stat_r_vs_nr_general(
        data = current_data,
        contrast = sprintf("%s R x NR", current_trial),
        variable = variable)
    
    current_variable_smd_estimates <- current_variable_smd_estimates %>%
      mutate(Trial = current_trial)
    general_smd_df <- general_smd_df %>% rbind(current_variable_smd_estimates)
  }
  general_smd_df <- general_smd_df %>%
  mutate(mean_delta = mean_responders - mean_nresponders,
         pooled_sd = sqrt((sd_responders^2 + sd_nresponders^2) / 2),
         effect_size = mean_delta / pooled_sd)
  
  general_smd_df <- general_smd_df %>%
  relocate(c(Trial, contrast, variable, n, n_responders, n_nresponders, 
             p.value, method), .before = statistic)
  general_smd_df
}
```


```{r}
smd_estimates_tmb <- get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials, variable = "TMB")

smd_estimates_neoantigens <- get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "pvactools_filtered_neoantigen_load")

smd_estimates_clonal_tmb <- get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "pyclone_clonal_TMB")

smd_estimates_subclonal_tmb <- get_trial_smd_estimates(
  data = wes_features_df,
  trial_list = compare_r_nr_trials,
  variable = "pyclone_subclonal_TMB")

smd_estimates_aneuploidy_score <- get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "Aneuploidy_score")

smd_estimates_scna_loss <-  get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "scna_loss")

smd_estimates_ploidy <-  get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "Ploidy")

# smd_estimates_purity <-  get_trial_smd_estimates(
#   data = facets_data_metrics, trial_list = compare_r_nr_trials,
#   variable = "Purity")

wes_features_df %>%
  mutate(Purity = ifelse((is.na(Purity) & ! is.na(Imputed_purity)),
    Imputed_purity, Purity)) %>%
              select(c(Purity, Trial, mean_VAF, median_VAF,
                     diploid_VAF_Variants, diploid_mean_VAF,
                     diploid_median_VAF, Imputed_purity, Purity_flag)) %>% view()

smd_estimates_purity <-  get_trial_smd_estimates(
  data = wes_features_df %>%
  mutate(Purity = ifelse((is.na(Purity) & ! is.na(Imputed_purity)),
    Imputed_purity, Purity)), trial_list = compare_r_nr_trials,
  variable = "Purity")

smd_estimates_wgII <-  get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "wgII_0.6_bases")

smd_estimates_frameshift <-  get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "TMB_frameshift")

# Cosmic
smd_estimates_cosmic_dmmr <- get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "dMMR_abs")

smd_estimates_cosmic_apobec <- get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "APOBEC_abs")

wes_features_df %>% filter(Trial == "NICHE-MSI") %>% select(HRD_abs) %>% summary()

smd_estimates_cosmic_hrd <- get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "HRD_abs")

smd_estimates_cosmic_uv <- get_trial_smd_estimates(
  data = wes_features_df, trial_list = compare_r_nr_trials,
  variable = "UV_abs")

smd_estimates_tmb %>%
  write_csv2(here(output_dir, "smd_estimates_tmb.csv"))
smd_estimates_aneuploidy_score %>%
  write_csv2(here(output_dir, "smd_estimates_aneuploidy_score.csv"))
smd_estimates_scna_loss %>%
  write_csv2(here(output_dir, "smd_estimates_scna_loss.csv"))
smd_estimates_ploidy %>%
  write_csv2(here(output_dir, "smd_estimates_ploidy.csv"))
smd_estimates_purity %>%
  write_csv2(here(output_dir, "smd_estimates_purity.csv"))
smd_estimates_wgII %>%
  write_csv2(here(output_dir, "smd_estimates_wgII.csv"))
smd_estimates_frameshift %>%
  write_csv2(here(output_dir, "smd_estimates_frameshift.csv"))

smd_estimates_cosmic_dmmr %>%
  write_csv2(here(output_dir, "smd_estimates_cosmic_dmmr.csv"))
smd_estimates_cosmic_apobec %>%
  write_csv2(here(output_dir, "smd_estimates_cosmic_apobec.csv"))
smd_estimates_cosmic_hrd %>%
  write_csv2(here(output_dir, "smd_estimates_cosmic_hrd.csv"))
smd_estimates_cosmic_uv %>%
  write_csv2(here(output_dir, "smd_estimates_cosmic_uv.csv"))

smd_estimates_clonal_tmb %>%
  write_csv2(here(output_dir, "smd_estimates_clonal_tmb.csv"))

smd_estimates_subclonal_tmb %>%
  write_csv2(here(output_dir, "smd_estimates_subclonal_tmb.csv"))

smd_estimates_neoantigens %>%
  write_csv2(here(output_dir, "smd_estimates_neoantigens.csv"))

combined_wes_smd <- smd_estimates_tmb %>%
  rbind(smd_estimates_aneuploidy_score) %>%
  rbind(smd_estimates_scna_loss) %>%
  rbind(smd_estimates_ploidy) %>%
  rbind(smd_estimates_purity) %>%
  rbind(smd_estimates_cosmic_apobec) %>%
  rbind(smd_estimates_cosmic_hrd) %>%
  rbind(smd_estimates_cosmic_uv) %>%
  rbind(smd_estimates_cosmic_dmmr) %>%
  rbind(smd_estimates_wgII) %>%
  rbind(smd_estimates_frameshift) %>%
  rbind(smd_estimates_clonal_tmb) %>%
  rbind(smd_estimates_subclonal_tmb) %>%
  rbind(smd_estimates_neoantigens)

combined_wes_smd <- combined_wes_smd %>%
  mutate(effect_size_cohens_d = effect_size) %>%
  mutate(weighted_pooled_sd = sqrt(
    ((sd_responders^2 * (n_responders - 1)) + (sd_nresponders^2 * (n_nresponders - 1))) / (n_responders + n_nresponders - 2)
    )
    ) %>%
  mutate(effect_size_hedges_g = mean_delta / weighted_pooled_sd) %>%
  mutate(effect_size_hedges_g_c = ifelse(
    n < 50, effect_size_hedges_g * ((n - 3) / (n - 2.25)) * sqrt((n - 2)/n), effect_size_hedges_g))

combined_wes_smd %>% write_csv2(here(output_dir, "combined_wes_smd.csv"))
combined_wes_smd <- read_csv2(here(output_dir, "combined_wes_smd.csv"))

compare_r_nr_trials
combined_wes_smd$Trial <- factor(
  combined_wes_smd$Trial, compare_r_nr_trials)

# c("Combined", "Combined_excl_dMMR", "BELLINI",
                            # "IMCISION", "MATISSE", "NABUCCO", "NICHE-MSS",
                            # "OpACIN_neo", "PRADO")

# Not resuts from meta-analysis call - individual SMD estimates
combined_wes_smd %>%
  mutate(label = cut(
    p.value,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = Trial,
    y = factor(
      variable, levels = rev(
        c("TMB", "pyclone_clonal_TMB", "pyclone_subclonal_TMB",
          "pvactools_filtered_neoantigen_load", "TMB_frameshift",
          "UV_abs",
          "dMMR_abs", "APOBEC_abs",
          "HRD_abs", "Aneuploidy_score", "wgII_0.6_bases", "Ploidy",
          "scna_loss", "Purity"))),
    fill = effect_size,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

ggsave(here(output_dir, "combined_wes_smd.png"), width = 7, height = 4)
```

# WES SMD meta_analysis
```{r, fig.height = 8, fig.width = 12}
safe_extract <- function(x) ifelse(is.null(x), NA, x)

run_smd_meta_analysis <- function(meta_analysis_df, variable_list,
                                  output_suffix, print_plot = TRUE,
                                  save_plot = TRUE) {
  
  meta_studies_results <- data.frame(matrix(nrow = 0, ncol = 0))
  meta_combined_effects <- data.frame(matrix(nrow = 0, ncol = 0))
  current_variable <- "TMB" 
  for (current_variable in unique(variable_list)) {
    print(current_variable)
    cat("  \n")
    
    current_contrast = "R x NR"
    
    meta_analysis_df_var <- meta_analysis_df %>%
      filter(variable == current_variable)
        
    meta_analysis_res <- metacont(n.e = meta_analysis_df_var$n_responders,
                              mean.e = meta_analysis_df_var$mean_responders,
                              sd.e = meta_analysis_df_var$sd_responders,
                              n.c = meta_analysis_df_var$n_nresponders,
                              mean.c = meta_analysis_df_var$mean_nresponders,
                              sd.c = meta_analysis_df_var$sd_nresponders,
                              studlab = meta_analysis_df_var$Trial,
                              sm = "SMD",
                              method.smd = "Hedges",
                              prediction = TRUE,
                              common = FALSE,
                              random = TRUE,
                              title = sprintf(
                                "%s %s", current_variable, current_contrast))
    
    meta_random_eff <- data.frame(
      Variable = current_variable,
      Contrast = current_contrast, 
      sm = meta_analysis_res$sm,
      level.ma = meta_analysis_res$level.ma,
      null.effect = meta_analysis_res$null.effect, 
      k = meta_analysis_res$k,
      method = meta_analysis_res$method,
      TE.random = meta_analysis_res$TE.random, # Random-effects model
      seTE.random = meta_analysis_res$seTE.random,
      statistic.random = meta_analysis_res$statistic.random,
      pval.random = meta_analysis_res$pval.random,
      method.random.ci = meta_analysis_res$method.random.ci,
      df.random = meta_analysis_res$df.random,
      lower.random = meta_analysis_res$lower.random,
      upper.random = meta_analysis_res$upper.random,
      level.predict = meta_analysis_res$level.predict,
      lower.predict = meta_analysis_res$lower.predict,
      upper.predict = meta_analysis_res$upper.predict,
      upper.predict = meta_analysis_res$upper.predict,
      method.tau = meta_analysis_res$method.tau,
      hakn = meta_analysis_res$hakn,
      TE.fixed = meta_analysis_res$TE.fixed, # Fixed-effect model (optional)
      seTE.fixed = meta_analysis_res$seTE.fixed,
      lower.fixed = meta_analysis_res$lower.fixed,
      upper.fixed = meta_analysis_res$upper.fixed,
      pval.fixed = meta_analysis_res$pval.fixed,
      Q = meta_analysis_res$Q, # Heterogeneity
      df.Q = meta_analysis_res$df.Q,
      pval.Q = meta_analysis_res$pval.Q,
      tau = meta_analysis_res$tau,
      tau2 = meta_analysis_res$tau^2,
      H = meta_analysis_res$H,
      H.lower = safe_extract(meta_analysis_res$H.lower),
      H.upper = safe_extract(meta_analysis_res$H.upper),
      I2 = meta_analysis_res$I2,
      I2.lower = safe_extract(meta_analysis_res$I2.lower),
      I2.upper = safe_extract(meta_analysis_res$I2.upper)
    )
    
    meta_analysis_signature_results <- as.data.frame(meta_analysis_res) %>%
      mutate(Variable = current_variable,
             Contrast = current_contrast)
    
    meta_studies_results <- meta_studies_results %>%
      rbind(meta_analysis_signature_results)
    
    meta_combined_effects <- meta_combined_effects %>%
      rbind(meta_random_eff)
    
    if (print_plot) {
      print(meta_analysis_res)

      forest(meta_analysis_res, studlab = TRUE, common = FALSE, random = TRUE,
       digits.mean = 1, digits.sd = 1,
       label.e = "R", label.c = "NR",
       leftlabs = c("Study", "N", "Mean", "SD", "N", "Mean", "SD"),
       allstudies = TRUE)
    }
    
    if (save_plot) {
        save_forestplot_pdf(forest_plot_data = meta_analysis_res,
       digits.mean = 1, digits.sd = 1,
       label.e = "R", label.c = "NR",
       leftlabs = c("Study", "N", "Mean", "SD", "N", "Mean", "SD"),
       allstudies = TRUE,
       title_text = sprintf("%s R vs. NR %s", current_variable, output_suffix),
                  filename = here(
                    output_dir,
                    sprintf("meta_analysis_wes_%s_%s.pdf",
                            current_variable, output_suffix)),
                  width = 12, height = 7)
        }
    
    cat("  \n")
  }
  
  meta_studies_results %>%
    write_csv2(here(output_dir, sprintf("meta_studies_results_%s.csv",
                                        output_suffix)))
  meta_combined_effects %>%
    write_csv2(here(output_dir, sprintf("meta_combined_effects_%s.csv",
                                        output_suffix)))
}

```

## Run SMD MA
```{r}
run_smd_meta_analysis(meta_analysis_df = combined_wes_smd %>%
      filter(Trial %in% c("Combined")),
      variable_list = unique(combined_wes_smd$variable),
      output_suffix = "combined",
      print_plot = FALSE, save_plot= FALSE
      )

run_smd_meta_analysis(meta_analysis_df = combined_wes_smd %>%
      filter(Trial %in% c("Combined_excl_dMMR")),
      variable_list = unique(combined_wes_smd$variable),
      output_suffix = "combined_excl_dMMR",
      print_plot = FALSE, save_plot= FALSE
      )

run_smd_meta_analysis(meta_analysis_df = combined_wes_smd %>%
      filter(!Trial %in% c("Combined", "Combined_excl_dMMR")),
      variable_list = unique(combined_wes_smd$variable),
      output_suffix = "random_effects"
      )

```

# Combine
```{r}
meta_studies_results_combined <- read_csv2(
  here(output_dir, "meta_studies_results_combined.csv"))

meta_studies_results_combined_excl_dMMR <- read_csv2(
    here(output_dir, "meta_studies_results_combined_excl_dMMR.csv"))

meta_studies_results_random_effects <- read_csv2(
  here(output_dir, "meta_studies_results_random_effects.csv"))

meta_combined_effects_random_effects <- read_csv2(
    here(wes_meta_dir, "meta_combined_effects_random_effects.csv"))

meta_studies_results_smd_all <- meta_studies_results_random_effects %>%
  full_join(meta_studies_results_combined) %>%
  full_join(meta_studies_results_combined_excl_dMMR)

meta_studies_results_smd_all %>%
  write_csv2(here(output_dir, "meta_studies_results_smd_all.csv"))

meta_studies_results_smd_all <- read_csv2(
  here(output_dir, "meta_studies_results_smd_all.csv"))
```

```{r}
meta_combined_effects_random_effects_f <- meta_combined_effects_random_effects %>%
  arrange(TE.random)

meta_combined_effects_random_effects %>%
  arrange(pval.random) %>%
  select(c(Variable, Contrast, TE.random, pval.random,
           lower.random, upper.random)) %>%
  mutate(pval.random = round(pval.random, 5)) %>%
  mutate_at(.vars = c("TE.random", "lower.random", "upper.random"),
            .funs = function(x){round(x, digits = 3)}) %>% view()

variable_order <- c("TMB", "pvactools_filtered_neoantigen_load",
                    "pyclone_clonal_TMB", "pyclone_subclonal_TMB", 
                    "APOBEC_abs", "UV_abs", "HRD_abs", "dMMR_abs",
                    "TMB_frameshift", "wgII_0.6_bases", 
                             "Aneuploidy_score", "Ploidy", "scna_loss",
                             "Purity")

meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = factor(
    Variable, levels = rev(c(variable_order))),
    y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip() + geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis SMD (95% CI)") +
  guides(size = guide_legend(nrow = 3))

ggsave(here(output_dir, "meta_combined_effects_random_effects_f.png"),
    height = 4, width = 5)
```


```{r, fig.width = 10, fig.height = 6}
meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = factor(
    Variable, levels = rev(c(variable_order))),
    y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip() + geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis SMD (95% CI)") +
  guides(size = guide_legend(nrow = 3)) +
  geom_text(
    aes(label = pval.random, y = 0.75, hjust = 0),
    data = meta_combined_effects_random_effects %>%
      mutate(pval.random = sprintf("P = %s", FormatDecimals(pval.random))))

```

```{r}
colnames(meta_studies_results_random_effects)

meta_studies_results_random_effects %>%
  arrange(pval) %>%
  select(c(Variable, Contrast, TE, seTE, pval, studlab, 
           lower, upper, n.e, n.c)) %>%
  mutate(pval = round(pval, 5)) %>%
  mutate_at(.vars = c("TE", "lower", "upper"),
            .funs = function(x){round(x, digits = 3)}) %>% view()

meta_studies_results_combined
meta_studies_results_combined_excl_dMMR


meta_studies_results_smd_all %>%
  mutate(label = cut(
    pval,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(studlab, levels = compare_r_nr_trials),
    y = factor(
      Variable, levels = rev(
        c(variable_order))),
    fill = TE,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

meta_combined_effects_random_effects %>%
  arrange(pval.random) %>%
  select(c(Variable, Contrast, TE.random, pval.random,
           lower.random, upper.random)) %>%
  mutate(pval.random = round(pval.random, 5)) %>%
  mutate_at(.vars = c("TE.random", "lower.random", "upper.random"),
            .funs = function(x){round(x, digits = 3)}) %>% view()
```

# Check Hedges G vs Cohen's d TE calculation
```{r}
test <- combined_wes_smd %>%
  left_join(meta_studies_results_smd_all %>%
               mutate(Trial = studlab, variable = Variable) %>%
  select(n.e, mean.e, sd.e, n.c, mean.c, sd.c, Trial, variable, TE, seTE, pval))

test %>%
  ggplot(aes(x = effect_size, y = TE, col = Trial, shape = variable)) +
  geom_point() + theme_Publication() +
  geom_abline(slope = 1, intercept = 0) +
  guides(shape = guide_legend(nrow = 4), col = guide_legend(nrow = 4))

test %>%
  ggplot(aes(x = effect_size_hedges_g, y = TE, col = Trial, shape = variable)) +
  geom_point() + theme_Publication() +
  geom_abline(slope = 1, intercept = 0) +
  guides(shape = guide_legend(nrow = 4), col = guide_legend(nrow = 4))

```

```{r}
meta_combined_effects_random_effects_f <- meta_combined_effects_random_effects %>%
  mutate(meta_FDR = p.adjust(pval.random, method = "BH")) %>%
  arrange(TE.random)

meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = pval.random, y = meta_FDR)) +
  geom_point() + theme_Publication()

meta_combined_effects_random_effects_f %>%
  filter(meta_FDR < 0.05) %>%
  ggplot(aes(x = pval.random, y = meta_FDR)) +
  geom_point() + theme_Publication()

meta_combined_effects_random_effects_f %>%
  arrange(-TE.random) %>% 
  select(Variable, pval.random, meta_FDR) %>%
  mutate(pval.random = round(pval.random, 2)) %>%
  mutate(meta_FDR = round(meta_FDR, 2))

```

# ----------
# Get log OR data
```{r}
get_trial_or_estimates <- function(data, trial_list, variable_list) {
  
  feature_odds_ratio_df_all <- data.frame(matrix(nrow = 0, ncol = 0))

  for (current_trial in trial_list) {
    current_data = data
    print(current_trial)

    if (! current_trial %in% c("Combined", "Combined_excl_dMMR")) {
      current_data <- data %>% filter(Trial == current_trial)
    }
    
    if (current_trial == "Combined_excl_dMMR") {
      current_data <- data %>% filter(Trial != c("NICHE-MSI"))
    }
    
  trial_feature_odds_ratio_df <- data.frame(matrix(nrow = 0, ncol = 0))
  
  for (current_feature in variable_list) {
    
    current_data_pos <- current_data[current_data[, current_feature, drop = TRUE], ]
    current_data_neg <- current_data[!current_data[, current_feature, drop = TRUE], ]
    print(
      sprintf("current_feature: %s, n_pos = %s, n_neg = %s",
              current_feature, nrow(current_data_pos), nrow(current_data_neg)))       
    
    if (nrow(current_data_pos) > 0 & nrow(current_data_neg) > 0) {
      test <- fisher.test(table(current_data[, current_feature, drop = TRUE],
                    current_data$Response))
    
    odds_ratio_df <- broom::tidy(test) %>%
      mutate(feature = current_feature,
             n_feature_pos = nrow(current_data_pos),
             n_feature_neg = nrow(current_data_neg),
             total = n_feature_pos + n_feature_neg,
             n_pos_r = nrow(current_data_pos %>%
                             filter(Response == "R")),
             n_pos_nr = nrow(current_data_pos %>%
                              filter(Response == "NR")),
             n_neg_r = nrow(current_data_neg %>%
                             filter(Response == "R")),
             n_neg_nr = nrow(current_data_neg %>%
                              filter(Response == "NR")),
             log_or = log(estimate),
             log_or_se = sqrt((1/n_pos_r + 1/n_pos_nr + 1/n_neg_r + 1/n_neg_nr))) %>%
      relocate(c(log_or, log_or_se), .after = estimate)
  
    } 
    else {
      odds_ratio_df <- data.frame(
        estimate = NA,
        p.value = NA,
        conf.low = NA,
        conf.high = NA,
        method = "Not tested",
        alternative = "NA",
        feature = current_feature,
        n_feature_pos = nrow(current_data_pos),
        n_feature_neg = nrow(current_data_neg),
        total = nrow(current_data_pos) + nrow(current_data_neg),
        n_pos_r = nrow(current_data_pos %>%
                        filter(Response == "R")),
        n_pos_nr = nrow(current_data_pos %>%
                         filter(Response == "NR")),
        n_neg_r = nrow(current_data_neg %>%
                        filter(Response == "R")),
        n_neg_nr = nrow(current_data_neg %>%
                         filter(Response == "NR")),
        log_or = NA,
        log_or_se = NA) %>%
        relocate(c(log_or, log_or_se), .after = estimate)
    }
  
    trial_feature_odds_ratio_df <- trial_feature_odds_ratio_df %>% rbind(odds_ratio_df) 
  }
  
  trial_feature_odds_ratio_df <- trial_feature_odds_ratio_df %>%
    mutate(Trial = current_trial) %>%
    relocate(c(Trial, feature), .before = estimate)
    
    # Add trial odds ratio data to others
    feature_odds_ratio_df_all <- feature_odds_ratio_df_all %>%
      rbind(trial_feature_odds_ratio_df)
  }
  
  feature_odds_ratio_df_all
}

colnames(pathway_level_mutation_df)[str_detect(colnames(pathway_level_mutation_df), "SMG")]
colnames(pathway_level_mutation_df)[str_detect(colnames(pathway_level_mutation_df), "SIGPW_")]

feature_odds_ratio_list <- c("RAS_RAF", "BRCA1_2",
                      "TP53", "GIE_AP", "GIE_IFNy", "SIGPW_Cell_cycle",
                      "SIGPW_Wnt_B_Cat",  "SIGPW_TGFB_signaling",
                      "SIGPW_PI3K_signaling", "HLA_alteration")

table(pathway_level_mutation_df$SETDB1)
table(pathway_level_mutation_df$CD58)
table(pathway_level_mutation_df$SIGPW_Cell_cycle)
table(pathway_level_mutation_df$SIGPW_TGFB_signaling)
table(pathway_level_mutation_df$SIGPW_PI3K_signaling)
table(pathway_level_mutation_df$SIGPW_TP53)
table(pathway_level_mutation_df$SIGPW_RTK_RAS)
table(pathway_level_mutation_df$SIGPW_Wnt_B_Cat)

wes_features_log_or <- wes_features_df %>%
  mutate(HLA_alteration = ifelse(HLA_grouped %in% c("HLA-LOH", "HLA-DEL"),
                                 "HLA-Altered", HLA_grouped)) %>%
  mutate(HLA_alteration = HLA_alteration == "HLA-Altered") %>%
  mutate(ploidy_WGD = ploidy_WGD == "WGD") %>%
  mutate(JAK1_JAK2 = JAK1 | JAK2,
         RAS_RAF = BRAF | KRAS | NRAS | HRAS,
         BRCA1_2 = BRCA1 | BRCA2
         )

table(wes_features_log_or$HLA_grouped, wes_features_log_or$HLA_alteration)

table(wes_features_log_or$HLA_alteration)

table(wes_features_log_or$B2M)
table(wes_features_log_or$JAK1_JAK2)
table(wes_features_log_or$BRCA1_2)
table(wes_features_log_or$RAS_RAF)


table(wes_features_log_or$Trial, wes_features_log_or$B2M)
table(wes_features_log_or$Trial, wes_features_log_or$JAK1_JAK2)
table(wes_features_log_or$Trial, wes_features_log_or$BRCA1_2)
table(wes_features_log_or$Trial, wes_features_log_or$RAS_RAF)

table(wes_features_log_or$Trial, wes_features_log_or$SIGPW_PI3K_signaling)
table(wes_features_log_or$Trial, wes_features_log_or$SIGPW_Wnt_B_Cat)
table(wes_features_log_or$Trial, wes_features_log_or$SIGPW_TGFB_signaling)
table(wes_features_log_or$Trial, wes_features_log_or$SIGPW_TP53)

table(wes_features_log_or$Trial, wes_features_log_or$SMG_Cell_cycle)
table(wes_features_log_or$Trial, wes_features_log_or$SMG_Genome_integrity)
table(wes_features_log_or$Trial, wes_features_log_or$SMG_MAPK_signaling)

facets_data_metrics$ploidy_WGD %>% unique()

trial_odds_ratio_df <- get_trial_or_estimates(data = wes_features_log_or,
                                              trial_list = compare_r_nr_trials,
                                              variable_list = feature_odds_ratio_list)


trial_odds_ratio_df %>%
  filter(Trial == "Combined") %>%
  mutate(estimate = round(estimate, 2), p.value = round(p.value, 3)) %>%
  arrange(-estimate) %>%
  forestplot(labeltext = c(feature, n_feature_pos, n_feature_neg, estimate, p.value),
             mean = "estimate", lower = "conf.low", upper = "conf.high",
             clip = c(0.1, 20),
             zero = 1,
             boxsize = 0.375,
             xlog = TRUE,
             xlab = "log(OR) (95% CI)") %>%
  fp_set_style(box = "royalblue",
               line = "darkblue",
               summary = "royalblue",
               txt_gp = fpTxtGp(ticks = gpar(cex = 1),
                                xlab  = gpar(cex = 1))
               ) %>% 
  fp_set_zebra_style("#EFEFEF") %>%
  fp_add_header(feature = c("Feature"),
                estimate = c("OR"),
                n_feature_pos = c("n pos"),
                n_feature_neg = c("n neg"),
                p.value = c("p.value")) %>%
  fp_add_lines()

trial_odds_ratio_df %>%
  filter(Trial == "Combined_excl_dMMR") %>%
  mutate(estimate = round(estimate, 2), p.value = round(p.value, 3)) %>%
  arrange(-estimate) %>%
  forestplot(labeltext = c(feature, n_feature_pos, n_feature_neg, estimate, p.value),
             mean = "estimate", lower = "conf.low", upper = "conf.high",
             clip = c(0.1, 20),
             zero = 1,
             boxsize = 0.375,
             xlog = TRUE,
             xlab = "log(OR) (95% CI)") %>%
  fp_set_style(box = "royalblue",
               line = "darkblue",
               summary = "royalblue",
               txt_gp = fpTxtGp(ticks = gpar(cex = 1),
                                xlab  = gpar(cex = 1))
               ) %>%
  fp_set_zebra_style("#EFEFEF") %>%
  fp_add_header(feature = c("Feature"),
                estimate = c("OR"),
                n_feature_pos = c("n pos"),
                n_feature_neg = c("n neg"),
                p.value = c("p.value")) %>%
  fp_add_lines()

trial_odds_ratio_df %>% write_csv2(here(output_dir, "trial_odds_ratio_df.csv"))
trial_odds_ratio_df <- read_csv2(here(output_dir, "trial_odds_ratio_df.csv"))
```

```{r}
prepare_response_or_data <- function(data, trial_list, variable) {
  
  response_variable_df_c <- data.frame(matrix(nrow = 0, ncol = 6))
  response_variable_df_c <- data.frame(Trial = "tmp")

  for (current_trial in trial_list) {
    current_data = data
    
    if (current_trial != "Combined") {
      current_data <- data %>% filter(Trial == current_trial)
    }
    
    response_variable_df <- table(
      current_data[, "Response", drop = TRUE],
      current_data[, variable, drop = TRUE]) %>%
      as.data.frame() %>%
      mutate(Var2 = ifelse(Var2 == TRUE, "variable_pos", "variable_neg")) %>% 
      setNames(c("Response", "Var2", "Freq")) %>%
      mutate(Trial = current_trial,
             Variable = variable,
             Response_variable = sprintf("%s_%s", Response, Var2)) %>%
      select(-c(Response, Var2)) %>%
      pivot_wider(names_from = Response_variable, values_from = Freq)
    
    response_variable_df_c <- response_variable_df_c %>%
      full_join(response_variable_df)
    
  }
  response_variable_df_c <- response_variable_df_c %>% filter(Trial != "tmp")
  response_variable_df_c
}

```

# WES log OR meta_analysis meta package
```{r}
run_or_meta_analysis <- function(meta_analysis_df, variable_list, output_suffix, print_plot = TRUE, save_plot = TRUE) {
  meta_log_or_results <- data.frame(matrix(nrow = 0, ncol = 0))
  meta_log_or_combined_effects <- data.frame(matrix(nrow = 0, ncol = 0))
  
  current_feature <- "TP53"
  for (current_feature in feature_odds_ratio_list) {
    print(current_feature)
    meta_analysis_res <- metagen(TE = log_or, seTE = log_or_se,
                               sm = "OR", random = TRUE, fixed = FALSE,
                               studlab = Trial,
                               data = meta_analysis_df %>%
                                 filter(feature == current_feature))
    if (print_plot) {
          print(meta_analysis_res)
          forest(meta_analysis_res, studlab = TRUE, common = FALSE, random = TRUE,
                 digits.TE = 2, digits.se = 2,)
    }
    
    if (save_plot) {
    save_forestplot_pdf(
      forest_plot_data = meta_analysis_res,
      digits.TE = 2, digits.se = 2,
      allstudies = TRUE,
      title_text = sprintf("%s Response OR %s", current_feature, output_suffix),
      filename = here(output_dir,
                      sprintf("meta_analysis_response_or_%s_%s.pdf", current_feature, output_suffix)),
      width = 12, height = 7)
    }
    
    meta_log_or_eff <- data.frame(
        feature = current_feature,
        Contrast = "Response_OR", 
        sm = meta_analysis_res$sm,
        level.ma = meta_analysis_res$level.ma,
        null.effect = meta_analysis_res$null.effect, 
        k = meta_analysis_res$k,
        method = meta_analysis_res$method,
        TE.random = meta_analysis_res$TE.random, # Random-effects model
        seTE.random = meta_analysis_res$seTE.random,
        statistic.random = meta_analysis_res$statistic.random,
        pval.random = meta_analysis_res$pval.random,
        method.random.ci = meta_analysis_res$method.random.ci,
        df.random = meta_analysis_res$df.random,
        lower.random = meta_analysis_res$lower.random,
        upper.random = meta_analysis_res$upper.random,
        level.predict = meta_analysis_res$level.predict,
        lower.predict = meta_analysis_res$lower.predict,
        upper.predict = meta_analysis_res$upper.predict,
        method.tau = meta_analysis_res$method.tau,
        hakn = meta_analysis_res$hakn,
        TE.fixed = meta_analysis_res$TE.fixed, # Fixed-effect model (optional)
        seTE.fixed = meta_analysis_res$seTE.fixed,
        lower.fixed = meta_analysis_res$lower.fixed,
        upper.fixed = meta_analysis_res$upper.fixed,
        pval.fixed = meta_analysis_res$pval.fixed,
        Q = meta_analysis_res$Q, # Heterogeneity
        df.Q = meta_analysis_res$df.Q,
        pval.Q = meta_analysis_res$pval.Q,
        tau = meta_analysis_res$tau,
        tau2 = meta_analysis_res$tau^2,
        H = meta_analysis_res$H,
        H.lower = safe_extract(meta_analysis_res$H.lower),
        H.upper = safe_extract(meta_analysis_res$H.upper),
        I2 = meta_analysis_res$I2,
        I2.lower = safe_extract(meta_analysis_res$I2.lower),
        I2.upper = safe_extract(meta_analysis_res$I2.upper)
      )
      
      meta_log_or_results <- meta_log_or_results %>%
        rbind(as.data.frame(meta_analysis_res) %>%
                mutate(feature = current_feature,
                       Contrast = "Response_OR"))
      
      meta_log_or_combined_effects <- meta_log_or_combined_effects %>%
        rbind(meta_log_or_eff)
  }
  
    meta_log_or_results %>%
        write_csv2(here(output_dir,
                        sprintf("meta_log_or_results_%s.csv",
                                output_suffix)))
    meta_log_or_combined_effects %>%
        write_csv2(here(output_dir,
                        sprintf("meta_log_or_combined_effects_%s.csv",
                                output_suffix)))
    
    meta_log_or_combined_effects %>%
      ggplot(aes(x = reorder(feature, TE.random), y = TE.random)) +
      geom_point(aes(size = -log10(pval.random)), shape = 22, 
                 fill = "darkgrey") +
      theme_Publication() +
      coord_flip() + geom_hline(yintercept = 0, linetype = "dashed") +
      geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
      labs(x = "", y = "Meta-analysis log OR (95% CI)")
    
    ggsave(here(
      output_dir, sprintf("meta_or_combined_effects_%s.png", output_suffix)),
      height = 8, width = 8)     
}

```

## Run Log OR MA
```{r}
run_or_meta_analysis(meta_analysis_df = trial_odds_ratio_df %>%
                               filter(!Trial %in% c("Combined",
                                                    "Combined_excl_dMMR")),
                     variable_list = feature_odds_ratio_list,
                     output_suffix = "random_effects")

run_or_meta_analysis(meta_analysis_df = trial_odds_ratio_df %>%
                               filter(Trial %in% c("Combined")),
                     variable_list = feature_odds_ratio_list,
                     output_suffix = "combined",
                     print_plot = FALSE, save_plot = FALSE)

run_or_meta_analysis(meta_analysis_df = trial_odds_ratio_df %>%
                               filter(Trial %in% c("Combined_excl_dMMR")),
                     variable_list = feature_odds_ratio_list,
                     output_suffix = "combined_excl_dmmr",
                     print_plot = FALSE, save_plot = FALSE)
  
```

# Combine
```{r}
meta_log_or_studies_results_combine <- read_csv2(
    here(output_dir, "meta_log_or_results_combined.csv"))

meta_log_or_studies_results_combined_excl_dMMR <- read_csv2(
    here(output_dir, "meta_log_or_results_combined_excl_dmmr.csv"))

meta_log_or_studies_results_random_effects <- read_csv2(
  here(output_dir, "meta_log_or_results_random_effects.csv"))

meta_log_or_combined_effects_random_effects <- read_csv2(
  here(output_dir, "meta_log_or_combined_effects_random_effects.csv")) %>%
  mutate(Variable = feature)

meta_studies_results_or_all <- meta_log_or_studies_results_random_effects %>%
  full_join(meta_log_or_studies_results_combined_excl_dMMR) %>%
  full_join(meta_log_or_studies_results_combine) %>%
  mutate(Variable = feature)

meta_studies_results_or_all %>%
  write_csv2(here(output_dir, "meta_studies_results_or_all.csv"))

meta_studies_results_or_all <- read_csv2(
  here(output_dir, "meta_studies_results_or_all.csv"))
```

```{r}
log_or_variable_order <- c("GIE_AP", "TP53", "BRCA1_2", "SIGPW_Wnt_B_Cat",
                           "SIGPW_PI3K_signaling", "SIGPW_Cell_cycle",
                           "SIGPW_TGFB_signaling", "GIE_IFNy", "RAS_RAF",
                           "HLA_alteration")
trial_odds_ratio_df %>%
  mutate(label = cut(
    p.value,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  )) %>%
  # filter(gene %in% meta_combined_effects_random_effects_f$gene) %>% 
  ggplot(aes(x = factor(Trial,
                        compare_r_nr_trials),
             y = factor(feature,
                        levels = rev(log_or_variable_order)),
             fill = log_or, 
             label = label)) +
  geom_tile() +
  geom_text() + 
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000",
                       na.value = "#CCCCCC") +
  labs(x = "", y = "", fill = "log(OR)") +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

meta_studies_results_or_all %>%
  mutate(label = cut(
    pval,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  )) %>%
  # filter(gene %in% meta_combined_effects_random_effects_f$gene) %>% 
  ggplot(aes(x = factor(studlab,
                        compare_r_nr_trials),
             y = factor(feature,
                        levels = rev(log_or_variable_order)),
             fill = TE, 
             label = label)) +
  geom_tile() +
  geom_text() + 
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000",
                       na.value = "#CCCCCC") +
  labs(x = "", y = "", fill = "log(OR)") +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


ggsave(here(output_dir, "response_log_or_combined_effects.png"),
       height = 10, width = 8)
```

```{r, fig.width = 9, fig.height = 4.5}
meta_log_or_combined_effects_random_effects %>%
  arrange(-TE.random) %>%
  pull(feature)

meta_log_or_combined_effects_random_effects  %>%
  ggplot(aes(x = factor(feature, levels = rev(log_or_variable_order)),
    y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip(clip = "off") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis SMD (95% CI)") +
  guides(size = guide_legend(nrow = 3)) +
  geom_text(
    aes(label = pval.random, y = 1.5, hjust = 0),
    data = meta_log_or_combined_effects_random_effects %>%
      mutate(pval.random = sprintf("P = %s", FormatDecimals(pval.random))))
```


# Combine OR and SMD WES

```{r}
meta_combined_effects_random_effects_wes <- meta_combined_effects_random_effects %>%
  arrange(TE.random) %>%
  full_join(meta_log_or_combined_effects_random_effects %>%
              select(-feature) %>%
  arrange(TE.random))

# "subclonal_TMB"
order_variables <- c("TMB", "clonal_TMB", "clonal_TMB_fraction", 
                     "APOBEC_abs", "UV_abs", "dMMR_abs",
  "TMB_frameshift", "HRD_abs", "Ploidy",
  "Aneuploidy_score", "wgII_0.6_bases",
  "scna_loss", "Purity", "GIE_AP", "TP53", "BRCA1_2", "SIGPW_Wnt_B_Cat",
  "SIGPW_PI3K_signaling", "SIGPW_Cell_cycle",
  "SIGPW_TGFB_signaling", "GIE_IFNy", "RAS_RAF", "HLA_alteration")

colnames(meta_combined_effects_random_effects)[!colnames(meta_combined_effects_random_effects) %in% colnames(meta_log_or_combined_effects_random_effects)]
colnames(meta_log_or_combined_effects_random_effects)[!colnames(meta_log_or_combined_effects_random_effects) %in% colnames(meta_combined_effects_random_effects)]

meta_combined_effects_random_effects_wes <- meta_combined_effects_random_effects_wes %>%
  mutate(meta_FDR = p.adjust(pval.random, method = "BH"))

meta_combined_effects_random_effects_wes %>%
  ggplot(aes(x = pval.random, y = meta_FDR)) +
  geom_point() + theme_Publication()

meta_combined_effects_random_effects_wes %>%
  filter(meta_FDR < 0.05) %>%
  ggplot(aes(x = pval.random, y = meta_FDR)) +
  geom_point() + theme_Publication()

meta_combined_effects_random_effects_wes %>%
  arrange(-TE.random) %>% 
  select(Variable, pval.random, meta_FDR) %>%
  mutate(pval.random = round(pval.random, 2)) %>%
  mutate(meta_FDR = round(meta_FDR, 2))
```

## Heterogeneity statistics
```{r}
meta_combined_effects_random_effects_wes %>%
  ggplot(aes(x = TE.random, y = TE.fixed)) + geom_point() +
  theme_Publication() + geom_abline(slope = 1, intercept = 0)

meta_combined_effects_random_effects_wes %>%
  ggplot(aes(x = Q, y = -log10(pval.Q))) + geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = Q, y = -log10(pval.Q), label = Variable),
                  data = meta_combined_effects_random_effects_wes)

meta_combined_effects_random_effects_wes %>%
  ggplot(aes(x = tau2, y = -log10(pval.Q))) + geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = tau2, y = -log10(pval.Q), label = Variable),
                  data = meta_combined_effects_random_effects_wes)

meta_combined_effects_random_effects_wes %>%
  ggplot(aes(x = I2, y = -log10(pval.Q))) + geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = I2, y = -log10(pval.Q), label = Variable),
                  data = meta_combined_effects_random_effects_wes)

meta_combined_effects_random_effects_wes %>%
  filter(pval.random < 0.05) %>%
  ggplot(aes(x = I2, y = -log10(pval.Q))) + geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = I2, y = -log10(pval.Q), label = Variable),
                  data = meta_combined_effects_random_effects_wes %>%
  filter(pval.random < 0.05))

meta_combined_effects_random_effects_wes %>%
  ggplot(aes(x = Q, y = I2)) + geom_point() + theme_Publication() +
  geom_text_repel(aes(x = Q, y = I2, label = Variable),
                  data = meta_combined_effects_random_effects_wes)

meta_combined_effects_random_effects_wes %>%
  ggplot(aes(x = tau2, y = I2)) + geom_point() + theme_Publication() +
  geom_text_repel(aes(x = tau2, y = I2, label = Variable),
                  data = meta_combined_effects_random_effects_wes)

meta_combined_effects_random_effects_wes %>%
  mutate(label = cut(
    pval.Q,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor("I2"), 
    y = factor(
      Variable, levels = rev(order_variables)),
    fill = I2,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

ggsave(here(output_dir, "meta_combined_effects_random_effects_het.png"),
    height = 9, width = 4.5)

heterogeneity_plot <- meta_combined_effects_random_effects_wes %>%
  mutate(label = cut(
    pval.Q,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor("I2"), 
    y = factor(
      Variable, levels = rev(order_variables)),
    fill = I2,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

graph2pdf(heterogeneity_plot,
          here(output_dir, "meta_combined_effects_random_effects_het.pdf"),
    height = 9, width = 4.5)
```

# ----------
# Lint script
```{r}
# lint("~/NKI_pancancer_neoadj_io/src/pancan_WES_meta.Rmd")
lint(filename = "~/NKI_pancancer_neoadj_io/src/pancan_WES_meta.Rmd",
  linters = unused_import_linter()
)
```

# ----------

# sessionInfo - Package Versions
```{r sessionInfo}
sessionInfo()
```
