---
title: "neoadjuvant pancancer parse WES pvactools and pyclone"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  export: TRUE
  output_dir: "pyclone_processed"
  metadata: ""
  filter_and_parse_vcfs: FALSE
---

Author comment:
File description comment:
The purpose of this script is to parse results from WES analysis
Additional tools, pycline VI runs, pvactools and HLA-LA typing info

# Loading Packages
```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# Install if necessary install.packages("tidyverse")
# Install if necessary BiocManager::install("DESeq2")

# library("devtools")
# devtools::install("~/Tools/TMBleR")
# BiocManager::install("VariantAnnotation")
# BiocManager::install("maftools")

library(tidyverse)
library(ICAMS)
source("helper_functions.R")
library(VariantAnnotation)
library(ggpubr)
library(corrplot)
library(dplyr)
library(readr)
library(stringr)
library(purrr)

library(here)
library(ggpubr)
MakeDataFrameFromVCF <- getFromNamespace("MakeDataFrameFromVCF", "ICAMS")

set.seed(42)
select <- dplyr::select
rename <- dplyr::rename
here <- here::here
```

# Setup output folder
```{r Setup output, echo = FALSE, include = FALSE}
export <- params$export
filter_and_parse_vcfs <- params$filter_and_parse_vcfs

# generate output directory
output_dir <- here("data", "processed", "WES_analysis")
metadata_dir <- here("data", "processed", "metadata")

print(output_dir)
# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)

} else {
  print("Output folder already exists")
}

```

# ----------
# Load metadata

```{r Load metadata}
dna_metadata <- read_csv2(here(metadata_dir, "combined_dna_metadata.csv"))

histology_df <- data.frame(
  Trial = c("BELLINI", "IMCISION", "NABUCCO", "NICHE-MSS", "NICHE-MSI",
            "OpACIN_neo", "PRADO", "MATISSE"),
  Histology = c("TBNC", "HNSCC", "Urothelial", "Colon", "Colon",
                "LN_melanoma", "LN_melanoma", "cSCC"))

dna_metadata$Response %>% table()
dna_metadata$Response %>% table() %>% prop.table() %>% round(2)
dna_metadata$Trial %>% table()

```

# Annotation color
```{r}
Color_NR <- "#F8766D"
Color_R <- "#00BA38" # "#619CFF"
Color_NA <- "#CD9600"
Color_Pre <- "#619CFF"
Color_Post <- "#F564E3"

# create colours for each group
annotation_color_list = list(
  Response = c(NR = Color_NR, R = Color_R),
  Treatment = c(Pre = Color_Pre, Post = Color_Post),
  Histology = c(LN_melanoma = "#00BFC4", Urothelial = "#F564E3",
                Colon = "#F8766D", HNSCC = "#00BA38", TBNC = "#619CFF",
                cSCC = "#CD9600"))

```

# -----------------------------
# Neoantigen_pipeline_runs updated pipeline
```{r}
pvactools_dir <- "/shared/data/niche/Neoantigen_pipeline_runs/pvactools_output"
list.files(pvactools_dir)

```

# Load metadata
```{r}
list.files(here("data", "processed", "metadata"))

patient_metadata_combined <- read_csv2(
  here("data", "processed", "metadata", "patient_metadata_combined.csv")) 

combined_dna_metadata <- read_csv2(
  here("data", "processed", "metadata", "combined_dna_metadata.csv")) 

combined_rna_metadata <- read_csv2(
  here("data", "processed", "metadata", "combined_rna_metadata.csv")) 

combined_rna_metadata_pre <- combined_rna_metadata %>%
  filter(Treatment == "Pre")

```

# Neoantigen_output
## BELLINI Neoantigen_output

```{r}

bellini_pvactools_dir <- here(pvactools_dir, "BELLINI", "pvactools")
list.files(bellini_pvactools_dir)

combined_filtered_df <- data.frame(matrix(ncol = 0, nrow = 0))
for (current_sample in list.files(bellini_pvactools_dir)) {
  print(current_sample)
  output_files <- list.files(here(bellini_pvactools_dir,
                                  current_sample, "MHC_Class_I"))
  
  output_files[str_detect(output_files, "filtered.tsv")]
  sample_filtered_df <- read_delim(
    here(bellini_pvactools_dir,current_sample, "MHC_Class_I",
         output_files[str_detect(output_files, "filtered.tsv")])) %>%
    mutate(pvactools_sample = current_sample)
  
  combined_filtered_df <- combined_filtered_df %>% rbind(sample_filtered_df)
}

n_distinct(combined_filtered_df$pvactools_sample)
table(combined_filtered_df$pvactools_sample)

bellini_combined_filtered_df <- combined_filtered_df

bellini_cf_samples <- bellini_combined_filtered_df$pvactools_sample
bellini_cf_samples <- gsub("-vs-.*", "", bellini_cf_samples)
bellini_cf_samples <- str_split(bellini_cf_samples, "_") %>% unlist()
bellini_cf_samples <- bellini_cf_samples[str_detect(bellini_cf_samples, "CF")]
bellini_combined_filtered_df$CF_sample <- bellini_cf_samples

table(str_detect(bellini_combined_filtered_df$pvactools_sample,
                 bellini_combined_filtered_df$CF_sample))
```

## MATISSE Neoantigen_output

```{r}
matisse_pvactools_dir <- here(pvactools_dir, "MATISSE", "pvactools")
list.files(matisse_pvactools_dir)

combined_filtered_df <- data.frame(matrix(ncol = 0, nrow = 0))
for (current_sample in list.files(matisse_pvactools_dir)) {
  print(current_sample)
  output_files <- list.files(
    here(matisse_pvactools_dir, current_sample, "MHC_Class_I"))
  
  output_files[str_detect(output_files, "filtered.tsv")]
  sample_filtered_df <- read_delim(
    here(matisse_pvactools_dir,
         current_sample, "MHC_Class_I",
         output_files[str_detect(output_files, "filtered.tsv")])) %>%
    mutate(pvactools_sample = current_sample)
  
  combined_filtered_df <- combined_filtered_df %>% rbind(sample_filtered_df)
}

n_distinct(combined_filtered_df$pvactools_sample)
table(combined_filtered_df$pvactools_sample)

matisse_combined_filtered_df <- combined_filtered_df

matisse_cf_samples <- matisse_combined_filtered_df$pvactools_sample
matisse_cf_samples <- gsub("-vs-.*", "", matisse_cf_samples)
matisse_cf_samples <- str_split(matisse_cf_samples, "_") %>% unlist()
matisse_cf_samples <- matisse_cf_samples[str_detect(matisse_cf_samples, "Nr")]
matisse_combined_filtered_df$CF_sample <- matisse_cf_samples

table(str_detect(matisse_combined_filtered_df$pvactools_sample,
                 matisse_combined_filtered_df$CF_sample))

```


## IMCISION Neoantigen_output

```{r}
imcision_pvactools_dir <- here(pvactools_dir, "IMCISION", "pvactools")
list.files(imcision_pvactools_dir)

combined_filtered_df <- data.frame(matrix(ncol = 0, nrow = 0))
for (current_sample in list.files(imcision_pvactools_dir)) {
  print(current_sample)
  output_files <- list.files(
    here(imcision_pvactools_dir, current_sample, "MHC_Class_I"))
  
  output_files[str_detect(output_files, "filtered.tsv")]
  sample_filtered_df <- read_delim(
    here(imcision_pvactools_dir, current_sample, "MHC_Class_I",
         output_files[str_detect(output_files, "filtered.tsv")])) %>%
    mutate(pvactools_sample = current_sample)
  
  combined_filtered_df <- combined_filtered_df %>% rbind(sample_filtered_df)
}

n_distinct(combined_filtered_df$pvactools_sample)
table(combined_filtered_df$pvactools_sample)
imcision_combined_filtered_df <- combined_filtered_df

imcision_cf_samples <- imcision_combined_filtered_df$pvactools_sample
imcision_cf_samples <- gsub("-vs-.*", "", imcision_cf_samples)
imcision_cf_samples <- str_split(imcision_cf_samples, "_") %>% unlist()
imcision_cf_samples <- imcision_cf_samples[
  str_detect(imcision_cf_samples, "Nr")]
imcision_combined_filtered_df$CF_sample <- imcision_cf_samples

table(
  str_detect(imcision_combined_filtered_df$pvactools_sample,
             imcision_combined_filtered_df$CF_sample))
```

## NABUCCO Neoantigen_output

```{r}
nabucco_pvactools_dir <- here(pvactools_dir, "NABUCCO", "pvactools") 
list.files(nabucco_pvactools_dir)

nabucco_samples <- list.files(nabucco_pvactools_dir)
nabucco_samples <- nabucco_samples[!str_detect(
  nabucco_samples, "backup_incomplete")]
length(nabucco_samples)

combined_filtered_df <- data.frame(matrix(ncol = 0, nrow = 0))
for (current_sample in nabucco_samples) {
  print(current_sample)
  output_files <- list.files(
    here(nabucco_pvactools_dir, current_sample, "MHC_Class_I"))
  
  tumor_sample <- gsub("-vs-.*", "", current_sample)
  output_files[str_detect(output_files, "filtered.tsv")]
  sample_filtered_df <- read_delim(
    here(nabucco_pvactools_dir, current_sample, "MHC_Class_I",
         output_files[str_detect(output_files, "filtered.tsv")])) %>%
    mutate(pvactools_sample = current_sample)
  
  combined_filtered_df <- combined_filtered_df %>% rbind(sample_filtered_df)
}

n_distinct(combined_filtered_df$pvactools_sample)
table(combined_filtered_df$pvactools_sample)

nabucco_combined_filtered_df <- combined_filtered_df

nabucco_cf_samples <- nabucco_combined_filtered_df$pvactools_sample
nabucco_cf_samples <- gsub("-vs-.*", "", nabucco_cf_samples)
nabucco_cf_samples <- str_split(nabucco_cf_samples, "_") %>% unlist()
nabucco_cf_samples <- nabucco_cf_samples[str_detect(nabucco_cf_samples, "CF")]
nabucco_combined_filtered_df$CF_sample <- nabucco_cf_samples

nabucco_combined_filtered_df %>% summary()

table(str_detect(nabucco_combined_filtered_df$pvactools_sample,
                 nabucco_combined_filtered_df$CF_sample))

```

## NICHE Neoantigen_output
```{r}
niche_pvactools_dir <- here(pvactools_dir, "NICHE", "pvactools") 
list.files(niche_pvactools_dir)

niche_samples <- list.files(niche_pvactools_dir)
niche_samples <- niche_samples[!str_detect(niche_samples, "backups_incomplete")]
length(niche_samples)

combined_filtered_df <- data.frame(matrix(ncol = 0, nrow = 0))
for (current_sample in niche_samples) {
  print(current_sample)
  output_files <- list.files(
    here(niche_pvactools_dir, current_sample, "MHC_Class_I"))
  
  tumor_sample <- gsub("-vs-.*", "", current_sample)
  output_files[str_detect(output_files, "filtered.tsv")]
  sample_filtered_df <- read_delim(
    here(niche_pvactools_dir, current_sample, "MHC_Class_I",
         output_files[str_detect(output_files, "filtered.tsv")])) %>%
    mutate(pvactools_sample = current_sample)
  
  combined_filtered_df <- combined_filtered_df %>% rbind(sample_filtered_df)
}

n_distinct(combined_filtered_df$pvactools_sample)
table(combined_filtered_df$pvactools_sample)

niche_combined_filtered_df <- combined_filtered_df

niche_cf_samples <- niche_combined_filtered_df$pvactools_sample
niche_cf_samples <- gsub("-vs-.*", "", niche_cf_samples)
niche_cf_samples <- str_split(niche_cf_samples, "_") %>% unlist()
niche_cf_samples <- niche_cf_samples[str_detect(niche_cf_samples, "CF")]
niche_combined_filtered_df$CF_sample <- niche_cf_samples

niche_combined_filtered_df %>% factor_summary()

table(
  str_detect(niche_combined_filtered_df$pvactools_sample,
             niche_combined_filtered_df$CF_sample))
```

## NICHE2 Neoantigen_output
```{r}
niche2_pvactools_dir <- here(pvactools_dir, "NICHE2", "pvactools") 
list.files(niche2_pvactools_dir)

niche2_samples <- list.files(niche2_pvactools_dir)
niche2_samples <- niche2_samples[!str_detect(niche2_samples,
                                             "backups_incomplete")]
length(niche2_samples)

combined_filtered_df <- data.frame(matrix(ncol = 0, nrow = 0))
for (current_sample in niche2_samples) {
  print(current_sample)
  output_files <- list.files(
    here(niche2_pvactools_dir, current_sample, "MHC_Class_I"))
  
  tumor_sample <- gsub("-vs-.*", "", current_sample)
  output_files[str_detect(output_files, "filtered.tsv")]
  sample_filtered_df <- read_delim(
    here(niche2_pvactools_dir, current_sample, "MHC_Class_I",
         output_files[str_detect(output_files, "filtered.tsv")])) %>%
    mutate(pvactools_sample = current_sample)
  
  combined_filtered_df <- combined_filtered_df %>% rbind(sample_filtered_df)
}

n_distinct(combined_filtered_df$pvactools_sample)
table(combined_filtered_df$pvactools_sample)

niche2_combined_filtered_df <- combined_filtered_df

niche2_cf_samples <- niche2_combined_filtered_df$pvactools_sample
niche2_cf_samples <- gsub("-vs-.*", "", niche2_cf_samples)
niche2_cf_samples <- str_split(niche2_cf_samples, "_") %>% unlist()
niche2_cf_samples <- niche2_cf_samples[str_detect(niche2_cf_samples, "CF")]
niche2_combined_filtered_df$CF_sample <- niche2_cf_samples

niche2_combined_filtered_df %>% factor_summary()

table(niche2_combined_filtered_df$CF_sample)
n_distinct(niche2_combined_filtered_df$CF_sample)
table(str_detect(niche2_combined_filtered_df$pvactools_sample,
                 niche2_combined_filtered_df$CF_sample))

```

## PRADO Neoantigen_output

```{r}
prado_pvactools_dir <- here(pvactools_dir, "PRADO", "MHC_run_relaxt") 

list.files(prado_pvactools_dir)

combined_filtered_df <- data.frame(matrix(ncol = 0, nrow = 0))
for (current_sample in list.files(prado_pvactools_dir)) {
  print(current_sample)
  output_files <- list.files(
    here(prado_pvactools_dir, current_sample, "MHC_Class_I"))
  
  output_files[str_detect(output_files, "filtered.tsv")]
  sample_filtered_df <- read_delim(
    here(prado_pvactools_dir, current_sample, "MHC_Class_I",
         output_files[str_detect(output_files, "filtered.tsv")])) %>%
    mutate(pvactools_sample = current_sample)
  
  combined_filtered_df <- combined_filtered_df %>% rbind(sample_filtered_df)
}

n_distinct(combined_filtered_df$pvactools_sample)
table(combined_filtered_df$pvactools_sample)
prado_combined_filtered_df <- combined_filtered_df

prado_cf_samples <- prado_combined_filtered_df$pvactools_sample
prado_cf_samples <- gsub("_.*", "", prado_cf_samples)
prado_combined_filtered_df$CF_sample <- prado_cf_samples

table(str_detect(prado_combined_filtered_df$pvactools_sample,
                 prado_combined_filtered_df$CF_sample))
```


## Opacin_neo Neoantigen_output

```{r}
opacin_neo_pvactools_dir <- here(pvactools_dir, "OpACIN_neo", "pvactools") 
list.files(opacin_neo_pvactools_dir)

combined_filtered_df <- data.frame(matrix(ncol = 0, nrow = 0))
for (current_sample in list.files(opacin_neo_pvactools_dir)) {
  print(current_sample)
  output_files <- list.files(
    here(opacin_neo_pvactools_dir, current_sample, "MHC_Class_I"))
  
  tumor_sample <- gsub("-vs-.*", "", current_sample)
  output_files[str_detect(output_files, "filtered.tsv")]
  sample_filtered_df <- read_delim(
    here(opacin_neo_pvactools_dir, current_sample, "MHC_Class_I",
         output_files[str_detect(output_files, "filtered.tsv")])) %>%
    mutate(pvactools_sample = current_sample)
  
  combined_filtered_df <- combined_filtered_df %>% rbind(sample_filtered_df)
}

n_distinct(combined_filtered_df$pvactools_sample)
table(combined_filtered_df$pvactools_sample)

opacin_neo_combined_filtered_df <- combined_filtered_df

opacin_neo_cf_samples <- opacin_neo_combined_filtered_df$pvactools_sample
opacin_neo_cf_samples <- gsub("-vs-.*", "", opacin_neo_cf_samples)
opacin_neo_combined_filtered_df$CF_sample <- opacin_neo_cf_samples

table(
  str_detect(opacin_neo_combined_filtered_df$pvactools_sample,
             opacin_neo_combined_filtered_df$CF_sample))

opacin_neo_combined_filtered_df %>% summary()

```


# Combine neoantigen
```{r}
n_distinct(prado_combined_filtered_df$pvactools_sample)
n_distinct(bellini_combined_filtered_df$pvactools_sample)
n_distinct(imcision_combined_filtered_df$pvactools_sample)
n_distinct(matisse_combined_filtered_df$pvactools_sample)
n_distinct(opacin_neo_combined_filtered_df$pvactools_sample)
n_distinct(nabucco_combined_filtered_df$pvactools_sample)
n_distinct(niche_combined_filtered_df$pvactools_sample)
n_distinct(niche2_combined_filtered_df$pvactools_sample)

neoantigen_output_combined_filtered <- prado_combined_filtered_df %>%
  full_join(bellini_combined_filtered_df) %>%
  full_join(imcision_combined_filtered_df) %>%
  full_join(matisse_combined_filtered_df) %>%
  full_join(opacin_neo_combined_filtered_df) %>%
  full_join(nabucco_combined_filtered_df) %>%
  full_join(niche_combined_filtered_df) %>%
  full_join(niche2_combined_filtered_df)

dim(neoantigen_output_combined_filtered)

# Neoantigen TMB, one per position
neoantigen_output_combined_filtered_unique_variants <- neoantigen_output_combined_filtered %>%
  select(Chromosome, Start, Stop, Reference, Variant, CF_sample) %>%
  unique()
dim(neoantigen_output_combined_filtered_unique_variants)

# Impute 0 for S1224Nr40 and S2697Nr53 
# no filtered predictions
pvactools_filtered_neoantigen_counts <- table(
  neoantigen_output_combined_filtered$CF_sample) %>%
  as.data.frame() %>%
  setNames(c("CF_sample_tumor", "pvactools_filtered_neoantigen_load")) %>%
  left_join(
    table(
      neoantigen_output_combined_filtered_unique_variants$CF_sample) %>%
      as.data.frame() %>%
      setNames(c("CF_sample_tumor", "pvactools_unique_neoantigen_sites"))) %>%
  add_row(CF_sample_tumor = "S1224Nr40",
          pvactools_filtered_neoantigen_load = 0,
          pvactools_unique_neoantigen_sites = 0) %>%
  add_row(CF_sample_tumor = "S2697Nr53",
          pvactools_filtered_neoantigen_load = 0,
          pvactools_unique_neoantigen_sites = 0) %>%
  left_join(combined_dna_metadata)

```


```{r}
pvactools_filtered_neoantigen_counts %>%
  filter(!is.na(Trial)) %>%
  ggplot(aes(x = Trial, y = pvactools_filtered_neoantigen_load)) +
  geom_boxplot() + theme_Publication()

pvactools_filtered_neoantigen_counts %>%
  filter(!is.na(Trial)) %>%
  ggplot(aes(x = Trial, y = pvactools_unique_neoantigen_sites)) +
  geom_boxplot() + theme_Publication()

pvactools_filtered_neoantigen_counts %>%
  filter(!is.na(Trial)) %>%
  ggplot(aes(x = pvactools_unique_neoantigen_sites,
             y = pvactools_filtered_neoantigen_load, col = Trial)) +
  geom_point() + theme_Publication() + geom_abline()

neoantigen_output_combined_filtered %>%
  write_csv2(here(output_dir,
                  "neoantigen_output_combined_filtered.csv"))

pvactools_filtered_neoantigen_counts %>%
  write_csv2(here(output_dir,
                  "pvactools_filtered_neoantigen_counts.csv"))

neoantigen_output_combined_filtered <- read_csv2(here(output_dir,
                  "neoantigen_output_combined_filtered.csv"))

pvactools_filtered_neoantigen_counts <- read_csv2(here(output_dir,
                  "pvactools_filtered_neoantigen_counts.csv"))

```

# -----------------------------
# Compare TMB variants

```{r}
ccf_vaf_clonal_summary_df <- read_csv2(
  here("data", "processed", "WES_analysis", "ccf_vaf_clonal_tmb_df.csv"))

table(ccf_vaf_clonal_summary_df$Trial)
table(pvactools_filtered_neoantigen_counts$Trial)

pvactools_filtered_neoantigen_counts_c <- pvactools_filtered_neoantigen_counts %>%
  left_join(ccf_vaf_clonal_summary_df %>%
              select(CF_sample_tumor, Variants, TMB, TMB_frameshift,
                     Frameshift_load, n_variants_CCF_0.80,
                     n_variants_CCF_0.90, n_variants_CCF_0.95)) %>%
  mutate(delta = Variants - pvactools_unique_neoantigen_sites) %>%
  mutate(ratio = pvactools_unique_neoantigen_sites / Variants)

pvactools_filtered_neoantigen_counts_c %>%
  select(Trial, CF_sample_tumor, Variants, pvactools_unique_neoantigen_sites,
         pvactools_filtered_neoantigen_load, delta, ratio) %>%
  view()

tapply(pvactools_filtered_neoantigen_counts_c$ratio,
       pvactools_filtered_neoantigen_counts_c$Trial,
       summary)
```

# neoantigen load (peptide level, multiple per site)

```{r}
pvactools_filtered_neoantigen_counts_c  %>%
  ggplot(aes(x = Trial, y = ratio)) +
  geom_boxplot() + theme_Publication() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

pvactools_filtered_neoantigen_counts_c %>%
  ggplot(aes(x = Variants, y = pvactools_filtered_neoantigen_load)) +
  geom_point(aes(col = Trial)) + theme_Publication() +
  stat_cor(method = "spearman") +
  geom_abline()

pvactools_filtered_neoantigen_counts_c  %>%
  ggplot(aes(x = Trial, y = delta)) +
  geom_boxplot() + theme_Publication() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

pvactools_filtered_neoantigen_counts_c %>%
  ggplot(aes(x = Trial, y = delta)) +
  geom_boxplot() + theme_Publication() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

pvactools_filtered_neoantigen_counts_c %>%
  filter(Trial %in% c("BELLINI", "IMCISION", "NICHE-MSS")) %>%
  ggplot(aes(x = Variants, y = pvactools_filtered_neoantigen_load,
             col = Trial)) +
  geom_point() + theme_Publication() + stat_cor(method = "spearman") +
  geom_abline()

pvactools_filtered_neoantigen_counts_c %>%
  filter(Trial %in% c("BELLINI", "IMCISION", "NICHE-MSS")) %>%
  mutate(delta = pvactools_filtered_neoantigen_load - Variants) %>%
  ggplot(aes(x = Trial, y = delta)) +
  geom_boxplot() + theme_Publication() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

pvactools_filtered_neoantigen_counts_c %>%
  ggplot(aes(x = TMB, y = pvactools_filtered_neoantigen_load, col = Trial)) +
  geom_point() + theme_Publication()
```

# Unique neoantigen sites/variants

```{r}
pvactools_filtered_neoantigen_counts_c %>%
  ggplot(aes(x = Variants, y = pvactools_unique_neoantigen_sites)) +
  geom_point(aes(col = Trial)) + theme_Publication() +
  stat_cor(method = "spearman") +
  geom_abline()

pvactools_filtered_neoantigen_counts_c %>%
  mutate(delta = Variants - pvactools_unique_neoantigen_sites) %>%
  ggplot(aes(x = Trial, y = delta)) +
  geom_boxplot() + theme_Publication() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

pvactools_filtered_neoantigen_counts_c %>%
  mutate(delta = pvactools_unique_neoantigen_sites - Variants) %>%
  ggplot(aes(x = Trial, y = delta)) +
  geom_boxplot() + theme_Publication() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

pvactools_filtered_neoantigen_counts_c %>%
  filter(Trial %in% c("BELLINI", "IMCISION", "NICHE-MSS")) %>%
  ggplot(aes(x = Variants, y = pvactools_unique_neoantigen_sites,
             col = Trial)) +
  geom_point() + theme_Publication() + stat_cor(method = "spearman") +
  geom_abline()

pvactools_filtered_neoantigen_counts_c %>%
  filter(Trial %in% c("BELLINI", "IMCISION", "NICHE-MSS")) %>%
  mutate(delta = pvactools_unique_neoantigen_sites - Variants) %>%
  ggplot(aes(x = Trial, y = delta)) +
  geom_boxplot() + theme_Publication() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

pvactools_filtered_neoantigen_counts_c %>%
  ggplot(aes(x = TMB, y = pvactools_unique_neoantigen_sites, col = Trial)) +
  geom_point() + theme_Publication()
```

```{r, fig.width = 10, fig.height = 10}

pvactools_filtered_neoantigen_counts_c %>%
  ggplot(aes(x = Variants, y = pvactools_filtered_neoantigen_load)) +
  geom_point(aes(col = Trial)) + theme_Publication() +
  stat_cor(method = "spearman") +
  geom_abline() + facet_wrap(~Trial, scales = "free")

pvactools_filtered_neoantigen_counts_c %>%
  ggplot(aes(x = Variants, y = pvactools_unique_neoantigen_sites)) +
  geom_point(aes(col = Trial)) + theme_Publication() +
  stat_cor(method = "spearman") +
  geom_abline() + facet_wrap(~Trial, scales = "free")
```

# -----------------------------

# Pyclone VI runs
```{r, eval = FALSE}
# Pyclone Input was exported with WES script
mkdir -p pyclone_output

for f in pyclone_input/*.tsv; do
    sample=$(basename "$f" .tsv)
    echo "Running PyClone-VI for $sample"

    pyclone-vi fit \
        --in-file "$f" \
        --out-file "pyclone_output/${sample}.h5"
done

for f in pyclone_output/*.h5; do
    sample=$(basename "$f" .h5)
    echo "Writing results for $sample"

    pyclone-vi write-results-file \
        --in-file "$f" \
        --out-file "pyclone_output/${sample}_results.tsv"
done

```

```{r}
input_dir <- "~/pyclone_output"
output_dir <- "pyclone_processed"

dir.create(output_dir, showWarnings = FALSE)

# List all PyClone-VI mutation-level result files
files <- c(list.files(input_dir, pattern = "_results.tsv$", full.names = TRUE),
           list.files("~/pyclone_output_missing/", pattern = "_results.tsv$",
                      full.names = TRUE))

length(files)

process_sample <- function(file) {
  
  df <- read_tsv(file, show_col_types = FALSE)
  
  # Extract sample name from filename
  sample_id <- str_replace(basename(file), "_results.tsv", "")
  
  # Compute cluster-level summaries
  cluster_summary <- df %>%
    group_by(cluster_id) %>%
    summarise(
      cluster_size = n(),
      mean_CCF = mean(cellular_prevalence),
      sd_CCF = sd(cellular_prevalence),
      median_CCF = median(cellular_prevalence),
      q05 = quantile(cellular_prevalence, 0.05),
      q95 = quantile(cellular_prevalence, 0.95)
    ) %>%
    arrange(desc(cluster_size)) %>%
    mutate(sample_id = sample_id)
  
  # Write cluster summary
  write_tsv(
    cluster_summary,
    file.path(output_dir, paste0(sample_id, "_cluster_summary.tsv"))
  )
  
  # Write mutation-level file (optional)
  write_tsv(
    df,
    file.path(output_dir, paste0(sample_id, "_mutation_summary.tsv"))
  )
  
  return(cluster_summary)
}

compile_pyclone_vi = FALSE
if (compile_pyclone_vi) {
  # Run for all samples
  all_cluster_summaries <- map_df(files, process_sample)
  
  all_cluster_summaries %>% write_csv2(
    here("data", "processed", "pyclone_processed", "all_cluster_summaries.csv"))
}

```

```{r}
all_cluster_summaries <- read_csv2(
  here("data", "processed", "pyclone_processed", "all_cluster_summaries.csv"))

all_cluster_summaries %>%
  ggplot(aes(x = cluster_id, y = cluster_size, col = sample_id)) +
  geom_point() + guides(col = FALSE) + theme_Publication()
all_cluster_summaries %>%
  ggplot(aes(x = cluster_id, y = mean_CCF, col = sample_id)) +
  geom_point() + guides(col = FALSE)

all_cluster_summaries_clonal <- all_cluster_summaries %>%
  mutate(clonal_status = ifelse(mean_CCF > 0.8, "Clonal", "Subclonal")) %>%
  group_by(sample_id) %>%
  add_count(name = "Total", wt = cluster_size) %>%
  ungroup()

all_cluster_summaries_clonal %>%
  mutate(cluster_prop = cluster_size / Total) %>%
  ggplot(aes(x = clonal_status, y = cluster_prop)) +
  geom_boxplot() +
  theme_Publication()

all_cluster_summaries_clonal %>%
  ggplot(aes(fill = clonal_status, x = mean_CCF)) + geom_density(alpha = 0.5) +
  theme_Publication()

sample_n_cluster_summary <- table(all_cluster_summaries_clonal$clonal_status,
                                  all_cluster_summaries_clonal$sample_id) %>%
  as.data.frame() %>%
  setNames(c("clonal_status", "sample_id", "Freq")) %>%
  pivot_wider(names_from = clonal_status, values_from = Freq) %>%
  setNames(c("sample_id", "n_clonal_clusters", y = "n_subclonal_clusters")) %>%
  mutate(n_clusters = n_clonal_clusters + n_subclonal_clusters)

sample_n_cluster_summary %>% summary()
sample_n_cluster_summary %>% filter(n_clonal_clusters == 0)
sample_n_cluster_summary %>% filter(n_subclonal_clusters == 0)
```

## Define pyclone_vi_clonal_summary_df
```{r}
pyclone_vi_clonal_summary_df <- sample_n_cluster_summary %>%
  left_join(
    all_cluster_summaries_clonal %>%
      group_by(clonal_status, sample_id, Total) %>%
      summarize(cluster_size = sum(cluster_size)) %>%
      mutate(Fraction = cluster_size/Total) %>%
      ungroup() %>%
      filter(clonal_status == "Clonal") %>%
      rename(clonal_fraction = Fraction, clonal_cluster_size = cluster_size) %>%
      select(sample_id, Total, clonal_cluster_size, clonal_fraction) 
    ) %>%
  left_join(
    all_cluster_summaries_clonal %>%
      group_by(clonal_status, sample_id, Total) %>%
      summarize(cluster_size = sum(cluster_size)) %>%
      mutate(Fraction = cluster_size/Total) %>%
      ungroup() %>%
      filter(clonal_status == "Subclonal") %>%
      rename(subclonal_fraction = Fraction,
             subclonal_cluster_size = cluster_size) %>%
      select(sample_id, subclonal_cluster_size, subclonal_fraction),
    by = "sample_id") %>%
  mutate(Total = ifelse(
    is.na(clonal_cluster_size) & !is.na(subclonal_cluster_size),
    subclonal_cluster_size, Total)) %>%
  mutate(Total = ifelse(
    is.na(subclonal_cluster_size) & !is.na(clonal_cluster_size),
    clonal_cluster_size, Total)) %>%
  mutate(clonal_cluster_size = ifelse(
    is.na(clonal_cluster_size), 0, clonal_cluster_size)) %>%
  mutate(clonal_fraction = ifelse(
    is.na(clonal_fraction), 0, clonal_fraction)) %>%
  mutate(subclonal_cluster_size = ifelse(
    is.na(subclonal_cluster_size), 0, subclonal_cluster_size)) %>%
  mutate(subclonal_fraction = ifelse(
    is.na(subclonal_fraction), 0, subclonal_fraction))
```

```{r}

patient_metadata_combined <- read_csv2(
  here("data", "processed", "metadata", "patient_metadata_combined.csv")) 

combined_dna_metadata <- read_csv2(
  here("data", "processed", "metadata", "combined_dna_metadata.csv")) 

combined_rna_metadata <- read_csv2(
  here("data", "processed", "metadata", "combined_rna_metadata.csv")) 

all_cluster_summaries_p <- all_cluster_summaries %>%
  mutate(CF_sample_tumor = sample_id) %>%
  left_join(combined_dna_metadata)

all_cluster_summaries_p %>%
  write_csv2(here("data", "processed", "pyclone_processed",
                  "all_cluster_summaries_p.csv"))

# WES target sizes
WES_target_sizes <- data.frame(
  Trial = c("IMCISION", "PRADO", "OpACIN_neo", "MATISSE",
            "NICHE-MSI", "NICHE-MSS", "NABUCCO", "BELLINI"),
  Kit = c(rep("Twist exome Refseq", 4), rep("Exome-IDT V1", 4)),
  WES_target_size = c(rep(36.72, 4), rep(39.02, 4)))

pyclone_vi_clonal_summary_df <- pyclone_vi_clonal_summary_df %>%
  mutate(CF_sample_tumor = sample_id) %>%
  left_join(combined_dna_metadata) %>%
  left_join(WES_target_sizes) %>%
  mutate(pyclone_clonal_TMB = clonal_cluster_size / WES_target_size,
         pyclone_subclonal_TMB = subclonal_cluster_size / WES_target_size)

pyclone_vi_clonal_summary_df %>%
  write_csv2(here(output_dir, "pyclone_vi_clonal_summary_df.csv"))

pyclone_vi_clonal_summary_df <- read_csv2(
  here(output_dir, "pyclone_vi_clonal_summary_df.csv"))

```

## Plots pyclone TMB

```{r}

pyclone_vi_clonal_summary_df %>%
  ggplot(aes(x = clonal_fraction, y = subclonal_fraction, col = n_clusters)) +
  geom_point() + theme_Publication()

pyclone_vi_clonal_summary_df %>%
  ggplot(aes(x = n_clonal_clusters, y = clonal_fraction, col = n_clusters)) +
  geom_point() + theme_Publication()

pyclone_vi_clonal_summary_df %>%
  ggplot(aes(x = n_subclonal_clusters, y = subclonal_fraction,
             col = n_clusters)) +
  geom_point() + theme_Publication()

pyclone_vi_clonal_summary_df %>%
  ggplot(aes(x = n_subclonal_clusters, y = n_clonal_clusters)) +
  geom_pointdensity() + theme_Publication()

pyclone_vi_clonal_summary_df %>%
  filter(subclonal_fraction == 1) %>% ggplot(aes(x = n_subclonal_clusters)) +
  geom_histogram() + theme_Publication()

pyclone_vi_clonal_summary_df %>%
  filter(clonal_fraction == 1) %>% ggplot(aes(x = n_clonal_clusters)) +
  geom_histogram() + theme_Publication()

pyclone_vi_clonal_summary_df %>%
  ggplot(aes(x = Trial, y = n_clusters)) +
  geom_boxplot() + theme_Publication() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
  
pyclone_vi_clonal_summary_df %>% ggplot(aes(x = Trial, y = clonal_fraction)) +
  geom_boxplot() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

```{r}
pyclone_vi_clonal_summary_df %>%
  ggplot(aes(x = Trial,
             y = pyclone_clonal_TMB, fill = Histology)) +
  geom_boxplot() + theme_Publication() +
  guides(fill = FALSE) +
  labs(x = "", y = "clonal TMB (mutations/Mb)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_fill_manual(values = c("#B73A3A", "#CD9600", "#23C26F", "#00BFC4",
                               "#009292", "#B66DFF", "#FEB5DA"))

pyclone_vi_clonal_summary_df %>%
  ggplot(aes(x = Trial,
             y = pyclone_subclonal_TMB, fill = Histology)) +
  geom_boxplot() + theme_Publication() +
  guides(fill = FALSE) +
  labs(x = "", y = "Subclonal TMB (mutations/Mb)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_fill_manual(values = c("#B73A3A", "#CD9600", "#23C26F", "#00BFC4",
                               "#009292", "#B66DFF", "#FEB5DA"))
pyclone_vi_clonal_summary_df %>%
  ggplot(aes(x = pyclone_clonal_TMB,
             y = pyclone_subclonal_TMB, color = Histology)) +
  geom_point() + theme_Publication() +
  guides(fill = FALSE) +
  labs(x = "pyclone_clonal_TMB", y = "pyclone_subclonal_TMB") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_fill_manual(values = c("#B73A3A", "#CD9600", "#23C26F", "#00BFC4",
                               "#009292", "#B66DFF", "#FEB5DA")) +
  geom_hline(yintercept = 10, linetype = "dashed") +
  geom_vline(xintercept = 10, linetype = "dashed")


```

```{r, fig.height = 12, fig.width = 12}
pyclone_vi_clonal_summary_df %>%
  ggplot(aes(x = pyclone_clonal_TMB,
             y = pyclone_subclonal_TMB, color = Histology)) +
  geom_point() + theme_Publication() +
  guides(fill = FALSE) +
  labs(x = "pyclone_clonal_TMB", y = "pyclone_subclonal_TMB") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_fill_manual(values = c("#B73A3A", "#CD9600", "#23C26F", "#00BFC4",
                               "#009292", "#B66DFF", "#FEB5DA")) +
  facet_wrap(~Trial, scales = "free") +
  geom_hline(yintercept = 10, linetype = "dashed") +
  geom_vline(xintercept = 10, linetype = "dashed")
```

# ----------
# Lint script
```{r}
# lint("~/NKI_pancancer_neoadj_io/src/pancan_parse_pyclone_pvactools.Rmd")
lint(
  filename = "~/NKI_pancancer_neoadj_io/src/pancan_parse_pyclone_pvactools.Rmd",
  linters = unused_import_linter()
)
```

# ----------
# sessionInfo - Package Versions
```{r sessionInfo}
sessionInfo()
```
