---
title: "neoadjuvant pancancer figures"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  export: TRUE
  output_dir: "_pancan_figures"
  metadata: "niche_gcf_metadata_exc_sens.csv"
  filter_and_parse_vcfs: FALSE
---

Author comment:
File description comment:
The purpose of this script is to produce figure panels, loading the source data

# Loading Packages
```{r}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

library(tidyverse)
library(readxl)
library(maftools)
source("helper_functions.R")
# source("oncomatrix.R")
library(VariantAnnotation)
library(ggpubr)
library(corrplot)
library(ggpointdensity)
library(export)
library(meta)
library(export)
library(ggpattern)
library(scales)
library(lintr)
library(pROC)
MakeDataFrameFromVCF <- getFromNamespace("MakeDataFrameFromVCF", "ICAMS")

set.seed(42)
select <- dplyr::select
rename <- dplyr::rename
here <- here::here
```

# Setup output folder
```{r Setup output, echo = FALSE, include = FALSE}
export <- params$export
filter_and_parse_vcfs <- params$filter_and_parse_vcfs

# generate output directory
output_dir <- here("data", "processed", "_pancan_figures")
source_data_dir <- here("data", "processed", "source_data")

print(output_dir)

# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)

    } else {
  print("Output folder already exists")
}

```


# ----------
# Define import directories
```{r}
# metadata_dir <- here("data", "processed", "metadata")
# wes_dir <- here("data", "processed", "WES_analysis")
# meta_wes_dir <- here("data", "processed", "WES_analysis_meta")
# meta_rna_dir <- here("data", "processed", "gene_signatures_meta") 
# wes_rna_dir <- here("data", "processed", "WES_RNA_integration")
# figures_dir <- here("data", "processed", "_pancan_figures")
# prediction_dir <- here("data", "processed", "response_prediction")
# differential_expression_meta_dir <- here(
#   "data", "processed", "differential_expression_meta")
# external_analysis_dir <- here("data", "processed",
#                               "pancan_external_data_analysis")
# Tissue_PDL1_TILs_dir <- here("data", "processed", "Tissue_PDL1_TILs") 

ssgsea_response_prediction_ssgsea <- read_csv2(
  here(source_data_dir, "pancan_response_ssgsea_results_df.csv"))

```

# ----------
# Load metadata

```{r Load metadata}

histology_df <- data.frame(
  Trial = c("BELLINI", "IMCISION", "NABUCCO", "NICHE-MSS", "NICHE-MSI",
            "OpACIN_neo", "PRADO", "MATISSE"),
  Histology = c("TBNC", "HNSCC", "Urothelial", "Colon", "Colon",
                "LN_melanoma", "LN_melanoma", "CSCC"))

dna_metadata <- read_csv2(here(source_data_dir, "combined_dna_metadata.csv"))

dna_metadata <- dna_metadata %>%
  filter(!(Trial %in% "NICHE-MSI" & Tumor_nr == "Tumor_2"))

all_cohort_patient_metadata <- read_csv2(
  here(source_data_dir, "all_cohort_patient_metadata.csv"))

all_cohort_sample_metadata <- read_csv2(
  here(source_data_dir, "all_cohort_sample_metadata.csv"))

all_cohort_patient_metadata <- all_cohort_patient_metadata %>%
  mutate(Cohort = ifelse(
    Cohort %in% c("Nivo_ipi_A", "Nivo_ipi_B", "Nivo_ipi_C"),
    "Nivo_ipi", Cohort))

all_cohort_sample_metadata <- all_cohort_sample_metadata %>%
  mutate(Cohort = ifelse(
    Cohort %in% c("Nivo_ipi_A", "Nivo_ipi_B", "Nivo_ipi_C"),
    "Nivo_ipi", Cohort))

table(all_cohort_sample_metadata$Patient) %>%
  as.data.frame() %>%
  filter(Freq > 1)

all_cohort_sample_metadata <- all_cohort_sample_metadata %>%
  filter(!(Trial %in% "NICHE-MSI" & Tumor_nr == "Tumor_2"))

# Exclude patient with only RNA post data
exclude_patients <- all_cohort_patient_metadata %>%
  filter(RNA_post) %>%
  filter(!RNA_pre) %>%
  filter(!DNA)

exclude_patients

all_cohort_patient_metadata <- all_cohort_patient_metadata %>%
  filter(!Patient %in% exclude_patients$Patient)

all_cohort_sample_metadata <- all_cohort_sample_metadata %>%
  filter(!Patient %in% exclude_patients$Patient)

factor_summary(all_cohort_patient_metadata)

patient_metadata_combined <- read_csv2(
  here(source_data_dir, "patient_metadata_combined.csv"))
```

# Annotation color
```{r}

# Define colours for each group
Color_NR <- "#F8766D"
Color_R <- "#00BA38" # "#619CFF"
Color_NA <- "#CD9600"
Color_Pre <- "#619CFF"
Color_Post <- "#F564E3"
Color_HNSCC <- "#009292"
Color_CSCC <- "#CD9600"
Color_Urothelial <- "#B66DFF"
Color_LN_melanoma <- "#00BFC4"
Color_Colon <- "#F8766D"
Color_Colon_MSI <- "#B73A3A"
Color_Colon_MSS <- "#23C26F"
Color_TNBC <- "#FEB5DA"
Color_BELLINI <- "#FEB5DA"
Color_IMCISION <- "#009292"
Color_MATISSE <- "#CD9600"
Color_NABUCCO <- "#B66DFF"
Color_BELLINI <- "#FEB5DA"
Color_NICHE_MSS <- "#23C26F"
Color_NICHE_MSI <- "#B73A3A"
Color_OpACIN_neo <- "#00BFC4"
Color_PRADO <- "#619CFF"

cosmic_colors <- c(Aflatoxin = "#d0e0e3", Aging = "#3d85c6", APOBEC = "#674ea7",
                   Aristolochic_acid = "#d9d2e9", dMMR = "#B73A3A",
                   HRD = "#e69138", Other = "#cccccc", POL1 = "#cc0000",
                   POLE_POLD1 = "#5b0f00", POLH = "#FEB5DA", ROS = "#6DB6FF",
                   SBS5 = "#ea9999", Smoking = "#38761d", TMZ = "#b5a7d6",
                   Tobacco_chewing = "#6aa84f", UV = "#f1c232")

annotation_color_list = list(
  Response = c(NR = Color_NR, R = Color_R), #R_NR = "#CD9600"
  Treatment = c(Pre = Color_Pre, Post = Color_Post),
  Histology = c(LN_melanoma = Color_LN_melanoma, Urothelial = Color_Urothelial,
                Colon = Color_Colon, HNSCC = Color_HNSCC, TBNC = Color_TNBC,
                cSCC = Color_CSCC),
  Trial = c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO))

trial_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_colors_combined <- c(Combined = "#999999", BELLINI = Color_BELLINI,
                           IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_Colon_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_label_dict <- data.frame(
  Trial = c("Combined", "BELLINI", "IMCISION", "MATISSE", "NABUCCO",
            "NICHE-MSS", "NICHE-MSI", "OpACIN_neo", "PRADO"),
  Trial_label = c("Pooled", "BELLINI", "IMCISION", "MATISSE", "NABUCCO",
                  "NICHE pMMR", "NICHE dMMR", "OpACIN neo", "PRADO")) %>%
  unique()

trial_label_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE pMMR` = Color_NICHE_MSS, `NICHE dMMR` = Color_NICHE_MSI,
            `OpACIN neo` = Color_OpACIN_neo, PRADO = Color_PRADO)

histology_msi_mss_colors <- c(`Colon-MSI`= Color_Colon_MSI,
                              `Colon-MSS`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              LN_melanoma = Color_LN_melanoma,
                              TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)

histology_msi_mss_label_dict <- data.frame(
  Histology_MSI_MSS = c("TBNC", "Colon-MSS", "HNSCC", "Urothelial",
                        "LN_melanoma", "LN_melanoma", "CSCC", "Colon-MSI"),
  histology_msi_mss_label = c("TBNC", "Colon pMMR", "HNSCC", "Urothelial",
                              "Melanoma", "Melanoma", "CSCC", "Colon dMMR")) %>%
  unique()

histology_msi_mss_label_colors <- c(`Colon dMMR`= Color_Colon_MSI,
                                    `Colon pMMR`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              Melanoma = Color_LN_melanoma, TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)


Color_HNSCC <- "#009292"
Color_CSCC <- "#CD9600"
Color_Urothelial <- "#B66DFF"
Color_LN_melanoma <- "#00BFC4"
Color_Colon <- "#F8766D"
Color_Colon_MSI <- "#B73A3A"
Color_Colon_MSS <- "#23C26F"
Color_Breast <- "#FEB5DA"
Color_BELLINI <- "#FEB5DA"
Color_IMCISION <- "#009292"
Color_MATISSE <- "#CD9600"
Color_NABUCCO <- "#B66DFF"
Color_NICHE_MSS <- "#23C26F"
Color_NICHE_MSI <- "#B73A3A"
Color_OpACIN_neo <- "#00BFC4"
Color_PRADO <- "#619CFF"

Color_Anders <- "#FEB5DA"
Color_imvigor <- "#B66DFF"
Color_Liu <- "#00BFC4"
Color_gide <- "darkslategray3"
Color_hugo <- "#619CFF"
Color_immotion <- "tan1"
Color_choueiri <- "tan3"
Color_prins <- "coral1"

itlas_trial_label_colors <- c(brca_iatlas_anders_2022 = Color_Anders,
                              imvigor210_2017 = Color_imvigor,
                               mel_iatlas_Liu_2019 = Color_Liu,
                              mel_iatlas_gide_2019 = Color_gide,
                               hugo_ucla_2016 = Color_hugo,
                              immotion150_2018 = Color_immotion,
                               ccrcc_iatlas_choueiri_2016 = Color_choueiri,
                              gbm_iatlas_prins_2019 = Color_prins)

```

# Gene sets
```{r}
dmmr_ptns <- c("MLH1", "MSH2", "MSH6", "PMS2")

# Immune gene signatures from https://doi.org/10.1038/s41588-023-01367-1
immune_escape_genes_df <- read.csv2(
  here("data", "external", "paper_supplementaries",
       "Martínez-Jiménez et al. - 2023 Supp Table 1.csv"))

immune_escape_genes_df <- immune_escape_genes_df %>%
  mutate(Pathway.general = ifelse(
    Pathway.general == "IFN-? pathway", "IFN-y pathway", Pathway.general))%>%
  mutate(Pathway.general = ifelse(
    Pathway.general == "Antigen Presentation ",
    "Antigen Presentation", Pathway.general))

immune_escape_genes_df %>% select(Gene, Pathway.specific, Pathway.general)

# Gene lists from PMID: 34873480
ddr_ptns <- c("MRE11", "RAD50", "MLH1", "MSH2", "MSH6", "PMS2", "POLD1",
              "ATM", "ATR", "BARD1", "BLM", "BRCA1", "BRCA2", "BRIP1", "CHEK1",
              "CHEK2", "FANCA", "FANCC", "FANCD2", "FANCG", "MUTYH", "PALB2",
              "POLE", "TP53", "PRKDC", "CUL3", "ERCC1", "FANCE", "FANCF",
              "FANCL", "RAD51", "WEE1")

crc_risk_ptns <- c("APC", "MLH1", "MSH2", "MSH6", "PMS2", "MUTYH", "PTEN",
                   "TP53", "STK11", "SMAD1", "SMAD2")

immune_escape_genes <- immune_escape_genes_df$Gene

# Expand with other drivers
crc_risk_ptns <- c(crc_risk_ptns,
                   c("KRAS", "NRAS", "HRAS", "BRAF", "EGFR", "PIK3CA",
                     "EPCAM"))

genes_to_assess <- c(ddr_ptns, crc_risk_ptns, immune_escape_genes) %>% unique()
length(genes_to_assess)

variant_eval_columns <- c("Symbol_Variant", "SYMBOL",
                          "CLIN_SIG", "SIFT", "PolyPhen",
                          "Variant_Classification",
                          "Variant_Type", "VARIANT_CLASS",
                          "IMPACT", "AF", "EUR_AF",
                          "gnomAD_AF", "gnomAD_NFE_AF",
                          "n_depth", "VAF")

signature_overview <- read_csv2(
  here(source_data_dir, 
       "_overview_of_signatures_20260705.csv")) 

signature_overview$Signature <- gsub(
  "B_catenin_signature_pancan_22_Luke_et_al",
  "B_catenin_signature_Luke_et_al", signature_overview$Signature)
```

# ----------
# Load data

```{r}
response_rates <- read_csv2(here(source_data_dir, "Trial_response.csv"))

dna_metadata_f <- read_csv2(here(source_data_dir, "dna_metadata_maf.csv"))

# Only a few patients do not have WES data
# dna_metadata$Patient[!dna_metadata$Patient %in% combined_maf_df$Patient]

combined_maf_dp20_vaf <- readRDS(
  here(source_data_dir, "combined_maf_dp20_vaf.rds"))

combined_maf_specific_f <- readRDS(
  here(source_data_dir, "combined_maf_specific_f.rds"))

wes_features_df <- read_csv2(here(source_data_dir, "wes_features_df.csv"))

wes_features_df$Purity %>% summary()

wes_features_df <- wes_features_df %>%
  left_join(histology_msi_mss_label_dict) %>%
      mutate(Purity = ifelse(is.na(Purity), Imputed_purity, Purity)) 

wes_features_df$Purity %>% summary()

rna_features_df_plus_wes <- read_csv2(here(source_data_dir,
                                      "wes_rna_features_df.csv"))
rna_features_df_plus_wes_pre <- rna_features_df_plus_wes %>%
  filter(Treatment == "Pre")

extra_columns <- colnames(
  rna_features_df_plus_wes)[!colnames(rna_features_df_plus_wes) %in%
                              colnames(wes_features_df)]

wes_rna_features_df <- wes_features_df %>%
  left_join(rna_features_df_plus_wes %>%
              select(all_of(c("Patient", "Treatment", extra_columns))))  %>%
  filter(Treatment == "Pre")%>%
  left_join(histology_msi_mss_label_dict)

wes_rna_features_df_pre <- wes_rna_features_df %>%
  filter(Treatment == "Pre")

wes_rna_features_df_pre %>% filter(!is.na(TMB)) %>% dim()

wes_rna_features_df_pre_f <- wes_rna_features_df_pre %>%
  filter(!is.na(TMB))

ssgsea_results_df <- read_csv2(here(source_data_dir, "ssgsea_results_df.csv"))
ssgsea_results_df_pre <- ssgsea_results_df %>%
  filter(Treatment == "Pre")

ssgsea_results_df_pre$Trial %>% table()

rna_features_df_plus_wes_pre
```

```{r}
tmb_response_rate <- read_csv2(here(source_data_dir, "tmb_response_rate.csv"))

tmb_response_rate <- tmb_response_rate %>%
  mutate(Response_rate_prop = Response_rate,
         Response_rate = Response_rate * 100,
         Yarchoan_lm_rr_prop = Yarchoan_lm_rr,
         Yarchoan_lm_rr = Yarchoan_lm_rr * 100)

tmb_response_rate <- tmb_response_rate %>%
  left_join(histology_msi_mss_label_dict)

n_variants_dp20_vaf_tmb <- read_csv2(
  here(source_data_dir, "n_variants_dp20_vaf_tmb.csv"))

n_variants_dp20_vaf_tmb <- n_variants_dp20_vaf_tmb %>%
  filter(!(Trial %in% "NICHE-MSI" & Tumor_nr == "Tumor_2")) %>%
  left_join(histology_msi_mss_label_dict) %>%
  unique()

tmb_group_trial_response <- read_csv2(
  here(source_data_dir, "tmb_group_trial_response.csv"))

```

# ----------
# Figure 1
## Fig.1a
### Study design

## Fig.1b
## Metadata plots
### metadata n patients histology
```{r}
table(all_cohort_patient_metadata$Trial)
metadata_n_patients_trial_plot <- table(all_cohort_patient_metadata$Trial) %>%
  as.data.frame() %>%
  setNames(c("Trial", "Count")) %>%
  left_join(histology_df) %>%
  ggplot(aes(x = reorder(Trial, Count), y = Count, fill = Histology)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "Trial", y = "Number of Patients") +
  scale_fill_manual(values = c("#B73A3A",  "#CD9600", "#009292",
                               "#00BFC4", "#FEB5DA", "#B66DFF")) +
  theme(text = element_text(family = "Arial"))

metadata_n_patients_trial_plot
  
graph2pdf(x = metadata_n_patients_trial_plot,
          file = here(output_dir, "_metadata_n_patients_trial.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4, width = 5)

table(all_cohort_patient_metadata$Trial, all_cohort_patient_metadata$Cohort)

# ggpattern versions - stripes
all_cohort_patient_cohort_metadata_plot <- table(
  all_cohort_patient_metadata$Trial,
  all_cohort_patient_metadata$Cohort) %>%
  as.data.frame() %>%
  setNames(c("Trial", "Cohort", "Count")) %>%
  left_join(histology_df) %>%
  ggplot(aes(x = reorder(Trial, Count), y = Count, fill = Histology)) +
  geom_col_pattern(
    aes(pattern = Cohort),
    pattern_spacing = 0.025,
    pattern_angle = 75,
    pattern_size = 0.5,
    pattern_alpha = 0.25,
    pattern_fill = "black",
    colour = "black",
    ) +
  coord_flip() +
  labs(x = "Trial", y = "Number of Patients") +
  scale_fill_manual(values = c("#B73A3A",  "#CD9600", "#009292",
                               "#00BFC4", "#FEB5DA", "#B66DFF")) +
  scale_pattern_manual(values = c("stripe", "none")) +
  guides(fill = guide_legend(override.aes = list(pattern = c("none")))) +
  guides(pattern = guide_legend(nrow = 2)) +
 theme_Publication()  +
  scale_y_continuous(breaks = seq(0, 120, by =  20))

all_cohort_patient_cohort_metadata_plot

graph2pdf(x = all_cohort_patient_cohort_metadata_plot,
          file = here(output_dir, "_metadata_n_patients_trial_cohort.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4.5, width = 6.5)

```
## Fig.1c
### metadata n sample response
```{r}
table(all_cohort_sample_metadata$Trial, all_cohort_sample_metadata$Cohort) 

table(all_cohort_sample_metadata$Trial, all_cohort_sample_metadata$Response,
      all_cohort_sample_metadata$Cohort) 

all_cohort_sample_response_metadata_plot <- table(
  all_cohort_sample_metadata$Trial,
  all_cohort_sample_metadata$Response,
  all_cohort_sample_metadata$Cohort) %>%
  as.data.frame() %>%
  setNames(c("Trial", "Response", "Cohort", "Count")) %>%
  ggplot(aes(x = reorder(Trial, Count), y = Count,
             fill = Response, group = Response)) +
  geom_col() +
  coord_flip() +
  labs(x = "Trial", y = "Number of Patients") +
  scale_fill_manual(values = c("#F8766D", "#00BA38")) +
  guides(fill = guide_legend(override.aes = list(pattern = c("none")))) +
  theme_Publication() 

all_cohort_sample_response_metadata_plot

graph2pdf(x = all_cohort_sample_response_metadata_plot,
          file = here(output_dir,
                      "_metadata_n_sample_trial_response_cohort.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4.5, width = 6.5)

# ggpattern versions - stripes
all_cohort_patient_response_metadata_plot <- table(
  all_cohort_patient_metadata$Trial,
  all_cohort_patient_metadata$Response,
  all_cohort_patient_metadata$Cohort) %>%
  as.data.frame() %>%
  setNames(c("Trial", "Response", "Cohort", "Count")) %>%
  ggplot(aes(x = reorder(Trial, Count), y = Count,
             fill = fct_rev(Response), group = fct_rev(Response))) +
  geom_col_pattern(
    aes(pattern = Cohort),
    pattern_spacing = 0.025,
    pattern_angle = 75,
    pattern_size = 0.5,
    pattern_alpha = 0.25,
    pattern_fill = "black",
    colour = "black",
    ) +
  coord_flip() +
  labs(x = "Trial", y = "Number of Patients", fill = "Response") +
  scale_fill_manual(values = c("#00BA38", "#F8766D")) + # "#CD9600", 
  scale_pattern_manual(values = c("stripe", "none")) +
  guides(fill = guide_legend(override.aes = list(pattern = c("none")))) +
  theme_Publication()  +
  scale_y_continuous(breaks = seq(0, 120, by =  20))

all_cohort_patient_response_metadata_plot

graph2pdf(x = all_cohort_patient_response_metadata_plot,
          file = here(
            output_dir, "_metadata_n_patient_trial_response_cohort.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4.5, width = 6.5)

```

## Fig.1d
### Combo vs Mono
```{r}
all_cohort_patient_metadata_combo_mono <- all_cohort_patient_metadata %>%
  filter(Trial %in% c("BELLINI", "IMCISION", "MATISSE"))

combo_mono_response_plot <- table(
  all_cohort_patient_metadata_combo_mono$Trial,
  all_cohort_patient_metadata_combo_mono$Response,
  all_cohort_patient_metadata_combo_mono$Cohort) %>%
  as.data.frame() %>%
  setNames(c("Trial", "Response", "Cohort", "Count")) %>%
  ggplot(aes(x = Cohort, y = Count,
             fill = Response)) +
  geom_col() +
  labs(x = "Treatment", y = "Number of Patients", fill = "Response") +
  facet_wrap(~Trial) +
  scale_fill_manual(values = c("#00BA38", "#F8766D"), breaks = c("R", "NR")) +
  theme_Publication()

combo_mono_response_plot

graph2pdf(x = combo_mono_response_plot,
          file = here(
            output_dir, "_combo_mono_response_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4, width = 6.5)

```

# ----------
# Extended Data Fig 1
## Extended Data Fig 1a 
### Interventions treatment schedule panorama
```{r, fig.height = 10, fig.width = 8}
interventions_format_p <- read_csv2(
  here(source_data_dir, "interventions_format_p.csv"))
patient_metadata_combined <- read_csv2(
  here(source_data_dir, "patient_metadata_combined.csv"))

cohort_id_n <- patient_metadata_combined$Cohort_id %>%
  table() %>%
  as.data.frame() %>%
  setNames(c("Cohort_id", "n_patients"))
cohort_id_n

# Adjust labels
interventions_format_p <- interventions_format_p %>%
  left_join(cohort_id_n)
interventions_format_p$Cohort_id <- gsub(
  "OpACIN_neo", "OpACIN-neo", interventions_format_p$Cohort_id)
interventions_format_p$Cohort_id <- gsub(
  "_", " ", interventions_format_p$Cohort_id)
interventions_format_p$Cohort_id <- gsub(
  "MSS", "pMMR", interventions_format_p$Cohort_id)
interventions_format_p$Cohort_id <- gsub(
  "MSI", "dMMR", interventions_format_p$Cohort_id)

treatment_schedule_panorama_plot <- interventions_format_p %>% 
  mutate(Dose = ifelse(is.na(Dose), Treatment, Dose)) %>%
  mutate(Group = as.numeric(as.factor(Group))) %>%
  ggplot(aes(x = Day, y = Group, col = Treatment))  +
  geom_point(aes(y = as.numeric(Group) + Treatment_Pos,
                 col = Treatment, shape = as.factor(Dose)), size = 3) +
  scale_shape_manual(values = c(16, 15, 18, 17, 25, 23)) +
  labs(title = "Neoadjuvant ICB Treatment Schedules",
       x = "Days",
       y = "",
       fill = "Treatment",
       shape = "Dose") +
  theme(panel.grid.major.y = element_blank()) +
  scale_y_continuous(breaks = interventions_format_p$Group + 0.5,
                     labels = interventions_format_p$Cohort_id,
                     trans = "reverse") +
  scale_x_continuous(breaks = c(0, 7, 14, 21, 28, 35, 42, 49, 56, 63, 70)) + 
  scale_colour_manual(values = c(Celecoxib = "#FFC1CB", Ipilimumab = "#118DFF",
                                 "Neoadj. Chemo. or Surgery" = "#744EC2",
                                 Nivolumab = "#E66C37", Surgery = "#0AAC00"))+
  theme(axis.title.y = element_blank()) +
  theme_Publication() +
  theme(legend.position = "right",
        legend.direction = "vertical")

treatment_schedule_panorama_plot
treatment_schedule_panorama_plot +
  coord_cartesian(clip = "off") +
  annotate("text", label = paste0("n = ",
                                  interventions_format_p$n_patients),
           x = max(interventions_format_p$Day) + 5,
           y = interventions_format_p$Group, hjust = 0)

graph2pdf(x = treatment_schedule_panorama_plot,
          file = here(output_dir, "_treatment_schedule_panorama.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 10,
          width = 9)
```

## Extended Data Fig 1b
### Data availability
```{r}
annot_df <- all_cohort_patient_metadata %>%
  select(Trial, Patient, Response) %>%
  column_to_rownames("Patient") %>%
  arrange(Trial, Response)

# test <- all_cohort_patient_metadata %>% filter(RNA_pre)
# test$Patient[!test$Patient %in% ssgsea_results_df_pre$Patient]

all_cohort_patient_metadata$RNA_pre <- all_cohort_patient_metadata$Patient %in% rna_features_df_plus_wes_pre$Patient
  
all_cohort_patient_metadata_m <- all_cohort_patient_metadata %>%
  select(-c(Trial, Response, Histology, Cohort)) %>%
  column_to_rownames("Patient") %>%
  mutate_all(as.numeric) %>%
  as.matrix() %>%
  t()

rowSums(all_cohort_patient_metadata_m)
dim(all_cohort_patient_metadata_m)

data_availability_heatmap <- pheatmap::pheatmap(
  all_cohort_patient_metadata_m[, rownames(annot_df)],
  annotation_col = annot_df,
  show_colnames = FALSE,
  fontsize_row = 10,
  annotation_colors = annotation_color_list,
  treeheight_row = 0,
  treeheight_col = 0,
  cluster_cols = FALSE,
  color = c("#D8D7BF", "#2F8AC3")
  )

data_availability_heatmap

all_cohort_patient_metadata_m %>% rowSums()

data_availability_heatmap <- pheatmap::pheatmap(
  all_cohort_patient_metadata_m[c("DNA", "RNA_pre"), rownames(annot_df)],
  annotation_col = annot_df,
  show_colnames = FALSE,
  fontsize_row = 10,
  annotation_colors = annotation_color_list,
  treeheight_row = 0,
  treeheight_col = 0,
  cluster_cols = FALSE,
  color = c("#D8D7BF", "#2F8AC3")
  )

all_cohort_patient_metadata_m %>% rowSums()
all_cohort_patient_metadata %>% filter(RNA_pre) %>% filter(DNA) %>% dim()

data_availability_heatmap

save_pheatmap_pdf(x = data_availability_heatmap,
                  filename = here(
                    output_dir,
                    "data_availability_heatmap.pdf"),
                  width = 8, height = 4)

```

## Extended Data Fig 1c
### Table on completeness

# ----------
# Figure 2
## Fig.2a
### Plot TMB histologies
```{r}
percentage_above_TMB_10 <- n_variants_dp20_vaf_tmb %>%
  group_by(histology_msi_mss_label) %>%
  summarise(proportion = mean(TMB > 10)) %>%
  mutate(percentage = round(proportion * 100))
            
TMB_histology_plot <- n_variants_dp20_vaf_tmb %>%
  ggplot(aes(x = factor(histology_msi_mss_label,
                        levels = unique(
                          tmb_response_rate$histology_msi_mss_label)),
             y = TMB, fill = histology_msi_mss_label)) + 
  stat_boxplot(geom = "errorbar", width = 0.1) +
  geom_boxplot() + theme_Publication() +
  guides(fill = FALSE) +
  labs(x = "", y = "TMB (mutations/Mb)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = histology_msi_mss_label_colors) + 
  geom_hline(yintercept = 10, linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.1))) +
  geom_text(data = percentage_above_TMB_10,
            aes(x = histology_msi_mss_label, y = 168,
                label = paste0(percentage, "%")), vjust = -1.5)

# Log transform
n_variants_dp20_vaf_tmb %>%
  ggplot(aes(x = factor(histology_msi_mss_label,
                        levels = unique(
                          tmb_response_rate$histology_msi_mss_label)),
             y = log2(TMB + 1), fill = histology_msi_mss_label)) + 
  stat_boxplot(geom = "errorbar", width = 0.1) +
  geom_boxplot() + theme_Publication() +
  guides(fill = FALSE) +
  labs(x = "", y = "log2(TMB +1) (mutations/Mb)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = histology_msi_mss_label_colors) + 
  geom_hline(yintercept = log2(10+1), linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.1))) +
  geom_text(data = percentage_above_TMB_10,
            aes(x = histology_msi_mss_label, y = 7,
                label = paste0(round(proportion * 100), "%")), vjust = -1.5)

TMB_histology_plot

graph2pdf(x = TMB_histology_plot,
          file = here(output_dir, "_TMB_histology_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 5)
# group = Response

Histology_Response_order <- data.frame(
  histology_msi_mss_label = rep(tmb_response_rate$histology_msi_mss_label, 2),
  Response = c("R", "NR"))

expanded_levels <- unique(
  as.vector(sapply(tmb_response_rate$histology_msi_mss_label,
                   function(x) c(paste0(x, " NR"), paste0(x, " R")))))

n_variants_dp20_vaf_tmb %>%
 mutate(Histology_Response = paste0(histology_msi_mss_label, " ", Response)) %>%
 ggplot(aes(x = factor(Histology_Response, levels = expanded_levels),
           y = log2(TMB + 1), fill = histology_msi_mss_label)) + 
  stat_boxplot(aes(), geom = "errorbar", width = 0.1) +
  geom_boxplot(aes()) + theme_Publication() +
  guides(fill = FALSE) +
  labs(x = "", y = "log2(TMB +1) (mutations/Mb)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = histology_msi_mss_label_colors) + 
  geom_hline(yintercept = log2(10+1), linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.1))) 

n_variants_dp20_vaf_tmb %>%
 mutate(Histology_Response = paste0(histology_msi_mss_label, " ", Response)) %>%
 ggplot(aes(x = factor(Histology_Response, levels = expanded_levels),
           y = TMB, fill = histology_msi_mss_label)) + 
  stat_boxplot(aes(), geom = "errorbar", width = 0.1) +
  geom_boxplot(aes()) + theme_Publication() +
  guides(fill = FALSE) +
  labs(x = "", y = "TMB (mutations/Mb)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = histology_msi_mss_label_colors) + 
  geom_hline(yintercept = 10, linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.1))) 


TMB_trial_response_plot <- n_variants_dp20_vaf_tmb %>%
  ggplot(aes(x = factor(Trial,
                        levels = unique(
                          tmb_response_rate$Trial)),
             y = TMB, fill = Response)) + 
  geom_boxplot(position = position_dodge()) + theme_Publication() +
  labs(x = "", y = "TMB (mutations/Mb)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = c(Color_NR, Color_R)) + 
  geom_hline(yintercept = 10, linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.1)))

TMB_trial_response_plot

graph2pdf(x = TMB_trial_response_plot,
          file = here(output_dir, "_TMB_trial_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 5)
```


### TMB High low groups
```{r}
n_variants_tmb_group <- n_variants_dp20_vaf_tmb %>%
  mutate(TMB_group = ifelse(TMB > 10, "TMB-H", "TMB-L")) %>%
  filter(!is.na(Response))

table(n_variants_tmb_group$TMB_group)
table(n_variants_tmb_group$Trial, n_variants_tmb_group$TMB_group)

table(n_variants_tmb_group$TMB_group, n_variants_tmb_group$Response)

table(n_variants_tmb_group$TMB_group, n_variants_tmb_group$Response) %>%
  as.data.frame() %>% setNames(c("TMB_group", "Response", "Freq")) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  mutate(Total = R + NR, Proportion = R / Total)

```


### Plot TMB response

```{r}
n_variants_dp20_vaf_tmb$Response %>% table()
tmb.stat.test <- compare_means(formula = TMB ~ Response,
                           method = "wilcox.test",
                           data = n_variants_dp20_vaf_tmb, paired = FALSE)
tmb.stat.test

n_labels <- n_variants_dp20_vaf_tmb %>%
group_by(Response) %>%
summarise(n = n()) %>%
mutate(label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Response)

tmb_response_plot <- n_variants_dp20_vaf_tmb %>%
  filter(!is.na(Response)) %>%
  group_by(Response) %>%
  ggplot(aes(x = Response, y = TMB)) + 
  stat_boxplot(geom = "errorbar", width = 0.1) +
  geom_boxplot(aes(fill = Response)) + #, outlier.shape = NA
  theme_Publication()  +
  stat_pvalue_manual(tmb.stat.test, y.position = 150,
                     label = "wilcoxon, p = {p.format}") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  expand_limits(y = 0) +
  labs(y = "TMB (mutations/Mb)") +
  guides(fill = FALSE) +
  theme(axis.title.x = element_blank()) +
  scale_x_discrete(labels = label_map) 

tmb_response_plot
graph2pdf(x = tmb_response_plot,
          file = here(output_dir, "_TMB_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 5)
```

### Plot TMB response excl. dMMR

```{r}
# Filtering out MSI dMMR
tmb.dmmr.stat.test <- compare_means(formula = TMB ~ Response,
                           method = "wilcox.test",
                           data = n_variants_dp20_vaf_tmb %>%
                             filter(!Trial %in% c("NICHE-MSI")),
                           paired = FALSE)
tmb.dmmr.stat.test

n_labels <- n_variants_dp20_vaf_tmb %>%
  filter(!Trial %in% c("NICHE-MSI")) %>%
  filter(!is.na(Response)) %>%
group_by(Response) %>%
summarise(n = n()) %>%
mutate(label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Response)

tmb_response_plot <- n_variants_dp20_vaf_tmb %>%
  filter(!Trial %in% c("NICHE-MSI")) %>%
  filter(!is.na(Response)) %>%
  ggplot(aes(x = Response, y = TMB)) + 
  stat_boxplot(geom = "errorbar", width = 0.1) +
  geom_boxplot(aes(fill = Response)) + #, outlier.shape = NA
  theme_Publication()  +
  stat_pvalue_manual(tmb.dmmr.stat.test, y.position = 150,
                     label = "wilcoxon, p = {p.format}") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  expand_limits(y = 0) +
  labs(y = "TMB (mutations/Mb)", title = "Excluding dMMR") +
  guides(fill = FALSE) +
  theme(axis.title.x = element_blank()) +
  scale_x_discrete(labels = label_map) 

tmb_response_plot
graph2pdf(x = tmb_response_plot,
          file = here(output_dir, "_TMB_response_plot_excl_dMMR.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 5)

```

### Filtering out MSI dMMR & cSCC

```{r}
# Filtering out MSI dMMR & cSCC
tmb.dmmr.stat.test <- compare_means(formula = TMB ~ Response,
                           method = "wilcox.test",
                           data = n_variants_dp20_vaf_tmb %>%
                             filter(!Trial %in% c("NICHE-MSI", "MATISSE")),
                           paired = FALSE)
tmb.dmmr.stat.test

n_labels <- n_variants_dp20_vaf_tmb %>%
  filter(!Trial %in% c("NICHE-MSI", "MATISSE")) %>%
  filter(!is.na(Response)) %>%
group_by(Response) %>%
summarise(n = n()) %>%
mutate(label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Response)

tmb_response_plot <- n_variants_dp20_vaf_tmb %>%
  filter(!Trial %in% c("NICHE-MSI", "MATISSE")) %>%
  filter(!is.na(Response)) %>%
  ggplot(aes(x = Response, y = TMB)) + 
  stat_boxplot(geom = "errorbar", width = 0.1) +
  geom_boxplot(aes(fill = Response)) + #, outlier.shape = NA
  # geom_point() +
  theme_Publication()  +
  stat_pvalue_manual(tmb.dmmr.stat.test, y.position = 150,
                     label = "wilcoxon, p = {p.format}") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  expand_limits(y = 0) +
  labs(y = "TMB (mutations/Mb)", title = "Excluding dMMR and cSCC") +
  guides(fill = FALSE) +
  theme(axis.title.x = element_blank()) +
  scale_x_discrete(labels = label_map) 

tmb_response_plot

graph2pdf(x = tmb_response_plot,
          file = here(output_dir, "_TMB_response_plot_excl_dMMR_cSCC.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 5)
```

## Fig.2b 

### Plot TMB response + sensitivity sets excl. dMMR & cSCC

```{r}
n_variants_dp20_vaf_tmb_sets <- n_variants_dp20_vaf_tmb %>%
  select(Patient, CF_sample_tumor, TMB, Trial, Response) %>%
  mutate(Cohort = "Pooled cohort") %>%
  rbind(
    n_variants_dp20_vaf_tmb %>%
      filter(!Trial %in% c("NICHE-MSI")) %>%
      select(Patient, CF_sample_tumor, TMB, Trial, Response) %>%
      mutate(Cohort = "Pooled cohort\nexcl. dMMR")  
  ) %>%
  rbind(
    n_variants_dp20_vaf_tmb %>%
      filter(!Trial %in% c("NICHE-MSI", "MATISSE")) %>%
      select(Patient, CF_sample_tumor, TMB, Trial, Response) %>%
      mutate(Cohort = "Pooled cohort\nexcl. dMMR & cSCC")  
  )

n_labels <- n_variants_dp20_vaf_tmb_sets %>%
group_by(Cohort, Response) %>%
summarise(n = n()) %>%
mutate(
  Cohort_Response = paste0(Cohort, " ", Response),
  label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Cohort_Response)

tmb_response_sensitivity_plot <- n_variants_dp20_vaf_tmb_sets %>%
  mutate(Cohort_Response = paste0(Cohort, " ", Response)) %>%
  ungroup() %>%
  ggplot(aes(x = Cohort_Response, y = TMB)) + 
  stat_boxplot(geom = "errorbar", width = 0.1) +
  geom_boxplot(aes(fill = Response)) + #, outlier.shape = NA
  # geom_point() +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  expand_limits(y = 0) +
  labs(y = "TMB (mutations/Mb)") +
  guides(fill = FALSE) +
  theme(axis.title.x = element_blank()) +
  scale_x_discrete(labels = label_map) +
  stat_compare_means(aes(group = Response), method = "wilcox") +
  facet_wrap(~Cohort, scale = "free_x") +
  scale_y_continuous(expand = expand_scale(add = 20)) 

tmb_response_sensitivity_plot

graph2pdf(x = tmb_response_sensitivity_plot,
          file = here(output_dir, "_TMB_response_plot_sensitivity.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 6.5)
```

## Fig.2c 

### Plot TMB response rate regression vs. Yarchoan regression

```{r}
tmb_response_rate %>%
  select(Trial, Response_rate, Yarchoan_lm_rr) %>%
  mutate(fold_diff = Response_rate / Yarchoan_lm_rr)

# Fit with percentage
response_rate_tmb_model <- lm(Response_rate ~ log(median_TMB),
                              data = tmb_response_rate)
summary(response_rate_tmb_model)
coef(response_rate_tmb_model) %>% round(1)

# Add CI
# Function to calculate proportion and CI
get_prop_ci <- function(x, n) {
  result <- prop.test(x, n, correct = FALSE)
  proportion <- (x / n)
  ci <- result$conf.int
  return(c(proportion, ci[1], ci[2]))
}

# Function to calculate percentage and CI
get_ci <- function(x, n) {
  result <- prop.test(x, n, correct = FALSE)
  percentage <- (x / n) * 100
  ci <- result$conf.int * 100
  return(c(percentage, ci[1], ci[2]))
}

ci_prop_results <- t(mapply(get_prop_ci, tmb_response_rate$R,
                            tmb_response_rate$Total_n))
tmb_response_rate$CI_lower_prop <- ci_prop_results[,2]
tmb_response_rate$CI_lower_prop <- ci_prop_results[,3]

ci_results <- t(mapply(get_ci, tmb_response_rate$R,
                       tmb_response_rate$Total_n))
tmb_response_rate$CI_lower <- ci_results[,2]
tmb_response_rate$CI_upper <- ci_results[,3]

min(tmb_response_rate$median_TMB)
max(tmb_response_rate$median_TMB)
tmb_range <- seq(min(tmb_response_rate$median_TMB),
                 max(tmb_response_rate$median_TMB), length.out = 100)
tmb_range <- seq(1.5, 51, length.out = 100)

lm_rr_prediction <- data.frame(median_TMB = tmb_range)

lm_rr_prediction$Response_rate <- predict(response_rate_tmb_model,
                                          lm_rr_prediction)

yarchoan_rr_prediction <-  data.frame(median_TMB = tmb_range) %>%
  mutate(Response_rate = (10.8 * log(median_TMB) - 0.7),
         Response_rate_prop = (10.8 * log(median_TMB) - 0.7)/100)

source_data_tmb_rr <- tmb_response_rate

tmb_rr_regression_plot <- ggplot(
  tmb_response_rate, aes(x = median_TMB,
                         y = Response_rate,
                         col = histology_msi_mss_label)) +
 geom_point() +
 geom_smooth(method = "lm",
             formula = Response_rate ~ log(median_TMB), se = TRUE) +
 scale_x_continuous(trans = log_trans(base = exp(1)),
                    breaks = c(1, 5, 10, 20, 30, 40, 50)) +
  expand_limits(y = 0) +
  scale_color_manual(values = histology_msi_mss_label_colors) +
  theme_Publication() +
  geom_line(data = lm_rr_prediction,
            aes(x = median_TMB, y = Response_rate),
            linetype = "dashed", color = "black")  +
  geom_line(data = yarchoan_rr_prediction,
            aes(x = median_TMB, y = Response_rate),
            linetype = "dashed", color = "gray") +
  labs(x = "median TMB (mutations/Mb)",
       y = "Response rate (%)", col = "") +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper),
                width = 0.02, alpha = 0.5)+
  geom_text(aes(x = 32, y = 20, label = "Yarchoan et al. 2017"),
            col = "gray", vjust = -1.5)

tmb_rr_regression_plot

graph2pdf(x = tmb_rr_regression_plot,
          file = here(output_dir, "_median_TMB_rr_regression_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 5)
```

## Fig.2d

### TMB quantile bins
```{r}
tmb_quantiles <- n_variants_dp20_vaf_tmb %>%
  arrange(TMB) %>%
  mutate(bin = ntile(TMB, n = 10)) %>%
  mutate(Response_n = ifelse(Response == "R", 1 ,0)) %>%
  left_join(trial_label_dict) %>%
  left_join(histology_msi_mss_label_dict)

tmb_quantiles %>%
  ggplot(aes(x = bin, y = log2(TMB), col = Trial_label)) +
  geom_point() + theme_Publication()

table(tmb_quantiles$bin)

# Summarize data by bin
tmb_quantile_response_df <- tmb_quantiles %>%
  group_by(bin) %>%
  summarise(
    response_rate = mean(Response_n, na.rm = TRUE),
    trial_distribution = list(table(Trial_label)),
    min_TMB = min(TMB),
    max_TMB = max(TMB)
  )
  
label_map <- setNames(c("[0-10%]", "(10-20%]", "(20-30%]", "(30-40%]",
                        "(40-50%]", "(50-60%]", "(60-70%]", "(70-80%]",
                        "(80-90%]", "(90-100%]"), c(1:10))    

label_map <- setNames(c("0-10", "10-20", "20-30", "30-40",
                        "40-50", "50-60", "60-70", "70-80",
                        "80-90", "90-100"), c(1:10))  

label_map <- setNames(c("10", "20", "30", "40",
                        "50", "60", "70", "80",
                        "90", "100"), c(1:10))

tmb_quantile_response_plot <- tmb_quantile_response_df %>%
  ggplot(aes(x = factor(bin), y = response_rate)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = "Response Rate by Quantile Bin",
         x = "TMB 10-quantile bin (%)", y = "Response Rate (%)", fill = "") +
    theme_Publication() +
   scale_x_discrete(labels = label_map)

tmb_quantile_response_plot

graph2pdf(x = tmb_quantile_response_plot,
          file = here(output_dir, "_tmb_quantile_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.5)

tmb_quantiles_histology_df <- tmb_quantiles %>%
    group_by(bin, histology_msi_mss_label) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(bin) %>%
    mutate(prop = count / sum(count))
    
tmb_quantiles_histology_plot <- tmb_quantiles_histology_df %>%
  ggplot(aes(x = factor(bin), y = prop, fill = histology_msi_mss_label )) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = "Trial Group Distribution by Bin",
         x = "TMB 10-quantile bin (%)", y = "Proportion", fill = "") +
    scale_fill_manual(values = histology_msi_mss_label_colors) +
    theme_Publication()+
   scale_x_discrete(labels = label_map)


tmb_quantiles_histology_plot
graph2pdf(x = tmb_quantiles_histology_plot,
          file = here(output_dir, "_tmb_quantiles_histology_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.5)

tmb_quantiles_trial_df <- tmb_quantiles %>%
    group_by(bin, Trial_label) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(bin) %>%
    mutate(prop = count / sum(count))
    
tmb_quantiles_trial_plot <- tmb_quantiles_trial_df %>%
  ggplot(aes(x = factor(bin), y = prop, fill = Trial_label )) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = "Trial Group Distribution by Bin",
         x = "TMB 10-quantile bin (%)", y = "Proportion", fill = "") +
    scale_fill_manual(values = trial_label_colors) +
    theme_Publication()+
   scale_x_discrete(labels = label_map)

graph2pdf(x = tmb_quantiles_trial_plot,
          file = here(output_dir, "_tmb_quantiles_trial_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.5)

process_and_plot <- function(data, variable, response_col,
                             trial_col, n_bins = 4) {
  # Create quantile bins
  data <- data %>%
    mutate(bin = ntile(.data[[variable]], n_bins))
  
  # Summarize data by bin
  summary_df <- data %>%
    group_by(bin) %>%
    summarise(
      response_rate = mean(.data[[response_col]], na.rm = TRUE),
      trial_distribution = list(table(.data[[trial_col]]))
    )
  
  # Expand trial distribution for plotting
  trial_df <- data %>%
    group_by(bin, trial = .data[[trial_col]]) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(bin) %>%
    mutate(prop = count / sum(count))
  
  # Plot response rate
  p1 <- ggplot(summary_df, aes(x = factor(bin), y = response_rate)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = "Response Rate by Quantile Bin",
         x = "TMB quantile bins", y = "Response Rate (%)") +
    theme_minimal()
  
  # Plot trial distribution
  p2 <- ggplot(trial_df, aes(x = factor(bin), y = prop, fill = trial)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = "Trial Group Distribution by Bin",
         x = "TMB quantile bins", y = "Proportion") +
    theme_minimal() +
    scale_fill_manual(values = trial_colors)
  
  print(p1)
  print(p2)
  
  return(list(summary = summary_df, trial_details = trial_df))
}

process_and_plot(tmb_quantiles %>%
                   mutate(Response_n = ifelse(Response == "R", 1 ,0)),
                 variable = "TMB", response_col = "Response_n",
                 trial_col = "Trial", n_bins = 10)

```

# ----------
# Figure 3

## Fig.3a
### WES Meta-analysis SMD continuous
```{r}
meta_studies_results <- read_csv2(
  here(source_data_dir, "wes_meta_studies_results_smd_all.csv"))
meta_combined_effects <- read_csv2(
  here(source_data_dir, "wes_meta_combined_effects_random_effects.csv"))

combined_wes_smd <- read_csv2(
  here(source_data_dir, "combined_wes_smd.csv"))

combined_wes_smd$Trial <- factor(
  combined_wes_smd$Trial,
  c("Combined", "Combined_excl_dMMR", "BELLINI", "IMCISION", "MATISSE",
    "NABUCCO", "NICHE-MSS", "OpACIN_neo", "PRADO"))

meta_studies_results %>%
  select(studlab, n.e, n.c) %>%
  mutate(total = n.e + n.c) %>%
  unique()
```
```{r}
meta_combined_effects <- meta_combined_effects %>%
  arrange(-TE.random)

meta_combined_effects %>%
    arrange(-TE.random) %>% pull(Variable)

smd_variable_order <- c("TMB", 
                        "pvactools_filtered_neoantigen_load",
                        "pyclone_subclonal_TMB", "pyclone_clonal_TMB", 
                        "APOBEC_abs", "UV_abs",
                        "dMMR_abs",
                        "TMB_frameshift", "HRD_abs", "Ploidy",
                        "Aneuploidy_score", "wgII_0.6_bases",
                        "scna_loss", "Purity")
smd_variable_labels = c("TMB", 
                        "neoantigen load",
                        "subclonal TMB", "clonal TMB",
                        "APOBEC abs", "UV abs",
                         "dMMR SBS",
                        "TMB frameshift", "HRD SBS", "Ploidy",
                        "Aneuploidy score", "wgII",
                        "SCNA loss", "Purity")
data.frame(smd_variable_order, smd_variable_labels)

meta_studies_results_sample_sizes <- meta_studies_results %>%
  select(studlab, n.e, n.c) %>%
  mutate(R = n.e, NR = n.c, total = R + NR) %>%
  unique()
meta_studies_results_sample_sizes

combined_smd_plot <- meta_studies_results %>%
  mutate(label = cut(
    pval,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(studlab, levels = c("Combined", "Combined_excl_dMMR", "BELLINI",
                                   "IMCISION", "MATISSE", "NABUCCO",
                                   "NICHE-MSS", "OpACIN_neo", "PRADO")),
    y = factor(
      Variable, levels = rev(smd_variable_order)),
    fill = TE,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "", fill = "SMD\n(Hedge's g)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "") +
  scale_x_discrete(labels = c("Combined", "Combined excl dMMR", "BELLINI",
                              "IMCISION", "MATISSE", "NABUCCO",
                              "NICHE pMMR", "OpACIN-neo", "PRADO")) +
  scale_y_discrete(labels = rev(smd_variable_labels))

combined_smd_plot
graph2pdf(x = combined_smd_plot,
          file = here(output_dir, "_combined_smd_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 7.7)

meta_combined_effects %>%
  select(Variable, k, pval.random) %>%
  mutate(pval.random = FormatDecimals(pval.random))

meta_combined_smd_plot <- meta_combined_effects %>%
  ggplot(aes(x = factor(
    Variable, levels = rev(smd_variable_order)),
    y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip(clip = "off") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis SMD (95% CI)") +
  guides(size = guide_legend(nrow = 3)) +
  geom_text(
    aes(label = pval.random, y = 0.9, hjust = 0),
    data = meta_combined_effects %>%
      mutate(pval.random = sprintf("P = %s", FormatDecimals(pval.random)))) +
  scale_x_discrete(labels = rev(c(smd_variable_labels))) + 
  theme(plot.margin = unit(c(1,4,1,1), "lines")) 
meta_combined_smd_plot

graph2pdf(x = meta_combined_smd_plot,
          file = here(output_dir, "_meta_combined_smd_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 5)

```

### WES Meta-analysis Log OR discrete
```{r}
meta_studies_results_or_all <- read_csv2(
  here(source_data_dir, "meta_studies_results_or_all.csv"))

meta_log_or_combined_effects_random_effects <- read_csv2(
  here(source_data_dir, "meta_log_or_combined_effects_random_effects.csv"))


log_or_variable_order <- c("GIE_AP", "TP53", "BRCA1_2", "SIGPW_Wnt_B_Cat",
                           "SIGPW_PI3K_signaling", "SIGPW_Cell_cycle",
                           "SIGPW_TGFB_signaling", "GIE_IFNy", "RAS_RAF",
                           "HLA_alteration")
log_or_variable_labels = c(
  "Antigen presentation mut", "TP53 mut", "BRCA 1/2 mut", "Wnt/B-cat mut",
  "PI3K signaling mut", "Cell cycle mut",
  "TGF-B signaling mut", "IFNy mut", "RAS/RAF mut", "HLA del/LOH")

meta_log_or_combined_effects_random_effects  %>%
  ggplot(aes(x = factor(feature, levels = rev(log_or_variable_order)),
    y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip(clip = "off") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis SMD (95% CI)") +
  guides(size = guide_legend(nrow = 3)) +
  geom_text(
    aes(label = pval.random, y = 1.5, hjust = 0),
    data = meta_log_or_combined_effects_random_effects %>%
      mutate(pval.random = sprintf("P = %s", FormatDecimals(pval.random))))


or_features_to_exclude <- c("JAK1_JAK2", "B2M")
trial_odds_ratio_df <- read_csv2(
  here(source_data_dir, "trial_odds_ratio_df.csv"))

trial_odds_ratio_df <- trial_odds_ratio_df %>%
  filter(!feature %in% or_features_to_exclude)

meta_log_or_results <- read_csv2(
  here(source_data_dir, "meta_log_or_results_combined.csv"))

meta_log_or_results <- meta_log_or_results %>%
  filter(!feature %in% or_features_to_exclude)

meta_log_or_combined_effects <- read_csv2(
  here(source_data_dir, "meta_log_or_combined_effects_random_effects.csv"))

meta_log_or_combined_effects <- meta_log_or_combined_effects %>%
  filter(!feature %in% or_features_to_exclude)

meta_log_or_combined_effects <- meta_log_or_combined_effects %>%
  arrange(TE.random)


combined_log_or_plot <- meta_studies_results_or_all %>%
  mutate(label = cut(
    pval,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  )) %>%
  ggplot(aes(x = factor(studlab,
                        c("Combined", "Combined_excl_dMMR", "BELLINI",
                          "IMCISION", "MATISSE", "NABUCCO",
                          "NICHE-MSS", "OpACIN_neo", "PRADO")),
             y = factor(feature,
                        levels = rev(log_or_variable_order)),
             fill = TE, 
             label = label)) +
  geom_tile() +
  geom_text() + 
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000",
                       na.value = "#CCCCCC"
                       ) +
  labs(x = "", y = "", fill = "log(OR)") +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_x_discrete(labels = c("Combined", "Combined excl dMMR", "BELLINI",
                              "IMCISION", "MATISSE", "NABUCCO",
                              "NICHE pMMR", "OpACIN-neo", "PRADO")) +
  scale_y_discrete(labels = rev(log_or_variable_labels))

combined_log_or_plot

graph2pdf(x = combined_log_or_plot,
          file = here(output_dir, "_combined_log_or_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 7)

k_studies <- trial_odds_ratio_df %>%
  filter(!is.na(log_or),
         !is.infinite(estimate),
         estimate != 0,
        !Trial %in% c("Combined", "Combined_excl_dMMR"))

effective_k_studies <- table(k_studies$feature) %>%
  as.data.frame() %>%
  setNames(c("feature", "effective_k"))
effective_k_studies

meta_log_or_combined_effects <- meta_log_or_combined_effects %>%
  left_join(effective_k_studies) %>%
  mutate(feature = sprintf("%s (%s)", feature, effective_k))

meta_combined_log_or_plot <- meta_log_or_combined_effects  %>%
  ggplot(aes(x = reorder(feature, TE.random), y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip(clip = "off") + geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis log(OR) (95% CI)") +
  guides(size = guide_legend(nrow = 3)) +
  geom_text(
    aes(label = pval.random, y = 2, hjust = 0),
    data = meta_log_or_combined_effects %>%
      mutate(pval.random = sprintf("P = %s", FormatDecimals(pval.random)))) + 
      theme(plot.margin = unit(c(1,4,1,1), "lines")) 

meta_log_or_combined_effects_random_effects %>%
  arrange(-TE.random) %>%
  pull(feature)

meta_combined_log_or_plot <- meta_log_or_combined_effects_random_effects  %>%
  ggplot(aes(x = factor(feature, levels = rev(log_or_variable_order)),
    y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip(clip = "off") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis log(OR)") +
  guides(size = guide_legend(nrow = 3)) +
  geom_text(
    aes(label = pval.random, y = 0.5, hjust = 0),
    data = meta_log_or_combined_effects_random_effects %>%
      left_join(effective_k_studies)%>%
      mutate(pval.random = sprintf("P = %s, k = %s",
                                   FormatDecimals(pval.random), effective_k))) +
  scale_x_discrete(labels = rev(log_or_variable_labels))

meta_combined_log_or_plot

meta_log_or_combined_effects %>%
  select(feature, pval.random) %>%
  mutate(pval.random = FormatDecimals(pval.random))

graph2pdf(x = meta_combined_log_or_plot,
          file = here(output_dir, "_meta_combined_log_or_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 5)
```
```{r}

meta_combined_effects_random_effects_wes <- meta_combined_effects %>%
  arrange(TE.random) %>%
  full_join(meta_log_or_combined_effects_random_effects %>%
              rename(Variable = feature) %>%
  arrange(TE.random)) %>%
  mutate(meta_FDR = p.adjust(pval.random, method = "BH"))

meta_combined_effects_random_effects_wes %>%
  select(Variable, pval.random, meta_FDR) %>%
  mutate(Variable = factor(Variable,
                           levels = c(fct_rev(smd_variable_order),
                                      fct_rev(log_or_variable_order)))) %>%
  arrange(Variable)  %>%
  mutate(pval.random = round(pval.random, 2))%>%
  mutate(meta_FDR = round(meta_FDR, 2)) %>%
  print(n = 30)
```

## Fig.3b
### WES correlation plot
```{r, fig.height = 8, fig.width = 8}
wes_features_m <- wes_features_df  %>%
  mutate(HLA_alteration = ifelse(HLA_grouped %in% c("HLA-LOH", "HLA-DEL"), "HLA-Altered", HLA_grouped)) %>%
  mutate(HLA_alteration = HLA_alteration == "HLA-Altered") %>%
  select(all_of(c(smd_variable_order, log_or_variable_order)))  %>%
 setNames(c(smd_variable_labels, log_or_variable_labels)) %>% 
  as.matrix()

wes_features_m %>%
  cor(use = "pairwise.complete.obs", method = "spearman") %>%
  corrplot(order = "hclust", type = "lower", tl.col = "black")

save_pheatmap_pdf(
  x = 
wes_features_m %>%
  cor(use = "pairwise.complete.obs", method = "spearman") %>%
  corrplot(order = "hclust", type = "lower", tl.col = "black"),
                  filename = here(
                    output_dir,
                    "wes_corplot.pdf"),
                  width = 6, height = 6)

wes_features_df_excl_dMMR <- wes_features_df %>%
  filter(!Trial %in% c("NICHE-MSI")) 

wes_features_m_excl_dMMR <- wes_features_df %>%
  filter(!Trial %in% c("NICHE-MSI")) %>%
  select(all_of(c(smd_variable_order, log_or_variable_order)))  %>%
 setNames(c(smd_variable_labels, log_or_variable_labels)) %>% 
  as.matrix()

wes_features_m_excl_dMMR %>%
  cor(use = "pairwise.complete.obs", method = "spearman") %>%
  corrplot(order = "hclust", type = "lower", tl.col = "black")

```

## Fig.3c
### Aggregate WES meta
#### Sensitivity subsets WES SMD
```{r}
wes_meta_aggregated <- read_csv2(
  here(source_data_dir, "wes_meta_aggregated.csv"))


smd_variable_order <- c("TMB", 
                        "pvactools_filtered_neoantigen_load",
                        "pyclone_subclonal_TMB", "pyclone_clonal_TMB", 
                        "APOBEC_abs", "UV_abs", "dMMR_abs",
                        "TMB_frameshift", "HRD_abs", "Ploidy",
                        "Aneuploidy_score", "wgII_0.6_bases",
                        "scna_loss", "Purity")

smd_variable_labels = c("TMB", 
                        "neoantigen load",
                        "subclonal TMB", "clonal TMB",
                        "APOBEC abs", "UV abs", "dMMR SBS",
                        "TMB frameshift", "HRD SBS", "Ploidy",
                        "Aneuploidy score", "wgII",
                        "SCNA loss", "Purity")

wes_smd_sensitivity_plot <- wes_meta_aggregated %>%
  mutate(Variable = factor(Variable, levels = fct_rev(smd_variable_order))) %>%
  mutate(label = cut(
    pval.random,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(
      Analysis_subset,
      levels = c("Original", "Excl_mono.", "Excl_clin.", "MPR", "pCR")), 
    y = fct_rev(Variable),
    fill = TE.random,
    label = label)
    ) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "", title = "Meta-analysis sensitivity", fill = "Meta-analysis\nSMD (Hedge's g)") +
  scale_y_discrete(labels = rev(smd_variable_labels))

wes_smd_sensitivity_plot

graph2pdf(x = wes_smd_sensitivity_plot,
          file = here(output_dir, "_wes_smd_sensitivity_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 7)
```

## Fig.3d
### Sensitivity subsets WES OR

```{r}
log_or_variable_order <- c("GIE_AP", "TP53", "BRCA1_2",
                           "SIGPW_Wnt_B_Cat",
                           "SIGPW_PI3K_signaling", "SIGPW_Cell_cycle",
                           "SIGPW_TGFB_signaling", "GIE_IFNy", "RAS_RAF",
                           "HLA_alteration")
log_or_variable_labels = c(
  "Antigen presentation mut", "TP53 mut", "BRCA 1/2 mut",
  "Wnt/B-cat mut",
  "PI3K signaling mut", "Cell cycle mut",
  "TGF-B signaling mut", "IFNy mut", "RAS/RAF mut",
  "HLA del/LOH")

wes_or_meta_aggregated <- read_csv2(
  here(source_data_dir, "wes_or_meta_aggregated.csv"))

wes_or_meta_aggregated$feature <- factor(wes_or_meta_aggregated$feature,
                                         levels = c(log_or_variable_order))

wes_or_meta_aggregated_plot <- wes_or_meta_aggregated %>%
  mutate(label = cut(
    pval.random,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(Analysis_subset,
               levels = c("Original", "Excl_mono.", "Excl_clin.", "MPR", "pCR")), 
    y = factor(
      feature, levels = fct_rev(as.factor(log_or_variable_order))),
    fill = TE.random,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "", title = "Meta-analysis OR sensitivity", fill = "Meta-analysis\nlog(OR)") 

wes_or_meta_aggregated_plot

graph2pdf(x = wes_or_meta_aggregated_plot,
          file = here(output_dir, "_wes_or_meta_aggregated_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 7)

```

# ----------
## Extended Data Fig 2 COSMIC
```{r}
cosmic_v2_aetiology <- read_excel(
  here(source_data_dir, "COSMIC_SBS_Aetiology_20250728.xlsx"),
  sheet = "v2.0")
cosmic_v2_aetiology_p <- cosmic_v2_aetiology %>%
  mutate(Aetiology_simplified = Aetiology_simplified_later,
         `Signature Number` = sprintf("Signature %s", parse_number(COSMIC_SBS))
         ) %>%
  select(COSMIC_SBS, `Signature Number`, Aetiology_simplified) %>%
  mutate(cosmic_signature = gsub(" ", ".", `Signature Number`))

cosmic_combined <- read_csv2(
  here(source_data_dir, "cosmic_combined_v2_GRCh38.csv"))
cosmic_perc_combined <- read_csv2(
  here(source_data_dir, "cosmic_perc_combined_v2_GRCh38.csv"))

cosmic_v2_aetiology_p

cosmic_combined <- cosmic_combined %>%
  filter(CF_sample_tumor %in% dna_metadata_f$CF_sample_tumor)

cosmic_perc_combined <- cosmic_perc_combined %>%
  filter(CF_sample_tumor %in% dna_metadata_f$CF_sample_tumor)

cosmic_combined_l <- cosmic_combined %>%
  pivot_longer(cols = c(paste0("Signature.", c(1:30))),
               names_to = "cosmic_signature",
               values_to = "cosmic_scores") %>%
  left_join(cosmic_v2_aetiology_p)

cosmic_perc_combined_l <- cosmic_perc_combined %>%
  pivot_longer(cols = c(paste0("Signature.", c(1:30))),
               names_to = "cosmic_signature",
               values_to = "percentage") %>%
  left_join(cosmic_v2_aetiology_p)

cosmic_combined_l$cosmic_signature <- factor(
  cosmic_combined_l$cosmic_signature,
  levels = c(paste0("Signature.", c(1:30))))
cosmic_perc_combined_l$cosmic_signature <- factor(
  cosmic_perc_combined_l$cosmic_signature,
  levels = c(paste0("Signature.", c(1:30))))

cosmic_colors <- c(Aflatoxin = "#d0e0e3", Aging = "#3d85c6",
                   APOBEC = "#674ea7",
                   Aristolochic_acid = "#d9d2e9", dMMR = "#B73A3A",
                   HRD = "#e69138", Other = "#cccccc", POL1 = "#cc0000",
                   POLE_POLD1 = "#5b0f00", POLH = "#FEB5DA", ROS = "#6DB6FF",
                   SBS5 = "#ea9999", Smoking = "#38761d", TMZ = "#b5a7d6",
                   Tobacco_chewing = "#6aa84f", UV = "#f1c232")

cosmic_combined_p <- cosmic_combined_l %>%
  group_by(CF_sample_tumor, Aetiology_simplified) %>%
  summarize(cosmic_scores = sum(cosmic_scores)) %>%
  pivot_wider(names_from = Aetiology_simplified,
              values_from = cosmic_scores) %>%
  left_join(dna_metadata_f) %>%
  left_join(histology_msi_mss_label_dict)

cosmic_perc_combined_p <- cosmic_perc_combined_l %>%
  group_by(CF_sample_tumor, Aetiology_simplified) %>%
  summarize(percentage = sum(percentage)) %>%
  pivot_wider(names_from = Aetiology_simplified, values_from = percentage) %>%
  left_join(dna_metadata_f) %>%
  left_join(histology_msi_mss_label_dict)

```

```{r, fig.width = 12, fig.height = 12}
mean_cosmic_abs <- read_csv2(here(source_data_dir, "mean_cosmic_abs.csv"))
mean_cosmic_percs <- read_csv2(here(source_data_dir, "mean_cosmic_percs.csv"))
```

## Extended Fig. 2a
### Perc COSMIC proportions
```{r, fig.width = 5.5, fig.height = 6}
mean_cosmic_perc_plot <- mean_cosmic_percs %>%
  left_join(cosmic_v2_aetiology_p) %>%
  mutate(Aetiology_simplified = ifelse(
    Aetiology_simplified %in% c("Aflatoxin", "Aristolochic_acid",
                                "POL1", "POLH"),
    "Other", Aetiology_simplified)) %>%
  group_by(Aetiology_simplified, Trial) %>%
  summarize(mean_percentage = sum(mean_percentage)) %>%
  ggplot(aes(x = Trial, y = mean_percentage, fill = Aetiology_simplified)) +
  geom_col() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = cosmic_colors) +
  labs(x = "", y = "Mean relative frequency", fill = "COSMIC aetiology") +
  guides(fill = guide_legend(nrow = 4)) +
  theme(axis.title.x = element_blank())

mean_cosmic_perc_plot
graph2pdf(x = mean_cosmic_perc_plot,
          file = here(output_dir, "_mean_cosmic_perc_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5.5,
          width = 6)
```

## Extended Fig. 2b-d

### TMB vs COSMIC
```{r}

TMB_UV_plot <- wes_features_df %>%
  ggplot(aes(x = TMB, y = UV_abs, col = Trial)) +
  geom_point() +
  theme_Publication() +
  scale_color_manual(values = trial_colors)

TMB_UV_plot 
graph2pdf(x = TMB_UV_plot,
          file = here(output_dir, "_TMB_UV_trial.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 6)

TMB_apobec_plot <- wes_features_df %>%
  ggplot(aes(x = TMB, y = APOBEC_abs, col = Trial)) +
  geom_point() +
  theme_Publication() +
  scale_color_manual(values = trial_colors)

TMB_apobec_plot
graph2pdf(x = TMB_apobec_plot,
          file = here(output_dir, "_TMB_APOBEC_trial.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 6)


TMB_dMMR_plot <- wes_features_df %>%
  ggplot(aes(x = TMB, y = dMMR_abs, col = Trial)) +
  geom_point() +
  theme_Publication() +
  scale_color_manual(values = trial_colors)

TMB_dMMR_plot
graph2pdf(x = TMB_dMMR_plot,
          file = here(output_dir, "_TMB_dMMR_trial.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 6)

wes_features_df %>%
  ggplot(aes(x = TMB, y = HRD_abs, col = Trial)) +
  geom_point() +
  theme_Publication() +
  scale_color_manual(values = trial_colors)
ggsave(here(output_dir, "TMB_HRD_trial.png"), height = 5, width = 6)

```

## Extended Fig. 2e

### TMB vs frameshift
```{r}
TMB_frameshift_plot <- n_variants_dp20_vaf_tmb %>%
  ggplot(aes(x = TMB, y = TMB_frameshift, col = Trial)) +
  geom_point() +
  theme_Publication() +
  scale_color_manual(values = trial_colors)

TMB_frameshift_plot 
graph2pdf(x = TMB_frameshift_plot,
          file = here(output_dir, "_TMB_frameshift_trial.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 6)
```


## Extended Fig. 2f
### Mean absolute COSMIC fits
```{r, fig.width = 12, fig.height = 12}
mean_cosmic_abs_plot <- mean_cosmic_abs %>%
  left_join(cosmic_v2_aetiology_p) %>%
  left_join(histology_df) %>%
  select(Trial, Aetiology_simplified, mean_abs) %>%
  group_by(Trial, Aetiology_simplified) %>%
  summarize(mean_abs = sum(mean_abs)) %>%
  ggplot(aes(x = Trial, y = mean_abs, fill = Aetiology_simplified)) +
  geom_col() +
  facet_wrap(~Aetiology_simplified, scales = "free") +
  scale_fill_manual(values = cosmic_colors) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(axis.title.x = element_blank()) +
  labs(y = "mean COSMIC SBS")

mean_cosmic_abs_plot
graph2pdf(x = mean_cosmic_abs_plot,
          file = here(output_dir, "_mean_cosmic_abs_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 12,
          width = 12)
```


# ----------
# Extended Data Fig 3
## Extended Data Fig 3a
### Oncoplot
```{r}
genes_to_assess_shortlist <- unique(c(ddr_ptns, crc_risk_ptns, immune_escape_genes,
                     "CD58", "CD274", "SETDB1", "CDKN2A", "TRAF2", "CCND1")) %>%
  unique() # leaving out Grasso for now grasso_2018_mss

# Dev off only works from the console, and not from the rmd chunk. Copy and run
# do not run from chunks and knit
if (FALSE) {
pdf(file = here(output_dir, "_oncoplot_genes.pdf"), width = 10, height = 16)
oncoplot(combined_maf_specific_f,
         genes = genes_to_assess_shortlist[!genes_to_assess_shortlist %in%
                                             c("RAD51", "CD274", "HLA−A",
                                               "HLA−B", "HLA−C",
                                               "POLE", "POLD1")], 
         clinicalFeatures = c("Response", "Trial"),
         annotationColor = annotation_color_list,
         topBarData = "TMB",
         drawColBar = TRUE,
         sortByAnnotation = TRUE,
         removeNonMutated = FALSE
         )
dev.off()
}

genes_to_assess_shortlist

# Dev off only works from the console, and not from the rmd chunk. Copy and run
# do not run from chunks and knit
if (FALSE) {
pdf(file = here(output_dir, "_oncoplot_ap_ifn_pathway.pdf"), width = 10, height = 8)
oncoplot(combined_maf_specific_f,
         pathways = immune_escape_genes_df %>%
           select(Gene, Pathway.general) %>%
           filter(Pathway.general %in% c("Antigen Presentation",
                                         "IFN-y pathway")),
         clinicalFeatures = c("Trial", "Response"),
         topBarData = "TMB",
         annotationColor = annotation_color_list,
         gene_mar = 8, fontSize = 0.6,
         sortByAnnotation = TRUE,
         removeNonMutated = FALSE)
dev.off()

png(file = here(output_dir, "_oncoplot_specific_genes_trial.png"),
    width = 20, height = 30)
}

```

## Extended Data Fig 3b
### Pathway data
```{r}
trial_list <- c("Combined", "BELLINI", "IMCISION", "MATISSE",
                "NABUCCO", "NICHE-MSS",  "NICHE-MSI", "OpACIN_neo", "PRADO")

trial_group_map <- data.frame(Trial = c(trial_list, trial_list, trial_list),
           Group = c(trial_list,
                     sprintf("%s_R", trial_list),
                     sprintf("%s_NR", trial_list)))

pathway_data_combined <- read.csv2(
  here(source_data_dir, "pathway_data_combined.csv")) %>%
  left_join(trial_group_map)
pathway_gene_data_combined <- read.csv2(
  here(source_data_dir, "pathway_gene_data_combined.csv")) %>%
  left_join(trial_group_map)

pathway_data_combined$Group <- factor(pathway_data_combined$Group,
                                      levels = c(trial_list,
                                                 sprintf("%s_R", trial_list),
                                                 sprintf("%s_NR", trial_list)))

pathway_data_combined <- pathway_data_combined %>%
  left_join(trial_group_map)

source_data_pathway_plot <- pathway_data_combined %>%
  mutate(Pathway = ifelse(Pathway == "Protein_homeostasis_ubiquitination",
                          "PTN Homeost. Ubiquitin.", Pathway)) %>%
  filter(Pathway %in% c("Antigen_Presentation", "IFN_y_pathway", "SIGPW_WNT",
                        "SIGPW_PI3K", "SIGPW_Cell_Cycle",
                        "SIGPW_TGF_Beta")) %>%
  filter(!str_detect(Group, "_R|_NR"))

pathway_data_combined$Trial <- factor(pathway_data_combined$Trial,
                                      levels = trial_list)
source_data_pathway_plot$Trial <- factor(source_data_pathway_plot$Trial,
                                      levels = trial_list)
source_data_pathway_plot$Group <- factor(source_data_pathway_plot$Group,
                                      levels = trial_list)
pathway_data_combined_plot <- source_data_pathway_plot %>%
  ggplot(aes(x = fct_rev(Group), y = Fraction_mutated_samples,
             fill = Group)) + geom_col() +
  facet_wrap(~Pathway) + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  coord_flip() + scale_fill_manual(values = trial_colors_combined) +
  guides(fill = FALSE) + labs(x = "", y = "Proportion of mutated samples")

pathway_data_combined_plot

graph2pdf(x = pathway_data_combined_plot,
          file = here(output_dir, "_pathway_data_combined_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 6, #9
          width = 8) #11


```

## Extended Data Fig 3c-h

### AP Trial Response

```{r, fig.height = 4, fig.width = 6}
# GIE_AP Genetic Immune evasion - Antigen Presentation

ap_mt_response_plot <- table(wes_features_df$Trial,
       wes_features_df$GIE_AP,
      wes_features_df$Response, useNA = "ifany") %>%
  as.data.frame() %>%
  setNames(c("Trial", "GIE_AP", "Response", "Freq")) %>%
  mutate(GIE_AP = ifelse(GIE_AP == TRUE,
                         "Antigen presentation mut", "Antigen presentation wt")) %>%
  ggplot(aes(x = Response, y = Freq, fill = Trial)) +
  geom_col() + theme_Publication() +
  scale_fill_manual(values = trial_colors) + facet_wrap(~GIE_AP)
 
ap_mt_response_plot

graph2pdf(x = ap_mt_response_plot,
          file = here(
            output_dir, "_ap_mt_response_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 5.5)

wes_features_df_excl_dmmr <- wes_features_df %>% filter(!Trial %in% c("NICHE-MSI"))

ap_mt_response_plot_excl_dmmr <- table(wes_features_df_excl_dmmr$Trial,
       wes_features_df_excl_dmmr$GIE_AP,
      wes_features_df_excl_dmmr$Response, useNA = "ifany") %>%
  as.data.frame() %>%
  setNames(c("Trial", "GIE_AP", "Response", "Freq")) %>%
  mutate(GIE_AP = ifelse(GIE_AP == TRUE,
                         "Antigen presentation mut", "Antigen presentation wt")) %>%
  ggplot(aes(x = Response, y = Freq, fill = Trial)) +
  geom_col() + theme_Publication() +
  scale_fill_manual(values = trial_colors) + facet_wrap(~GIE_AP) +
  labs(title = "Excluding dMMR")

ap_mt_response_plot_excl_dmmr

graph2pdf(x = ap_mt_response_plot_excl_dmmr,
          file = here(
            output_dir, "_ap_mt_response_plot_excl_dmmr.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 5.5)

table(wes_features_df$Response,
      wes_features_df$GIE_AP) %>%
  `colnames<-`( c("AP wt", "AP mt"))

table(wes_features_df$Response,
      wes_features_df$GIE_AP) %>%
  fisher.test()

table(wes_features_df$Response,
      wes_features_df$GIE_AP) %>%
  prop.table(margin = 1) %>%
  round(2)  %>%
  `colnames<-`( c("AP wt", "AP mt"))

table(wes_features_df$Response,
      wes_features_df$GIE_AP) %>%
  prop.table(margin = 2) %>%
  round(2) %>%
  `colnames<-`( c("AP wt", "AP mt"))

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$GIE_AP) %>%
  `colnames<-`( c("AP wt", "AP mt"))

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$HLA_alteration) %>%
  fisher.test()

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$GIE_AP) %>%
  prop.table(margin = 1) %>%
  round(2)  %>%
  `colnames<-`( c("AP wt", "AP mt"))

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$GIE_AP) %>%
  prop.table(margin = 2) %>%
  round(2)  %>%
  `colnames<-`( c("AP wt", "AP mt"))
```
### IFNy Trial Response
```{r, fig.height = 4, fig.width = 6}
ifny_mt_response_plot <- table(wes_features_df$Trial,
       wes_features_df$GIE_IFNy,
      wes_features_df$Response, useNA = "ifany") %>%
  as.data.frame() %>%
  setNames(c("Trial", "GIE_IFNy", "Response", "Freq")) %>%
  mutate(GIE_IFNy = ifelse(GIE_IFNy == TRUE,
                         "IFNy pathway mut", "IFNy pathway wt")) %>%
  ggplot(aes(x = Response, y = Freq, fill = Trial)) +
  geom_col() + theme_Publication() +
  scale_fill_manual(values = trial_colors) + facet_wrap(~GIE_IFNy)
 
ifny_mt_response_plot

graph2pdf(x = ifny_mt_response_plot,
          file = here(
            output_dir, "_ifny_mt_response_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 5.5)

ifny_mt_response_plot_excl_dmmr <- table(wes_features_df_excl_dmmr$Trial,
       wes_features_df_excl_dmmr$GIE_IFNy,
      wes_features_df_excl_dmmr$Response, useNA = "ifany") %>%
  as.data.frame() %>%
  setNames(c("Trial", "GIE_IFNy", "Response", "Freq")) %>%
  mutate(GIE_IFNy = ifelse(GIE_IFNy == TRUE,
                         "IFNy pathway mut", "IFNy pathway wt")) %>%
  ggplot(aes(x = Response, y = Freq, fill = Trial)) +
  geom_col() + theme_Publication() +
  scale_fill_manual(values = trial_colors) + facet_wrap(~GIE_IFNy) +
  labs(title = "Excluding dMMR")

ifny_mt_response_plot_excl_dmmr

graph2pdf(x = ifny_mt_response_plot_excl_dmmr,
          file = here(
            output_dir, "_ifny_mt_response_plot_excl_dmmr.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 5.5)

table(wes_features_df$Response,
      wes_features_df$GIE_IFNy) %>%
  `colnames<-`( c("IFNy wt", "IFNy mt"))

table(wes_features_df$Response,
      wes_features_df$GIE_IFNy) %>%
  fisher.test()

table(wes_features_df$Response,
      wes_features_df$GIE_IFNy) %>%
  prop.table(margin = 1) %>%
  round(2) %>%
  `colnames<-`( c("IFNy wt", "IFNy mt"))

table(wes_features_df$Response,
      wes_features_df$GIE_IFNy) %>%
  prop.table(margin = 2) %>%
  round(2) %>%
  `colnames<-`( c("IFNy wt", "IFNy mt"))

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$GIE_IFNy) %>%
  `colnames<-`( c("IFNy wt", "IFNy mt"))

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$HLA_alteration) %>%
  fisher.test()

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$GIE_IFNy) %>%
  prop.table(margin = 1) %>%
  round(2) %>%
  `colnames<-`( c("IFNy wt", "IFNy mt"))

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$GIE_IFNy) %>%
  prop.table(margin = 2) %>%
  round(2) %>%
  `colnames<-`( c("IFNy wt", "IFNy mt"))
```

### HLA Alteration Response
```{r}
hla_alteration_response_plot <- table(wes_features_df$Trial,
       wes_features_df$HLA_alteration,
      wes_features_df$Response, useNA = "ifany") %>%
  as.data.frame() %>%
  setNames(c("Trial", "HLA_alteration", "Response", "Freq")) %>%
  mutate(HLA_alteration = ifelse(HLA_alteration == TRUE,
                         "HLA altered", "HLA intact")) %>%
  ggplot(aes(x = Response, y = Freq, fill = Trial)) +
  geom_col() + theme_Publication() +
  scale_fill_manual(values = trial_colors) + facet_wrap(~HLA_alteration)

hla_alteration_response_plot

graph2pdf(x = hla_alteration_response_plot,
          file = here(
            output_dir, "_hla_alteration_response_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 5.5)

hla_alteration_response_plot_excl_dmmr <-  table(wes_features_df_excl_dmmr$Trial,
       wes_features_df_excl_dmmr$HLA_alteration,
      wes_features_df_excl_dmmr$Response, useNA = "ifany") %>%
  as.data.frame() %>%
  setNames(c("Trial", "HLA_alteration", "Response", "Freq"))  %>%
  mutate(HLA_alteration = ifelse(HLA_alteration == TRUE,
                         "HLA altered", "HLA intact")) %>%
  ggplot(aes(x = Response, y = Freq, fill = Trial)) +
  geom_col() + theme_Publication() +
  scale_fill_manual(values = trial_colors) + facet_wrap(~HLA_alteration) +
  labs(title = "Excluding dMMR")

hla_alteration_response_plot_excl_dmmr

graph2pdf(x = hla_alteration_response_plot_excl_dmmr,
          file = here(
            output_dir, "_hla_alteration_response_plot_excl_dmmr.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 5.5)
table(wes_features_df$Response,
      wes_features_df$HLA_alteration) %>%
  `colnames<-`( c("HLA_normal", "HLA_alteration"))

table(wes_features_df$Response,
      wes_features_df$HLA_alteration) %>%
  fisher.test()

table(wes_features_df$Response,
      wes_features_df$HLA_alteration) %>%
  prop.table(margin = 1) %>%
  round(2) %>%
  `colnames<-`( c("HLA_normal", "HLA_alteration"))

table(wes_features_df$Response,
      wes_features_df$HLA_alteration) %>%
  prop.table(margin = 2) %>%
  round(2) %>%
  `colnames<-`( c("HLA_normal", "HLA_alteration"))

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$HLA_alteration) %>%
  `colnames<-`( c("HLA_normal", "HLA_alteration"))

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$HLA_alteration) %>%
  fisher.test()

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$HLA_alteration) %>%
  prop.table(margin = 1) %>%
  round(2) %>%
  `colnames<-`( c("HLA_normal", "HLA_alteration"))

table(wes_features_df_excl_dmmr$Response,
      wes_features_df_excl_dmmr$HLA_alteration) %>%
  prop.table(margin = 2) %>%
  round(2) %>%
  `colnames<-`( c("HLA_normal", "HLA_alteration"))

# `rownames<-`(c("TMB-L", "TMB-H")) %>%
# `colnames<-`( c("mTIL_low", "mTIL_high"))
```

# ----------
## Extended Data Fig 4 FACETS

## Extended Data Fig 4a
```{r}
aneuploidy_trial_response_plot <- wes_features_df %>%
  ggplot(aes(x = factor(Trial,
                        levels = c("NICHE-MSI", "MATISSE",
                                   "OpACIN_neo", "PRADO", "BELLINI" ,"NICHE-MSS", "IMCISION", "NABUCCO")), y = Aneuploidy_score, fill = Response)) +
  geom_boxplot(position = position_dodge()) +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(x = "")

aneuploidy_trial_response_plot
graph2pdf(x = aneuploidy_trial_response_plot,
          file = here(output_dir, "_aneuploidy_trial_response_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 5)

```

## Extended Data Fig 4b
```{r}
scna_trial_response_plot <- wes_features_df %>%
  ggplot(aes(x = factor(Trial,
                        levels = c("NICHE-MSI", "MATISSE",
                                   "OpACIN_neo", "PRADO", "BELLINI",
                                   "NICHE-MSS", "IMCISION", "NABUCCO")),
             y = scna_loss, fill = Response)) +
  geom_boxplot(position = position_dodge()) +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(x = "")

scna_trial_response_plot
graph2pdf(x = scna_trial_response_plot,
          file = here(output_dir, "_scna_trial_response_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 5)
```

## Extended Data Fig 4c
### Facets Aneuplody vs scna
```{r, fig.height = 6, fig.width = 6}

wes_features_df %>%
  ggplot(aes(x = scna_loss, y = Aneuploidy_score,
             col = Trial, shape = Response)) +
  geom_point() +
  theme_Publication() +
  guides(alpha = FALSE) +
  facet_wrap(~Response) +
  scale_color_manual(values = trial_colors) +
  geom_abline(slope = 1, intercept = 0)

wes_features_df %>%
  ggplot(aes(x = scna_loss, y = Aneuploidy_score,
             col = Trial, shape = Response)) +
  geom_point() +
  theme_Publication() +
  guides(alpha = FALSE) +
  facet_wrap(~Trial) +
  scale_color_manual(values = trial_colors) +
  geom_abline(slope = 1, intercept = 0)

aneuploidy_scna_plot <- wes_features_df %>%
  ggplot(aes(x = scna_loss, y = Aneuploidy_score,
             col = Trial, shape = Response)) +
  geom_pointdensity() +
  theme_Publication() +
  guides(alpha = FALSE) +
  facet_wrap(~Trial) +
  scale_color_manual(values = trial_colors) +
  geom_abline(slope = 1, intercept = 0)

graph2pdf(x = aneuploidy_scna_plot,
          file = here(output_dir, "_aneuploidy_scna_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 7,
          width = 7)
```



## Extended Data Fig 4d
### Facets cytobands

```{r}
facets_c_cytobands_excl_dMMR_del_freq <- read_csv2(
  here(source_data_dir, "facets_c_cytobands_excl_dMMR_del_freq.csv"))

facets_c_cytobands_excl_dMMR_del_freq_plot <- facets_c_cytobands_excl_dMMR_del_freq[
  select_top_extremes(facets_c_cytobands_excl_dMMR_del_freq$Delta_del_hemizyg, top_n = 10) |
    facets_c_cytobands_excl_dMMR_del_freq$chr_arm_cytoband %in% c("chr9_p21.3", "chr9_q34.3"), ] %>%
  ggplot(aes(x = reorder(chr_arm_cytoband, Delta_del_hemizyg), y = Delta_del_hemizyg)) +
  geom_col() + coord_flip() + theme_Publication() +
  labs(x = "", y = "Difference in proportions (R - NR)",
       title = "Delta in cytoband losses")

facets_c_cytobands_excl_dMMR_del_freq_plot

graph2pdf(x = facets_c_cytobands_excl_dMMR_del_freq_plot,
          file = here(
            output_dir, "_facets_c_cytobands_excl_dMMR_del_freq_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 7,
          width = 6)
```


## Extended Data Fig 4e


```{r}
cytoband_df <- read.delim(
  here(source_data_dir, "cytoBand.txt"), header = F) %>%
  setNames(c("chr", "startpos", "endpos", "cytoband", "giemsa"))

cytoband_df <- cytoband_df %>%
  mutate(chr_arm = paste0(chr, ifelse(str_detect(cytoband, "p"), "p", "q")))

facets_c_cytobands_excl_dMMR_mean_cn  <- read_csv2(
  here(source_data_dir, "facets_c_cytobands_excl_dMMR_mean_cn.csv"))

cytoband_order <- cytoband_df %>%
  filter(!chr %in% c("chrX", "chrY")) %>%
  mutate(chr_arm_cytoband = paste0(chr, "_", cytoband))
cytoband_order$chr <- as.factor(cytoband_order$chr,
                                levels = c("chr1", "chr2", "chr3", "chr4",
                                           "chr5", "chr6", "chr7", "chr8",
                                           "chr9", "chr10", "chr11", "chr12",
                                           "chr13", "chr14", "chr15", "chr16",
                                           "chr17", "chr18", "chr19", "chr20",
                                           "chr21", "chr22"))

cytoband_order$chrom_num <- as.numeric(gsub("chr", "", cytoband_order$chr))

cytoband_order <- cytoband_order[order(cytoband_order$chrom_num), ]

cytoband_order_f <- cytoband_order %>%
  filter(
    chr_arm_cytoband %in% facets_c_cytobands_excl_dMMR_mean_cn$chr_arm_cytoband)

facets_c_cytobands_excl_dMMR_mean_cn$chr_arm_cytoband <- factor(
  facets_c_cytobands_excl_dMMR_mean_cn$chr_arm_cytoband,
  levels = cytoband_order_f$chr_arm_cytoband)

chr_label_positions <- facets_c_cytobands_excl_dMMR_mean_cn %>%
 group_by(chr) %>%
 slice(1) %>%
 ungroup()

chr_label_positions <- facets_c_cytobands_excl_dMMR_mean_cn[
  !duplicated(facets_c_cytobands_excl_dMMR_mean_cn$chr), ] %>%
  filter(!chr %in% c("chrX", "chrY"))
```

```{r}
facets_cytoband_cn_plot <- facets_c_cytobands_excl_dMMR_mean_cn %>%
  filter(!chr %in% c("chrX", "chrY")) %>%
  ggplot(aes(x = chr_arm_cytoband, y = mean_cn, col = Response)) +
  geom_point() + theme_Publication() +
  labs(x = "", y = "Average cytoband copy number") +
  scale_color_manual(values = c(Color_NR, Color_R)) +
  theme(axis.text.x = element_blank()) +
geom_text(data = chr_label_positions,
          aes(y = 4, label = chr), col = "black", vjust = -1.5, size = 3.5) 

facets_cytoband_cn_plot 
graph2pdf(x = facets_cytoband_cn_plot,
          file = here(output_dir, "_facets_cytoband_cn_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 20)
```

# ----------
# Figure 4
## Fig.4a
### Transcriptomics Meta-analysis plots

```{r, fig.height = 12, fig.width = 10}
meta_rna_studies_results <- read_csv2(
  here(source_data_dir, "meta_studies_results.csv"))
meta_rna_combined_effects <- read_csv2(
  here(source_data_dir, "meta_combined_effects.csv"))
meta_rna_combined_effects <- meta_rna_combined_effects %>%
  arrange(-TE.random)

meta_studies_results_random_effects <- read_csv2(
  here(source_data_dir, "meta_studies_results_random_effects.csv"))

meta_combined_effects_random_effects <- read_csv2(
  here(source_data_dir, "meta_combined_effects_random_effects.csv"))

meta_studies_results_rna_all <- read_csv2(
  here(source_data_dir, "meta_studies_results_smd_all.csv"))

meta_combined_effects_random_effects_f <- meta_combined_effects_random_effects %>%
  arrange(TE.random)

rna_meta_studies_results_sample_sizes <- meta_studies_results_rna_all %>%
  select(studlab, n.e, n.c) %>%
  mutate(R = n.e, NR = n.c, total = R + NR) %>%
  unique()
rna_meta_studies_results_sample_sizes

meta_studies_results_rna_all$studlab %>% unique()

combined_rna_smd_plot <- meta_studies_results_rna_all %>%
  mutate(label = cut(
    pval,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(studlab, levels = c("Combined", "Combined_excl_dMMR", "BELLINI",
                                   "IMCISION", "MATISSE",
                                   "NABUCCO", "NICHE-MSS",
                                   "OpACIN_neo", "PRADO")),
    y = factor(
      Variable, levels = meta_combined_effects_random_effects_f$Variable),
    fill = TE,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

graph2pdf(x = combined_rna_smd_plot,
          file = here(output_dir, "_combined_rna_smd_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 9,
          width = 7.7)

meta_combined_effects_random_effects_f


random_effects_meta_rna_plot <- meta_combined_effects_random_effects_f %>%
  ggplot(aes(x = reorder(Variable, TE.random),
    y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip(clip = "off") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis SMD (95% CI)") +
  guides(size = guide_legend(nrow = 3)) +
  geom_text(
    aes(label = pval.random, y = 0.9, hjust = 0),
    data = meta_combined_effects_random_effects_f %>%
      mutate(pval.random = sprintf("P = %s", FormatDecimals(pval.random)))) + 
      theme(plot.margin = unit(c(1,4,1,1), "lines")) # This keeps the labels from disappearing
  
random_effects_meta_rna_plot

graph2pdf(x = random_effects_meta_rna_plot,
          file = here(output_dir, "_random_effects_meta_rna_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 9,
          width = 6)

```

```{r, fig.height = 12, fig.width = 10}
combined_rna_smd_plot <- signature_response_c %>%
  filter(contrast == "Pre R x NR") %>%
  filter(signature %in% meta_rna_combined_effects$Signature) %>%
  ggplot(aes(x = Trial,
             y = factor(signature,
                        levels = rev(meta_rna_combined_effects$Signature)),
             fill = effect_size_hedges_g)) +
  geom_tile() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication()  +
  labs(y = "", fill = "SMD\n(Hedge's g)") + # \n(Hedge's g) (Cohen's d)
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")
combined_rna_smd_plot

meta_rna_combined_effects %>%
  select(Signature, k, pval.random) %>%
  mutate(pval.random = FormatDecimals(pval.random))

meta_rna_combined_smd_plot <- meta_rna_combined_effects %>%
  ggplot(aes(x = reorder(Signature, TE.random),
    y = TE.random)) +
  geom_point(aes(size = -log10(pval.random)), shape = 22, fill = "darkgrey") +
  theme_Publication() +
  coord_flip(clip = "off") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower.random, ymax = upper.random), width=.2) +
  labs(x = "", y = "Meta-analysis SMD (95% CI)") +
  guides(size = guide_legend(nrow = 3)) +
  geom_text(
    aes(label = pval.random, y = 0.9, hjust = 0),
    data = meta_rna_combined_effects %>%
      mutate(pval.random = sprintf("P = %s", FormatDecimals(pval.random)))) + 
      theme(plot.margin = unit(c(1,4,1,1), "lines")) # This keeps the labels from disappearing
  
meta_rna_combined_smd_plot

graph2pdf(x = meta_rna_combined_smd_plot,
          file = here(output_dir, "_meta_rna_combined_smd_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 9,
          width = 6)

```

## Fig.4b

### RNA corr_plot
```{r, fig.height = 20, fig.width = 20}
tmb_rna_features_pre_m <- rna_features_df_plus_wes_pre %>%
  select(all_of(c(meta_combined_effects_random_effects_f$Variable, "TMB"))) %>%
  as.matrix()

tmb_rna_features_pre_m %>%
  cor(use = "pairwise.complete.obs", method = "spearman") %>%
  corrplot(order = "hclust", type = "lower", tl.col = "black")

save_pheatmap_pdf(
  x = tmb_rna_features_pre_m %>%
    cor(use = "pairwise.complete.obs", method = "spearman") %>%
    corrplot(order = "hclust", type = "lower", tl.col = "black"),
  filename = here(
    output_dir,
    "rna_dna_pre_corplot.pdf"),
  width = 9, height = 9)

# RNA DNA
rna_dna_features_pre_m <- rna_features_df_plus_wes_pre %>%
  select(all_of(c(meta_combined_effects_random_effects_f$Variable,
                  "TMB", "APOBEC_abs", "UV_abs", "Aneuploidy_score"))) %>%
  as.matrix() 

rna_dna_features_pre_m %>%
  cor(use = "pairwise.complete.obs", method = "spearman") %>%
  corrplot(order = "hclust", type = "lower", tl.col = "black")

# save_pheatmap_pdf(
#   x =  rna_dna_features_pre_m %>%
#   cor(use = "pairwise.complete.obs", method = "spearman") %>%
#   corrplot(order = "hclust", type = "lower", tl.col = "black"),
#   filename = here(
#     output_dir,
#     "rna_dna_pre_corplot.pdf"),
#   width = 9, heighst = 9)

```

## Fig.4c
### PD-L1 combined

```{r}
PDL1_cat_combined <- read_csv2(here(source_data_dir, "PDL1_cat_combined.csv"))

n_labels <- PDL1_cat_combined %>%
group_by(Trial, PDL1_group) %>%
summarise(n = n()) %>%
mutate(
  Trial_PDL1_group = paste0(Trial, " ", PDL1_group),
  label = paste0(PDL1_group, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Trial_PDL1_group)

PDL1_category_response_plot <- table(PDL1_cat_combined$PDL1_group,
      PDL1_cat_combined$Response,
      PDL1_cat_combined$Trial) %>%
  as.data.frame() %>%
  setNames(c("PDL1_group", "Response", "Trial", "Freq")) %>%
  filter(! (Trial %in% c("BELLINI", "IMCISION", "NABUCCO") & PDL1_group %in% c("TPS < 1", "TPS >= 1"))) %>%
  filter(! (Trial %in% c("PRADO", "OpACIN_neo") & PDL1_group %in% c("CPS < 10", "CPS >= 10")))  %>%
  pivot_wider(values_from = Freq, names_from = Response) %>%
  mutate(Total = NR + R, Response_rate = R / Total) %>%
  mutate(Trial_PDL1_group = paste0(Trial, " ", PDL1_group)) %>%
  ggplot(aes(Trial_PDL1_group, y = Response_rate)) + geom_col() +
  theme_Publication() + facet_wrap(~Trial, scales = "free")+
  labs(x = "", y = "Response rate")  +
  scale_x_discrete(labels = label_map) 

PDL1_category_response_plot

graph2pdf(x = PDL1_category_response_plot,
          file = here(output_dir, "_PDL1_category_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 6.75)
```

## Fig.4d
### CD8_combined
```{r}
CD8_combined <- read_csv2(
  here(source_data_dir, "CD8_combined.csv"))
```

```{r}
table(CD8_combined$Trial, CD8_combined$Response)
table(CD8_combined$Response)

n_labels <- CD8_combined %>%
  group_by(Response) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Response)

CD8_combined_response_plot <- CD8_combined %>%
  ggplot(aes(x = Response, y = CD8_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE) +
  labs(title = "Pooled Cohort (TNBC, HNSCC, \nUrothelial, Colon pMMR)",
       x = "", y = "CD8+ cells/mm2") +
  scale_x_discrete(labels = label_map)
  
CD8_combined_response_plot
graph2pdf(x = CD8_combined_response_plot,
          file = here(output_dir, "_CD8_combined_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.5)
```


## Fig.4e
### CD8_PD1_combined
```{r}
CD8_PD1_combined  <- read_csv2(
  here(source_data_dir, "CD8_PD1_combined.csv"))
```


```{r}
table(CD8_PD1_combined$Trial, CD8_PD1_combined$Response)
table(CD8_PD1_combined$Response)

n_labels <- CD8_PD1_combined %>%
  group_by(Response) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Response)

CD8_PD1_combined_response_plot <- CD8_PD1_combined %>%
  ggplot(aes(x = Response, y = CD8_PD1_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE) +
  labs(title = "Pooled Cohort (TNBC, \nUrothelial, Colon pMMR)",
       x = "", y = "CD8+ PD-1+ cells/mm2") +
  scale_x_discrete(labels = label_map)
  
CD8_PD1_combined_response_plot
graph2pdf(x = CD8_PD1_combined_response_plot,
          file = here(output_dir, "_CD8_PD1_combined_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.5)
```


```{r}
CD8_PD1_combined %>%
  ggplot(aes(x = Response, y = CD8_PD1_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)
```

```{r, fig.height = 5, fig.width = 10}

table(CD8_PD1_combined$Trial, CD8_PD1_combined$Response,
      is.na(CD8_PD1_combined$CD8_PD1_per_mm2))

CD8_PD1_combined %>%
  ggplot(aes(x = Response, y = CD8_PD1_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() + facet_wrap(~Trial) +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

```


# ----------
# Extended data Fig. 5
## Extended data Fig. 5a
### RNA Meta analysis sensitivity

```{r, fig.height = 8, fig.width = 7}

rna_meta_random_effects <- read_csv2(
    here(source_data_dir,
         "meta_combined_effects_random_effects.csv"))

rna_meta_aggregated <- read_csv2(here(source_data_dir,
                                      "rna_meta_aggregated_all_statistics.csv"))

signature_variable_order <- rna_meta_random_effects %>%
  mutate(Variable = ifelse(Variable == "Gamma_delta_TCR_Yu",
                           "Gamma_delta_TCRs", Variable)) %>%
  arrange(TE.random) %>%
  pull("Variable")

rna_smd_sensitivity_plot <- rna_meta_aggregated %>%
  mutate(label = cut(
    pval.random,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(Analysis_subset,
               levels = c("Original", "Excl_mono.", "Excl_clin.", "MPR", "pCR")), 
    y = factor(
      Variable, levels = signature_variable_order),
    fill = TE.random,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "", title = "Meta-analysis sensitivity",
       fill = "Meta-analysis\nSMD (Hedge's g)")

rna_smd_sensitivity_plot

graph2pdf(x = rna_smd_sensitivity_plot,
          file = here(output_dir, "_rna_smd_sensitivity_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 9,
          width = 7.7)
```

## Extended data Fig. 5b
### CD8_combined per cohort

```{r, fig.height = 4, fig.width = 10}
table(CD8_combined$Trial, CD8_combined$Response)

CD8_combined %>%
  ggplot(aes(x = Response, y = CD8_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() +
  facet_wrap(~Trial, scales = "free", nrow = 1) +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE) +
  labs(title = "Combined")

n_labels <- CD8_combined %>%
  group_by(Response, Trial) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"),
  Trial_response = paste0(Trial, "_", Response))
label_map <- setNames(n_labels$label, n_labels$Trial_response)

CD8_per_cohort_response_plot <- CD8_combined %>%
  mutate(Trial_response = paste0(Trial, "_", Response)) %>%
  ggplot(aes(x = Trial_response, y = CD8_per_mm2)) +
  geom_boxplot(aes(fill = Response)) + geom_point() +
  facet_wrap(~Trial, scales = "free", nrow = 1) + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE) +
  labs(x = "", y = "CD8+ cells/mm2") +
  scale_x_discrete(labels = label_map)

CD8_per_cohort_response_plot
graph2pdf(x = CD8_per_cohort_response_plot,
          file = here(output_dir, "_CD8_per_cohort_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4,
          width = 8)
```

## Extended data Fig. 5c
### CD8_PD1_combined per cohort

```{r, fig.height = 5, fig.width = 10}
table(CD8_PD1_combined$Trial, CD8_PD1_combined$Response)

CD8_PD1_combined %>%
  ggplot(aes(x = Response, y = CD8_PD1_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() + facet_wrap(~Trial, scales = "free") +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE) +
  labs(title = "Combined")

n_labels <- CD8_PD1_combined %>%
  group_by(Response, Trial) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"),
  Trial_response = paste0(Trial, "_", Response))
label_map <- setNames(n_labels$label, n_labels$Trial_response)

CD8_PD1_per_cohort_response_plot <- CD8_PD1_combined %>%
  mutate(Trial_response = paste0(Trial, "_", Response)) %>%
  ggplot(aes(x = Trial_response, y = CD8_PD1_per_mm2)) +
  geom_boxplot(aes(fill = Response)) + geom_point() +
  facet_wrap(~Trial, scales = "free") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE) +
  labs(x = "", y = "CD8+ PD-1+ cells/mm2") +
  scale_x_discrete(labels = label_map)

CD8_PD1_per_cohort_response_plot

graph2pdf(x = CD8_PD1_per_cohort_response_plot,
          file = here(output_dir, "_CD8_PD1_per_cohort_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4,
          width = 5.5)
```


## Extended Data Fig. 5d
### B catenin TIL score correlations per trial

```{r, fig.height = 8, fig.width = 8}
B_cat_til_trial_corr_plot <- rna_features_df_plus_wes_pre %>%
  ggplot(aes(x = B_catenin_signature_Luke_et_al, y = TIL_score, col = Trial)) +
  geom_point(aes()) + theme_Publication() + stat_cor(method = "pearson") +
  geom_smooth(method = "lm") +
  scale_color_manual(values = trial_colors) + facet_wrap(~Trial) +
  guides(col = FALSE) +
  labs(x = "B-catenin signature ssGSEA", y = "TIL score ssGSEA")

B_cat_til_trial_corr_plot

graph2pdf(x = B_cat_til_trial_corr_plot,
          file = here(output_dir, "_B_cat_til_trial_corr_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 7,
          width = 7)
```

## Extended Data Fig. 5e
### B catenin per trial
```{r, fig.height = 4, fig.width = 5}
B_cat_trial_plot <- rna_features_df_plus_wes_pre %>%
  ggplot(aes(x = Trial, y = B_catenin_signature_Luke_et_al)) +
  geom_boxplot() +
  geom_point(aes(col = Trial)) + theme_Publication() +
  scale_color_manual(values = trial_colors) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  labs(y = "B-catenin signature ssGSEA") +
  theme(axis.title.x = element_blank()) +
  guides(col = FALSE)

B_cat_trial_plot

graph2pdf(x = B_cat_trial_plot,
          file = here(output_dir, "_B_cat_trial_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 5)
```

# ----------
# Figure 5
## Fig.5a
### ML RF CV Schemes

```{r}
metastatic_wes_rna_features_df_pre <- read_csv2(
  here(source_data_dir, "metastatic_wes_rna_features_df_pre.csv"))
```

```{r}
outer_fold_results <- readRDS(here(source_data_dir, "outer_fold_results.rds"))
data_partition <- read_csv2(here(source_data_dir, "data_partition.csv"))

all_outer <- bind_rows(outer_fold_results) %>%
  mutate(ground_truth = truth) %>%
  mutate(truth = ifelse(truth %in% c(1, "R"), 1, 0)) %>%
  mutate(outer_fold = paste0("fold_", outer_fold)) %>%
  mutate(row_index = as.character(row_index)) %>%
  left_join(data_partition %>%
              mutate(row_index  = as.character(1:nrow(data_partition))))

table(all_outer$ground_truth, all_outer$truth)
# table(all_outer$outer_fold, all_outer$trial_response)

varimp_summary <- read_csv2(here(source_data_dir, "varimp_summary.csv"))

```

## Fig.5b
### ROC plot Outer Folds CV
```{r}
folds <- unique(all_outer$outer_fold)
roc_list <- list()

# --- Compute ROC per fold ---
roc_list <- list()

for (f in folds) {
  df_fold <- all_outer[all_outer$outer_fold == f, ]
  
  truth <- factor(df_fold$ground_truth, levels = c("NR", "R"))
  pred   <- df_fold$pred
  
  if (length(unique(truth)) < 2) {
    cat("Skipping", f, "→ only one class present\n")
    next
  }
  
  roc_list[[f]] <- roc(truth, pred, quiet = TRUE)
}

# Common FPR grid
fpr_grid <- seq(0, 1, length.out = 200)

# Interpolate each fold onto the grid
interp_df <- map_df(names(roc_list), function(f) {
  r <- roc_list[[f]]
  tibble(
    fold = f,
    fpr  = fpr_grid,
    tpr  = approx(
      x = 1 - r$specificities,
      y = r$sensitivities,
      xout = fpr_grid
    )$y
  )
})

# Mean and SD across folds
summary_df <- interp_df %>%
  group_by(fpr) %>%
  summarise(
    mean_tpr = mean(tpr, na.rm = TRUE),
    sd_tpr   = sd(tpr, na.rm = TRUE),
    upper    = pmin(mean_tpr + sd_tpr, 1),
    lower    = pmax(mean_tpr - sd_tpr, 0)
  )

# --- ggplot ---
Outer_fold_ROC_plot <- ggplot() +
  # Per-fold ROC curves
  geom_line(
    data = interp_df,
    aes(x = fpr, y = tpr, group = fold),
    color = "grey80"
  ) +
  # SD ribbon
  geom_ribbon(
    data = summary_df,
    aes(x = fpr, ymin = lower, ymax = upper),
    fill = "skyblue",
    alpha = 0.3
  ) +
  # Mean ROC curve
  geom_line(
    data = summary_df,
    aes(x = fpr, y = mean_tpr),
    color = "blue",
    linewidth = 1.2
  ) +
  
  # Diagonal reference line
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  
  labs( x = "False Positive Rate", y = "True Positive Rate"
  ) +coord_equal() +
  theme_Publication()

Outer_fold_ROC_plot

graph2pdf(x = Outer_fold_ROC_plot,
          file = here(output_dir, "_Outer_fold_ROC_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 5.5)
```

### Performance metrics outer fold summaries
```{r}

auc_per_fold <- all_outer %>%
  group_by(outer_fold) %>%
  summarise(
    auc = as.numeric(pROC::auc(pROC::roc(truth, pred)))
  )

summary_auc <- auc_per_fold %>%
  summarise(
    mean_auc = mean(auc),
    sd_auc = sd(auc)
  )

summary_auc %>% round(2)

safe_sens <- function(truth, pred) {
  if (length(unique(truth)) < 2 || length(unique(pred)) < 2) return(NA_real_)
  caret::sensitivity(truth, pred, positive = "R")
}

safe_spec <- function(truth, pred) {
  if (length(unique(truth)) < 2 || length(unique(pred)) < 2) return(NA_real_)
  caret::specificity(truth, pred, negative = "NR")
}

results <- list()

for (folds in unique(all_outer$outer_fold)) {
  
  cat("Processing:", folds, "\n")
  
  df_fold <- all_outer[all_outer$outer_fold == folds, ]
  
  truth_f <- factor(df_fold$truth == 1, levels = c(FALSE, TRUE))
  pred_f  <- factor(df_fold$pred > 0.5, levels = c(FALSE, TRUE))
  
  # Check class presence
  if (length(unique(truth_f)) < 2) {
    cat("  Skipped:", folds, "→ only one class present in ground truth\n")
    sens <- NA_real_
    spec <- NA_real_
  } else if (length(unique(pred_f)) < 2) {
    cat("  Warning:", folds, "→ model predicted only one class\n")
    sens <- NA_real_
    spec <- NA_real_
  } else {
    sens <- caret::sensitivity(truth_f, pred_f, positive = TRUE)
    spec <- caret::specificity(truth_f, pred_f, negative = FALSE)
  }
  
  results[[folds]] <- data.frame(
    outer_fold = folds,
    sens = sens,
    spec = spec
  )
}

sens_spec_per_fold <- do.call(rbind, results)

summary_sens_spec <- sens_spec_per_fold %>%
  summarise(
    mean_sens = mean(sens),
    sd_sens = sd(sens),
    mean_spec = mean(spec),
    sd_spec = sd(spec)
  )

summary_sens_spec %>% round(2)


```

```{r}
folds <- unique(all_outer$outer_fold)
roc_list <- list()

# postscript("roc_plot.ps", width = 6, height = 6, horizontal = FALSE, paper = "special")
for (f in folds) {
  df_fold <- all_outer[all_outer$outer_fold == f, ]
  
  truth <- factor(df_fold$ground_truth, levels = c("NR", "R"))
  pred   <- df_fold$pred
  
  # Skip folds with only one class
  if (length(unique(truth)) < 2) {
    cat("Skipping", f, "→ only one class present\n")
    next
  }
  
  roc_list[[f]] <- roc(truth, pred, quiet = TRUE)
}

# Common grid of FPR values
fpr_grid <- seq(0, 1, length.out = 200)

# Interpolate TPR for each fold
tpr_mat <- sapply(roc_list, function(r) {
  approx(
    x = 1 - r$specificities,
    y = r$sensitivities,
    xout = fpr_grid
  )$y
})

mean_tpr <- rowMeans(tpr_mat, na.rm = TRUE)
sd_tpr   <- apply(tpr_mat, 1, sd, na.rm = TRUE)

plot(NULL, xlim=c(0,1), ylim=c(0,1),
     xlab="False Positive Rate", ylab="True Positive Rate")

# Per-fold ROC curves
for (r in roc_list) {
  lines(1 - r$specificities, r$sensitivities, col="grey80")
}

# Shaded SD band
upper <- pmin(mean_tpr + sd_tpr, 1)
lower <- pmax(mean_tpr - sd_tpr, 0)
polygon(
  c(fpr_grid, rev(fpr_grid)),
  c(upper, rev(lower)),
  col=adjustcolor("skyblue", alpha.f=0.3),
  border=NA
)

# Mean ROC curve
lines(fpr_grid, mean_tpr, col="blue", lwd=2)

abline(0,1,lty=2)

# Close the device and write the file
dev.off()
```

## Fig.5c
### ROC LOOCV and combined outer folds

```{r}
auc_comparison <- read_csv2(here(source_data_dir, "auc_comparison.csv"))

auc_comparison_plot <- auc_comparison %>%
  left_join(summary_auc %>%
              mutate(Trial = "Combined_outer_folds")) %>%
  ggplot(aes(x = reorder(Trial, AUC), y = AUC, fill = Trial)) +
  geom_col() +
  geom_errorbar(
    aes(ymin = AUC - sd_auc, ymax = AUC + sd_auc),
    width = 0.2,
    linewidth = 0.5
  )  + theme_Publication() + coord_flip() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_fill_manual(values = trial_colors) +
  labs(x = "") +
  guides(fill = FALSE)

auc_comparison_plot

graph2pdf(x = auc_comparison_plot,
          file = here(output_dir, "_auc_comparison_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4.5,
          width = 6.5)

```

```{r}

auc_comparison_aggregate_all <- read_csv2(
  here(source_data_dir, "auc_comparison_aggregate_all.csv"))

auc_comparison_aggregate_all_plot <- auc_comparison_aggregate_all %>%
  filter(!(Training_set %in% c("Pancancer Nested CV") & outer_fold
           %in% c("fold_1", "fold_2", "fold_3", "fold_4", "fold_5"))) %>%
  filter(!Training_set %in% c("Pancancer LOSO")) %>%
  ggplot(aes(x = reorder(Trial, AUC), y = AUC, fill = Trial,
             shape = Training_set)) +
  geom_col() + theme_Publication() + coord_flip() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_fill_manual(values = trial_colors)  +
  scale_color_manual(values = trial_colors)+
  geom_errorbar(
    aes(ymin = AUC - sd_auc, ymax = AUC + sd_auc),
    width = 0.2,
    linewidth = 0.5
  ) + geom_point(aes(x = Trial, y = AUC), data = auc_comparison_aggregate_all) +
  geom_point(aes(x = Trial, y = AUC, col = "black"),
             data = auc_comparison_aggregate_all %>%
               filter(Training_set == "Pancancer LOSO")) +
  labs(x = "") +
  guides(shape = guide_legend(nrow = 3), size = NULL) +
  # guides(fill = guide_legend(nrow = 3), size = NULL) +
  guides(fill = NULL, col = NULL)

auc_comparison_aggregate_all_plot

graph2pdf(x = auc_comparison_aggregate_all_plot,
          file = here(output_dir, "_auc_comparison_aggregate_all_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4.5,
          width = 6.5)

```

## Fig.5d
### Feature importance

```{r, fig.height = 16, fig.width = 8}
importance_df <- read_csv2(here(source_data_dir, "importance_df.csv"))
dim(importance_df)
n_distinct(importance_df$Variable)

importance_df_summary <- importance_df %>%
  group_by(Variable) %>%
  summarize(mean = mean(Overall_var_imp)) %>%
  arrange(-mean)

importance_df_summary

feature_importance_plot <- importance_df %>%
  ggplot(aes(x = factor(Variable, levels = rev(varimp_summary$Variable)),
             y = Overall_var_imp, col = model)) +
  geom_point() +
  coord_flip() + theme_Publication() +
  labs(x = "") +
  scale_color_manual(values = trial_colors_combined)

feature_importance_plot

graph2pdf(x = feature_importance_plot,
          file = here(output_dir, "_feature_importance_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 12,
          width = 6)
```

## Fig.5e
### Neoadjuvant quadrant response
```{r}
median_til_score_pre <- median(
  rna_features_df_plus_wes_pre$TIL_score)
median_til_score_pre

wes_rna_features_df_pre_f %>%
  ggplot(aes(x = log2(TMB), y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication()

# Quadrant RR
table(wes_rna_features_df_pre_f$TMB >= 10,
      wes_rna_features_df_pre_f$TIL_score >= median_til_score_pre) %>%
  `rownames<-`(c("TMB-L", "TMB-H")) %>%
  `colnames<-`( c("mTIL_low", "mTIL_high"))

quadrant_response_proportions <- table(
  wes_rna_features_df_pre_f$TMB >= 10,
  wes_rna_features_df_pre_f$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre_f$Response) %>%
  as.data.frame() %>%
  setNames(c("TMB", "mTIL_group", "Response", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB == TRUE, "TMB_high", "TMB_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions

quadrant_response_proportions %>%
  write_csv2(here(output_dir, "quadrant_response_proportions_tmb_til.csv"))

TMB_TIL_quadrant_response_proportions <- read_csv2(
  here(output_dir, "quadrant_response_proportions_tmb_til.csv")) 

TMB_TIL_quadrant_chisq <- quadrant_response_proportions %>%
  select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()
TMB_TIL_quadrant_chisq

TMB_TIL_quadrant_response_proportions_plot <- quadrant_response_proportions %>%
  filter(Response == "R") %>%
  rename(Response_rate = quadrant_prop) %>%
  ggplot(aes(x = quadrant, y = Response_rate)) + geom_col() +
  theme_Publication() + labs(x = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text(aes(x = 3, y = 0.8, label = label),
                data = broom::tidy(TMB_TIL_quadrant_chisq) %>%
                  mutate(label = sprintf("X2 = %s; P = %s",
                                         round(statistic, 1), round(p.value, 9)))
                ) +
  labs(y = "Response rate") + theme(axis.title.x = element_blank())

TMB_TIL_quadrant_response_proportions_plot
graph2pdf(x = TMB_TIL_quadrant_response_proportions_plot,
          file = here(
            output_dir, "_TMB_TIL_quadrant_response_proportions_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 4)
```

## Fig.5g
## Neoadjuvant Quadrant plots Proliferation TIL score

```{r}
# Not filtering for TMB paired samples
dim(rna_features_df_plus_wes_pre)

median_til_score_pre <- median(
  rna_features_df_plus_wes_pre$TIL_score)
median_til_score_pre

median_Tumor_proliferation_rate_pre <- rna_features_df_plus_wes_pre %>%
  filter(Treatment == "Pre") %>%
  pull(Tumor_proliferation_rate) %>%
  median()
median_Tumor_proliferation_rate_pre

rna_features_df_plus_wes_pre %>%
  filter(Treatment == "Pre") %>% 
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Trial)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = median_Tumor_proliferation_rate_pre,
             linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  scale_color_manual(values = trial_colors) +
  guides(col = guide_legend(nrow = 3))
```


```{r}
quadrant_response_proportions <- table(
  rna_features_df_plus_wes_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  rna_features_df_plus_wes_pre$TIL_score > median_til_score_pre,
  rna_features_df_plus_wes_pre$Response) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Response", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>%
  add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions

quadrant_response_proportions %>% select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()

```

```{r}
trial_quadrant_proportions <- table(
  rna_features_df_plus_wes_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  rna_features_df_plus_wes_pre$TIL_score > median_til_score_pre,
  rna_features_df_plus_wes_pre$Trial) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE, "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
trial_quadrant_proportions

trial_quadrant_proportions %>%
  select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

quadrant_trial_proportions <- table(
  rna_features_df_plus_wes_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  rna_features_df_plus_wes_pre$TIL_score > median_til_score_pre,
  rna_features_df_plus_wes_pre$Trial) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = Freq / Total)

quadrant_response_proportions <- table(
  rna_features_df_plus_wes_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  rna_features_df_plus_wes_pre$TIL_score > median_til_score_pre,
  rna_features_df_plus_wes_pre$Response) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Response", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions
quadrant_response_proportions %>%
  write_csv2(here(output_dir, "quadrant_response_proportions_prolif_til.csv"))


Prolif_TIL_quadrant_response_proportions <- read_csv2(
  here(output_dir, "quadrant_response_proportions_prolif_til.csv")) 

Prolif_TIL_quadrant_chisq <- Prolif_TIL_quadrant_response_proportions %>%
  select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()
Prolif_TIL_quadrant_chisq

Prolif_TIL_quadrant_response_proportions_plot <- Prolif_TIL_quadrant_response_proportions %>%
  filter(Response == "R") %>%
  rename(Response_rate = quadrant_prop) %>%
  mutate(quadrant = factor(quadrant,
                           levels = c("Prolif_high_mTIL_high",
                                      "Prolif_low_mTIL_high",
                                      "Prolif_high_mTIL_low",
                                      "Prolif_low_mTIL_low"))) %>%
  ggplot(aes(x = quadrant, y = Response_rate)) + geom_col() +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text(aes(x = 3, y = 0.7, label = label),
                data = broom::tidy(Prolif_TIL_quadrant_chisq) %>%
                  mutate(label = sprintf("X2 = %s; P = %s",
                                         round(statistic, 1),
                                         round(p.value, 3)))
                ) +
  labs(y = "Response rate") + theme(axis.title.x = element_blank())

Prolif_TIL_quadrant_response_proportions_plot
graph2pdf(x = Prolif_TIL_quadrant_response_proportions_plot,
          file = here(
            output_dir, "_Prolif_TIL_quadrant_response_proportions_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 4)


prolif_TIL_quadrant_trial_proportions_plot <- quadrant_trial_proportions %>%
  ggplot(aes(x = quadrant, y = quadrant_prop, fill = Trial)) + geom_col() +
  scale_fill_manual(values = trial_colors) + theme_Publication() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 3)) +
  coord_flip()
  
graph2pdf(x = prolif_TIL_quadrant_trial_proportions_plot,
          file = here(
            output_dir, "_prolif_TIL_quadrant_trial_proportions_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 6)

trial_quadrant_proportions %>% select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

rna_features_df_plus_wes_pre %>%
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Trial)) +
  geom_point() + theme_Publication() + facet_wrap(~Response)


```

```{r}
wes_rna_features_df_pre_f %>% 
  ggplot(aes(x = TMB, y = TIL_score)) +
  geom_point(aes(col = Trial, shape = Response)) + theme_Publication() +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  scale_x_continuous(trans = "log2") +
  scale_color_manual(values = trial_colors) +
  guides(col = guide_legend(nrow = 3)) +
  geom_smooth(method = "lm")

rna_features_df_plus_wes_pre%>% 
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score)) +
  geom_point(aes(col = Trial, shape = Response)) + theme_Publication() +
  geom_vline(xintercept = median_Tumor_proliferation_rate_pre,
             linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  scale_x_continuous(trans = "log2") +
  scale_color_manual(values = trial_colors) +
  guides(col = guide_legend(nrow = 3)) +
  geom_smooth(method = "lm")
```

## Fig.5f
### Metastatic Quadrant plots TMB TIL score

```{r}
median_metastatic_til_score_pre <- median(
  metastatic_wes_rna_features_df_pre$TIL_score)

# Quadrant RR
table(metastatic_wes_rna_features_df_pre$TMB_NONSYNONYMOUS >= 10,
      metastatic_wes_rna_features_df_pre$TIL_score >= median_metastatic_til_score_pre) %>%
  `rownames<-`(c("TMB-L", "TMB-H")) %>%
  `colnames<-`( c("mTIL_low", "mTIL_high"))

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = log2(TMB), y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication()

# Quadrant RR
table(metastatic_wes_rna_features_df_pre$TMB >= 10,
      metastatic_wes_rna_features_df_pre$TIL_score >= median_metastatic_til_score_pre) %>%
  `rownames<-`(c("TMB-L", "TMB-H")) %>%
  `colnames<-`( c("mTIL_low", "mTIL_high"))

metastatic_quadrant_response_proportions <- table(
  metastatic_wes_rna_features_df_pre$TMB >= 10,
  metastatic_wes_rna_features_df_pre$TIL_score > median_metastatic_til_score_pre,
  metastatic_wes_rna_features_df_pre$Response) %>%
  as.data.frame() %>%
  setNames(c("TMB", "mTIL_group", "Response", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB == TRUE, "TMB_high", "TMB_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

metastatic_quadrant_response_proportions

metastatic_TMB_TIL_quadrant_response_proportions_plot <- metastatic_quadrant_response_proportions %>%
  filter(Response == "R") %>%
  rename(Response_rate = quadrant_prop) %>%
  ggplot(aes(x = quadrant, y = Response_rate)) + geom_col() +
  theme_Publication() + labs(x = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

metastatic_TMB_TIL_quadrant_response_proportions_plot

graph2pdf(x = metastatic_TMB_TIL_quadrant_response_proportions_plot,
          file = here(
            output_dir, "_metastatic_TMB_TIL_quadrant_response_proportions_plott.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 4)

metastatic_quadrant_response_proportions %>%
  select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()
```


## Fig 5h
### Metastatic Quadrant plots Proliferation TIL score

```{r}
median_metastatic_Tumor_proliferation_rate_pre <- metastatic_wes_rna_features_df_pre %>%
  pull(Tumor_proliferation_rate) %>%
  median()
median_metastatic_Tumor_proliferation_rate_pre

median_metastatic_til_score_pre <- median(
  metastatic_wes_rna_features_df_pre$TIL_score)
median_metastatic_til_score_pre
```


```{r}
trial_quadrant_proportions <- table(
  metastatic_wes_rna_features_df_pre$Tumor_proliferation_rate > median_metastatic_Tumor_proliferation_rate_pre, 
  metastatic_wes_rna_features_df_pre$TIL_score > median_metastatic_til_score_pre,
  metastatic_wes_rna_features_df_pre$Study) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
dim(metastatic_wes_rna_features_df_pre)

table(metastatic_wes_rna_features_df_pre$Response)

quadrant_response_proportions <- table(
  metastatic_wes_rna_features_df_pre$Tumor_proliferation_rate > median_metastatic_Tumor_proliferation_rate_pre,
  metastatic_wes_rna_features_df_pre$TIL_score > median_metastatic_til_score_pre,
  metastatic_wes_rna_features_df_pre$Response) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Response", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions

quadrant_response_proportions %>%
  select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()

metastatic_prolif_TIL_quadrant_response_proportions_plot <- quadrant_response_proportions %>%
  filter(Response == "R") %>%
  rename(Response_rate = quadrant_prop) %>%
  mutate(quadrant = factor(quadrant,
                           levels = c("Prolif_high_mTIL_high",
                                      "Prolif_low_mTIL_high",
                                      "Prolif_high_mTIL_low",
                                      "Prolif_low_mTIL_low"))) %>%
  ggplot(aes(x = quadrant, y = Response_rate)) + geom_col() +
  theme_Publication() + labs(x = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
metastatic_prolif_TIL_quadrant_response_proportions_plot

graph2pdf(x = metastatic_prolif_TIL_quadrant_response_proportions_plot,
          file = here(
            output_dir, "_metastatic_prolif_TIL_quadrant_response_proportions_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 4)
```




## Fig.5i
### Novel response signature plots
### Positive component
```{r}
pancan_response_ssgsea_results_df <- read_csv2(
  here(source_data_dir, "pancan_response_ssgsea_results_df.csv"))

dim(pancan_response_ssgsea_results_df)
table(pancan_response_ssgsea_results_df$Trial)

n_labels <- pancan_response_ssgsea_results_df %>%
  group_by(Response) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Response)

neoadjuvant_pancan_positive_response_plot <- pancan_response_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_sig_1.25, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) + guides(fill = FALSE) +
  labs(x = "", y = "Pancancer Positive Response\ncomponent ssGSEA score",
       title = "Neoadjuvant Pancancer") +
  scale_x_discrete(labels = label_map)+
  guides(fill = FALSE)

neoadjuvant_pancan_positive_response_plot 
graph2pdf(x = neoadjuvant_pancan_positive_response_plot,
          file = here(output_dir,
                      "_neoadjuvant_pancan_positive_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.25)
```

## Fig.5j
### Negative component

```{r}
neoadjuvant_pancan_negative_response_plot <- pancan_response_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_nonresponse_sig_1.25,
             fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Negative Response\ncomponent ssGSEA score",
       title = "Neoadjuvant Pancancer") +
  scale_x_discrete(labels = label_map)+
  guides(fill = FALSE)
neoadjuvant_pancan_negative_response_plot

graph2pdf(x = neoadjuvant_pancan_negative_response_plot,
          file = here(output_dir,
                      "_neoadjuvant_pancan_negative_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.25)
```

## Fig.5k
### Novel signature score

```{r}
neoadjuvant_pancan_response_score_plot <- pancan_response_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Response score\nssGSEA delta",
       title = "Neoadjuvant Pancancer") +
  scale_x_discrete(labels = label_map) +
  guides(fill = FALSE)

neoadjuvant_pancan_response_score_plot

graph2pdf(x = neoadjuvant_pancan_response_score_plot,
          file = here(output_dir,
                      "_neoadjuvant_pancan_response_score_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.25)
```


## Fig.5l
### Metastatic External iAtlas Response Sigs
### Positive component
```{r}
dim(metastatic_wes_rna_features_df_pre)
table(metastatic_wes_rna_features_df_pre$Trial)
table(metastatic_wes_rna_features_df_pre$Trial,
      metastatic_wes_rna_features_df_pre$CANCER_TYPE)

n_labels <- metastatic_wes_rna_features_df_pre %>%
  group_by(Response) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Response)

metastatic_pancan_positive_response_plot <- metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = pancancer_response_sig_1.25, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) + guides(fill = FALSE) +
  labs(x = "", y = "Pancancer Positive Response\ncomponent ssGSEA score",
       title = "iATLAS ICB Pancancer") +
  scale_x_discrete(labels = label_map) +
  guides(fill = FALSE)

metastatic_pancan_positive_response_plot 
graph2pdf(x = metastatic_pancan_positive_response_plot,
          file = here(output_dir,
                      "_metastatic_pancan_positive_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.25)
```

## Fig.5m
### Negative component

```{r}
metastatic_pancan_negative_response_plot <- metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = pancan_nonresponse_signature_1.25,
             fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Negative Response\ncomponent ssGSEA score",
       title = "iATLAS ICB Pancancer") +
  scale_x_discrete(labels = label_map)+
  guides(fill = FALSE)

metastatic_pancan_negative_response_plot

graph2pdf(x = metastatic_pancan_negative_response_plot,
          file = here(output_dir,
                      "_metastatic_pancan_negative_response_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.25)
```


## Fig.5n
### Novel signature score

```{r}
metastatic_pancan_response_score_plot <- metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Response score\nssGSEA delta",
       title = "iATLAS ICB Pancancer") +
  scale_x_discrete(labels = label_map) +
  guides(fill = FALSE)

metastatic_pancan_response_score_plot

graph2pdf(x = metastatic_pancan_response_score_plot,
          file = here(output_dir,
                      "_metastatic_pancan_response_score_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.25)
```

```{r}
neoadjuvant_ssgsea_c <- read_csv2(
  here(source_data_dir, "neoadjuvant_ssgsea_c.csv"))

dim(neoadjuvant_ssgsea_c)
table(neoadjuvant_ssgsea_c$Trial)
table(neoadjuvant_ssgsea_c$Trial,
      neoadjuvant_ssgsea_c$Response)

n_labels <- neoadjuvant_ssgsea_c %>%
  group_by(Response) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Response)

neoadjuvant_ssgsea_c
neoadjuvant_ssgsea_c_pancan_response_score_plot <- neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "NKI pancancer signature\nssGSEA score delta",
       title = "nICB+/-chemo\n(ABACUS, I-SPY2, PANDA)") +
  scale_x_discrete(labels = label_map) +
  guides(fill = FALSE)

neoadjuvant_ssgsea_c_pancan_response_score_plot

graph2pdf(x = neoadjuvant_ssgsea_c_pancan_response_score_plot,
          file = here(output_dir,
                      "_neoadjuvant_ssgsea_c_pancan_response_score_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.25)

```

## Fig.5k
### Novel signature score

```{r}
neoadjuvant_external_pancan_response_score_plot <- neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Response score\nssGSEA delta",
       title = "Neoadjuvant Pancancer") +
  scale_x_discrete(labels = label_map) +
  guides(fill = FALSE)

neoadjuvant_pancan_response_score_plot

graph2pdf(x = neoadjuvant_pancan_response_score_plot,
          file = here(output_dir,
                      "_neoadjuvant_pancan_response_score_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.5)
```


# ----------
# Extended Data Fig. 6


## Extended Data Fig. 6a
### PCA feature sets RNA
```{r}
pca_data <- read_csv2(here(source_data_dir, "pca_data.csv"))

pc1_var <- 44
pc2_var <- 18

dim(pca_data)

pca_rna_trial_plot <- pca_data %>%
  ggplot(aes(x = PC1, y = PC2, col = Trial, shape = Response)) +
  geom_point() +
  theme_Publication()  +
  guides(col = guide_legend(nrow = 3, byrow = TRUE),
         shape = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = sprintf("PC1: %s%% variance", pc1_var),
       y = sprintf("PC2: %s%% variance", pc2_var)) +
  scale_color_manual(values = trial_colors) +
  labs(col = "") +
  theme(legend.spacing.x = unit(0.3, "cm"))

pca_rna_trial_plot
graph2pdf(x = pca_rna_trial_plot,
          file = here(output_dir, "_pca_rna_trial_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 6,
          width = 6)
```

## Extended Data Fig. 6b

###  PCA feature sets RNA signatures 
```{r}
pca_rna_features <- c(unique(meta_rna_combined_effects$Signature))
dim(wes_rna_features_df_pre)
rna_features_m <- wes_rna_features_df_pre %>%
  select(Patient, pca_rna_features) %>%
  column_to_rownames("Patient") %>% 
  mutate_all(as.numeric) %>%
  as.matrix()

rna_features_m[is.na(rna_features_m)] <- 0 
rna_features_z_norm <- rna_features_m %>% scale()

dim(rna_features_z_norm)
pca <- prcomp(rna_features_z_norm)

percent_var <- pca$sdev^2 / sum(pca$sdev^2)

percent_var <- percent_var %>%
  as.data.frame() %>%
  setNames("Var_exp") %>%
  rownames_to_column("PC") %>%
  mutate(Cumvar = cumsum(Var_exp), PC = as.numeric(PC))
percent_var_pc1 <- round(percent_var[1, 2], 2) * 100
percent_var_pc2 <- round(percent_var[2, 2], 2) * 100

pca_data <- as.data.frame(pca$x) %>%
  rownames_to_column("Patient") %>%
  left_join(wes_rna_features_df_pre %>%
              select(Patient, Response, Trial, Treatment))
  
pca_rna_sig_trial_plot <- pca_data %>%
  ggplot(aes(x = PC1, y = PC2, col = Trial, shape = Response)) +
  geom_point() +
  theme_Publication()  +
  guides(col = guide_legend(nrow = 3, byrow = TRUE),
         shape = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = sprintf("PC1: %s%% variance", percent_var_pc1),
       y = sprintf("PC2: %s%% variance", percent_var_pc2)) +
  scale_color_manual(values = trial_colors) +
  labs(col = "") +
  theme(legend.spacing.x = unit(0.3, "cm"))

pca_rna_sig_trial_plot
graph2pdf(x = pca_rna_sig_trial_plot,
          file = here(output_dir, "_pca_rna_sig_trial_plot.pdf"),
          font = "Arial",
          height = 6,
          width = 6)

```


## Extended Data Fig. 6c

###  PCA feature sets WES RNA
```{r}
# pca_wes_rna_features <- c("TMB",
#                           colnames(pathway_level_mutation_m)[
#                             str_detect(colnames(pathway_level_mutation_m),
#                                        "SIGPW|GIE_IFNy|GIE_AP")],
#                           unique(meta_rna_combined_effects$Signature))

pca_wes_rna_features <- unique(
  c(smd_variable_order, log_or_variable_order, signature_variable_order))
length(pca_wes_rna_features)

wes_rna_features_m <- wes_rna_features_df_pre_f %>%
  # left_join(pathway_level_mutation_df %>%
  #             select(Patient,
  #                    colnames(pathway_level_mutation_m)[
  #                      str_detect(colnames(pathway_level_mutation_m),
  #                                 "SIGPW|GIE_IFNy|GIE_AP")])) %>%
  select(Patient, pca_wes_rna_features) %>%
  column_to_rownames("Patient") %>% 
  mutate_all(as.numeric) %>%
  as.matrix()

dim(wes_rna_features_m)

wes_rna_features_m[is.na(wes_rna_features_m)] <- 0 
wes_rna_features_z_norm <- wes_rna_features_m %>% scale()

dim(wes_rna_features_z_norm)
pca <- prcomp(wes_rna_features_z_norm)

percent_var <- pca$sdev^2 / sum(pca$sdev^2)

percent_var <- percent_var %>%
  as.data.frame() %>%
  setNames("Var_exp") %>%
  rownames_to_column("PC") %>%
  mutate(Cumvar = cumsum(Var_exp), PC = as.numeric(PC))
percent_var_pc1 <- round(percent_var[1, 2], 2) * 100
percent_var_pc2 <- round(percent_var[2, 2], 2) * 100

pca_data <- as.data.frame(pca$x) %>%
  rownames_to_column("Patient") %>%
  left_join(dna_metadata_f)
  
pca_wes_rna_trial_plot <- pca_data %>%
  ggplot(aes(x = PC1, y = PC2, col = Trial, shape = Response)) +
  geom_point() +
  theme_Publication()  +
  guides(col = guide_legend(nrow = 3, byrow = TRUE),
         shape = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = sprintf("PC1: %s%% variance", percent_var_pc1),
       y = sprintf("PC2: %s%% variance", percent_var_pc2)) +
  scale_color_manual(values = trial_colors) +
  labs(col = "") +
  theme(legend.spacing.x = unit(0.3, "cm"))

pca_wes_rna_trial_plot
graph2pdf(x = pca_wes_rna_trial_plot,
          file = here(output_dir, "_pca_wes_rna_trial_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 6,
          width = 6)

```


## Extended Data Fig. 6d
### External iATLAS metastatic metadata plots
```{r}

external_metadata_n_patients_trial_plot <- table(
  metastatic_wes_rna_features_df_pre$Study,
  metastatic_wes_rna_features_df_pre$CANCER_TYPE) %>%
  as.data.frame()  %>%
  setNames(c("Study", "CANCER_TYPE", "Freq"))  %>%
  ggplot(aes(x = reorder(CANCER_TYPE, Freq), y = Freq, fill = Study)) +
  geom_col() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  guides(fill = guide_legend(nrow = 3)) +
  coord_flip() +
  labs(x = "", y = "Number of patients",
       title = "iATLAS ICB Pancancer") +
  theme(text = element_text(family = "Arial")) +
  scale_fill_manual(values = itlas_trial_label_colors)

external_metadata_n_patients_trial_plot

graph2pdf(x = external_metadata_n_patients_trial_plot,
          file = here(output_dir,
                      "_external_metadata_n_patients_trial_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4.5, width = 6.5)

```


## Extended Fig. 6e (6h)
### Neoadjuvant TMB TIL continuous distribution

```{r}
median_til_score_pre <- median(
  rna_features_df_plus_wes_pre$TIL_score)
median_til_score_pre

wes_rna_features_df_pre_f %>% 
  ggplot(aes(x = TMB, y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  scale_x_continuous(trans = "log2") +
  scale_color_manual(values = trial_colors) +
  guides(col = guide_legend(nrow = 3))

TMB_TIL_quadrant_trial_plot <- wes_rna_features_df_pre_f %>% 
  ggplot(aes(x = TMB, y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() + facet_wrap(~Response) +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed")  +
  scale_x_continuous(trans = "log2") +
  scale_color_manual(values = trial_colors) +
  guides(col = guide_legend(nrow = 3))

TMB_TIL_quadrant_trial_plot

graph2pdf(x = TMB_TIL_quadrant_trial_plot,
          file = here(output_dir, "_TMB_TIL_quadrant_trial_continuous_distribution_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 7)

```

## Extended Data Fig. 6f (6x)
### Neoadjuvant Prolif TIL continuous distribution plot
```{r}
prolif_TIL_quadrant_trial_plot <- rna_features_df_plus_wes_pre %>%
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Trial)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = median_Tumor_proliferation_rate_pre,
             linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  scale_color_manual(values = trial_colors) +
  guides(col = guide_legend(nrow = 3)) + facet_wrap(~Response)

prolif_TIL_quadrant_trial_plot

graph2pdf(x = prolif_TIL_quadrant_trial_plot,
          file = here(
            output_dir,
            "_prolif_TIL_quadrant_trial_continuous_distribution_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 7)
```

## Extended Data Fig. 6g (6x)
### Neoadjuvant TMB TIL quadrant trial proportions

```{r}
trial_quadrant_proportions <- table(
  wes_rna_features_df_pre_f$TMB >= 10,
  wes_rna_features_df_pre_f$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre_f$Trial) %>%
  as.data.frame() %>%
  setNames(c("TMB_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB_group == TRUE, "TMB-H", "TMB-L")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
trial_quadrant_proportions %>% select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

quadrant_trial_proportions <- table(
  wes_rna_features_df_pre_f$TMB >= 10,
  wes_rna_features_df_pre_f$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre_f$Trial) %>%
  as.data.frame() %>%
  setNames(c("TMB_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB_group == TRUE, "TMB-H", "TMB-L")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group)) %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = Freq / Total)

TMB_TIL_quadrant_trial_proportions_plot <- quadrant_trial_proportions %>%
  ggplot(aes(x = quadrant, y = quadrant_prop, fill = Trial)) + geom_col() +
  scale_fill_manual(values = trial_colors) + theme_Publication() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 3)) +
  coord_flip()

TMB_TIL_quadrant_trial_proportions_plot

graph2pdf(x = TMB_TIL_quadrant_trial_proportions_plot,
          file = here(
            output_dir, "_TMB_TIL_quadrant_trial_proportions_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 6)
```

## Extended Data Fig. 6h (6x)
### Neoadjuvant Prolif TIL quadrant trial proportions

```{r}
median_til_score_pre <- median(
  rna_features_df_plus_wes_pre$TIL_score)
median_til_score_pre

median_Tumor_proliferation_rate_pre <- rna_features_df_plus_wes_pre %>%
  filter(Treatment == "Pre") %>%
  pull(Tumor_proliferation_rate) %>%
  median()
median_Tumor_proliferation_rate_pre

trial_quadrant_proportions <- table(
  rna_features_df_plus_wes_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  rna_features_df_plus_wes_pre$TIL_score > median_til_score_pre,
  rna_features_df_plus_wes_pre$Trial) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>%
  add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
dim(rna_features_df_plus_wes_pre)
table(rna_features_df_plus_wes_pre$Response)
```


## Add Extended Fig 6x (6g) metastatic TMB TIL continuous distributions

```{r}
metastatic_wes_rna_features_df_pre$Response %>% table()
metastatic_wes_rna_features_df_pre_f <- metastatic_wes_rna_features_df_pre %>% filter(!is.na(TMB))
metastatic_wes_rna_features_df_pre_f$Response %>% table()

metastatic_wes_rna_features_df_pre %>% 
  ggplot(aes(x = TMB, y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_metastatic_til_score_pre, linetype = "dashed") +
  scale_x_continuous(trans = "log2") +
  guides(col = guide_legend(nrow = 3)) +
  scale_color_manual(values = itlas_trial_label_colors)

metastatic_TMB_TIL_quadrant_trial_plot <- metastatic_wes_rna_features_df_pre %>% 
  ggplot(aes(x = TMB, y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() + facet_wrap(~Response) +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_metastatic_til_score_pre, linetype = "dashed")  +
  scale_x_continuous(trans = "log2") +
  guides(col = guide_legend(nrow = 3)) +
  scale_color_manual(values = itlas_trial_label_colors)

metastatic_TMB_TIL_quadrant_trial_plot

graph2pdf(x = metastatic_TMB_TIL_quadrant_trial_plot,
          file = here(output_dir, "_metastatic_TMB_TIL_quadrant_trial_distribution_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 7)
```

## Add Extended Data Fig. 6x (6h) metastatic prolif TIL continuous distribution plot
```{r}
metastatic_wes_rna_features_df_pre$Response %>% table()

metastatic_prolif_TIL_quadrant_trial_plot <- metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Study)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = median_metastatic_Tumor_proliferation_rate_pre,
             linetype = "dashed") +
  geom_hline(yintercept = median_metastatic_til_score_pre, linetype = "dashed") +
  guides(col = guide_legend(nrow = 3)) + facet_wrap(~Response) +
    scale_color_manual(values = itlas_trial_label_colors)

metastatic_prolif_TIL_quadrant_trial_plot

graph2pdf(x = metastatic_prolif_TIL_quadrant_trial_plot,
          file = here(output_dir, "_metastatic_prolif_TIL_quadrant_trial_distribution_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 7)
```


## Add Extended Data Fig. 6x metastatic trial proportions
```{r}
trial_quadrant_proportions <- table(
  metastatic_wes_rna_features_df_pre$Tumor_proliferation_rate > median_metastatic_Tumor_proliferation_rate_pre,
  metastatic_wes_rna_features_df_pre$TIL_score > median_metastatic_til_score_pre,
  metastatic_wes_rna_features_df_pre$Study) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
trial_quadrant_proportions

trial_quadrant_proportions %>%
  select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

metastatic_prolif_TIL_quadrant_trial_proportions_plot <- trial_quadrant_proportions %>%
  select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

metastatic_prolif_TIL_quadrant_trial_proportions_plot

graph2pdf(x = metastatic_prolif_TIL_quadrant_trial_proportions_plot,
          file = here(
            output_dir,
            "_metastatic_prolif_TIL_quadrant_trial_proportions_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 6)

trial_quadrant_proportions <- table(
  metastatic_wes_rna_features_df_pre$Tumor_proliferation_rate > median_metastatic_Tumor_proliferation_rate_pre,
  metastatic_wes_rna_features_df_pre$TIL_score > median_metastatic_til_score_pre,
  metastatic_wes_rna_features_df_pre$Study) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
quadrant_trial_proportions <- table(
  metastatic_wes_rna_features_df_pre$Tumor_proliferation_rate > median_metastatic_Tumor_proliferation_rate_pre,
  metastatic_wes_rna_features_df_pre$TIL_score > median_metastatic_til_score_pre,
  metastatic_wes_rna_features_df_pre$Study) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group)) %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = Freq / Total)


metastatic_prolif_TIL_quadrant_trial_proportions_plot <- quadrant_trial_proportions %>%
  ggplot(aes(x = quadrant, y = quadrant_prop, fill = Trial)) + geom_col() +
  theme_Publication() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 3)) +
  coord_flip() +
  scale_fill_manual(values = itlas_trial_label_colors)

metastatic_prolif_TIL_quadrant_trial_proportions_plot
graph2pdf(x = metastatic_prolif_TIL_quadrant_trial_proportions_plot,
          file = here(
            output_dir,
            "_metastatic_prolif_TIL_quadrant_trial_proportions_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 6)
```


## Add Extended Fig 6x metastatic TMB TIL trial proportions
```{r}
trial_quadrant_proportions <- table(
  metastatic_wes_rna_features_df_pre$TMB >= 10,
  metastatic_wes_rna_features_df_pre$TIL_score > median_metastatic_til_score_pre,
  metastatic_wes_rna_features_df_pre$Trial) %>%
  as.data.frame() %>%
  setNames(c("TMB_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB_group == TRUE, "TMB-H", "TMB-L")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
trial_quadrant_proportions %>%
  select(quadrant, quadrant_prop, Trial) %>%
  filter(!Trial %in% c("gbm_iatlas_prins_2019", "mel_iatlas_gide_2019")) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

metastatic_quadrant_trial_proportions <- table(
  metastatic_wes_rna_features_df_pre$TMB >= 10,
  metastatic_wes_rna_features_df_pre$TIL_score > median_metastatic_til_score_pre,
  metastatic_wes_rna_features_df_pre$Trial) %>%
  as.data.frame() %>%
  setNames(c("TMB_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB_group == TRUE, "TMB-H", "TMB-L")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group)) %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = Freq / Total)

metastatic_TMB_TIL_quadrant_trial_proportions_plot <- metastatic_quadrant_trial_proportions %>%
  filter(!Trial %in% c("gbm_iatlas_prins_2019", "mel_iatlas_gide_2019")) %>%
  ggplot(aes(x = quadrant, y = quadrant_prop, fill = Trial)) + geom_col() +
  theme_Publication() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 3)) +
  coord_flip() +
  scale_fill_manual(values = itlas_trial_label_colors)

metastatic_TMB_TIL_quadrant_trial_proportions_plot

metastatic_TMB_TIL_quadrant_trial_proportions_plot
graph2pdf(x = metastatic_TMB_TIL_quadrant_trial_proportions_plot,
          file = here(
            output_dir,
            "_metastatic_TMB_TIL_quadrant_trial_proportions_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 6)
```

# ----------
# Extended Data Fig. 7
## Extended Data Fig. 7a
### Positive Response Degs
```{r, fig.height = 8, fig.width = 8}

deseq_r1_c <- read_csv2(here(source_data_dir, "deseq_r1_c.csv"))
deseq_r1_meta_df <- read_csv2(here(source_data_dir, "deseq_r1_meta_df.csv"))

pancan_response_signature_1.25 <- read_csv2(
  here(source_data_dir, "pancan_response_signature_1.25.csv"))

pancan_nonresponse_signature_1.25 <- read_csv2(
  here(source_data_dir, "pancan_nonresponse_signature_1.25.csv"))

positive_response_component_deg_plot <- deseq_r1_c %>%
  filter(gene %in% pancan_response_signature_1.25$gene) %>%
  left_join(deseq_r1_meta_df %>%
              select(gene, n_responder_enrichment,
                     n_nresponder_enrichment, mean_log2fc)) %>%
  ggplot(aes(x = reorder(gene, mean_log2fc),
             y = log2FoldChange, col = Trial)) +
  geom_point() +
  geom_point(aes(x = gene, y = mean_log2fc, col = "black"),
             shape = 5,
             data = deseq_r1_meta_df %>%
               filter(gene %in% pancan_response_signature_1.25$gene)) +
  geom_hline(yintercept = 0) +
  theme_Publication() + coord_flip() +
  scale_color_manual(values = trial_colors_combined) +
  geom_hline(yintercept = log2(1.25), linetype = "dashed") +
  labs(x = "")

positive_response_component_deg_plot

graph2pdf(x = positive_response_component_deg_plot,
          file = here(
            output_dir,
            "_positive_response_component_deg_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 8,
          width = 6)
```

## Extended Data Fig. 7b
### Negative Response Degs

```{r, fig.height = 8, fig.width = 8}

negative_response_component_deg_plot <- deseq_r1_c %>%
  filter(gene %in% pancan_nonresponse_signature_1.25$gene) %>%
  left_join(deseq_r1_meta_df %>%
              select(gene, n_responder_enrichment,
                     n_nresponder_enrichment, mean_log2fc)) %>%
  ggplot(aes(x = reorder(gene, mean_log2fc),
             y = log2FoldChange, col = Trial)) +
  geom_point() +
  geom_point(aes(x = gene, y = mean_log2fc, col = "black"),
             shape = 5,
             data = deseq_r1_meta_df %>%
               filter(gene %in% pancan_nonresponse_signature_1.25$gene)) +
  geom_hline(yintercept = 0) +
  theme_Publication() + coord_flip() +
  scale_color_manual(values = trial_colors_combined) +
  geom_hline(yintercept = -log2(1.25), linetype = "dashed") +
  labs(x = "")

negative_response_component_deg_plot

graph2pdf(x = negative_response_component_deg_plot,
          file = here(
            output_dir,
            "_negative_response_component_deg_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 8,
          width = 6)
```

## Extended Data Fig. 7e
### Per trial
```{r, fig.height = 12, fig.width = 12}

n_labels <- pancan_response_ssgsea_results_df %>%
  group_by(Trial, Response) %>%
  summarise(n = n()) %>%
mutate(
  Trial_response =  paste0(Trial, " ", Response),
  label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Trial_response)
label_map

pancan_response_ssgsea_results_df %>%
  mutate(
    Trial_response =  paste0(Trial, " ", Response)) %>%
  ggplot(aes(x = Trial_response,
             y = pancancer_response_sig_1.25, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  facet_wrap(~Trial, scales = "free_x") + 
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R))+
  labs(x = "", y = "Pancancer Positive Response\ncomponent ssGSEA score",
       title = "Neoadjuvant Pancancer") +
  guides(fill = FALSE)

neoadjuvant_pancan_positive_response_per_trial_plot <- pancan_response_ssgsea_results_df %>%
  mutate(
    Trial_response =  paste0(Trial, " ", Response)) %>%
  ggplot(aes(x = Trial_response,
             y = pancancer_response_sig_1.25, fill = Response)) +
  geom_boxplot() +
  facet_wrap(~Trial, scales = "free_x") + 
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Positive Response\ncomponent ssGSEA score",
       title = "Neoadjuvant Pancancer") +
  scale_x_discrete(labels = label_map)  +
  guides(fill = FALSE)

neoadjuvant_pancan_positive_response_per_trial_plot 
graph2pdf(x = neoadjuvant_pancan_positive_response_per_trial_plot,
          file = here(output_dir, "_neoadjuvant_pancan_positive_response_per_trial_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 7.5,
          width = 7.5)

pancan_response_ssgsea_results_df %>%
  mutate(
    Trial_response =  paste0(Trial, " ", Response)) %>%
  ggplot(aes(x = Response, y = pancancer_nonresponse_sig_1.25,
             fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") + facet_wrap(~Trial) + theme_Publication()+
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Negative Response\ncomponent ssGSEA score",
       title = "Neoadjuvant Pancancer") +
  guides(fill = FALSE)

neoadjuvant_pancan_negative_response_per_trial_plot <- pancan_response_ssgsea_results_df %>%
  mutate(
    Trial_response =  paste0(Trial, " ", Response)) %>%
  ggplot(aes(x = Trial_response, y = pancancer_nonresponse_sig_1.25,
             fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") + facet_wrap(~Trial, scales = "free_x") +
  theme_Publication()+
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Negative Response\ncomponent ssGSEA score",
       title = "Neoadjuvant Pancancer") + 
  scale_x_discrete(labels = label_map)  +
  guides(fill = FALSE)

neoadjuvant_pancan_negative_response_per_trial_plot 
graph2pdf(x = neoadjuvant_pancan_negative_response_per_trial_plot,
          file = here(output_dir, "_neoadjuvant_pancan_negative_response_per_trial_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 7.5,
          width = 7.5)

pancan_response_ssgsea_results_df %>%
  mutate(
    Trial_response =  paste0(Trial, " ", Response)) %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + facet_wrap(~Trial, scales = "free_x") +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Response score\nssGSEA delta",
       title = "Neoadjuvant Pancancer") +
  guides(fill = FALSE)

neoadjuvant_pancan_response_score_per_trial_plot <- pancan_response_ssgsea_results_df %>%
  mutate(
    Trial_response =  paste0(Trial, " ", Response)) %>%
  ggplot(aes(x = Trial_response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + facet_wrap(~Trial, scales = "free_x") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Response score\nssGSEA delta",
       title = "Neoadjuvant Pancancer") + 
  scale_x_discrete(labels = label_map)  +
  guides(fill = FALSE)

neoadjuvant_pancan_response_score_per_trial_plot 

graph2pdf(x = neoadjuvant_pancan_response_score_per_trial_plot,
          file = here(output_dir, "_neoadjuvant_pancan_response_score_per_trial_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 7.5,
          width = 7.5)
```

## Extended Data Fig. 7f metastatic response sig  Per trial
```{r, fig.height = 12, fig.width = 12}

n_labels <- metastatic_wes_rna_features_df_pre %>%
  group_by(Trial, Response) %>%
  summarise(n = n()) %>%
mutate(
  Trial_response =  paste0(Trial, " ", Response),
  label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Trial_response)
label_map


metastatic_wes_rna_features_df_pre %>%
  mutate(
    Trial_response =  paste0(Trial, " ", Response)) %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + facet_wrap(~Trial, scales = "free_x") +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Response score\nssGSEA delta",
       title = "iATLAS metastatic Pancancer") +
  guides(fill = FALSE)

metastatic_pancan_response_score_per_trial_plot <- metastatic_wes_rna_features_df_pre %>%
  mutate(
    Trial_response =  paste0(Trial, " ", Response)) %>%
  ggplot(aes(x = Trial_response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + facet_wrap(~Trial, scales = "free_x") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Pancancer Response score\nssGSEA delta",
       title = "iATLAS metastatic Pancancer") + 
  scale_x_discrete(labels = label_map)  +
  guides(fill = FALSE)

metastatic_pancan_response_score_per_trial_plot 

graph2pdf(x = metastatic_pancan_response_score_per_trial_plot,
          file = here(output_dir, "_metastatic_pancan_response_score_per_trial_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 7.5,
          width = 7.5)
```

### Extended Data Fig. 7 correlations
### corr Neoadjuvant
```{r, fig.height = 10, fig.width = 8}
neoadjuvant_ssgsea_correlations <- wes_rna_features_df_pre %>%
  left_join(pancan_response_ssgsea_results_df %>%
              select(c("CF_sample", "pancancer_nonresponse_sig_1.25",
                     "pancancer_response_sig_1.25",
                     "pancancer_response_score"))) %>%
  select(c(signature_variable_order,
           "pancancer_nonresponse_sig_1.25",
           "pancancer_response_sig_1.25",
           "pancancer_response_score")
         ) %>%
  cor(method = "spearman")

neoadjuvant_ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancancer_response_sig_1.25) %>%
  setNames(c("pancan_response_sig_corr")) %>%
  rownames_to_column("Signature") %>%
  filter(!Signature %in% c("pancancer_response_sig_1.25",
                           "pancancer_nonresponse_sig_1.25",
                           "pancancer_response_score")) %>%
  filter(!is.na(pancan_response_sig_corr)) %>%
  left_join(signature_overview %>% select(Signature, Signature_abbreviation)) %>%
  ggplot(aes(x = reorder(Signature_abbreviation, pancan_response_sig_corr),
                         y = pancan_response_sig_corr)) +
  geom_col() + theme_Publication() + coord_flip() +
  labs(x = "",
       y = "Correlations Positive Response component",
       title = "Neoadjuvant Cohort")

neoadjuvant_ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancancer_nonresponse_sig_1.25) %>%
  setNames(c("pancancer_nonresponse_sig_1.25")) %>%
  rownames_to_column("Signature") %>%
  filter(!Signature %in% c("pancancer_nonresponse_sig_1.25",
                           "pancancer_response_sig_1.25",
                           "pancancer_response_score")) %>%
  filter(!is.na(pancancer_nonresponse_sig_1.25)) %>%
  left_join(signature_overview %>% select(Signature, Signature_abbreviation)) %>%
  ggplot(aes(x = reorder(Signature_abbreviation, pancancer_nonresponse_sig_1.25),
                         y = pancancer_nonresponse_sig_1.25)) +
  geom_col() + theme_Publication() + coord_flip() +
  labs(x = "",
       y = "Correlations Negative Response component",
       title = "Neoadjuvant Cohort")
```

### Hypoxia and TIL score

```{r}

TIL_score_positive_response_component_plot <- rna_features_df_plus_wes_pre %>%
  left_join(pancan_response_ssgsea_results_df) %>%
  ggplot(aes(x = TIL_score, y = pancancer_response_sig_1.25)) +
  geom_point(aes(col = Trial)) + stat_cor(method = "spearman") +
  theme_Publication() +
  scale_color_manual(values = trial_colors) +
  labs(x = "TIL score ssGSEA",
       y = "Pancancer Positive Response\ncomponent ssGSEA score")
       
TIL_score_positive_response_component_plot
graph2pdf(x = TIL_score_positive_response_component_plot,
          file = here(output_dir, 
                    "_TIL_score_positive_response_component_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 5.5)

hypoxia_negative_response_component_plot <- rna_features_df_plus_wes_pre %>%
  left_join(pancan_response_ssgsea_results_df) %>%
  ggplot(aes(x = H_Hypoxia, y = pancancer_nonresponse_sig_1.25)) +
  geom_point(aes(col = Trial)) + stat_cor(method = "spearman") +
  theme_Publication() +
  scale_color_manual(values = trial_colors) +
  labs(x = "H_Hypoxia ssGSEA",
       y = "Pancancer Negative Response\ncomponent ssGSEA score")

hypoxia_negative_response_component_plot

graph2pdf(x = hypoxia_negative_response_component_plot,
          file = here(output_dir,
                      "_hypoxia_negative_response_component_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 5,
          width = 5.5)

```

## Extended Data Fig. 7c

```{r, fig.height = 10, fig.width = 8}
metastatic_ssgsea_correlations <- metastatic_wes_rna_features_df_pre %>%
  select(c(signature_variable_order,
           "pancan_nonresponse_signature_1.25",
           "pancancer_response_sig_1.25",
           "pancancer_response_score")
         ) %>%
  cor(method = "spearman")
```

### Positive correlation components metastatic

```{r, fig.height = 10, fig.width = 8}
metastatic_positive_component_correlation_plot <- metastatic_ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancancer_response_sig_1.25) %>%
  setNames(c("pancan_response_sig_corr")) %>%
  rownames_to_column("Signature") %>%
  filter(!Signature %in% c("pancancer_response_sig_1.25",
                           "pancan_nonresponse_signature_1.25",
                           "pancancer_response_score")) %>%
  filter(!is.na(pancan_response_sig_corr)) %>%
  left_join(signature_overview %>% select(Signature, Signature_abbreviation)) %>%
  ggplot(aes(x = reorder(Signature_abbreviation, pancan_response_sig_corr),
                         y = pancan_response_sig_corr)) +
  geom_col() + theme_Publication() + coord_flip() +
  labs(x = "",
       y = "Correlations Positive Response component",
       title = "iATLAS metastatic Pancancer")

metastatic_positive_component_correlation_plot
graph2pdf(x = metastatic_positive_component_correlation_plot,
          file = here(output_dir, "_metastatic_positive_component_correlation_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 8,
          width = 5)
```

## Extended Data Fig. 7d
### Negative correlation components metastatic

```{r, fig.height = 10, fig.width = 8}
metastatic_negative_component_correlation_plot <- metastatic_ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancan_nonresponse_signature_1.25) %>%
  setNames(c("pancan_nonresponse_signature_1.25")) %>%
  rownames_to_column("Signature") %>%
  filter(!Signature %in% c("pancan_nonresponse_signature_1.25",
                           "pancancer_response_sig_1.25",
                           "pancancer_response_score")) %>%
  filter(!is.na(pancan_nonresponse_signature_1.25)) %>%
  left_join(signature_overview %>% select(Signature, Signature_abbreviation)) %>%
  ggplot(aes(x = reorder(Signature_abbreviation, pancan_nonresponse_signature_1.25),
                         y = pancan_nonresponse_signature_1.25)) +
  geom_col() + theme_Publication() + coord_flip() +
  labs(x = "",
       y = "Correlations Negative Response component",
       title = "iATLAS metastatic Pancancer")

metastatic_negative_component_correlation_plot
graph2pdf(x = metastatic_negative_component_correlation_plot,
          file = here(output_dir, "_metastatic_negative_component_correlation_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 8,
          width = 5)

metastatic_ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancancer_response_score) %>%
  setNames(c("pancancer_response_score")) %>%
  rownames_to_column("Signature") %>%
  filter(!is.na(pancancer_response_score))%>%
  filter(!Signature %in% c("pancan_nonresponse_signature_1.25",
                           "pancancer_response_sig_1.25",
                           "pancancer_response_score")) %>%
  ggplot(aes(x = reorder(Signature, pancancer_response_score),
                         y = pancancer_response_score)) +
  geom_col() + theme_Publication() + coord_flip() +
  labs(x = "",
       y = "Signature Correlations Pancancer Response score",
       title = "iATLAS metastatic Pancancer")
```


## Fig.7g
###  Novel signature score ABACUS

```{r}
table(neoadjuvant_ssgsea_c$Trial)

n_labels <- neoadjuvant_ssgsea_c  %>%
  filter(Trial == "ABACUS") %>%
  group_by(Response) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"))
label_map <- setNames(n_labels$label, n_labels$Response)

ABACUS_pancan_response_score_plot <- neoadjuvant_ssgsea_c %>%
  filter(Trial == "ABACUS") %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "NKI pancancer signature\nssGSEA score delta",
       title = "ABACUS (nICB bladder)") +
  scale_x_discrete(labels = label_map) +
  guides(fill = FALSE)

ABACUS_pancan_response_score_plot

graph2pdf(x = ABACUS_pancan_response_score_plot,
          file = here(output_dir,
                      "_ABACUS_pancan_response_score_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.5)
```


## Fig.7h
###  Novel signature score I-SPY2

```{r}

n_labels <- neoadjuvant_ssgsea_c  %>%
  filter(Study == "ISPY2") %>%
  group_by(Response) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"))

label_map <- setNames(n_labels$label, n_labels$Response)

ISPY2_pancan_response_score_plot <- neoadjuvant_ssgsea_c %>%
  filter(Study == "ISPY2") %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "NKI pancancer signature\nssGSEA score delta",
       title = "I-SPY2 (nICB+chemo breast)") +
  scale_x_discrete(labels = label_map) +
  guides(fill = FALSE)

ISPY2_pancan_response_score_plot

graph2pdf(x = ISPY2_pancan_response_score_plot,
          file = here(output_dir,
                      "_ISPY2_pancan_response_score_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.5)
```



## Fig.7i
###  Novel signature score PANDA

```{r}

n_labels <- neoadjuvant_ssgsea_c  %>%
  filter(Study == "PANDA") %>%
  group_by(Response) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"))

label_map <- setNames(n_labels$label, n_labels$Response)

PANDA_pancan_response_score_plot <- neoadjuvant_ssgsea_c %>%
  filter(Study == "PANDA") %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "NKI pancancer signature\nssGSEA score delta",
       title = "PANDA (nICB+chemo G/GEJ)") +
  scale_x_discrete(labels = label_map) +
  guides(fill = FALSE)

PANDA_pancan_response_score_plot

graph2pdf(x = PANDA_pancan_response_score_plot,
          file = here(output_dir,
                      "_PANDA_pancan_response_score_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.5)


n_labels <- neoadjuvant_ssgsea_c  %>%
  filter(Study == "PANDA") %>%
  group_by(Response) %>%
  summarise(n = n()) %>%
mutate(
  label = paste0(Response, "\n(n = ", n, ")"))

label_map <- setNames(n_labels$label, n_labels$Response)

adapted_PANDA_pancan_response_score_plot <- neoadjuvant_ssgsea_c %>%
  filter(Study == "PANDA") %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "NKI pancancer signature\nssGSEA score delta",
       title = "PANDA (nICB+chemo G/GEJ)") +
  scale_x_discrete(labels = label_map) +
  guides(fill = FALSE)

adapted_PANDA_pancan_response_score_plot

graph2pdf(x = adapted_PANDA_pancan_response_score_plot,
          file = here(output_dir,
                      "_adapted_PANDA_pancan_response_score_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 4.5)
```

# ----------

# Supporting figures revision
## Signature AUC
```{r, fig.height = 8, fig.width = 7}

signature_prediction_data <- read_csv2(
  here(source_data_dir, "signature_prediction_data_with_msi.csv"))

signature_prediction_data %>%
  mutate(AUC = as.numeric(AUC)) %>%
  ggplot(aes(x = reorder(Signature, AUC), y = AUC)) +
  geom_col() +
  expand_limits(y = 0) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  geom_hline(yintercept = 0.5, col = "red", linetype = "dashed") +
  labs(x = "") +
  coord_flip() +
  scale_y_continuous(breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7))


signature_prediction_data  %>%
  filter(!Signature %in% c("pancancer_response_sig_1.25",
                           "pancancer_nonresponse_sig_1.25")) %>%
  mutate(Signature = ifelse(Signature == "pancancer_response_score",
                            "NKI pan-cancer signature", Signature)) %>%
  mutate(AUC = as.numeric(AUC))  %>%
  ggplot(aes(x = reorder(Signature, AUC), y = AUC)) +
  geom_col() +
  expand_limits(y = 0) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  geom_hline(yintercept = 0.5, col = "orange", linetype = "longdash", width = 2) +
  labs(x = "") +
  coord_flip() +
  scale_y_continuous(breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9))

signature_mixed_effects_covariate_prediction_data  %>%
  filter(!Signature %in% c("pancancer_response_sig_1.25",
                           "pancancer_nonresponse_sig_1.25")) %>%
  mutate(Signature = ifelse(Signature == "pancancer_response_score",
                            "NKI pan-cancer signature", Signature)) %>%
  mutate(AUC = as.numeric(AUC))  %>%
  ggplot(aes(x = reorder(Signature, AUC), y = AUC)) +
  geom_col() +
  expand_limits(y = 0) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  geom_hline(yintercept = 0.5, col = "orange", linetype = "longdash", width = 2) +
  labs(x = "") +
  coord_flip() +
  scale_y_continuous(breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9))

```

## HLA alteration vs TMB and response
```{r, fig.width = 5, fig.height = 4}
wes_features_df %>%
  mutate(HLA_Intact = ifelse(
    HLA_Intact == "HLA-Altered", "HLA del/LOH", HLA_Intact)) %>%
  ggplot(aes(x = HLA_Intact, y = TMB, fill = HLA_Intact)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  geom_hline(yintercept = 10, linetype = "dashed") +
  guides(fill = FALSE) +
  scale_fill_manual(values = c("darkorange", "lightblue")) +
  labs(x = "", y = "TMB (Mutations/Mb)")

wes_features_df %>%
  filter(!Trial %in% "NICHE-MSI") %>%
  mutate(HLA_Intact = ifelse(
    HLA_Intact == "HLA-Altered", "HLA del/LOH", HLA_Intact)) %>%
  ggplot(aes(x = HLA_Intact, y = TMB, fill = HLA_Intact)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  geom_hline(yintercept = 10, linetype = "dashed") +
  guides(fill = FALSE) +
  scale_fill_manual(values = c("darkorange", "lightblue")) +
  labs(x = "", y = "TMB (Mutations/Mb)")
```

```{r}
wes_features_df  %>%
  mutate(HLA_Intact = ifelse(
    HLA_Intact == "HLA-Altered", "HLA del/LOH", HLA_Intact))%>%
  ggplot(aes(x = HLA_Intact, y = TMB)) + facet_wrap(~Response) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  geom_hline(yintercept = 10, linetype = "dashed")

wes_features_df %>%
  mutate(HLA_Intact = ifelse(
    HLA_Intact == "HLA-Altered", "HLA del/LOH", HLA_Intact)) %>%
  group_by(HLA_Intact, Response) %>%
  add_count() %>%
  mutate(Label = sprintf("%s\n(n = %s)", Response, n)) %>%
  ggplot(aes(x = Label, y = TMB, fill = Response)) +
  facet_wrap(~HLA_Intact, scales = "free_x") +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  geom_hline(yintercept = 10, linetype = "dashed") +
  labs(x = "", y = "TMB (Mutations/Mb)") + guides(fill = FALSE)

wes_features_df %>%
  mutate(HLA_Intact = ifelse(
    HLA_Intact == "HLA-Altered", "HLA del/LOH", HLA_Intact)) %>%
  group_by(HLA_Intact, Response) %>%
  add_count() %>%
  mutate(Label = sprintf("%s\n(n = %s)", Response, n)) %>%
  ggplot(aes(x = Response, y = TMB, fill = Response)) +
  facet_wrap(~HLA_Intact, scales = "free_x") +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  geom_hline(yintercept = 10, linetype = "dashed") +
  labs(x = "", y = "TMB (Mutations/Mb)") + guides(fill = FALSE)

```


## Median TMB vs. Alteration frequency
```{r, fig.width = 5, fig.height = 5}
immune_hits_ap_freq <- table(wes_features_df$Trial,
                             wes_features_df$immune_hits_ap > 0) %>%
  as.data.frame() %>%
  setNames(c("Trial", "immune_hits_ap", "Freq")) %>%
  mutate(immune_hits_ap = ifelse(
    as.logical(immune_hits_ap), "AP_mut", "AP_wt")) %>%
  pivot_wider(names_from = immune_hits_ap, values_from = Freq) %>%
  mutate(Total = AP_wt + AP_mut,
         AP_mut_perc = AP_mut / Total * 100) %>%
  left_join(tmb_response_rate)

immune_hits_ap_freq

immune_hits_ap_freq %>%
  ggplot(aes(x = median_TMB, y = AP_mut_perc, col = Histology_MSI_MSS)) +
  geom_point(size = 3) +
  labs(x = "median TMB (mutations/Mb)",
       y = "Antigen presentation mut (%)",  col = "") +
  scale_x_continuous(trans = "log", breaks = c(1, 10, 20, 30, 40, 50)) +
  theme_Publication() +
  scale_color_manual(values = histology_msi_mss_colors) +
  expand_limits(y = 0)+
  geom_vline(xintercept = 10, linetype = "dashed") +
  guides(size = "none")

immune_hits_ifny_freq <- table(
  wes_features_df$Trial,
  wes_features_df$immune_hits_ifny > 0) %>%
  as.data.frame() %>%
  setNames(c("Trial", "immune_hits_ifny", "Freq")) %>%
  mutate(immune_hits_ifny = ifelse(
    as.logical(immune_hits_ifny), "pIFNy_mut", "pIFNy_wt")) %>%
  pivot_wider(names_from = immune_hits_ifny, values_from = Freq) %>%
  mutate(Total = pIFNy_mut + pIFNy_wt,
         pIFNy_mut_perc = pIFNy_mut / Total * 100) %>%
  left_join(tmb_response_rate)

immune_hits_ifny_freq

immune_hits_ifny_freq %>%
  ggplot(aes(x = median_TMB, y = pIFNy_mut_perc, col = Histology_MSI_MSS)) +
  geom_point(size = 3) +
  labs(x = "median TMB (mutations/Mb)",
       y = "IFNy pathway mut (%)",  col = "") +
  theme_Publication() +
  scale_x_continuous(trans = "log", breaks = c(1, 10, 20, 30, 40, 50)) +
  scale_color_manual(values = histology_msi_mss_colors) +
  expand_limits(y = 0) +
  geom_vline(xintercept = 10, linetype = "dashed") +
  guides(size = "none")


immune_hits_hla_alteration_freq <- table(
  wes_features_df$Trial,
  wes_features_df$HLA_alteration > 0) %>%
  as.data.frame() %>%
  setNames(c("Trial", "HLA_alteration", "Freq")) %>%
  mutate(HLA_alteration = ifelse(as.logical(HLA_alteration),
                                 "HLA_altered", "HLA_intact")) %>%
  pivot_wider(names_from = HLA_alteration, values_from = Freq) %>%
  mutate(Total = HLA_altered + HLA_intact,
         HLA_altered_perc = HLA_altered / Total * 100) %>%
  left_join(tmb_response_rate)

immune_hits_hla_alteration_freq

immune_hits_hla_alteration_freq %>%
  ggplot(aes(x = median_TMB, y = HLA_altered_perc, col = Histology_MSI_MSS)) +
  geom_point(size = 3) +
  labs(x = "median TMB (mutations/Mb)", y = "HLA del/LOH (%)",  col = "") +
  theme_Publication(base_size = 14) +
  scale_x_continuous(trans = "log", breaks = c(1, 10, 20, 30, 40, 50)) +
  scale_color_manual(values = histology_msi_mss_colors) +
  expand_limits(y = 0) +
  geom_vline(xintercept = 10, linetype = "dashed") +
  guides(size = "none")
```

## Neoantigen vs TMB TMB groups
```{r}
wes_features_df %>%
  ggplot(aes(x = Response, y = pvactools_filtered_neoantigen_load)) +
  facet_wrap(~ifelse(TMB >= 10, "TMB-high", "TMB-low"), scales = "free") +
  geom_boxplot() +
  theme_Publication() + stat_compare_means()

wes_features_df %>%
  ggplot(aes(x = Response, y = TMB)) + facet_wrap(~ifelse(TMB >= 10, "TMB-high", "TMB-low"), scales = "free") +
  geom_boxplot() + theme_Publication() + stat_compare_means()

wes_features_df %>% ggplot(aes(x = TMB, y = pvactools_filtered_neoantigen_load, col = Trial)) + geom_point() + theme_Publication() +
  scale_color_manual(values = trial_colors)

TMB_high_low_neoantigen_tmb_correlation_plot <-  wes_features_df %>%
  ggplot(aes(x = TMB, y = pvactools_filtered_neoantigen_load, col = Trial)) +
  geom_point() + theme_Publication() +
  facet_wrap(~ifelse(TMB >= 10, "TMB-high", "TMB-low"), scales = "free") +
  scale_color_manual(values = trial_colors) + labs(x = "TMB (Mutations/Mb)",
                                                   y = "Predicted Neoantigen load")

TMB_high_low_neoantigen_tmb_correlation_plot

graph2pdf(x = TMB_high_low_neoantigen_tmb_correlation_plot,
          file = here(output_dir,
                      "_TMB_high_low_neoantigen_tmb_correlation_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 7)

TMB_low_response_neoantigen_tmb_correlation_plot <-  wes_features_df %>%
  filter(TMB < 10) %>%
  ggplot(aes(x = TMB, y = pvactools_filtered_neoantigen_load, col = Trial)) +
  geom_point() + theme_Publication() +
  facet_wrap(~Response) +
  scale_color_manual(values = trial_colors) +
  labs(x = "TMB (Mutations/Mb)", y = "Predicted Neoantigen load", title = "TMB-low")
TMB_low_response_neoantigen_tmb_correlation_plot

graph2pdf(x = TMB_low_response_neoantigen_tmb_correlation_plot,
          file = here(output_dir,
                      "_TMB_low_response_neoantigen_tmb_correlation_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 7)


TMB_high_response_neoantigen_tmb_correlation_plot <-  wes_features_df %>%
  filter(TMB >= 10) %>%
  ggplot(aes(x = TMB, y = pvactools_filtered_neoantigen_load, col = Trial)) +
  geom_point() + theme_Publication() +
  facet_wrap(~Response) +
  scale_color_manual(values = trial_colors) +
  labs(x = "TMB (Mutations/Mb)", y = "Predicted Neoantigen load", title = "TMB-high")
TMB_high_response_neoantigen_tmb_correlation_plot

graph2pdf(x = TMB_high_response_neoantigen_tmb_correlation_plot,
          file = here(output_dir,
                      "_TMB_high_response_neoantigen_tmb_correlation_plot.pdf"),
          colormodel = "rgb",
          # font = "Arial",
          height = 4.5,
          width = 7)
```


```{r, fig.height = 10, fig.width = 10}
wes_features_df %>%
  filter(TMB < 10) %>%
  ggplot(aes(x = Response, y = TMB)) +
  facet_wrap(~Trial, scales = "free") +
  geom_boxplot(outlier.shape = NA) + geom_point() +
  theme_Publication() + stat_compare_means()

wes_features_df %>%
  filter(TMB < 10) %>%
  ggplot(aes(x = Response, y = pvactools_filtered_neoantigen_load)) +
  facet_wrap(~Trial, scales = "free") +
  geom_boxplot(outlier.shape = NA) + geom_point() +
  theme_Publication() + stat_compare_means()

```

## MATISSE proliferation subset combination vs monotherapy

```{r}
wes_rna_features_df_pre %>%
  filter(Histology == "CSCC")%>%
  group_by(Cohort, Response) %>%
  ggplot(aes(x = Response, y = Tumor_proliferation_rate)) +
  geom_boxplot(outlier.shape = NA) + geom_point() +
  facet_wrap(~Cohort, scales = "free_x") +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  labs(title = "MATISSE cSCC", y = "Proliferation ssGSEA", x = "") +
  geom_hline(yintercept = median_Tumor_proliferation_rate_pre,
             linetype = "dashed")

wes_rna_features_df_pre %>%
  filter(Histology == "CSCC")%>%
  group_by(Cohort, Response) %>%
  select(Histology, Trial, Response, Cohort, Tumor_proliferation_rate) %>%
  add_count() %>%
  mutate(Label = sprintf("%s\n(n = %s)", Response, n)) %>%
  ggplot(aes(x = Label, y = Tumor_proliferation_rate, fill = Response)) +
  geom_boxplot(outlier.shape = NA) + geom_point() +
  facet_wrap(~Cohort, scales = "free_x") +
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  labs(title = "MATISSE cSCC", y = "Proliferation ssGSEA", x = "") +
  geom_hline(yintercept = median_Tumor_proliferation_rate_pre,
             linetype = "dashed") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

wes_rna_features_df_pre %>%
  filter(Histology == "CSCC") %>%
  group_by(Response) %>%
  select(Histology, Trial, Response, Cohort, Tumor_proliferation_rate) %>%
  add_count() %>%
  mutate(Label = sprintf("%s\n(n = %s)", Response, n)) %>%
  ggplot(aes(x = Label, y = Tumor_proliferation_rate)) +
  geom_boxplot(outlier.shape = NA) + geom_point()  + 
  stat_compare_means(method = "wilcox") +
  theme_Publication() +
  labs(title = "MATISSE cSCC", y = "Proliferation ssGSEA", x = "") +
  geom_hline(yintercept = median_Tumor_proliferation_rate_pre,
             linetype = "dashed")
```


```{r}
wes_features_df_dmmr <- wes_features_df %>% filter(Trial == "NICHE-MSI")
wes_features_df_nabucco <- wes_features_df %>% filter(Trial == "NABUCCO")

table(wes_features_df_dmmr$BRCA1_2,
      wes_features_df_dmmr$Response_pCR,
      wes_features_df_dmmr$Trial)


fisher.test(wes_features_df_dmmr$BRCA1_2,
      wes_features_df_dmmr$Response_pCR)

table(wes_features_df_dmmr$BRCA1_2,
      wes_features_df_dmmr$Response_pCR,
      wes_features_df_dmmr$Trial) %>%
  as.data.frame() %>%
  setNames(c("BRCA1_2", "Response", "Trial", "Freq")) %>%
  pivot_wider(values_from = Freq, names_from = Response) %>%
  mutate(Total = non_pCR + pCR, pCR_response_rate = pCR / Total) %>%
  ggplot(aes(BRCA1_2, y = pCR_response_rate)) + geom_col() +
  theme_Publication() + facet_wrap(~Trial, scales = "free")

table(wes_features_df_nabucco$Response,
      wes_features_df_nabucco$Response_pCR)

table(wes_features_df_nabucco$BRCA1_2,
      wes_features_df_nabucco$Response_pCR,
      wes_features_df_nabucco$Trial)

fisher.test(wes_features_df_nabucco$BRCA1_2,
      wes_features_df_nabucco$Response_pCR)

table(wes_features_df_nabucco$BRCA1_2,
      wes_features_df_nabucco$Response_pCR,
      wes_features_df_nabucco$Trial) %>%
  as.data.frame() %>%
  setNames(c("BRCA1_2", "Response", "Trial", "Freq")) %>%
  pivot_wider(values_from = Freq, names_from = Response) %>%
  mutate(Total = non_pCR + pCR, pCR_response_rate = pCR / Total) %>%
  ggplot(aes(BRCA1_2, y = pCR_response_rate)) + geom_col() +
  theme_Publication() + facet_wrap(~Trial, scales = "free")

```


## MIXCR
```{r, fig.height = 8.5, fig.width = 12}
all_cohort_metrics_p <- read_csv2(here(
 source_data_dir, "all_cohort_mixcr_metrics_p.csv"))

all_cohort_metrics_pre <- all_cohort_metrics_p %>%
  filter(Treatment == "Pre")

all_cohort_metrics_pre %>%
  ggplot(aes(x = Response, y = total_clones, fill = Response)) +
  geom_boxplot() +
  facet_wrap(~chain, scales = "free") +
  theme_Publication()  +
  stat_compare_means(method = "wilcox") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(x = "", y = "Total number of reconstructed clones") +
  guides(fill = FALSE)
```


```{r, fig.height = 4, fig.width = 5}
all_cohort_metrics_pre_IGH <- all_cohort_metrics_pre %>%
  filter(chain == "IGH")  

all_cohort_metrics_pre_IGH_f <- all_cohort_metrics_pre_IGH %>%
  filter(total_clones > 50)

table(all_cohort_metrics_pre_IGH_f$Trial)
table(all_cohort_metrics_pre_IGH_f$Trial,
      all_cohort_metrics_pre_IGH_f$Response)

all_cohort_metrics_pre_IGH_f %>%
  ggplot(aes(x = Response, y = clonality, fill = Response)) +
  geom_boxplot() +
  theme_Publication()   +
  stat_compare_means(method = "wilcox") +
  scale_fill_manual(values = c(Color_NR, Color_R))+
  labs(title = "MIXCR IGH chains") +
  expand_limits(y = 0) +
  labs(x = "") + guides(fill = FALSE)
```


### UV and APOBEC vs TMB or neoantigens
```{r, fig.height = 8, fig.width = 8}

wes_features_df %>%
  ggplot(aes(x = UV_abs, y = pvactools_filtered_neoantigen_load)) +
  geom_point(aes(col = Trial)) +
  stat_cor(method = "spearman") + theme_Publication() +
  scale_color_manual(values = trial_colors) +
  facet_wrap(~Trial, scales = "free")

wes_features_df %>%
  ggplot(aes(x = APOBEC_abs, y = pvactools_filtered_neoantigen_load)) +
  geom_point(aes(col = Trial)) +
  stat_cor(method = "spearman") + theme_Publication() +
  scale_color_manual(values = trial_colors) +
  facet_wrap(~Trial, scales = "free")

wes_features_df %>%
  ggplot(aes(x = UV_abs, y = TMB)) +
  geom_point(aes(col = Trial)) +
  stat_cor(method = "spearman") + theme_Publication() +
  scale_color_manual(values = trial_colors) +
  facet_wrap(~Trial, scales = "free")

wes_features_df %>%
  ggplot(aes(x = APOBEC_abs, y = TMB)) +
  geom_point(aes(col = Trial)) +
  stat_cor(method = "spearman") + theme_Publication() +
  scale_color_manual(values = trial_colors) +
  facet_wrap(~Trial, scales = "free")
```

# ----------
# Lint script
```{r}
# lint("~/NKI_pancancer_neoadj_io/src/pancan_figures.Rmd")
lint(filename = "~/NKI_pancancer_neoadj_io/src/pancan_figures.Rmd",
  linters = unused_import_linter()
)
```

# ----------
# sessionInfo - Package Versions
```{r sessionInfo}
sessionInfo()
```
