---
title: "Pancan External cohort analysis"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  export: TRUE
  output_dir: "pancan_external_data_analysis"
  metadata: ""
  filter_and_parse_vcfs: FALSE
  load_all_germline: FALSE
---

Author comment:
File description comment:
The purpose of this script is to parse results from WES analysis
vcf files, quality logs, comparing with RNAseq data
filter_and_parse_vcfs

It receives parsed WES results and metadata as inputs
filters variants in germline and somatic
Studies mutation associations with response

# Loading Packages
```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# Install if necessary install.packages("BiocManager")
# Install if necessary BiocManager::install("GenVisR")

# library("devtools")
# devtools::install("~/Tools/TMBleR")
# BiocManager::install("VariantAnnotation")

library(tidyverse)
library(readxl)
source("helper_functions.R")
library(VariantAnnotation)
library(ggpubr)
library(ggrepel)
library(corrplot)
library(reshape2)
library(here)
library(scales)
library(lme4)
library(org.Hs.eg.db)
library("recount3")
library(GEOquery)
library("TCGAbiolinks")
library(ROCit)
library(PRROC)
library(pheatmap)
library(GSVA)
library(DESeq2)
library(rstatix)
library(biomaRt)

MakeDataFrameFromVCF <- getFromNamespace("MakeDataFrameFromVCF", "ICAMS")

set.seed(42)
select <- dplyr::select
rename <- dplyr::rename
summarize <- dplyr::summarize
here <- here::here
```

# Setup output folder
```{r Setup output, echo = FALSE, include = FALSE}
export <- params$export
filter_and_parse_vcfs <- params$filter_and_parse_vcfs

filter_and_parse_vcfs <- FALSE

output_dir <- here("data", "processed", params$output_dir)
print(output_dir)

external_analysis_dir <- here("data", "processed", params$output_dir)
differential_expression_meta_dir <- here("data", "processed",
                                         "differential_expression_meta")

# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)
    dir.create(here(output_dir, "filtered_vcfs"))
    # dir.create(here(output_dir, "missing_germline_vcfs"))

} else {
  print("Output folder already exists")
}

```


# Annotation color
```{r}

# Define colours for each group
Color_NR <- "#F8766D"
Color_R <- "#00BA38" # "#619CFF"
Color_NA <- "#CD9600"
Color_Pre <- "#619CFF"
Color_Post <- "#F564E3"

Color_HNSCC <- "#009292"
Color_CSCC <- "#CD9600"
Color_Urothelial <- "#B66DFF"
Color_LN_melanoma <- "#00BFC4"
Color_Colon <- "#F8766D"
Color_Colon_MSI <- "#B73A3A"
Color_Colon_MSS <- "#23C26F"
Color_TNBC <- "#FEB5DA"
Color_BELLINI <- "#FEB5DA"
Color_IMCISION <- "#009292"
Color_MATISSE <- "#CD9600"
Color_NABUCCO <- "#B66DFF"
Color_BELLINI <- "#FEB5DA"
Color_NICHE_MSS <- "#23C26F"
Color_NICHE_MSI <- "#B73A3A"
Color_OpACIN_neo <- "#00BFC4"
Color_PRADO <- "#619CFF"

annotation_color_list = list(
  Response = c(NR = Color_NR, R = Color_R),
  Treatment = c(Pre = Color_Pre, Post = Color_Post),
  Histology = c(LN_melanoma = Color_LN_melanoma, Urothelial = Color_Urothelial,
                Colon = Color_Colon, HNSCC = Color_HNSCC, TBNC = Color_TNBC,
                CSCC = Color_CSCC),
  Trial = c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO))

trial_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_colors_combined <- c(Combined = "#999999", BELLINI = Color_BELLINI,
                           IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_Colon_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

histology_msi_mss_colors <- c(`Colon-MSI`= Color_Colon_MSI,
                              `Colon-MSS`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              LN_melanoma = Color_LN_melanoma,
                              TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)

```

# -----------------------------

# Gene sets

```{r}
gene_signatures <- read_csv2(
  here("data", "processed", "signatures", "gene_signatures.csv")) %>%
  filter(!is.na(Gene))

signature_overview <- read_csv2(
  here("data", "external", "signatures",
       "_overview_of_signatures_20260212.csv")) 

gene_signatures_meta_analysis <- signature_overview %>%
  filter(Meta_analysis == "Yes")

# Load NKI pancancer signature components
pancan_response_signature_1.25 <- read_csv2(
  here(differential_expression_meta_dir,
       "pancan_response_signature_1.25.csv"))

pancan_nonresponse_signature_1.25 <- read_csv2(
  here(differential_expression_meta_dir,
       "pancan_nonresponse_signature_1.25.csv"))
```

## Combine
```{r}
gene_signatures <- gene_signatures %>%
  full_join(pancan_response_signature_1.25 %>%
              select(gene) %>%
              setNames("Gene") %>%
              mutate(Signature = "pancancer_response_sig_1.25")
            ) %>%
  full_join(pancan_nonresponse_signature_1.25 %>%
              select(gene) %>%
              setNames("Gene") %>%
              mutate(Signature = "pancan_nonresponse_signature_1.25")
            )

gene_signatures_l <- split(
  gene_signatures$Gene, gene_signatures$Signature)
gene_signatures_l <- gene_signatures_l[gene_signatures$Signature %>% unique()]

print(knitr::kable(gene_signatures %>%
        select(Signature, Class, Geneset_size) %>%
        unique()), n = n_distinct(gene_signatures$Signature))

gene_features <- read.csv2(
  file = here("data", "external", "biomaRt_gene_features.csv"))

gene_descriptions <- gene_features %>%
              select(hgnc_symbol, description) %>%
              rename("gene" = "hgnc_symbol") %>%
              unique()

```

# -----------------------------
# iAtlas datasets

##  blca_iatlas_imvigor210_2017
```{r}
imvigor210_dir <- "/shared/data/chalabi_group/cBioportal_datasets/blca_iatlas_imvigor210_2017"
list.files(imvigor210_dir)

imvigor210_clinical_patient <- read_delim(
  here(imvigor210_dir, "data_clinical_patient.txt"), skip = 4)
imvigor210_clinical_sample <- read_delim(
  here(imvigor210_dir, "data_clinical_sample.txt"), skip = 4) %>%
  mutate(RNA_SAMPLE_ID = paste0("X", SAMPLE_ID))

imvigor210_clinical_patient %>% mutate_if(is.character, as.factor) %>% summary()
imvigor210_clinical_sample %>% mutate_if(is.character, as.factor) %>% summary()

nrow(imvigor210_clinical_patient)
nrow(imvigor210_clinical_sample)
table(
  imvigor210_clinical_sample$PATIENT_ID == imvigor210_clinical_sample$SAMPLE_ID)

imvigor210_data_mutations <- read.delim(
  here(imvigor210_dir, "data_mutations.txt"))
imvigor210_data_mutations <- imvigor210_data_mutations %>%
  mutate(RNA_SAMPLE_ID = paste0("X", Tumor_Sample_Barcode))
imvigor210_data_mutations %>% colnames()

# 242 samples have genomics
n_distinct(imvigor210_data_mutations$Tumor_Sample_Barcode)
imvigor210_data_mutations$Tumor_Sample_Barcode %>% unique() %>% head()
imvigor210_data_gene_signatures <- read.delim(
  here(imvigor210_dir, "data_gene_signatures.txt"))

# RNA
imvigor210_mrna_seq_tpm <- read.delim(
  here(imvigor210_dir, "data_mrna_seq_tpm.txt"))

dim(imvigor210_mrna_seq_tpm)

table(
  colnames(imvigor210_mrna_seq_tpm) %in% paste0(
    "X", unique(imvigor210_data_mutations$Tumor_Sample_Barcode)))

imvigor210_clinical_sample %>%
  filter(SAMPLE_ID %in% imvigor210_data_mutations)

imvigor210_clinical_sample %>%
  filter(RNA_SAMPLE_ID %in% colnames(imvigor210_mrna_seq_tpm))

imvigor210_clinical_patient

imvigor210_clinical_sample %>% filter(is.na(RESPONSE)) %>% factor_summary()

imvigor210_metadata <- imvigor210_clinical_patient %>%
  left_join(imvigor210_clinical_sample) %>%
  filter(SAMPLE_ID %in% imvigor210_data_mutations$Tumor_Sample_Barcode) %>%
  filter(RNA_SAMPLE_ID %in% colnames(imvigor210_mrna_seq_tpm)) %>%
  mutate(has_RNA = TRUE, has_DNA = TRUE) %>%
  filter(!is.na(RESPONSE)) %>%
  mutate(PATIENT_ID = as.character(PATIENT_ID))%>%
  mutate(SAMPLE_ID = as.character(SAMPLE_ID)) %>%
  mutate(Study = "imvigor210_2017") %>%
  mutate(CLINICAL_BENEFIT = as.character(CLINICAL_BENEFIT))

imvigor210_metadata %>% factor_summary()
imvigor210_metadata %>% write_csv2(here(output_dir, "imvigor210_metadata.csv"))
```


## mel_iatlas_hugo_ucla_2016

```{r}
mel_iatlas_hugo_ucla_2016_dir <- "/shared/data/chalabi_group/cBioportal_datasets/mel_iatlas_hugo_ucla_2016"
list.files(mel_iatlas_hugo_ucla_2016_dir)

mel_iatlas_hugo_ucla_2016_clinical_patient <- read_delim(
  here(mel_iatlas_hugo_ucla_2016_dir, "data_clinical_patient.txt"), skip = 4)
mel_iatlas_hugo_ucla_2016_clinical_sample <- read_delim(
  here(mel_iatlas_hugo_ucla_2016_dir, "data_clinical_sample.txt"), skip = 4)

mel_iatlas_hugo_ucla_2016_clinical_patient %>% factor_summary()
mel_iatlas_hugo_ucla_2016_clinical_sample %>% factor_summary()

nrow(mel_iatlas_hugo_ucla_2016_clinical_patient)
nrow(mel_iatlas_hugo_ucla_2016_clinical_sample)

table(mel_iatlas_hugo_ucla_2016_clinical_sample$PATIENT_ID ==
        mel_iatlas_hugo_ucla_2016_clinical_sample$SAMPLE_ID)

mel_iatlas_hugo_ucla_2016_data_mutations <- read.delim(
  here(mel_iatlas_hugo_ucla_2016_dir, "data_mutations.txt"))
# str(mel_iatlas_hugo_ucla_2016_data_mutations)
mel_iatlas_hugo_ucla_2016_data_mutations %>% colnames()

n_distinct(mel_iatlas_hugo_ucla_2016_data_mutations$Tumor_Sample_Barcode)

mel_iatlas_hugo_ucla_2016_mrna_seq <- read.delim(
  here(mel_iatlas_hugo_ucla_2016_dir, "data_mrna_seq_tpm.txt"))

mel_iatlas_hugo_ucla_2016_signatures <- read.delim(
  here(mel_iatlas_hugo_ucla_2016_dir, "data_gene_signatures.txt"))

dim(mel_iatlas_hugo_ucla_2016_mrna_seq)
table(colnames(mel_iatlas_hugo_ucla_2016_mrna_seq) %in%
        unique(mel_iatlas_hugo_ucla_2016_data_mutations$Tumor_Sample_Barcode))

mel_iatlas_hugo_ucla_2016_metadata <- mel_iatlas_hugo_ucla_2016_clinical_patient %>%
  left_join(mel_iatlas_hugo_ucla_2016_clinical_sample) %>%
  filter(SAMPLE_ID %in% mel_iatlas_hugo_ucla_2016_data_mutations$Tumor_Sample_Barcode)  %>%
  filter(SAMPLE_ID %in% colnames(mel_iatlas_hugo_ucla_2016_mrna_seq)) %>%
  mutate(has_RNA = TRUE, has_DNA = TRUE) %>%
  mutate(Study = "hugo_ucla_2016")  %>%
  mutate(CLINICAL_BENEFIT = as.character(CLINICAL_BENEFIT))

mel_iatlas_hugo_ucla_2016_metadata %>% factor_summary()
mel_iatlas_hugo_ucla_2016_metadata %>%
  write_csv2(here(output_dir, "mel_iatlas_hugo_ucla_2016_metadata.csv"))
```


## rcc_iatlas_immotion150_2018

```{r}
# mcdermott et al
rcc_iatlas_immotion150_2018_dir <- "/shared/data/chalabi_group/cBioportal_datasets/rcc_iatlas_immotion150_2018"
list.files(rcc_iatlas_immotion150_2018_dir)

## No sex information
rcc_iatlas_immotion150_2018_clinical_patient <- read_delim(
  here(rcc_iatlas_immotion150_2018_dir, "data_clinical_patient.txt"), skip = 4)
rcc_iatlas_immotion150_2018_clinical_sample <- read_delim(
  here(rcc_iatlas_immotion150_2018_dir, "data_clinical_sample.txt"), skip = 4)

rcc_iatlas_immotion150_2018_clinical_patient %>% factor_summary()
rcc_iatlas_immotion150_2018_clinical_sample %>% factor_summary()

nrow(rcc_iatlas_immotion150_2018_clinical_patient)
nrow(rcc_iatlas_immotion150_2018_clinical_sample)
table(rcc_iatlas_immotion150_2018_clinical_sample$PATIENT_ID ==
        rcc_iatlas_immotion150_2018_clinical_sample$SAMPLE_ID)

rcc_iatlas_immotion150_2018_data_mutations <- read.delim(
  here(rcc_iatlas_immotion150_2018_dir, "data_mutations.txt"))

# str(rcc_iatlas_immotion150_2018_data_mutations)
rcc_iatlas_immotion150_2018_data_mutations %>% colnames()

n_distinct(rcc_iatlas_immotion150_2018_data_mutations$Tumor_Sample_Barcode)

rcc_iatlas_immotion150_2018_data_mutations$Tumor_Sample_Barcode %>%
  unique() %>%
  head()

rcc_iatlas_immotion150_2018_mrna_seq <- read.delim(
  here(rcc_iatlas_immotion150_2018_dir, "data_mrna_seq_tpm.txt"))

rcc_iatlas_immotion150_2018_signatures <- read.delim(
  here(rcc_iatlas_immotion150_2018_dir, "data_gene_signatures.txt"))

dim(rcc_iatlas_immotion150_2018_mrna_seq)
table(colnames(rcc_iatlas_immotion150_2018_mrna_seq) %in%
        unique(rcc_iatlas_immotion150_2018_data_mutations$Tumor_Sample_Barcode))

rcc_iatlas_immotion150_2018_metadata <- rcc_iatlas_immotion150_2018_clinical_patient %>%
  left_join(rcc_iatlas_immotion150_2018_clinical_sample) %>%
  filter(SAMPLE_ID %in%
           rcc_iatlas_immotion150_2018_data_mutations$Tumor_Sample_Barcode)  %>%
  filter(SAMPLE_ID %in% colnames(rcc_iatlas_immotion150_2018_mrna_seq)) %>%
  filter(!is.na(CLINICAL_BENEFIT)) %>%
  mutate(has_RNA = TRUE, has_DNA = TRUE) %>%
  mutate(Study = "immotion150_2018") %>%
  mutate(CLINICAL_BENEFIT = as.character(CLINICAL_BENEFIT))

rcc_iatlas_immotion150_2018_metadata %>% factor_summary()

rcc_iatlas_immotion150_2018_metadata %>%
  write_csv2(here(output_dir, "rcc_iatlas_immotion150_2018_metadata.csv"))
```


## mel_iatlas_liu_2019

```{r}

mel_iatlas_liu_2019_dir <- "/shared/data/chalabi_group/cBioportal_datasets/mel_iatlas_liu_2019"
list.files(mel_iatlas_liu_2019_dir)

mel_iatlas_liu_2019_clinical_patient <- read_delim(
  here(mel_iatlas_liu_2019_dir, "data_clinical_patient.txt"), skip = 4)
mel_iatlas_liu_2019_clinical_sample <- read_delim(
  here(mel_iatlas_liu_2019_dir, "data_clinical_sample.txt"), skip = 4)

mel_iatlas_liu_2019_clinical_patient %>% factor_summary()
mel_iatlas_liu_2019_clinical_sample %>% factor_summary()

nrow(mel_iatlas_liu_2019_clinical_patient)
nrow(mel_iatlas_liu_2019_clinical_sample)
table(mel_iatlas_liu_2019_clinical_sample$PATIENT_ID == mel_iatlas_liu_2019_clinical_sample$SAMPLE_ID)

mel_iatlas_liu_2019_data_mutations <- read.delim(
  here(mel_iatlas_liu_2019_dir, "data_mutations.txt"))

# str(mel_iatlas_liu_2019_data_mutations)
mel_iatlas_liu_2019_data_mutations %>% colnames()

n_distinct(mel_iatlas_liu_2019_data_mutations$Tumor_Sample_Barcode)
mel_iatlas_liu_2019_data_mutations$Tumor_Sample_Barcode %>% unique() %>% head()

mel_iatlas_liu_2019_mrna_seq <- read.delim(
  here(mel_iatlas_liu_2019_dir, "data_mrna_seq_tpm.txt"))

mel_iatlas_liu_2019_signatures <- read.delim(
  here(mel_iatlas_liu_2019_dir, "data_gene_signatures.txt"))

dim(mel_iatlas_liu_2019_mrna_seq)
table(colnames(mel_iatlas_liu_2019_mrna_seq) %in%
        unique(mel_iatlas_liu_2019_data_mutations$Tumor_Sample_Barcode))

mel_iatlas_liu_2019_metadata <- mel_iatlas_liu_2019_clinical_patient %>%
  left_join(mel_iatlas_liu_2019_clinical_sample) %>%
  filter(SAMPLE_ID %in% mel_iatlas_liu_2019_data_mutations$Tumor_Sample_Barcode)  %>%
  filter(SAMPLE_ID %in% colnames(mel_iatlas_liu_2019_mrna_seq)) %>%
  filter(!is.na(CLINICAL_BENEFIT)) %>%
  mutate(has_RNA = TRUE, has_DNA = TRUE) %>%
  mutate(Study = "mel_iatlas_Liu_2019") %>%
  mutate(CLINICAL_BENEFIT = as.character(CLINICAL_BENEFIT))

mel_iatlas_liu_2019_metadata %>% factor_summary()
mel_iatlas_liu_2019_metadata %>%
  write_csv2(here(output_dir, "mel_iatlas_liu_2019_metadata.csv"))
```


## mel_iatlas_gide_2019

### RNA only, pre and on treatment
```{r}

list.files("/shared/data/chalabi_group/cBioportal_datasets/mel_iatlas_gide_2019")

mel_iatlas_gide_2019_dir <- "/shared/data/chalabi_group/cBioportal_datasets/mel_iatlas_gide_2019"
list.files(mel_iatlas_gide_2019_dir)

mel_iatlas_gide_2019_clinical_patient <- read_delim(
  here(mel_iatlas_gide_2019_dir, "data_clinical_patient.txt"), skip = 4)
mel_iatlas_gide_2019_clinical_sample <- read_delim(
  here(mel_iatlas_gide_2019_dir, "data_clinical_sample.txt"), skip = 4)

mel_iatlas_gide_2019_clinical_patient %>% factor_summary()
mel_iatlas_gide_2019_clinical_sample %>% factor_summary()

nrow(mel_iatlas_gide_2019_clinical_patient)
nrow(mel_iatlas_gide_2019_clinical_sample)
table(mel_iatlas_gide_2019_clinical_sample$PATIENT_ID ==
        mel_iatlas_gide_2019_clinical_sample$SAMPLE_ID)

# No DNA data
# str(mel_iatlas_gide_2019_data_mutations)
# n_distinct(mel_iatlas_gide_2019_data_mutations$Tumor_Sample_Barcode)
# mel_iatlas_gide_2019_data_mutations$Tumor_Sample_Barcode %>% unique() %>% head()

mel_iatlas_gide_2019_mrna_seq <- read.delim(
  here(mel_iatlas_gide_2019_dir, "data_mrna_seq_expression.txt"))

dim(mel_iatlas_gide_2019_mrna_seq)
table(mel_iatlas_gide_2019_mrna_seq$Hugo_Symbol) %>%
  as.data.frame() %>%
  filter(Freq > 1)

mel_iatlas_gide_2019_metadata <- mel_iatlas_gide_2019_clinical_patient %>%
  left_join(mel_iatlas_gide_2019_clinical_sample) %>%
  filter(SAMPLE_ID %in% colnames(mel_iatlas_gide_2019_mrna_seq)) %>%
  filter(!is.na(CLINICAL_BENEFIT)) %>%
  mutate(has_RNA = TRUE, has_DNA = FALSE) %>%
  mutate(Study = "mel_iatlas_gide_2019") %>%
  mutate(CLINICAL_BENEFIT = as.character(CLINICAL_BENEFIT))

mel_iatlas_gide_2019_metadata %>% factor_summary()
mel_iatlas_gide_2019_metadata %>%
  write_csv2(here(output_dir, "mel_iatlas_gide_2019_metadata.csv"))
```


## brca_iatlas_anders_2022
```{r}
list.files("/shared/data/chalabi_group/cBioportal_datasets/brca_iatlas_anders_2022")

brca_iatlas_anders_2022_dir <- "/shared/data/chalabi_group/cBioportal_datasets/brca_iatlas_anders_2022"
list.files(brca_iatlas_anders_2022_dir)

brca_iatlas_anders_2022_clinical_patient <- read_delim(
  here(brca_iatlas_anders_2022_dir, "data_clinical_patient.txt"), skip = 4)
brca_iatlas_anders_2022_clinical_sample <- read_delim(
  here(brca_iatlas_anders_2022_dir, "data_clinical_sample.txt"), skip = 4) %>%
    mutate(RNA_SAMPLE_ID = paste0("X", SAMPLE_ID))

brca_iatlas_anders_2022_clinical_patient %>% factor_summary()
brca_iatlas_anders_2022_clinical_sample %>% factor_summary()

nrow(brca_iatlas_anders_2022_clinical_patient)
nrow(brca_iatlas_anders_2022_clinical_sample)
table(brca_iatlas_anders_2022_clinical_sample$PATIENT_ID ==
        brca_iatlas_anders_2022_clinical_sample$SAMPLE_ID)

brca_iatlas_anders_2022_data_mutations <- read.delim(
  here(brca_iatlas_anders_2022_dir, "data_mutations.txt")) %>%
  mutate(RNA_SAMPLE_ID = paste0("X", Tumor_Sample_Barcode))

# str(brca_iatlas_anders_2022_data_mutations)
brca_iatlas_anders_2022_data_mutations %>% colnames()

n_distinct(brca_iatlas_anders_2022_data_mutations$Tumor_Sample_Barcode)

imvigor210_data_mutations <- imvigor210_data_mutations %>%
  mutate(RNA_SAMPLE_ID = paste0("X", Tumor_Sample_Barcode))

brca_iatlas_anders_2022_mrna_seq <- read.delim(
  here(brca_iatlas_anders_2022_dir, "data_mrna_seq_tpm.txt"))

dim(brca_iatlas_anders_2022_mrna_seq)
table(colnames(brca_iatlas_anders_2022_mrna_seq) %in%
        unique(brca_iatlas_anders_2022_data_mutations$RNA_SAMPLE_ID))

brca_iatlas_anders_2022_metadata <- brca_iatlas_anders_2022_clinical_patient %>%
  left_join(brca_iatlas_anders_2022_clinical_sample) %>%
  filter(SAMPLE_ID %in%
           brca_iatlas_anders_2022_data_mutations$Tumor_Sample_Barcode)  %>%
  filter(RNA_SAMPLE_ID %in% colnames(brca_iatlas_anders_2022_mrna_seq)) %>%
  filter(!is.na(CLINICAL_BENEFIT)) %>%
  mutate(has_RNA = TRUE, has_DNA = TRUE) %>%
  mutate(Study = "brca_iatlas_anders_2022") %>%
  mutate(PATIENT_ID = as.character(PATIENT_ID))%>%
  mutate(SAMPLE_ID = as.character(SAMPLE_ID)) %>%
  mutate(CLINICAL_BENEFIT = as.character(CLINICAL_BENEFIT)) %>%
  mutate(SPECIMEN_PRESERVATION_TYPE = as.character(SPECIMEN_PRESERVATION_TYPE))

brca_iatlas_anders_2022_metadata %>% factor_summary()
brca_iatlas_anders_2022_metadata %>%
  write_csv2(here(output_dir, "brca_iatlas_anders_2022_metadata.csv"))

```

## gbm_iatlas_prins_2019

### RNA only
```{r}
gbm_iatlas_prins_2019_dir <- "/shared/data/chalabi_group/cBioportal_datasets/gbm_iatlas_prins_2019"
list.files(gbm_iatlas_prins_2019_dir)

gbm_iatlas_prins_2019_clinical_patient <- read_delim(
  here(gbm_iatlas_prins_2019_dir, "data_clinical_patient.txt"), skip = 4)
gbm_iatlas_prins_2019_clinical_sample <- read_delim(
  here(gbm_iatlas_prins_2019_dir, "data_clinical_sample.txt"), skip = 4)

gbm_iatlas_prins_2019_clinical_patient %>% factor_summary()
gbm_iatlas_prins_2019_clinical_sample %>% factor_summary()

nrow(gbm_iatlas_prins_2019_clinical_patient)
nrow(gbm_iatlas_prins_2019_clinical_sample)
table(
  gbm_iatlas_prins_2019_clinical_sample$PATIENT_ID ==
    gbm_iatlas_prins_2019_clinical_sample$SAMPLE_ID)

gbm_iatlas_prins_2019_mrna_seq <- read.delim(
  here(gbm_iatlas_prins_2019_dir, "data_mrna_seq_expression.txt"))

dim(gbm_iatlas_prins_2019_mrna_seq)
table(colnames(gbm_iatlas_prins_2019_mrna_seq) %in%
        unique(gbm_iatlas_prins_2019_clinical_sample$SAMPLE_ID))

gbm_iatlas_prins_2019_metadata <- gbm_iatlas_prins_2019_clinical_patient %>%
  left_join(gbm_iatlas_prins_2019_clinical_sample) %>%
  filter(SAMPLE_ID %in% colnames(gbm_iatlas_prins_2019_mrna_seq)) %>%
  filter(!is.na(CLINICAL_BENEFIT)) %>%
  mutate(has_RNA = TRUE, has_DNA = FALSE) %>%
  mutate(Study = "gbm_iatlas_prins_2019") %>%
  mutate(CLINICAL_BENEFIT = as.character(CLINICAL_BENEFIT))

gbm_iatlas_prins_2019_metadata %>% factor_summary()
gbm_iatlas_prins_2019_metadata %>%
  write_csv2(here(output_dir, "gbm_iatlas_prins_2019_metadata.csv"))
```

## ccrcc_iatlas_choueiri_2016

```{r}
ccrcc_iatlas_choueiri_2016_dir <- "/shared/data/chalabi_group/cBioportal_datasets/ccrcc_iatlas_choueiri_2016"
list.files(ccrcc_iatlas_choueiri_2016_dir)

ccrcc_iatlas_choueiri_2016_clinical_patient <- read_delim(
  here(ccrcc_iatlas_choueiri_2016_dir, "data_clinical_patient.txt"), skip = 4)
ccrcc_iatlas_choueiri_2016_clinical_sample <- read_delim(
  here(ccrcc_iatlas_choueiri_2016_dir, "data_clinical_sample.txt"), skip = 4)

ccrcc_iatlas_choueiri_2016_clinical_patient %>% factor_summary()
ccrcc_iatlas_choueiri_2016_clinical_sample %>% factor_summary()

nrow(ccrcc_iatlas_choueiri_2016_clinical_patient)
nrow(ccrcc_iatlas_choueiri_2016_clinical_sample)
table(ccrcc_iatlas_choueiri_2016_clinical_sample$PATIENT_ID ==
        ccrcc_iatlas_choueiri_2016_clinical_sample$SAMPLE_ID)

ccrcc_iatlas_choueiri_2016_data_mutations <- read.delim(
  here(ccrcc_iatlas_choueiri_2016_dir, "data_mutations.txt"))

ccrcc_iatlas_choueiri_2016_data_mutations %>% colnames()

n_distinct(ccrcc_iatlas_choueiri_2016_data_mutations$Tumor_Sample_Barcode)
ccrcc_iatlas_choueiri_2016_data_mutations$Tumor_Sample_Barcode %>% unique() %>% head()

ccrcc_iatlas_choueiri_2016_mrna_seq <- read.delim(
  here(ccrcc_iatlas_choueiri_2016_dir, "data_mrna_seq_expression.txt"))

dim(ccrcc_iatlas_choueiri_2016_mrna_seq)
table(colnames(ccrcc_iatlas_choueiri_2016_mrna_seq) %in%
        unique(ccrcc_iatlas_choueiri_2016_data_mutations$Tumor_Sample_Barcode))

ccrcc_iatlas_choueiri_2016_metadata <- ccrcc_iatlas_choueiri_2016_clinical_patient %>%
  left_join(ccrcc_iatlas_choueiri_2016_clinical_sample) %>%
  filter(SAMPLE_ID %in%
           ccrcc_iatlas_choueiri_2016_data_mutations$Tumor_Sample_Barcode)  %>%
  filter(SAMPLE_ID %in% colnames(ccrcc_iatlas_choueiri_2016_mrna_seq)) %>%
  filter(!is.na(CLINICAL_BENEFIT)) %>%
  mutate(has_RNA = TRUE, has_DNA = TRUE) %>%
  mutate(Study = "ccrcc_iatlas_choueiri_2016") %>%
  mutate(CLINICAL_BENEFIT = as.character(CLINICAL_BENEFIT))

ccrcc_iatlas_choueiri_2016_metadata %>% factor_summary()
ccrcc_iatlas_choueiri_2016_metadata %>%
  write_csv2(here(output_dir, "ccrcc_iatlas_choueiri_2016_metadata.csv"))
```


# Combine metadata
```{r}
combined_metadata <- imvigor210_metadata %>%
  full_join(mel_iatlas_hugo_ucla_2016_metadata) %>%
  full_join(rcc_iatlas_immotion150_2018_metadata ) %>%
  full_join(brca_iatlas_anders_2022_metadata) %>%
  full_join(mel_iatlas_liu_2019_metadata) %>%
  full_join(ccrcc_iatlas_choueiri_2016_metadata) %>%
  full_join(gbm_iatlas_prins_2019_metadata) %>%
  full_join(mel_iatlas_gide_2019_metadata)

combined_metadata <- combined_metadata %>%
  mutate(RNA_SAMPLE_ID = ifelse(is.na(RNA_SAMPLE_ID),
                                SAMPLE_ID, RNA_SAMPLE_ID)) %>%
  mutate(CANCER_TYPE = ifelse(Study == "gbm_iatlas_prins_2019",
                              "Glioblastoma", CANCER_TYPE))

metastatic_metastatic_wes_rna_features_df_pre %>%
  filter(is.na(SEX)) %>%
  factor_summary()

dim(combined_metadata)
combined_metadata %>% filter(ICI_PATHWAY == "None") %>%
  select(Study) %>%
  factor_summary()

combined_metadata %>%
  filter(ICI_PATHWAY == "None") %>%
  factor_summary()

combined_metadata %>%
  filter(is.na(ICI_PATHWAY)) %>%
  select(Study) %>%
  factor_summary()

combined_metadata %>%
  filter(is.na(ICI_TARGET)) %>%
  select(Study) %>%
  factor_summary()

combined_metadata <- combined_metadata %>% filter(ICI_PATHWAY != "None")

combined_metadata %>% factor_summary()

combined_metadata %>% select(CANCER_TYPE, Study) %>% unique()

table(combined_metadata$ICI_PATHWAY, combined_metadata$NEOICI_RX)
table(combined_metadata$Study, combined_metadata$NEOICI_RX)
table(combined_metadata$Study, combined_metadata$ICI_RX)
table(combined_metadata$Study, combined_metadata$ICI_PATHWAY)
table(combined_metadata$Study, combined_metadata$CANCER_TYPE)

table(combined_metadata$Study, combined_metadata$CLINICAL_STAGE)

```

##  Harmonize Response/CB annotation
```{r}
# Harmonize CB annotation
# combined_metadata %>% filter(is.na(RESPONSE)) %>% select(Study, RESPONSE, CLINICAL_BENEFIT)
table(combined_metadata$RESPONDER, combined_metadata$RESPONSE, useNA = "ifany")
table(combined_metadata$Study, combined_metadata$RESPONSE, useNA = "ifany")

# combined_metadata %>% filter(Study %in% c("Van_allen_dfci_2015")) %>% factor_summary()
# combined_metadata %>% filter(Study %in% c("Snyder_mskcc_2014")) %>% factor_summary()

table(combined_metadata$Study, combined_metadata$CLINICAL_BENEFIT)
combined_metadata$Response <- combined_metadata$RESPONSE
combined_metadata <- combined_metadata %>%
  mutate(Response = ifelse(Response %in% c("Complete Response", "CR", "LB",
                                           "Partial Response", "PR"), "R",
                           Response)) %>%
  mutate(Response = ifelse(Response %in% c("NB", "PD", "Progressive Disease",
                                           "SD", "Stable Disease"), "NR", 
                           Response))

table(combined_metadata$Response)
table(combined_metadata$Study, combined_metadata$Response)
table(combined_metadata$Study, combined_metadata$Response) %>%
  prop.table(margin = 1) %>% round(2)

study_response_rate <- table(combined_metadata$Study,
                             combined_metadata$Response) %>%
  prop.table(margin = 1) %>%
  round(2) %>%
  as.data.frame() %>%
  setNames(c("Study", "Response", "Response_rate")) %>%
  filter(Response == "R")

study_response_rate

combined_metadata %>%
  write_csv2(here(output_dir, "combined_metadata.csv"))

table(combined_metadata$Study, is.na(combined_metadata$RNA_SAMPLE_ID))
table(combined_metadata$Study, is.na(combined_metadata$RNA_SAMPLE_ID))

```

# Plot number of studies

```{r, fig.height = 6}
table(combined_metadata$Study, combined_metadata$CANCER_TYPE) %>%
  as.data.frame() %>%
  setNames(c("Study", "CANCER_TYPE", "Freq")) %>%
  ggplot(aes(x = Study, y = Freq, fill = CANCER_TYPE)) +
  geom_col() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  guides(fill = guide_legend(nrow = 2))
```


```{r, fig.height = 6}
table(combined_metadata$Study, combined_metadata$CANCER_TYPE) %>%
  as.data.frame()  %>%
  setNames(c("Study", "CANCER_TYPE", "Freq"))  %>%
  ggplot(aes(x = CANCER_TYPE, y = Freq, fill = Study)) +
  geom_col() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  guides(fill = guide_legend(nrow = 3))
```

# -----------------------------

# TMB vs neoantigen
```{r}
combined_metadata  %>%
  ggplot(aes(x = TMB_NONSYNONYMOUS, y = SNV_NEOANTIGEN)) +
  geom_point(aes(col = CANCER_TYPE)) +
  theme_Publication() + stat_cor(method = "spearman") +
  labs(title = "cBioportal metastatic cohort") +
  guides(col = guide_legend(nrow = 3))

combined_metadata %>%
  ggplot(aes(x = TMB_NONSYNONYMOUS, y = INDEL_NEOANTIGEN)) +
  geom_point(aes(col = CANCER_TYPE)) +
  theme_Publication() +  stat_cor(method = "spearman") +
    guides(col = guide_legend(nrow = 4)) +
  labs(title = "cBioportal metastatic cohort") +
  guides(col = guide_legend(nrow = 3))

combined_metadata %>%
  filter(!Study %in% c("gbm_iatlas_prins_2019", "mel_iatlas_gide_2019")) %>%
  mutate(TOTAL_NEOANTIGEN = SNV_NEOANTIGEN + INDEL_NEOANTIGEN +
           VIRUS_NEOANTIGEN + FUSION_NEOANTIGEN + CTA_SELF_NEOANTIGEN +
           ERV_NEOANTIGEN ) %>%
  ggplot(aes(x = TMB_NONSYNONYMOUS, y = TOTAL_NEOANTIGEN)) +
  stat_cor(method = "spearman") +
  geom_point(aes(col = CANCER_TYPE)) + theme_Publication() +
  labs(title = "cBioportal metastatic cohort") +
  guides(col = guide_legend(nrow = 3))

```

```{r}
combined_metadata %>%
  filter(!Study %in% c("gbm_iatlas_prins_2019", "mel_iatlas_gide_2019")) %>%
  mutate(TOTAL_NEOANTIGEN = SNV_NEOANTIGEN + INDEL_NEOANTIGEN +
           VIRUS_NEOANTIGEN + FUSION_NEOANTIGEN + CTA_SELF_NEOANTIGEN +
           ERV_NEOANTIGEN ) %>%
  ggplot(aes(x = TMB_NONSYNONYMOUS, y = TOTAL_NEOANTIGEN)) +
  stat_cor(method = "spearman") +
  geom_point(aes(col = CANCER_TYPE)) + theme_Publication() +
  facet_wrap(~CANCER_TYPE) +
  guides(col = guide_legend(nrow = 2))

```

# -----------------------------
# Combine RNA
```{r}
# Check they genes are in the same order
table(imvigor210_mrna_seq_tpm$Hugo_Symbol == mel_iatlas_hugo_ucla_2016_mrna_seq$Hugo_Symbol)
table(imvigor210_mrna_seq_tpm$Hugo_Symbol == rcc_iatlas_immotion150_2018_mrna_seq$Hugo_Symbol)
table(imvigor210_mrna_seq_tpm$Hugo_Symbol == mel_iatlas_liu_2019_mrna_seq$Hugo_Symbol)
table(imvigor210_mrna_seq_tpm$Hugo_Symbol == brca_iatlas_anders_2022_mrna_seq$Hugo_Symbol)
table(imvigor210_mrna_seq_tpm$Hugo_Symbol == mel_iatlas_gide_2019_mrna_seq$Hugo_Symbol)
table(imvigor210_mrna_seq_tpm$Hugo_Symbol == ccrcc_iatlas_choueiri_2016_mrna_seq$Hugo_Symbol)
table(imvigor210_mrna_seq_tpm$Hugo_Symbol == gbm_iatlas_prins_2019_mrna_seq$Hugo_Symbol)

iatlas_combined_RNA <- imvigor210_mrna_seq_tpm %>%
  cbind(mel_iatlas_hugo_ucla_2016_mrna_seq[, -1]) %>%
  cbind(rcc_iatlas_immotion150_2018_mrna_seq[, -1]) %>%
  cbind(mel_iatlas_liu_2019_mrna_seq[, -1]) %>%
  cbind(brca_iatlas_anders_2022_mrna_seq[, -1]) %>%
  cbind(mel_iatlas_gide_2019_mrna_seq[, -1])%>%
  cbind(ccrcc_iatlas_choueiri_2016_mrna_seq[, -1]) %>%
  cbind(gbm_iatlas_prins_2019_mrna_seq[, -1]) %>%
  column_to_rownames("Hugo_Symbol") %>%
  as.matrix()

colSums(iatlas_combined_RNA) %>% summary()

dim(iatlas_combined_RNA)
class(iatlas_combined_RNA)
length(unique(colnames(iatlas_combined_RNA)))
```

# TPM ssGSEA

```{r}
packageVersion("GSVA")
run_ssgsea <- FALSE
if (run_ssgsea) {
    
  tpm_ssgsea_results <- gsva(
    as.matrix(iatlas_combined_RNA),
    gene_signatures_l,
    method = "ssgsea",
    kcdf = "Gaussian",
    # Minimum gene set size
    min.sz = 1,
    # Compute Gaussian-distributed scores
    mx.diff = TRUE,
    # Don't print out the progress bar
    verbose = FALSE
    )

  tpm_ssgsea_results %>%
    as.data.frame() %>%
    rownames_to_column("Signature") %>%
    write_csv2(here(output_dir, "tpm_ssgsea_results.csv"))
  
}

tpm_ssgsea_results <- read_csv2(here(output_dir, "tpm_ssgsea_results.csv"))

table(colnames(tpm_ssgsea_results) %in% combined_metadata$RNA_SAMPLE_ID)
table(colnames(tpm_ssgsea_results) %in% combined_metadata$SAMPLE_ID)

tpm_ssgsea_results_df <- tpm_ssgsea_results %>%
  column_to_rownames("Signature") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("RNA_SAMPLE_ID") %>%
  left_join(combined_metadata, by = "RNA_SAMPLE_ID") %>%
  filter(!is.na(Study))

# Check NA values
tpm_ssgsea_NAs <- tpm_ssgsea_results %>%
  column_to_rownames("Signature") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("RNA_SAMPLE_ID") %>%
  left_join(combined_metadata, by = "RNA_SAMPLE_ID") %>%
  filter(is.na(Study))

dim(tpm_ssgsea_NAs)
tpm_ssgsea_NAs$RNA_SAMPLE_ID
```

# -----------------------------
# WES RNA features df
```{r}
# TPM iatlas
metastatic_wes_rna_features_df_pre <- tpm_ssgsea_results_df %>%
  mutate(Response_n = ifelse(
    Response == "R",
    1, ifelse(
      Response == "NR", 0, NA))) %>%
  mutate(pancancer_response_score =
             pancancer_response_sig_1.25 - pancan_nonresponse_signature_1.25,
         Trial = Study) %>%
  mutate(
    TMB = TMB_NONSYNONYMOUS,
    TMB_scaled = scale(TMB),
    TIL_score_scaled = scale(TIL_score),
    Proliferation_scaled = scale(Tumor_proliferation_rate),
    B_cat_scaled = scale(B_catenin_signature_Luke_et_al),
    pancancer_response_score_scaled = scale(pancancer_response_score))

metastatic_wes_rna_features_df_pre[1:5, c(284, 285, 286, 287)]

metastatic_wes_rna_features_df_pre$TIL_score_scaled <- as.vector(
  metastatic_wes_rna_features_df_pre$TIL_score_scaled)
metastatic_wes_rna_features_df_pre$Proliferation_scaled <- as.vector(
  metastatic_wes_rna_features_df_pre$Proliferation_scaled)
metastatic_wes_rna_features_df_pre$B_cat_scaled <- as.vector(
  metastatic_wes_rna_features_df_pre$B_cat_scaled)
metastatic_wes_rna_features_df_pre$TMB_scaled <- as.vector(
  metastatic_wes_rna_features_df_pre$TMB_scaled)
metastatic_wes_rna_features_df_pre$pancancer_response_score_scaled <- as.vector(
  metastatic_wes_rna_features_df_pre$pancancer_response_score_scaled)

```


# Correlations
```{r, fig.height = 30, fig.width = 6}
# TPM
ssgsea_correlations <- tpm_ssgsea_results_df %>%
  select(tpm_ssgsea_results$Signature) %>%
  cor()

ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancancer_response_sig_1.25) %>%
  setNames(c("pancan_response_sig_corr")) %>%
  rownames_to_column("Signature") %>%
  filter(!is.na(pancan_response_sig_corr)) %>%
  ggplot(aes(x = reorder(Signature, pancan_response_sig_corr),
                         y = pancan_response_sig_corr)) +
  geom_col() + theme_Publication() + coord_flip()

ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancan_nonresponse_signature_1.25) %>%
  setNames(c("pancan_nonresponse_signature_1.25")) %>%
  rownames_to_column("Signature") %>%
  filter(!is.na(pancan_nonresponse_signature_1.25)) %>%
  ggplot(aes(x = reorder(Signature, pancan_nonresponse_signature_1.25),
                         y = pancan_nonresponse_signature_1.25)) +
  geom_col() + theme_Publication() + coord_flip()
```

```{r, fig.height = 12, fig.width = 12}
metastatic_wes_rna_features_df_pre$B_catenin_signature_Luke_et_al %>% summary()
gene_signatures_meta_analysis$Signature

metastatic_wes_rna_features_df_pre %>%
  select(all_of(c(gene_signatures_meta_analysis$Signature[
    gene_signatures_meta_analysis$Signature %in% colnames(ssgsea_results_df)],
    "pancancer_response_sig_1.25", "pancan_nonresponse_signature_1.25",
    "pancancer_response_score"))) %>%
  as.matrix() %>%
  cor() %>%
  corrplot(order = "AOE")

```

```{r, fig.height = 8, fig.width = 6}
ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancancer_response_sig_1.25) %>%
  setNames(c("pancan_response_sig_corr")) %>%
  rownames_to_column("Signature") %>%
  filter(pancan_response_sig_corr > 0.7) %>%
  ggplot(aes(x = reorder(Signature, pancan_response_sig_corr),
                         y = pancan_response_sig_corr)) +
  geom_col() + theme_Publication() + coord_flip() + labs(x = "")

ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancancer_response_sig_1.25) %>%
  setNames(c("pancan_response_sig_corr")) %>%
  rownames_to_column("Signature") %>%
  filter(Signature %in% gene_signatures_meta_analysis$Signature) %>%
  ggplot(aes(x = reorder(Signature, pancan_response_sig_corr),
                         y = pancan_response_sig_corr)) +
  geom_col() + theme_Publication() + coord_flip() + labs(x = "") +
  labs(title = "External metastatic cohorts\nResponse sig ssGSEA corr")

ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancan_nonresponse_signature_1.25) %>%
  setNames(c("pancan_nonresponse_sig_corr")) %>%
  rownames_to_column("Signature")%>%
  filter(Signature %in% gene_signatures_meta_analysis$Signature)%>%
  ggplot(aes(x = reorder(Signature, pancan_nonresponse_sig_corr),
                         y = pancan_nonresponse_sig_corr)) +
  geom_col() + theme_Publication() + coord_flip() + labs(x = "") +
  labs(title = "External metastatic cohorts\nNon-response sig ssGSEA corr")

```

# ------------

# GLMer Models
```{r}
dim(metastatic_wes_rna_features_df_pre)

Trial_model <- glmer(Response_n ~ (1 | Trial),
                     data = metastatic_wes_rna_features_df_pre,
                     family = binomial)
summary(Trial_model)

Proliferation_model <- glmer(Response_n ~ Proliferation_scaled + (1 | Trial),
                             data = metastatic_wes_rna_features_df_pre,
                             family = binomial)
summary(Proliferation_model)

TIL_model <- glmer(Response_n ~ TIL_score_scaled + (1 | Trial),
                   data = metastatic_wes_rna_features_df_pre, family = binomial)
summary(TIL_model)

TMB_model <- glmer(Response_n ~ TMB_scaled + (1 | Trial),
                   data = metastatic_wes_rna_features_df_pre, family = binomial)
summary(TMB_model)

```

# GLMer TMB tumour type

## Fitting trials
```{r}
metastatic_wes_TMB_rna_features_df_pre_f <- metastatic_wes_rna_features_df_pre %>%
  filter(!Study %in% c("gbm_iatlas_prins_2019", "mel_iatlas_gide_2019")) %>%
  filter(!is.na(TMB)) 

TMB_model <- glm(Response_n ~ TMB_scaled,
                 data = metastatic_wes_TMB_rna_features_df_pre_f,
                 family = binomial)
summary(TMB_model)

Trial_model <- glmer(Response_n ~ (1 | Trial),
                     data = metastatic_wes_TMB_rna_features_df_pre_f,
                     family = binomial)
summary(Trial_model)

TMB_trial_model <- glmer(Response_n ~ TMB_scaled + (1 | Trial),
                         data = metastatic_wes_TMB_rna_features_df_pre_f,
                         family = binomial)
summary(TMB_trial_model)

trial_predictions <- predict(Trial_model, type = "response")
tmb_predictions <- predict(TMB_model, type = "response")
tmb_trial_predictions <- predict(TMB_trial_model, type = "response")
```

## Likelihood ratio test
```{r}
anova(TMB_trial_model, Trial_model) # These are nested
AIC(TMB_trial_model, TMB_model)
BIC(TMB_trial_model, TMB_model)
```

# Boxplots

```{r}
# Incliuding GBM and mel_itlas
table(metastatic_wes_rna_features_df_pre$Study,
      metastatic_wes_rna_features_df_pre$CANCER_TYPE)

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = Tumor_proliferation_rate, fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "External metastatic data") +
  scale_fill_manual(values = c(Color_NR, Color_R))

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = TIL_score, fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = TMB_NONSYNONYMOUS, fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = factor(Response),
             y = log2(TMB), fill = Response)) +
  geom_boxplot() + theme_Publication() +
  # guides(fill = FALSE) +
  labs(x = "", y = "log2(TMB) (mutations/Mb)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  geom_hline(yintercept = log2(10), linetype = "dashed", alpha = 0.5) +
  stat_compare_means(method = "wilcox") +
  scale_fill_manual(values = c(Color_NR, Color_R))
```

```{r, fig.height = 12, fig.width= 12}
metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = B_catenin_signature_Luke_et_al,
             fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") + facet_wrap(~Study) +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 

metastatic_wes_TMB_rna_features_df_pre_f %>%
  ggplot(aes(x = Response, y = TMB_NONSYNONYMOUS, fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  facet_wrap(~Study, scale = "free") + expand_limits(y = 0) +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data")
```


## Pancan Response Signature
```{r}
table(metastatic_wes_rna_features_df_pre$Study,
      metastatic_wes_rna_features_df_pre$CANCER_TYPE)

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = pancancer_response_sig_1.25, fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = pancan_nonresponse_signature_1.25,
             fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 
```

```{r, fig.height = 12, fig.width= 12}

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = pancancer_response_sig_1.25,
             fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") + facet_wrap(~Study) +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = pancan_nonresponse_signature_1.25,
             fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") + facet_wrap(~Study) +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = pancancer_response_score,
             fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") + facet_wrap(~Study) +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 
```


# Proliferation TIL score

```{r, fig.height = 12, fig.width = 12}
metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = Tumor_proliferation_rate, fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  facet_wrap(~Study) +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Response, y = TIL_score, fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  facet_wrap(~Study) +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  labs(title = "External metastatic data") 

```

```{r}
Trial_model <- glmer(Response_n ~ (1 | Trial),
                     data = metastatic_wes_rna_features_df_pre,
                     family = binomial)
summary(Trial_model)

median_Tumor_proliferation_rate_pre <- metastatic_wes_rna_features_df_pre %>%
  pull(Tumor_proliferation_rate) %>%
  median()
median_Tumor_proliferation_rate_pre

median_til_score_pre <- median(
  metastatic_wes_rna_features_df_pre$TIL_score)
median_til_score_pre

prolif_TIL_quadrant_trial_plot <- metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Study)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = median_Tumor_proliferation_rate_pre,
             linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  guides(col = guide_legend(nrow = 3)) + facet_wrap(~Response)

prolif_TIL_quadrant_trial_plot

trial_quadrant_proportions <- table(
  metastatic_wes_rna_features_df_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre, 
  metastatic_wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  metastatic_wes_rna_features_df_pre$Study) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
dim(metastatic_wes_rna_features_df_pre)

table(metastatic_wes_rna_features_df_pre$Response)

quadrant_response_proportions <- table(
  metastatic_wes_rna_features_df_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  metastatic_wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  metastatic_wes_rna_features_df_pre$Response) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Response", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions

quadrant_response_proportions %>%
  select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()

quadrant_response_proportions %>%
  filter(Response == "R") %>%
  rename(Response_rate = quadrant_prop) %>%
  ggplot(aes(x = quadrant, y = Response_rate)) + geom_col() +
  theme_Publication() + labs(x = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

trial_quadrant_proportions <- table(
  metastatic_wes_rna_features_df_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  metastatic_wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  metastatic_wes_rna_features_df_pre$Study) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
trial_quadrant_proportions

trial_quadrant_proportions %>% select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

metastatic_wes_rna_features_df_pre %>% 
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Study)) +
  geom_point() + theme_Publication() + facet_wrap(~Response) 

```

```{r}
Proliferation_TIL_score_model <- glmer(
  Response_n ~ Proliferation_scaled * TIL_score_scaled + (1 | Trial),
  data = metastatic_wes_rna_features_df_pre, family = binomial)
summary(Proliferation_TIL_score_model)

ranef(Proliferation_TIL_score_model)
summary(Proliferation_TIL_score_model)$coefficients %>% round(2)
summary(Proliferation_TIL_score_model)$coefficients[, 4] %>% round(4)

Proliferation_TIL_int_score_model <- glmer(
  Response_n ~ Proliferation_scaled * TIL_score_scaled + (1 | Trial),
  data = metastatic_wes_rna_features_df_pre, family = binomial)
summary(Proliferation_TIL_int_score_model)
ranef(Proliferation_TIL_int_score_model)
```


```{r}
TMB_TIL_score_model <- glmer(
  Response_n ~ TMB_scaled * TIL_score_scaled + (1 | Trial),
  data = metastatic_wes_rna_features_df_pre, family = binomial)
summary(TMB_TIL_score_model)

ranef(TMB_TIL_score_model)
summary(TMB_TIL_score_model)$coefficients %>% round(2)
summary(TMB_TIL_score_model)$coefficients[, 4] %>% round(4)
```

# TMB vs TIL score

```{r}
# Quadrant RR
table(metastatic_wes_rna_features_df_pre$TMB_NONSYNONYMOUS >= 10,
      metastatic_wes_rna_features_df_pre$TIL_score >= median_til_score_pre) %>%
  `rownames<-`(c("TMB-L", "TMB-H")) %>%
  `colnames<-`( c("mTIL_low", "mTIL_high"))

metastatic_wes_rna_features_df_pre %>%
  ggplot(aes(x = log2(TMB), y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication()

# Quadrant RR
table(metastatic_wes_rna_features_df_pre$TMB >= 10,
      metastatic_wes_rna_features_df_pre$TIL_score >= median_til_score_pre) %>%
  `rownames<-`(c("TMB-L", "TMB-H")) %>%
  `colnames<-`( c("mTIL_low", "mTIL_high"))

quadrant_response_proportions <- table(
  metastatic_wes_rna_features_df_pre$TMB >= 10,
  metastatic_wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  metastatic_wes_rna_features_df_pre$Response) %>%
  as.data.frame() %>%
  setNames(c("TMB", "mTIL_group", "Response", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB == TRUE, "TMB_high", "TMB_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions

quadrant_response_proportions %>%
  filter(Response == "R") %>%
  rename(Response_rate = quadrant_prop) %>%
  ggplot(aes(x = quadrant, y = Response_rate)) + geom_col() +
  theme_Publication() + labs(x = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

quadrant_response_proportions

quadrant_response_proportions %>%
  select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()

trial_quadrant_proportions <- table(
  metastatic_wes_rna_features_df_pre$TMB >= 10,
  metastatic_wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  metastatic_wes_rna_features_df_pre$Trial) %>%
  as.data.frame() %>%
  setNames(c("TMB_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB_group == TRUE, "TMB-H", "TMB-L")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
trial_quadrant_proportions %>%
  select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

quadrant_trial_proportions <- table(
  metastatic_wes_rna_features_df_pre$TMB >= 10,
  metastatic_wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  metastatic_wes_rna_features_df_pre$Trial) %>%
  as.data.frame() %>%
  setNames(c("TMB_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB_group == TRUE, "TMB-H", "TMB-L")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group)) %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = Freq / Total)

TMB_TIL_quadrant_trial_proportions_plot <- quadrant_trial_proportions %>%
  ggplot(aes(x = quadrant, y = quadrant_prop, fill = Trial)) + geom_col() +
  theme_Publication() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 3)) +
  coord_flip()

TMB_TIL_quadrant_trial_proportions_plot

```

```{r}

metastatic_wes_rna_features_df_pre %>% 
  ggplot(aes(x = TMB, y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  scale_x_continuous(trans = "log2") +
  guides(col = guide_legend(nrow = 3))

TMB_TIL_quadrant_trial_plot <- metastatic_wes_rna_features_df_pre %>% 
  ggplot(aes(x = TMB, y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() + facet_wrap(~Response) +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed")  +
  scale_x_continuous(trans = "log2") +
  guides(col = guide_legend(nrow = 3))

TMB_TIL_quadrant_trial_plot

```

# metastatic GLMER model fits plus pancancer signature
```{r}

table(metastatic_wes_rna_features_df_pre$Trial)
table(metastatic_wes_rna_features_df_pre$Trial,
      metastatic_wes_rna_features_df_pre$CANCER_TYPE)

metastatic_response_score_model <- glmer(
  Response_n ~ pancancer_response_score + (1 | Trial),
  data = metastatic_wes_rna_features_df_pre, family = binomial)
summary(metastatic_response_score_model)
metastatic_response_score_model
```

```{r}
table(metastatic_wes_rna_features_df_pre$Study,
      metastatic_wes_rna_features_df_pre$Response)

metastatic_model1 <- glmer(Response_n ~ TMB_scaled * TIL_score_scaled + (1 | Trial),
                data = metastatic_wes_rna_features_df_pre %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
                  filter(!is.na(TMB)),
                family = binomial)

metastatic_model2 <- glmer(Response_n ~ TMB_scaled * TIL_score_scaled + pancancer_response_score_scaled + (1 | Trial),
                data = metastatic_wes_rna_features_df_pre %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
                  filter(!is.na(TMB)),
                family = binomial)

summary(metastatic_model1)
summary(metastatic_model2)
anova(metastatic_model1, metastatic_model2)

table(metastatic_wes_rna_features_df_pre$Study)
dim(metastatic_wes_rna_features_df_pre)

metastatic_model3 <- glmer(Response_n ~ Proliferation_scaled * TIL_score_scaled + (1 | Trial),
                data = metastatic_wes_rna_features_df_pre %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)),
                family = binomial)

metastatic_model4 <- glmer(Response_n ~ Proliferation_scaled * TIL_score_scaled + pancancer_response_score_scaled + (1 | Trial),
                data = metastatic_wes_rna_features_df_pre %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) ,
                family = binomial)

summary(metastatic_model3)
summary(metastatic_model4)
anova(metastatic_model3, metastatic_model4)


metastatic_model5 <- glmer(Response_n ~ Proliferation_scaled + pancancer_response_score_scaled + (1 | Trial),
                data = metastatic_wes_rna_features_df_pre %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)),
                family = binomial)

metastatic_model6 <- glmer(Response_n ~ pancancer_response_score_scaled + (1 | Trial),
                data = metastatic_wes_rna_features_df_pre %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) ,
                family = binomial)

metastatic_model7 <- glmer(Response_n ~ TIL_score_scaled + (1 | Trial),
                data = metastatic_wes_rna_features_df_pre %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) ,
                family = binomial)

summary(metastatic_model5)
summary(metastatic_model6)
summary(metastatic_model7)

anova(metastatic_model5, metastatic_model6)
anova(metastatic_model7, metastatic_model6)

```


```{r}
get_multivariate_auc_models <- FALSE

if (get_multivariate_auc_models) {
  metastatic_model_auc_data <- data.frame(
    matrix(nrow = 0, ncol = 2))
  
  metastatic_model_predictions_combined <- data.frame(
    matrix(nrow = 0, ncol = 2))
  
  model_formulas <- c(
    "Response_n ~ TMB_scaled",
    "Response_n ~ TIL_score_scaled",
    "Response_n ~ Proliferation_scaled",
    "Response_n ~ TMB_scaled * TIL_score_scaled",
    "Response_n ~ TMB_scaled * TIL_score_scaled + pancancer_response_score_scaled",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled + pancancer_response_score_scaled",
    "Response_n ~ pancancer_response_score_scaled")
    
  for (current_formula in model_formulas) {
    
    print(current_formula)
    model <- glm(as.formula(current_formula), family = "binomial",
               data = metastatic_wes_rna_features_df_pre)
  
    print(summary(model))
  
    measure <- measureit(score =  model$fitted.values, class = model$y,
                         measure = c("ACC", "SENS", "SPEC", "FSCR"))
  
    predicted <- predict(model, metastatic_wes_rna_features_df_pre %>%
                           column_to_rownames("RNA_SAMPLE_ID"),
                         type = "response")
  
    roc_empirical <- rocit(score = model$fitted.values, class = model$y)
    print(summary(roc_empirical))
  
    AUC <- summary(roc_empirical)[4, 1, drop = TRUE]
    AUC <- str_split(AUC, ":")[[1]][2]
    print(plot(roc_empirical, values = FALSE))
  
    model_auc <- tibble(
      formula = current_formula,
      AUC = AUC,
      sensitivity = sum(model$fitted.values >= 0.5 & model$y == 1) / sum(model$y == 1), # Evaluated at 0.5, not youden optimal point
      specificity = sum(model$fitted.values < 0.5 & model$y == 0) / sum(model$y == 0)) # Evaluated at 0.5, not youden optimal point
    
    model_predictions <- as.data.frame(predicted) %>%
      setNames("Predictions") %>%
      rownames_to_column("CF_sample") %>%
      mutate(formula = current_formula)
    
    metastatic_model_auc_data <- metastatic_model_auc_data %>%
      rbind(model_auc)
    
    metastatic_model_predictions_combined <- metastatic_model_predictions_combined %>%
      rbind(model_predictions)
    }

  ranef_model_formulas <- c(
    "Response_n ~ (1 | Trial)",
    "Response_n ~ TMB_scaled + (1 | Trial)",
    "Response_n ~ TIL_score_scaled + (1 | Trial)",
    "Response_n ~ Proliferation_scaled + (1 | Trial)",
    "Response_n ~ TMB_scaled * TIL_score_scaled + (1 | Trial)",
    "Response_n ~ TMB_scaled * TIL_score_scaled + pancancer_response_score_scaled + (1 | Trial)",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled + (1 | Trial)",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled + pancancer_response_score_scaled + (1 | Trial)",
    "Response_n ~ pancancer_response_score_scaled + (1 | Trial)")
    
  for (current_formula_covariates in ranef_model_formulas) {
    
    print(current_formula_covariates)

    # Glmer
    mixed_model <- glmer(as.formula(current_formula_covariates),
                         family = "binomial",
               data = metastatic_wes_rna_features_df_pre %>%
                 column_to_rownames("RNA_SAMPLE_ID"))
  
    print(summary(mixed_model))
    coefs <- fixef(mixed_model)
    coef_table <-summary(mixed_model)$coefficients
    print(coef_table)
    
    if(str_detect(current_formula_covariates, "\\|")) {
      print(ranef(mixed_model)) 
    }
    
    fitted_vals <- fitted(mixed_model)
    pred_class <- as.integer(fitted_vals >= 0.5)
    true_class <- mixed_model@resp$y
    
    plot <- data.frame(true_class, fitted_vals) %>%
      ggplot(aes(x = as.factor(true_class), y = fitted_vals)) + geom_boxplot() +
      labs(title = sprintf("Covariate model %s", current_gene_signature)) +
      theme_Publication()
    
    print(plot)
    
    roc_empirical <- rocit(score = fitted_vals, class = true_class)
    print(summary(roc_empirical))
    
    AUC <- summary(roc_empirical)[4, 1, drop = TRUE]
    AUC <- str_split(AUC, ":")[[1]][2]
    print(plot(roc_empirical, values = FALSE))
  
    mixed_model_auc <- tibble(
      formula = current_formula_covariates,
      AUC = AUC,
      sensitivity = sum(pred_class == 1 & true_class == 1) / sum(true_class == 1), # Evaluated at 0.5, not youden optimal point
      specificity = sum(pred_class == 0 & true_class == 0) / sum(true_class == 0) # Evaluated at 0.5, not youden optimal point)
      ) 
  
    model_predictions <- as.data.frame(fitted_vals) %>%
      setNames("Predictions") %>%
      rownames_to_column("CF_sample") %>%
      mutate(formula = current_formula_covariates)
    
    mixed_effects_covariate_auc_data <- mixed_effects_covariate_auc_data %>%
      rbind(mixed_model_auc)
  
    metastatic_model_auc_data <- metastatic_model_auc_data %>%
      rbind(mixed_model_auc)
    
    metastatic_model_predictions_combined <- metastatic_model_predictions_combined %>%
      rbind(model_predictions)
  }
  
  metastatic_model_auc_data %>%
    write_csv2(here(external_analysis_dir,
                    "metastatic_model_auc_data.csv"))
  
  metastatic_model_predictions_combined %>%
    write_csv2(here(external_analysis_dir,
                    "metastatic_model_predictions_combined.csv"))

}

metastatic_model_auc_data <- read_csv2(
    here(external_analysis_dir, "metastatic_model_auc_data.csv"))

metastatic_model_predictions_combined <- read_csv2(
    here(external_analysis_dir, "metastatic_model_predictions_combined.csv"))

metastatic_model_auc_data %>% arrange(AUC)

metastatic_model_auc_data %>%
  mutate(AUC = as.numeric(AUC)) %>%
  arrange(AUC) %>%
  mutate(AUC = round(AUC, 2),
         sensitivity = round(sensitivity, 2),
         specificity = round(specificity, 2))

```

```{r, fig.height = 14, fig.width = 16}
metastatic_model_predictions_combined$formula %>% unique()

metastatic_model_predictions_combined %>%
  left_join(metastatic_wes_rna_features_df_pre %>%
              mutate(CF_sample = RNA_SAMPLE_ID)) %>%
  ggplot(aes(x = Response, y = Predictions)) +
  geom_hline(yintercept = 0.5, linetype = "dashed") + facet_wrap(~formula) +
  geom_boxplot() + stat_compare_means(method = "wilcox")
```

# Export metastatic
```{r}
metastatic_wes_rna_features_df_pre %>%
  write_csv2(here(output_dir, "metastatic_wes_rna_features_df_pre"))
```

# -----------------------------

# Neoadjuvant cohorts

# ABACUS
```{r}
abacus_dir <- "/shared/data/niche/abacus"
list.files(abacus_dir)
ABACUS_survival_data <- read_csv(here(abacus_dir, "ABACUS_survival_data.csv"))
table(ABACUS_survival_data$PCR, ABACUS_survival_data$OUTCOME)

ABACUS_tmb_data <- read_csv(here(abacus_dir,
                                 "anon_abacus.fmi.data.csv"))
ABACUS_metadata <- read_csv(here(abacus_dir,
                                 "anon_abacus.RNAseq.pdata.public.csv"))
ABACUS_log_tpm <- read_csv(here(abacus_dir,
                                "anon_abacus.RNAseq.TPM.log2.148.30727.csv"))

```

```{r}
table(ABACUS_survival_data$subjectId %in% ABACUS_metadata$subjectId)
table(unique(ABACUS_metadata$subjectId) %in% ABACUS_survival_data$subjectId)

colnames(ABACUS_log_tpm)
n_distinct(ABACUS_metadata$subjectId)

ABACUS_metadata %>%
  select(subjectId, VISIT, alias) %>%
  pivot_wider(names_from = VISIT, values_from = alias) %>%
  view()

dim(ABACUS_log_tpm)

ABACUS_metadata %>% factor_summary()
ABACUS_metadata_pre <- ABACUS_metadata %>%
  filter(VISIT == "PRE")

ABACUS_metadata_pre_f <- ABACUS_metadata_pre %>%
  filter(alias %in% colnames(ABACUS_log_tpm))

ABACUS_metadata_pre_f %>% factor_summary()

table(colnames(ABACUS_log_tpm) %in% ABACUS_metadata$alias)
table(colnames(ABACUS_log_tpm) %in% ABACUS_metadata_pre$alias)
```


```{r}
ABACUS_log_tpm$Gene %>% head()
abacus_gene_ids <- gsub("GeneID:", "", ABACUS_log_tpm$Gene)
abacus_gene_ids <- as.character(abacus_gene_ids)

abacus_gene_symbols <- mapIds(
  org.Hs.eg.db,
  keys = abacus_gene_ids,
  column = "SYMBOL",
  keytype = "ENTREZID",
  multiVals = "first"
)
table(is.na(abacus_gene_symbols))

ABACUS_log_tpm$Symbol <- abacus_gene_symbols
table(ABACUS_log_tpm$Symbol)%>% as.data.frame() %>% filter(Freq > 1)

ABACUS_log_tpm_m <- ABACUS_log_tpm %>%
  select(-Gene) %>%
  filter(!is.na(Symbol)) %>%
  column_to_rownames("Symbol") %>%
  mutate_all(as.numeric) %>%
  as.matrix()

# Impute with 0
ABACUS_log_tpm_m[is.na(ABACUS_log_tpm_m)] <- 0
```

## ssGSEA
```{r}
run_ssgsea <- FALSE
if (run_ssgsea) {
   
  ABACUS_tpm_ssgsea_results <- gsva(
    as.matrix(ABACUS_log_tpm_m),
    gene_signatures_l,
    method = "ssgsea",
    kcdf = "Gaussian",
    # Minimum gene set size
    min.sz = 1,
    # Compute Gaussian-distributed scores
    mx.diff = TRUE,
    # Don't print out the progress bar
    verbose = FALSE
    )
  
  ABACUS_tpm_ssgsea_results %>%
    as.data.frame() %>%
    rownames_to_column("Signature") %>%
    write_csv2(here(output_dir, "abacus_tpm_ssgsea_results.csv"))
  
}

ABACUS_tpm_ssgsea_results <- read_csv2(
  here(output_dir, "abacus_tpm_ssgsea_results.csv"))

table(colnames(ABACUS_tpm_ssgsea_results) %in% ABACUS_metadata_pre$alias)
table(colnames(ABACUS_tpm_ssgsea_results) %in% ABACUS_metadata_pre$subjectId)

ABACUS_tpm_ssgsea_results_df <- ABACUS_tpm_ssgsea_results %>%
  column_to_rownames("Signature") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("RNA_SAMPLE_ID") %>%
  left_join(ABACUS_metadata_pre %>%
              mutate(RNA_SAMPLE_ID = alias)) %>%
  mutate(Study = "ABACUS") %>%
  filter(!is.na(subjectId)) %>%
  mutate(Response = ifelse(PCR == "Yes", "R", "NR")) %>%
    mutate(pancancer_response_score =
             pancancer_response_sig_1.25 - pancan_nonresponse_signature_1.25)
```

```{r}
ABACUS_metadata_pre_f %>% filter(!is.na(TMB)) %>% factor_summary()

ABACUS_metadata_pre_f %>% filter(!is.na(TMB)) %>%
  ggplot(aes(x = PCR, y = TMB, fill = PCR)) +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "ABACUS TMB") +
  scale_fill_manual(values = c(Color_NR, Color_R))

```

```{r}
ABACUS_tpm_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = Tumor_proliferation_rate, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "ABACUS ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

ABACUS_tpm_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = TIL_score, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "ABACUS ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

ABACUS_tpm_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_sig_1.25, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "ABACUS ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

ABACUS_tpm_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancan_nonresponse_signature_1.25,
             fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "ABACUS ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

ABACUS_tpm_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "ABACUS ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

ABACUS_tpm_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = B_catenin_signature_Luke_et_al,
             fill = Response)) +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "ABACUS ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

```

# -----------------------------

# I-SPY2
```{r}
# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE196096
# install.packages("GEOquery")
ISPY2_gse <- getGEO("GSE196096", GSEMatrix = TRUE)[[1]]
ISPY2_metadata <- pData(ISPY2_gse)

ISPY2_metadata %>% factor_summary()

ISPY2_metadata_f <- ISPY2_metadata %>%
  filter(`arm:ch1` == "Paclitaxel + Pembrolizumab")
ISPY2_metadata_f %>% factor_summary()

table(ISPY2_metadata_f$characteristics_ch1.2,
      ISPY2_metadata_f$characteristics_ch1.3)

ISPY2_RNA_df <- read.delim(
  gzfile(
    "/shared/data/niche/ISPY2/GSE194040_ISPY2ResID_AgilentGeneExp_990_FrshFrzn_meanCol_geneLevel_n988.txt.gz")
  )
dim(ISPY2_RNA_df)

# Impute missing with 0
ISPY2_RNA_df[is.na(ISPY2_RNA_df)] <- 0
table(ISPY2_metadata$title %in% gsub("X", "ISPY2_", colnames(ISPY2_RNA_df)))

```

## ssGSEA
```{r}
run_ssgsea <- FALSE
if (run_ssgsea) {
   
  ISPY2_ssgsea_results <- gsva(
    as.matrix(ISPY2_RNA_df),
    gene_signatures_l,
    method = "ssgsea",
    kcdf = "Gaussian",
    # Minimum gene set size
    min.sz = 1,
    # Compute Gaussian-distributed scores
    mx.diff = TRUE,
    # Don't print out the progress bar
    verbose = FALSE
    )
  
  ISPY2_ssgsea_results %>%
    as.data.frame() %>%
    rownames_to_column("Signature") %>%
    write_csv2(here(output_dir, "ISPY2_ssgsea_results.csv"))
  
}

ISPY2_ssgsea_results <- read_csv2(
  here(output_dir, "ISPY2_ssgsea_results.csv"))

table(colnames(ISPY2_ssgsea_results) %in% ISPY2_metadata$RNA_SAMPLE_ID)

ISPY2_metadata$RNA_SAMPLE_ID <- gsub("ISPY2_", "X", ISPY2_metadata$title)
table(ISPY2_metadata$title %in% gsub("X", "ISPY2_", colnames(ISPY2_RNA_df)))

ISPY2_ssgsea_results_df <- ISPY2_ssgsea_results %>%
  column_to_rownames("Signature") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("RNA_SAMPLE_ID") %>%
  left_join(ISPY2_metadata, by = "RNA_SAMPLE_ID") %>%
  filter(`arm:ch1` == "Paclitaxel + Pembrolizumab") %>%
  mutate(Response = characteristics_ch1.5) %>%
  mutate(Response = ifelse(Response == "pcr: 1", "R", Response)) %>%
  mutate(Response = ifelse(Response == "pcr: 0", "NR", Response)) %>%
  mutate(Study = "ISPY2") %>%
    mutate(pancancer_response_score =
             pancancer_response_sig_1.25 - pancan_nonresponse_signature_1.25)
```


```{r}
dim(ISPY2_ssgsea_results_df)
table(ISPY2_ssgsea_results_df$Response)

ISPY2_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = Tumor_proliferation_rate, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "ISPY2 ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

ISPY2_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = TIL_score, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "ISPY2 ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

ISPY2_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_sig_1.25, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "ISPY2 ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

ISPY2_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancan_nonresponse_signature_1.25,
             fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "ISPY2 ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

ISPY2_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "ISPY2 ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))
```

# -----------------------------

# PANDA
```{r}
panda_sample_metadata <- read_csv2(
  here(here("/shared/data/chalabi_group/PANDA/NKI_PANDA/data/processed/PANDA_samples.csv")))
panda_sample_metadata_pre <- panda_sample_metadata %>%
  filter(Timepoint == "PND-1")
table(panda_sample_metadata_pre$Response_PA)
table(panda_sample_metadata_pre$Response)

panda_metadata <- read_excel(
  here(here("/shared/data/chalabi_group/PANDA/NKI_PANDA/data/external/PND_overview data seq_YV.xlsx")))

panda_counts <- read_csv2(
  here("/shared/data/chalabi_group/PANDA/NKI_PANDA/data/processed/rnaseq_data/rnaseq_counts_p.csv"))
dim(panda_counts)

table(panda_sample_metadata_pre$CFSample %in% colnames(panda_counts))
panda_sample_metadata_pre$CFSample[
  !panda_sample_metadata_pre$CFSample %in% colnames(panda_counts)]

panda_sample_metadata_pre <- panda_sample_metadata_pre %>%
  filter(CFSample %in% colnames(panda_counts))

panda_counts_pre <- panda_counts %>%
  select("gene", all_of(panda_sample_metadata_pre$CFSample))

dim(panda_counts_pre)

panda_counts_m <- panda_counts_pre %>%
  column_to_rownames("gene") %>%
  as.matrix()

# Create a `DESeqDataSet` object
panda_dds <- DESeqDataSetFromMatrix(
  countData = panda_counts_m,
  colData = panda_sample_metadata_pre,
  design = ~1
)

# Normalize and transform the data in the `DESeqDataSet` object using `vst()`
panda_dds_norm <- vst(panda_dds)
panda_vst_data <- assay(panda_dds_norm)

run_ssgsea <- FALSE
if (run_ssgsea) {
  # Run ssGSEA
  
  PANDA_ssgsea_results <- gsva(
    as.matrix(panda_vst_data),
    gene_signatures_l,
    method = "ssgsea",
    # Appropriate for vst transformed data
    kcdf = "Gaussian",
    # Minimum gene set size
    min.sz = 1,
    # Compute Gaussian-distributed scores
    mx.diff = TRUE,
    # Don't print out the progress bar
    verbose = FALSE
    )
  
  PANDA_ssgsea_results %>%
    as.data.frame() %>%
    rownames_to_column("Signature") %>%
    write_csv2(here(output_dir, "PANDA_ssgsea_results.csv"))
  
}

PANDA_ssgsea_results <- read_csv2(
  here(output_dir, "PANDA_ssgsea_results.csv"))

PANDA_ssgsea_results_df <- PANDA_ssgsea_results %>%
  column_to_rownames("Signature") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("RNA_SAMPLE_ID") %>%
  left_join(panda_sample_metadata_pre %>%
              mutate(RNA_SAMPLE_ID = CFSample), by = "RNA_SAMPLE_ID") %>%
  mutate(Response = ifelse(Response %in% c("Responder", "R"), "R", "NR")) %>%
  mutate(Study = "PANDA") %>%
  filter(Timepoint == "PND-1") %>%
      mutate(pancancer_response_score =
               pancancer_response_sig_1.25 - pancan_nonresponse_signature_1.25)

dim(PANDA_ssgsea_results_df)
table(PANDA_ssgsea_results_df$Response)

panda_tmb <- read_csv2(
  "/shared/data/niche/PANDA_vcf_n_variants_nonsyn_frameshift.csv")

PANDA_TMB_ssgsea_results_df <- PANDA_ssgsea_results_df %>%
  mutate(Patient = paste0("PND_", TrialNumber)) %>%
  left_join(panda_tmb %>%
  select(Patient, TMB, CF_sample))

table(PANDA_TMB_ssgsea_results_df$Patient)
table(PANDA_TMB_ssgsea_results_df$TMB, useNA = "ifany")

```

```{r}
PANDA_TMB_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = TMB,
             fill = Response)) +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "PANDA TMB") +
  scale_fill_manual(values = c(Color_NR, Color_R))
```

```{r}
PANDA_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = Tumor_proliferation_rate,
             fill = Response)) +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "PANDA ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

PANDA_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = H_E2F_Targets, fill = Response)) +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "PANDA ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

PANDA_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = TIL_score, fill = Response)) +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "PANDA ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

PANDA_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_sig_1.25,
             fill = Response)) +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "PANDA ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

PANDA_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancan_nonresponse_signature_1.25,
             fill = Response)) +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "PANDA ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

PANDA_ssgsea_results_df %>%
  ggplot(aes(x = Response, y = pancancer_response_score,
             fill = Response)) +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "PANDA ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

PANDA_ssgsea_results_df %>%
  ggplot(aes(x = Response_PA, y = pancancer_response_score,
             fill = Response)) +
  geom_boxplot() + theme_Publication() +
  geom_point() + 
  labs(title = "PANDA (nICB+chemo G/GEJ)",
       y = "NKI pancancer signature\nssGSEA score delta") +
  scale_fill_manual(values = c(Color_NR, Color_R))

PANDA_ssgsea_results_df %>%
  mutate(Response_pCR = ifelse(Response_PA %in% c("CR"), "CR", "non-CR")) %>%
  ggplot(aes(x = Response_pCR, y = pancancer_response_score,
             fill = Response_pCR)) +
  stat_compare_means(method = "wilcox") +
  geom_boxplot() + theme_Publication() +
  geom_point() + 
  labs(title = "PANDA (nICB+chemo G/GEJ)",
       y = "NKI pancancer signature\nssGSEA score delta") +
  scale_fill_manual(values = c(Color_R, Color_NR))
```
# -----------------------------

# Neoadjuvant combined RNA
```{r}
neoadjuvant_ssgsea_c <- PANDA_ssgsea_results_df %>%
  mutate(Response = ifelse(
    Study == "PANDA" & Response_PA %in% c("CR"), "R", "NR")) %>%
  select(all_of(c("RNA_SAMPLE_ID", "Response", "Study",
                  names(gene_signatures_l)))) %>%
  rbind(
    ABACUS_tpm_ssgsea_results_df %>%
          mutate(Gamma_delta_TCR_Yu = 0, Gamma_delta_TCRs = 0) %>%
      select(all_of(c("RNA_SAMPLE_ID", "Response", "Study",
                      names(gene_signatures_l))))) %>%
  rbind(
    ISPY2_ssgsea_results_df %>%
      select(all_of(c("RNA_SAMPLE_ID", "Response", "Study",
                      names(gene_signatures_l))))) %>%
  mutate(Response_n = ifelse(Response == "R", 1, 0),
    Trial = Study) %>%
  mutate(
    TIL_score_scaled = scale(TIL_score),
    Proliferation_scaled = scale(Tumor_proliferation_rate),
    B_cat_scaled = scale(B_catenin_signature_Luke_et_al)) %>%
      mutate(pancancer_response_score =
               pancancer_response_sig_1.25 - pancan_nonresponse_signature_1.25)

neoadjuvant_ssgsea_c$TIL_score_scaled <- as.vector(
  neoadjuvant_ssgsea_c$TIL_score_scaled)
neoadjuvant_ssgsea_c$Proliferation_scaled <- as.vector(
  neoadjuvant_ssgsea_c$Proliferation_scaled)
neoadjuvant_ssgsea_c$B_cat_scaled <- as.vector(
  neoadjuvant_ssgsea_c$B_cat_scaled)

table(neoadjuvant_ssgsea_c$Response,
      neoadjuvant_ssgsea_c$Study)

neoadjuvant_ssgsea_c %>%
  write_csv2(here(output_dir, "neoadjuvant_ssgsea_c.csv"))

```

## Define TMB + ssgsea

```{r}
ABACUS_TMB_tpm_ssgsea_results_df <- ABACUS_tpm_ssgsea_results_df %>%
  left_join(ABACUS_metadata_pre_f %>%
              select(subjectId, TMB)) %>%
  filter(!is.na(TMB))

neoadjuvant_TMB_ssgsea_c <- PANDA_TMB_ssgsea_results_df %>%
  select(all_of(c("RNA_SAMPLE_ID", "Response", "Study", "TMB",
                  names(gene_signatures_l)))) %>%
  full_join(
    ABACUS_TMB_tpm_ssgsea_results_df %>%
      mutate(Gamma_delta_TCR_Yu = 0, Gamma_delta_TCRs = 0) %>%
      select(all_of(c("RNA_SAMPLE_ID", "Response", "Study", "TMB",
                      names(gene_signatures_l))))
    )  %>%
  mutate(Response_n = ifelse(
    Response == "R",
    1, ifelse(
      Response == "NR", 0, NA))) %>%
  mutate(
    TMB_scaled = scale(TMB),
    TIL_score_scaled = scale(TIL_score),
    Proliferation_scaled = scale(Tumor_proliferation_rate),
    B_cat_scaled = scale(B_catenin_signature_Luke_et_al),
      Trial = Study) %>%
    mutate(pancancer_response_score =
             pancancer_response_sig_1.25 - pancan_nonresponse_signature_1.25)

table(neoadjuvant_TMB_ssgsea_c$Study, neoadjuvant_TMB_ssgsea_c$Response)
```


```{r}

neoadjuvant_TMB_ssgsea_c %>% filter(!is.na(TMB)) %>%
  ggplot(aes(x = Response, y = TMB, fill = Response)) +
  geom_boxplot() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "Neoadjuvant cohort (PANDA+ABACUS) TMB") +
  scale_fill_manual(values = c(Color_NR, Color_R))
```

```{r}
neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Response, y = Tumor_proliferation_rate, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "Neoadjuvant cohort ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Response, y = H_E2F_Targets, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "Neoadjuvant cohort ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Response, y = TIL_score, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "Neoadjuvant cohort ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Response, y = B_catenin_signature_Luke_et_al,
             fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "Neoadjuvant cohort ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))

```


```{r}
table(neoadjuvant_ssgsea_c$Study)

neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Response, y = pancancer_response_sig_1.25, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "Neoadjuvant cohort ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Response, y = pancan_nonresponse_signature_1.25, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "Neoadjuvant cohort ssGSEA") +
  scale_fill_manual(values = c(Color_NR, Color_R))+
  guides(fill = FALSE)

neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "Neoadjuvant cohort ssGSEA\n(complete ABACUS + ISPY2 + PANDA)") +
  scale_fill_manual(values = c(Color_NR, Color_R))+
  guides(fill = FALSE)

neoadjuvant_ssgsea_c %>%
  filter(!Study %in% "PANDA") %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + theme_Publication() + stat_compare_means(method = "wilcox") +
  labs(title = "Neoadjuvant cohort ssGSEA\n(complete ABACUS + ISPY2)") +
  scale_fill_manual(values = c(Color_NR, Color_R))+
  guides(fill = FALSE)

```

## Prolif + TIL

```{r}
Trial_model <- glmer(Response_n ~ (1 | Trial),
                     data = neoadjuvant_ssgsea_c, family = binomial)
summary(Trial_model)

median_Tumor_proliferation_rate_pre <- neoadjuvant_ssgsea_c %>%
  pull(Tumor_proliferation_rate) %>%
  median()
median_Tumor_proliferation_rate_pre

median_til_score_pre <- median(
  neoadjuvant_ssgsea_c$TIL_score)
median_til_score_pre

prolif_TIL_quadrant_trial_plot <- neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Study)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = median_Tumor_proliferation_rate_pre,
             linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  guides(col = guide_legend(nrow = 3)) + facet_wrap(~Response)

prolif_TIL_quadrant_trial_plot

trial_quadrant_proportions <- table(
  neoadjuvant_ssgsea_c$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre, 
  neoadjuvant_ssgsea_c$TIL_score > median_til_score_pre,
  neoadjuvant_ssgsea_c$Study) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
dim(neoadjuvant_ssgsea_c)

table(neoadjuvant_ssgsea_c$Response)

quadrant_response_proportions <- table(
  neoadjuvant_ssgsea_c$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  neoadjuvant_ssgsea_c$TIL_score > median_til_score_pre,
  neoadjuvant_ssgsea_c$Response) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Response", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions

quadrant_response_proportions %>%
  select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()

quadrant_response_proportions %>%
  filter(Response == "R") %>%
  rename(Response_rate = quadrant_prop) %>%
  ggplot(aes(x = quadrant, y = Response_rate)) + geom_col() +
  theme_Publication() + labs(x = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

trial_quadrant_proportions <- table(
  neoadjuvant_ssgsea_c$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  neoadjuvant_ssgsea_c$TIL_score > median_til_score_pre,
  neoadjuvant_ssgsea_c$Study) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
trial_quadrant_proportions

trial_quadrant_proportions %>% select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

neoadjuvant_ssgsea_c %>% 
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Study)) +
  geom_point() + theme_Publication() + facet_wrap(~Response) 

```

```{r}
Proliferation_model <- glmer(
  Response_n ~ Proliferation_scaled + (1 | Trial),
  data = neoadjuvant_ssgsea_c, family = binomial)
summary(Proliferation_model)

TIL_score_model <- glmer(
  Response_n ~ TIL_score_scaled + (1 | Trial),
  data = neoadjuvant_ssgsea_c, family = binomial)
summary(TIL_score_model)

Proliferation_TIL_score_model <- glmer(
  Response_n ~ Proliferation_scaled + TIL_score_scaled + (1 | Trial),
  data = neoadjuvant_ssgsea_c, family = binomial)
summary(Proliferation_TIL_score_model)

ranef(Proliferation_TIL_score_model)
summary(Proliferation_TIL_score_model)$coefficients %>% round(2)
summary(Proliferation_TIL_score_model)$coefficients[, 4] %>% round(4)

Proliferation_TIL_int_score_model <- glmer(
  Response_n ~ Proliferation_scaled * TIL_score_scaled + (1 | Trial),
  data = neoadjuvant_ssgsea_c, family = binomial)
summary(Proliferation_TIL_int_score_model)
ranef(Proliferation_TIL_int_score_model)

```


## TMB + TIL

```{r, fig.height = 6, fig.width = 6}
percentage_above_TMB_10 <- neoadjuvant_TMB_ssgsea_c %>%
  group_by(Trial) %>%
  summarise(proportion = mean(TMB > 10)) %>%
  mutate(percentage = round(proportion * 100))
percentage_above_TMB_10

tmb_trial_plot <- neoadjuvant_TMB_ssgsea_c %>%
  ggplot(aes(x = factor(Trial),
             y = log2(TMB), fill = Response)) +
  geom_boxplot() + theme_Publication() +
  # guides(fill = FALSE) +
  labs(x = "", y = "log2(TMB) (mutations/Mb)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  geom_hline(yintercept = log2(10), linetype = "dashed", alpha = 0.5) +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.2))) +
  geom_text(data = percentage_above_TMB_10 %>%
              mutate(Response = "NR"),
            aes(x = Trial, y = log2(400), label = paste0(percentage, "%")),
            vjust = -1.5) +
  scale_fill_manual(values = c(Color_NR, Color_R))

tmb_trial_plot

```

```{r}
table(neoadjuvant_TMB_ssgsea_c$Trial)
TMB_TIL_score_model <- glmer(
  Response_n ~ TMB_scaled + TIL_score_scaled + (1 | Trial),
  data = neoadjuvant_TMB_ssgsea_c, family = binomial)
summary(TMB_TIL_score_model)

ranef(TMB_TIL_score_model)
summary(TMB_TIL_score_model)$coefficients %>% round(2)
summary(TMB_TIL_score_model)$coefficients[, 4] %>% round(4)

# median_til_score_pre <- median(
#   neoadjuvant_TMB_ssgsea_c$TIL_score)
# median_til_score_pre

# Quadrant RR
table(neoadjuvant_TMB_ssgsea_c$TMB >= 10,
      neoadjuvant_TMB_ssgsea_c$TIL_score >= median_til_score_pre) %>%
  `rownames<-`(c("TMB-L", "TMB-H")) %>%
  `colnames<-`( c("mTIL_low", "mTIL_high"))

neoadjuvant_TMB_ssgsea_c %>%
  ggplot(aes(x = log2(TMB), y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = log2(10), linetype = "dashed")

# # Quadrant RR
table(neoadjuvant_TMB_ssgsea_c$TMB >= 10,
      neoadjuvant_TMB_ssgsea_c$TIL_score >= median_til_score_pre) %>%
  `rownames<-`(c("TMB-L", "TMB-H")) %>%
  `colnames<-`( c("mTIL_low", "mTIL_high"))

quadrant_response_proportions <- table(
  neoadjuvant_TMB_ssgsea_c$TMB >= 10,
  neoadjuvant_TMB_ssgsea_c$TIL_score > median_til_score_pre,
  neoadjuvant_TMB_ssgsea_c$Response) %>%
  as.data.frame() %>%
  setNames(c("TMB", "mTIL_group", "Response", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB == TRUE, "TMB_high", "TMB_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions

quadrant_response_proportions %>%
  filter(Response == "R") %>%
  rename(Response_rate = quadrant_prop) %>%
  ggplot(aes(x = quadrant, y = Response_rate)) + geom_col() +
  theme_Publication() + labs(x = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

quadrant_response_proportions

quadrant_response_proportions %>%
  select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()

trial_quadrant_proportions <- table(
  neoadjuvant_TMB_ssgsea_c$TMB >= 10,
  neoadjuvant_TMB_ssgsea_c$TIL_score > median_til_score_pre,
  neoadjuvant_TMB_ssgsea_c$Trial) %>%
  as.data.frame() %>%
  setNames(c("TMB_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB_group == TRUE, "TMB-H", "TMB-L")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)

trial_quadrant_proportions %>%
  select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

quadrant_trial_proportions <- table(
  neoadjuvant_TMB_ssgsea_c$TMB >= 10,
  neoadjuvant_TMB_ssgsea_c$TIL_score > median_til_score_pre,
  neoadjuvant_TMB_ssgsea_c$Trial) %>%
  as.data.frame() %>%
  setNames(c("TMB_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB_group == TRUE, "TMB-H", "TMB-L")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group)) %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = Freq / Total)

TMB_TIL_quadrant_trial_proportions_plot <- quadrant_trial_proportions %>%
  ggplot(aes(x = quadrant, y = quadrant_prop, fill = Trial)) + geom_col() +
  scale_fill_manual(values = trial_colors) +
  theme_Publication() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 3)) +
  coord_flip()

TMB_TIL_quadrant_trial_proportions_plot

```

```{r}

neoadjuvant_TMB_ssgsea_c %>%
  ggplot(aes(x = TMB, y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  scale_x_continuous(trans = "log2") +
  guides(col = guide_legend(nrow = 3))

TMB_TIL_quadrant_trial_plot <- neoadjuvant_TMB_ssgsea_c %>%
  ggplot(aes(x = TMB, y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() + facet_wrap(~Response) +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed")  +
  scale_x_continuous(trans = "log2") +
  guides(col = guide_legend(nrow = 3))

TMB_TIL_quadrant_trial_plot

```

## Check Response Score models
```{r}
table(neoadjuvant_ssgsea_c$Trial)
table(neoadjuvant_ssgsea_c$Trial, neoadjuvant_ssgsea_c$Response)

neoadjuvant_response_score_model <- glmer(
  Response_n ~ pancancer_response_score + (1 | Trial),
  data = neoadjuvant_ssgsea_c, family = binomial)
summary(neoadjuvant_response_score_model)

Proliferation_TIL_int_score_model_adapted_panda <- glmer(
  Response_n ~ Proliferation_scaled * TIL_score_scaled + (1 | Trial),
  data = neoadjuvant_ssgsea_c, family = binomial)
summary(Proliferation_TIL_int_score_model)
ranef(Proliferation_TIL_int_score_model)

neoadjuvant_ssgsea_c %>%
  ggplot(aes(x = Response, y = pancancer_response_score, fill = Response)) +
  geom_boxplot() + theme_Publication() +
  facet_wrap(~Trial) +
  stat_compare_means(method = "wilcox") +
  guides(fill = FALSE) +
  scale_fill_manual(values = c(Color_NR, Color_R))

```

## Correlations
```{r, fig.height = 10, fig.width = 8}
# TPM
neoadjuvant_ssgsea_correlations <- neoadjuvant_ssgsea_c %>%
  select(tpm_ssgsea_results$Signature) %>% cor()

neoadjuvant_ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancancer_response_sig_1.25) %>%
  setNames(c("pancancer_response_sig_1.25")) %>%
  rownames_to_column("Signature") %>%
  filter(!is.na(pancancer_response_sig_1.25)) %>%
  filter(Signature %in% c(gene_signatures_meta_analysis$Signature,
           "pancancer_nonresponse_sig_1.25",
           "pancancer_response_sig_1.25",
           "pancancer_response_score")) %>%
  filter(!Signature %in% c("pancancer_response_sig_1.25")) %>%
  ggplot(aes(x = reorder(Signature, pancancer_response_sig_1.25),
                         y = pancancer_response_sig_1.25)) +
  geom_col() + theme_Publication() + coord_flip()

neoadjuvant_ssgsea_correlations %>%
  as.data.frame() %>%
  select(pancan_nonresponse_signature_1.25) %>%
  setNames(c("pancan_nonresponse_signature_1.25")) %>%
  rownames_to_column("Signature")  %>%
  filter(Signature %in% c(gene_signatures_meta_analysis$Signature,
           "pancan_nonresponse_signature_1.25",
           "pancancer_response_sig_1.25",
           "pancancer_response_score")) %>%
  filter(!is.na(pancan_nonresponse_signature_1.25))  %>%
  filter(!Signature %in% c("pancan_nonresponse_signature_1.25")) %>%
  ggplot(aes(x = reorder(Signature, pancan_nonresponse_signature_1.25),
                         y = pancan_nonresponse_signature_1.25)) +
  geom_col() + theme_Publication() + coord_flip()

```

## neoadjuvant GLMER model fits previous combinations plus pancancer signature
```{r}
neoadjuvant_ssgsea_prediction_df <- neoadjuvant_ssgsea_c %>%
  left_join(neoadjuvant_TMB_ssgsea_c %>%
              select(TMB, RNA_SAMPLE_ID)) %>%
  mutate(pancancer_response_score_scaled = scale(pancancer_response_score)) %>%
  mutate(TMB_scaled = scale(TMB)) 

neoadjuvant_ssgsea_prediction_df$TIL_score_scaled <- as.vector(
  neoadjuvant_ssgsea_prediction_df$TIL_score_scaled)
neoadjuvant_ssgsea_prediction_df$Proliferation_scaled <- as.vector(
  neoadjuvant_ssgsea_prediction_df$Proliferation_scaled)
neoadjuvant_ssgsea_prediction_df$TMB_scaled <- as.vector(
  neoadjuvant_ssgsea_prediction_df$TMB_scaled)
neoadjuvant_ssgsea_prediction_df$pancancer_response_score_scaled <- as.vector(
  neoadjuvant_ssgsea_prediction_df$pancancer_response_score_scaled)
```


```{r}
external_nICB_model1 <- glmer(Response_n ~ TMB_scaled * TIL_score_scaled + (1 | Trial),
                data = neoadjuvant_ssgsea_prediction_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
                  filter(!is.na(TMB)),
                family = binomial)

external_nICB_model2 <- glmer(Response_n ~ TMB_scaled * TIL_score_scaled + pancancer_response_score_scaled + (1 | Trial),
                data = neoadjuvant_ssgsea_prediction_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) %>%
                  filter(!is.na(TMB)),
                family = binomial)

summary(external_nICB_model1)
summary(external_nICB_model2)
anova(external_nICB_model1, external_nICB_model2)

table(neoadjuvant_ssgsea_prediction_df$Study)
dim(neoadjuvant_ssgsea_prediction_df)

external_nICB_model3 <- glmer(Response_n ~ Proliferation_scaled * TIL_score_scaled + (1 | Trial),
                data = neoadjuvant_ssgsea_prediction_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)),
                family = binomial)

external_nICB_model4 <- glmer(Response_n ~ Proliferation_scaled * TIL_score_scaled + pancancer_response_score_scaled + (1 | Trial),
                data = neoadjuvant_ssgsea_prediction_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) ,
                family = binomial)

summary(external_nICB_model3)
summary(external_nICB_model4)
anova(external_nICB_model3, external_nICB_model4)


external_nICB_model5 <- glmer(Response_n ~ Proliferation_scaled + pancancer_response_score_scaled + (1 | Trial),
                data = neoadjuvant_ssgsea_prediction_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)),
                family = binomial)

external_nICB_model6 <- glmer(Response_n ~ pancancer_response_score_scaled + (1 | Trial),
                data = neoadjuvant_ssgsea_prediction_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) ,
                family = binomial)

external_nICB_model7 <- glmer(Response_n ~ TIL_score_scaled + (1 | Trial),
                data = neoadjuvant_ssgsea_prediction_df %>%
                  mutate(Response_n = ifelse(Response == "R", 1, 0)) ,
                family = binomial)

summary(external_nICB_model5)
summary(external_nICB_model6)
summary(external_nICB_model7)

anova(external_nICB_model5, external_nICB_model6)
anova(external_nICB_model7, external_nICB_model6)

table(neoadjuvant_ssgsea_prediction_df$Study, neoadjuvant_ssgsea_prediction_df$Response)

neoadjuvant_ssgsea_prediction_df %>%
  ggplot(aes(x = TIL_score, y = pancancer_response_score, color = Trial)) +
  geom_point() + stat_cor()
```


```{r}
get_multivariate_auc_models <- TRUE

if (get_multivariate_auc_models) {
  external_nICB_model_auc_data <- data.frame(
    matrix(nrow = 0, ncol = 2))
  
  external_nICB_model_predictions_combined <- data.frame(
    matrix(nrow = 0, ncol = 2))
  
  model_formulas <- c(
    "Response_n ~ TMB_scaled",
    "Response_n ~ TIL_score_scaled",
    "Response_n ~ Proliferation_scaled",
    "Response_n ~ TMB_scaled * TIL_score_scaled",
    "Response_n ~ TMB_scaled * TIL_score_scaled + pancancer_response_score_scaled",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled + pancancer_response_score_scaled",
    "Response_n ~ pancancer_response_score_scaled")
    
  for (current_formula in model_formulas) {
    
    print(current_formula)
    model <- glm(as.formula(current_formula), family = "binomial",
               data = neoadjuvant_ssgsea_prediction_df)
  
    print(summary(model))
  
    measure <- measureit(score =  model$fitted.values, class = model$y,
                         measure = c("ACC", "SENS", "SPEC", "FSCR"))
  
    predicted <- predict(model, neoadjuvant_ssgsea_prediction_df %>%
                           column_to_rownames("RNA_SAMPLE_ID"),
                         type = "response")
  
    roc_empirical <- rocit(score = model$fitted.values, class = model$y)
    print(summary(roc_empirical))
  
    AUC <- summary(roc_empirical)[4, 1, drop = TRUE]
    AUC <- str_split(AUC, ":")[[1]][2]
    print(plot(roc_empirical, values = FALSE))
  
    model_auc <- tibble(
      formula = current_formula,
      AUC = AUC,
      sensitivity = sum(model$fitted.values >= 0.5 & model$y == 1) / sum(model$y == 1), # Evaluated at 0.5, not youden optimal point
      specificity = sum(model$fitted.values < 0.5 & model$y == 0) / sum(model$y == 0)) # Evaluated at 0.5, not youden optimal point
    
    model_predictions <- as.data.frame(predicted) %>%
      setNames("Predictions") %>%
      rownames_to_column("CF_sample") %>%
      mutate(formula = current_formula)
    
    external_nICB_model_auc_data <- external_nICB_model_auc_data %>%
      rbind(model_auc)
    
    external_nICB_model_predictions_combined <- external_nICB_model_predictions_combined %>%
      rbind(model_predictions)
    }

  ranef_model_formulas <- c(
    "Response_n ~ (1 | Trial)",
    "Response_n ~ TMB_scaled + (1 | Trial)",
    "Response_n ~ TIL_score_scaled + (1 | Trial)",
    "Response_n ~ Proliferation_scaled + (1 | Trial)",
    "Response_n ~ TMB_scaled * TIL_score_scaled + (1 | Trial)",
    "Response_n ~ TMB_scaled * TIL_score_scaled + pancancer_response_score_scaled + (1 | Trial)",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled + (1 | Trial)",
    "Response_n ~ Proliferation_scaled * TIL_score_scaled + pancancer_response_score_scaled + (1 | Trial)",
    "Response_n ~ pancancer_response_score_scaled + (1 | Trial)")
    
  for (current_formula_covariates in ranef_model_formulas) {
    
    print(current_formula_covariates)

    # Glmer
    mixed_model <- glmer(as.formula(current_formula_covariates), family = "binomial",
               data = neoadjuvant_ssgsea_prediction_df %>% column_to_rownames("RNA_SAMPLE_ID"))
  
    print(summary(mixed_model))
    coefs <- fixef(mixed_model)
    coef_table <-summary(mixed_model)$coefficients
    print(coef_table)
    
    if(str_detect(current_formula_covariates, "\\|")) {
      print(ranef(mixed_model)) 
    }
    
    fitted_vals <- fitted(mixed_model)
    pred_class <- as.integer(fitted_vals >= 0.5)
    true_class <- mixed_model@resp$y
    
    plot <- data.frame(true_class, fitted_vals) %>%
      ggplot(aes(x = as.factor(true_class), y = fitted_vals)) + geom_boxplot() +
      labs(title = sprintf("Covariate model %s", current_gene_signature)) +
      theme_Publication()
    
    print(plot)
    
    roc_empirical <- rocit(score = fitted_vals, class = true_class)
    print(summary(roc_empirical))
    
    AUC <- summary(roc_empirical)[4, 1, drop = TRUE]
    AUC <- str_split(AUC, ":")[[1]][2]
    print(plot(roc_empirical, values = FALSE))
  
    mixed_model_auc <- tibble(
      formula = current_formula_covariates,
      AUC = AUC,
      sensitivity = sum(pred_class == 1 & true_class == 1) / sum(true_class == 1), # Evaluated at 0.5, not youden optimal point
      specificity = sum(pred_class == 0 & true_class == 0) / sum(true_class == 0) # Evaluated at 0.5, not youden optimal point)
      ) 
  
    model_predictions <- as.data.frame(fitted_vals) %>%
      setNames("Predictions") %>%
      rownames_to_column("CF_sample") %>%
      mutate(formula = current_formula_covariates)
    
    mixed_effects_covariate_auc_data <- mixed_effects_covariate_auc_data %>%
      rbind(mixed_model_auc)
  
    external_nICB_model_auc_data <- external_nICB_model_auc_data %>%
      rbind(mixed_model_auc)
    
    external_nICB_model_predictions_combined <- external_nICB_model_predictions_combined %>%
      rbind(model_predictions)
  }
  
  external_nICB_model_auc_data %>%
    write_csv2(here(external_analysis_dir, "external_nICB_model_auc_data.csv"))
  
  external_nICB_model_predictions_combined %>%
    write_csv2(here(external_analysis_dir, "external_nICB_model_predictions_combined.csv"))
  

}

external_nICB_model_predictions_combined <- read_csv2(
  here(external_analysis_dir, "external_nICB_model_predictions_combined.csv"))

external_nICB_model_auc_data <- read_csv2(
  here(external_analysis_dir, "external_nICB_model_auc_data.csv"))
  
external_nICB_model_auc_data %>% arrange(AUC)


external_nICB_model_auc_data %>%
  mutate(AUC = as.numeric(AUC)) %>%
  arrange(AUC) %>%
  mutate(AUC = round(AUC, 2),
         sensitivity = round(sensitivity, 2),
         specificity = round(specificity, 2))
```


```{r, fig.height = 14, fig.width = 16}
external_nICB_model_predictions_combined$formula %>% unique()

external_nICB_model_predictions_combined %>%
  left_join(neoadjuvant_ssgsea_prediction_df %>%
              mutate(CF_sample = RNA_SAMPLE_ID)) %>%
  ggplot(aes(x = Response, y = Predictions)) +
  geom_hline(yintercept = 0.5, linetype = "dashed") + facet_wrap(~formula) +
  geom_boxplot() + stat_compare_means(method = "wilcox")
```

# -----------------------------

# Lint script
```{r}
library(lintr)
# lint("~/NKI_pancancer_neoadj_io/src/pancan_external_data_analysis.Rmd")
lint(
  filename = "~/NKI_pancancer_neoadj_io/src/pancan_external_data_analysis.Rmd",
  linters = unused_import_linter()
)
```

# ----------
# sessionInfo - Package Versions
```{r sessionInfo}
sessionInfo()
```
