---
title: "neoadjuvant pancancer GLMer"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  export: TRUE
  output_dir: "GLMM_analysis"
  metadata: ""
  filter_and_parse_vcfs: FALSE
  exclude_niche_msi: TRUE
  run_meta_analysis: TRUE

---

Author comment:
File description comment:
The purpose of this script is to run generalized linear mixed models (GLMM) and sensitivity analysis and adjusting for tumor type by fitting trials as random effects, also using different data subsets.

# Loading Packages
```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

library(tidyverse)
source("helper_functions.R")
source("oncomatrix.R")
library(VariantAnnotation)
library(ggpubr)
library(export)
library(lme4)
library(export)
library(lintr)

MakeDataFrameFromVCF <- getFromNamespace("MakeDataFrameFromVCF", "ICAMS")

set.seed(42)
select <- dplyr::select
rename <- dplyr::rename
here <- here::here
```

# Setup output folder

```{r Setup output, echo = FALSE, include = FALSE}
export <- params$export
filter_and_parse_vcfs <- params$filter_and_parse_vcfs

# generate output directory
wes_dir <- here("data", "processed", "WES_analysis")
metadata_dir <- here("data", "processed", "metadata")
output_dir <- here("data", "processed", params$output_dir)
wes_meta_dir <- here("data", "processed", "WES_analysis_meta")
wes_rna_dir <- here("data", "processed", "WES_RNA_integration")

print(output_dir)

# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)

    } else {
  print("Output folder already exists")
}

```

# ----------
# Load metadata

```{r Load metadata}
dna_metadata <- read_csv2(here(metadata_dir, "combined_dna_metadata.csv"))

histology_df <- data.frame(
  Trial = c("BELLINI", "IMCISION", "NABUCCO", "NICHE-MSS", "NICHE-MSI",
            "OpACIN_neo", "PRADO", "MATISSE"),
  Histology = c("TBNC", "HNSCC", "Urothelial", "Colon", "Colon",
                "LN_melanoma", "LN_melanoma", "CSCC"))

```

# Annotation color
```{r}

# Define colours for each group
Color_NR <- "#F8766D"
Color_R <- "#00BA38" # "#619CFF"
Color_NA <- "#CD9600"
Color_Pre <- "#619CFF"
Color_Post <- "#F564E3"

Color_HNSCC <- "#009292"
Color_CSCC <- "#CD9600"
Color_Urothelial <- "#B66DFF"
Color_LN_melanoma <- "#00BFC4"
Color_Colon <- "#F8766D"
Color_Colon_MSI <- "#B73A3A"
Color_Colon_MSS <- "#23C26F"
Color_TNBC <- "#FEB5DA"
Color_BELLINI <- "#FEB5DA"
Color_IMCISION <- "#009292"
Color_MATISSE <- "#CD9600"
Color_NABUCCO <- "#B66DFF"
Color_BELLINI <- "#FEB5DA"
Color_NICHE_MSS <- "#23C26F"
Color_NICHE_MSI <- "#B73A3A"
Color_OpACIN_neo <- "#00BFC4"
Color_PRADO <- "#619CFF"

annotation_color_list = list(
  Response = c(NR = Color_NR, R = Color_R),
  Treatment = c(Pre = Color_Pre, Post = Color_Post),
  Histology = c(LN_melanoma = Color_LN_melanoma, Urothelial = Color_Urothelial,
                Colon = Color_Colon, HNSCC = Color_HNSCC, TBNC = Color_TNBC,
                CSCC = Color_CSCC),
  Trial = c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO))

trial_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_colors_combined <- c(Combined = "#999999", BELLINI = Color_BELLINI,
                           IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_Colon_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

histology_msi_mss_colors <- c(`Colon-MSI`= Color_Colon_MSI,
                              `Colon-MSS`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              LN_melanoma = Color_LN_melanoma,
                              TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)

```

# ----------
# Load data

```{r}
n_variants_tmb <- read_csv2(here(wes_dir, "n_variants_dp20_vaf_tmb.csv"))

wes_features_df <- read_csv2(here(wes_dir, "wes_features_df.csv"))
wes_features_df <- wes_features_df %>%
  filter(!(Trial %in% "NICHE-MSI" & Tumor_nr == "Tumor_2"))

pathway_level_mutation_df <- read_csv2(here(wes_dir,
                                            "pathway_level_mutation_df.csv"))
pathway_level_mutation_df <- pathway_level_mutation_df %>%
  filter(!(Trial %in% "NICHE-MSI" & Tumor_nr == "Tumor_2"))

facets_data_metrics <- read_csv2(here(wes_dir, "facets_data_metrics.csv"))
facets_data_metrics <- facets_data_metrics %>%
  filter(Tumor_Sample_Barcode %in% wes_features_df$Tumor_Sample_Barcode)

facets_aneuploidy_score_df <- read_csv2(
  here(wes_dir, "facets_aneuploidy_score_df.csv"))
facets_aneuploidy_score_df <- facets_aneuploidy_score_df %>%
  filter(CF_sample_tumor %in% wes_features_df$CF_sample_tumor)

combined_wes_smd <- read_csv2(here(wes_meta_dir, "combined_wes_smd.csv"))
trial_odds_ratio_df <- read_csv2(here(wes_meta_dir, "trial_odds_ratio_df.csv"))

meta_studies_results_smd_all <- read_csv2(
  here(wes_meta_dir, "meta_studies_results_smd_all.csv"))

meta_studies_results_or_all <- read_csv2(
  here(wes_meta_dir, "meta_studies_results_or_all.csv"))

patient_metadata_combined <- read_csv2(
  here("data", "processed", "metadata", "patient_metadata_combined.csv"))

# Add patient N stage and sex information
wes_features_df <- wes_features_df %>%
  left_join(patient_metadata_combined %>%
              select(Patient, Sex, Baseline_N_h)) %>%
  mutate(Baseline_N_h = ifelse(Baseline_N_h == "Nx", "N+", Baseline_N_h))

wes_features_df <- wes_features_df %>%
  mutate(Response_n = ifelse(
    Response == "R",
    1, ifelse(
      Response == "NR", 0, NA))) %>%
  mutate(TMB_scaled = scale(TMB),
         UV_scaled = scale(UV_abs),
         APOBEC_scaled = scale(APOBEC_abs)) 

```


# ----------

```{r}
table(wes_features_df$Response)
wes_features_df$Response_n %>% mean() %>% round(2)

wes_features_df$TMB_scaled %>% summary()
mean(wes_features_df$TMB_scaled)
sd(wes_features_df$TMB_scaled)

wes_features_df_excl_dMMR <- wes_features_df %>%
  filter(!Trial %in% c("NICHE-MSI"))
wes_features_df_excl_dMMR_UV <- wes_features_df %>%
  filter(!Trial %in% c("NICHE-MSI",  "PRADO", "OpACIN_neo", "MATISSE"))

```

# Sex Stage Cohort
```{r}
wes_features_df 

# Mixed-effects linear model, treating trial as a random effect
patient_metadata_combined_model <- patient_metadata_combined %>%
  mutate(Response_n = ifelse(
    Response == "R",
    1, ifelse(
      Response == "NR", 0, NA)))%>%
  mutate(Baseline_N_h = ifelse(Baseline_N_h == "Nx", "N+", Baseline_N_h))

sex_stage_treatment_mixed_effects_model <- glmer(
  Response_n ~ Sex + Baseline_N_h + Cohort + (1 | Trial),
  data = patient_metadata_combined_model , family = binomial)
summary(sex_stage_treatment_mixed_effects_model)

table(wes_features_df$Cohort, wes_features_df$Trial)
```

# Supplementary Table 1.  GLMM analysis Nivo Ipi combination vs Nivo Monotherapy

```{r}
sex_stage_treatment_mixed_effects_model_f <- glmer(
  Response_n ~ Sex + Baseline_N_h + Cohort + (1 | Trial),
  data = patient_metadata_combined_model %>%
    filter(Trial %in% c("MATISSE", "BELLINI", "IMCISION")),
  family = binomial)
summary(sex_stage_treatment_mixed_effects_model_f)

table(patient_metadata_combined$Cohort,
      patient_metadata_combined$Response, patient_metadata_combined$Trial) %>%
  as.data.frame() %>%
  setNames(c("Cohort", "Response", "Trial", "Freq")) %>%
  filter(Trial %in% c("MATISSE", "BELLINI", "IMCISION")) %>%
  ggplot(aes(x = Cohort, y = Freq, fill = Response)) +
  facet_wrap(~Trial) + geom_col() + theme_Publication()

```

# GLMer TMB tumour type
```{r}
# Mixed-effects linear model, treating trial as a random effect
tmb_mixed_effects_model <- glmer(Response_n ~ TMB_scaled + (1 | Trial),
                                 data = wes_features_df, family = binomial)
summary(tmb_mixed_effects_model)

tmb_mixed_effects_model_excl_dmmr <- glmer(
  Response_n ~ TMB_scaled + (1 | Trial),
  data = wes_features_df_excl_dMMR, family = binomial)
summary(tmb_mixed_effects_model_excl_dmmr)

```

## Fitting trials
```{r}
TMB_model <- glm(Response_n ~ TMB_scaled,
                 data = wes_features_df, family = binomial)
summary(TMB_model)

Trial_model <- glmer(Response_n ~ (1 | Trial),
                     data = wes_features_df, family = binomial)
summary(Trial_model)

TMB_trial_model <- glmer(Response_n ~ TMB_scaled + (1 | Trial),
                         data = wes_features_df, family = binomial)
summary(TMB_trial_model)

# Cohort Nivo Ipi Mono
TMB_trial_sex_stage_treatment_model <- glmer(
  Response_n ~ TMB_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
  data = wes_features_df, family = binomial)
TMB_trial_sex_stage_treatment_model
summary(TMB_trial_sex_stage_treatment_model)

```

## Likelihood ratio test
```{r}
anova(TMB_trial_model, Trial_model) # These are nested
# anova(TMB_trial_model, TMB_model) # This comparison is not valid, not nested
AIC(TMB_trial_model, TMB_model)
BIC(TMB_trial_model, TMB_model)
```

## Excl dMMR
```{r}
TMB_model <- glm(Response_n ~ TMB_scaled,
                 data = wes_features_df_excl_dMMR, family = binomial)
summary(TMB_model)
        
Trial_model <- glmer(Response_n ~ (1 | Trial),
                     data = wes_features_df_excl_dMMR, family = binomial)
summary(Trial_model)

TMB_trial_model <- glmer(Response_n ~ TMB_scaled + (1 | Trial),
                         data = wes_features_df_excl_dMMR, family = binomial)
summary(TMB_trial_model)

```

## Likelihood ratio test
```{r}
anova(TMB_trial_model, Trial_model) # These are nested
# anova(TMB_trial_model, TMB_model) # This comparison is not valid, not nested
AIC(TMB_trial_model, TMB_model)
BIC(TMB_trial_model, TMB_model)
```

# ----------
# UV vs TMB
## Visualizations
```{r}
wes_features_df %>%
  ggplot(aes(x = Trial, y = log2(UV_abs + 1), fill = Response)) +
  geom_boxplot(outlier.shape = NA) + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))  +
  scale_fill_manual(values = c(Color_NR, Color_R))

wes_features_df %>%
  ggplot(aes(x = Response, y = UV_abs)) +
  geom_boxplot(outlier.shape = NA) + geom_point(aes(col = Trial)) + 
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  stat_compare_means(method = "wilcox") +
  scale_color_manual(values = trial_colors)

wes_features_df_excl_dMMR %>%
  ggplot(aes(x = Response, y = UV_abs)) +
  geom_boxplot(outlier.shape = NA)  + geom_point(aes(col = Trial)) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  stat_compare_means(method = "wilcox") +
  labs(title = "Excl. dMMR") +
  scale_color_manual(values = trial_colors)

wes_features_df %>%
  filter(TMB < 10) %>%
  ggplot(aes(x = Response, y = UV_abs)) +
  geom_boxplot(outlier.shape = NA)  + geom_point(aes(col = Trial)) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  stat_compare_means(method = "wilcox") +
  labs(title = "TMB < 10 mut/Mb") +
  scale_color_manual(values = trial_colors)

wes_features_df_excl_dMMR_UV %>%
  filter(!Trial %in% c("NICHE-MSI", "PRADO", "OpACIN_neo", "MATISSE")) %>%
  ggplot(aes(x = Response, y = UV_abs)) +
  geom_boxplot(outlier.shape = NA) + geom_point(aes(col = Trial)) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  stat_compare_means(method = "wilcox") +
  labs(title = "Excl. dMMR and UV cohorts") +
  scale_color_manual(values = trial_colors)

wes_features_df_excl_dMMR_UV %>%
  filter(TMB < 10) %>%
  ggplot(aes(x = Response, y = UV_abs)) +
  geom_boxplot(outlier.shape = NA) + geom_point(aes(col = Trial)) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  stat_compare_means(method = "wilcox") +
  labs(title = "Excl. dMMR and UV cohorts and TMB > 10") +
  scale_color_manual(values = trial_colors)
```

```{r}
# Check collinearity
cor(wes_features_df$TMB, wes_features_df$UV_abs)
cor(wes_features_df_excl_dMMR$TMB, wes_features_df_excl_dMMR$UV_abs)
cor(wes_features_df_excl_dMMR_UV$TMB, wes_features_df_excl_dMMR_UV$UV_abs)

wes_features_df_sensitivity <- wes_features_df %>%
  mutate(cohorts = "Pooled") %>%
  rbind(wes_features_df_excl_dMMR %>%
          mutate(cohorts = "Pooled excl. dMMR")) %>%
  rbind(wes_features_df_excl_dMMR_UV %>%
          mutate(cohorts = "Pooled excl. dMMR & UV"))

wes_features_df_sensitivity %>%
  ggplot(aes(x = TMB, y = UV_abs)) +
  geom_point(aes(col = Trial)) +
  theme_Publication() + facet_wrap(~cohorts, scales = "free") + stat_cor() +
  scale_color_manual(values = trial_colors)
```


## Models

### Supplementary Table 4. GLMM Sensitivity Analysis UV + TMB

```{r}
# Cohort Nivo Ipi Mono

model1 <- glmer(Response_n ~ UV_scaled  + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_features_df, family = binomial)
summary(model1)

model2 <- glmer(Response_n ~ TMB_scaled + UV_scaled  + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_features_df, family = binomial)
summary(model2)

model3 <- glmer(Response_n ~ TMB_scaled  + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_features_df, family = binomial)
summary(model12)

anova(model2, model1)
anova(model2, model3)

```

```{r}
# Excluding dMMR
model4 <- glmer(Response_n ~ TMB_scaled + (1 | Trial),
                data = wes_features_df_excl_dMMR, family = binomial)
summary(model4)

model5 <- glmer(Response_n ~ UV_scaled + (1 | Trial),
                data = wes_features_df_excl_dMMR, family = binomial)
summary(model5)

model6 <- glmer(Response_n ~ TMB_scaled + UV_scaled + (1 | Trial),
                data = wes_features_df_excl_dMMR, family = binomial)
summary(model6)

anova(model5, model6)

```

## Excl dMMR and UV

### Supplementary Table 5. GLMM Sensitivity Analysis UV excluding dMMR and UV cohorts
### Supplementary Table 6. GLMM Sensitivity Analysis UV excluding dMMR and UV cohorts,  accounting for TMB

```{r}
# Cohort Nivo Ipi Mono
table(wes_features_df_excl_dMMR_UV$Cohort, wes_features_df_excl_dMMR_UV$Trial)

model10 <- glmer(Response_n ~ TMB_scaled + UV_scaled  + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_features_df_excl_dMMR_UV, family = binomial)

summary(model10)

model11 <- glmer(Response_n ~ TMB_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_features_df_excl_dMMR_UV, family = binomial)
summary(model11)

model12 <- glmer(Response_n ~ UV_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_features_df_excl_dMMR_UV, family = binomial)
summary(model12)

anova(model10, model11)
anova(model10, model12)
```

## TMB low histologies

# ----------
# APOBEC vs TMB
## Visualizations
```{r}
wes_features_df %>%
  ggplot(aes(x = Response, y = log2(APOBEC_abs + 1))) +
  geom_boxplot(outlier.shape = NA) + geom_point(aes(col = Trial)) + 
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  stat_compare_means(method = "wilcox") +
  scale_color_manual(values = trial_colors)

wes_features_df_excl_dMMR %>%
  ggplot(aes(x = Response, y = log2(APOBEC_abs + 1))) +
  geom_boxplot(outlier.shape = NA) + geom_point(aes(col = Trial)) + 
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  stat_compare_means(method = "wilcox") +
  scale_color_manual(values = trial_colors)

wes_features_df %>%
  ggplot(aes(x = Trial, y = log2(APOBEC_abs + 1), fill = Response)) +
  geom_boxplot(outlier.shape = NA) + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = c(Color_NR, Color_R))
```

### Supplementary Table 7. GLMM Sensitivity Analysis APOBEC + TMB

```{r}
# Cohort Nivo Ipi Mono
table(wes_features_df_excl_dMMR_UV$Cohort, wes_features_df_excl_dMMR_UV$Trial)

model10 <- glmer(Response_n ~ TMB_scaled + APOBEC_scaled  + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_features_df, family = binomial)

summary(model10)

model11 <- glmer(Response_n ~ TMB_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_features_df, family = binomial)
summary(model11)

model12 <- glmer(Response_n ~ APOBEC_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_features_df, family = binomial)
summary(model12)

anova(model10, model11)
anova(model10, model12)
```

## TMB low histologies

### Supplementary Table 8. GLMM Sensitivity Analysis APOBEC in TMB low cohorts
### Supplementary Table 9. GLMM Sensitivity Analysis APOBEC in TMB low cohorts,  accounting for TMB
```{r}

# Cohort Nivo Ipi Mono
table(wes_features_df_excl_dMMR_UV$Cohort, wes_features_df_excl_dMMR_UV$Trial)

model10 <- glmer(Response_n ~ TMB_scaled + APOBEC_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_features_df_excl_dMMR %>%
                  filter(Trial %in% c("BELLINI", "IMCISION", "NICHE-MSS")),
                family = binomial)

summary(model10)

model11 <- glmer(Response_n ~ TMB_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_features_df_excl_dMMR %>%
                  filter(Trial %in% c("BELLINI", "IMCISION", "NICHE-MSS")),
                family = binomial)
summary(model11)

model12 <- glmer(Response_n ~ APOBEC_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_features_df_excl_dMMR %>%
                  filter(Trial %in% c("BELLINI", "IMCISION", "NICHE-MSS")),
                family = binomial)
summary(model12)

anova(model10, model11)
anova(model11, model12)

```

# ----------
# WES RNA
```{r}
wes_rna_features_df <- read_csv2(here(wes_rna_dir, "wes_rna_features_df.csv"))

wes_rna_features_df <- wes_rna_features_df %>%
  mutate(Response_n = ifelse(
    Response == "R",
    1, ifelse(
      Response == "NR", 0, NA))) %>%
  mutate(TMB_scaled = scale(TMB),
         UV_scaled = scale(UV_abs),
         APOBEC_scaled = scale(APOBEC_abs),
         TIL_score_scaled = scale(TIL_score),
         Proliferation_scaled = scale(Tumor_proliferation_rate),
         B_cat_scaled = scale(B_catenin_signature_Luke_et_al))

# Add patient N stage and sex information
wes_rna_features_df <- wes_rna_features_df %>%
  left_join(patient_metadata_combined %>%
              select(Patient, Sex, Baseline_N_h)) %>%
  mutate(Baseline_N_h = ifelse(Baseline_N_h == "Nx", "N+", Baseline_N_h)) %>%
  mutate(Cohort = ifelse(Cohort %in% c("Nivo_ipi_A", "Nivo_ipi_B", "Nivo_ipi_C"),
                         "Nivo_ipi", Cohort))

table(wes_rna_features_df$Cohort)

wes_rna_features_df_pre <- wes_rna_features_df %>%
  filter(Treatment == "Pre")

wes_rna_features_df_pre_f <- wes_rna_features_df_pre %>% filter(!is.na(TMB)) 
wes_rna_features_df_pre_f %>% dim()
```

# ----------
# Quadrant plots TMB TIL score
```{r}
median_til_score_pre <- median(
  wes_rna_features_df_pre$TIL_score)
median_til_score_pre

wes_rna_features_df_pre_f %>% 
  ggplot(aes(x = TMB, y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  scale_x_continuous(trans = "log2") +
  scale_color_manual(values = trial_colors) +
  guides(col = guide_legend(nrow = 3))

TMB_TIL_quadrant_trial_plot <- wes_rna_features_df_pre_f %>% 
  ggplot(aes(x = TMB, y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication() + facet_wrap(~Response) +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed")  +
  scale_x_continuous(trans = "log2") +
  scale_color_manual(values = trial_colors) +
  guides(col = guide_legend(nrow = 3))

TMB_TIL_quadrant_trial_plot

graph2pdf(x = TMB_TIL_quadrant_trial_plot,
          file = here(output_dir, "_TMB_TIL_quadrant_trial_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 5,
          width = 7)

quadrant_response_proportions

# Quadrant RR
table(wes_rna_features_df_pre_f$TMB >= 10,
      wes_rna_features_df_pre_f$TIL_score >= median_til_score_pre) %>%
  `rownames<-`(c("TMB-L", "TMB-H")) %>%
  `colnames<-`( c("mTIL_low", "mTIL_high"))

```

```{r}
wes_rna_features_df_pre_f %>%
  ggplot(aes(x = log2(TMB), y = TIL_score, shape = Response, col = Trial)) +
  geom_point() + theme_Publication()

# Quadrant RR
table(wes_rna_features_df_pre_f$TMB >= 10,
      wes_rna_features_df_pre_f$TIL_score >= median_til_score_pre) %>%
  `rownames<-`(c("TMB-L", "TMB-H")) %>%
  `colnames<-`( c("mTIL_low", "mTIL_high"))

quadrant_response_proportions <- table(
  wes_rna_features_df_pre_f$TMB >= 10,
  wes_rna_features_df_pre_f$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre_f$Response) %>%
  as.data.frame() %>%
  setNames(c("TMB", "mTIL_group", "Response", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB == TRUE, "TMB_high", "TMB_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions %>%
  write_csv2(here(output_dir, "quadrant_response_proportions_tmb_til.csv"))

quadrant_response_proportions %>%
  filter(Response == "R") %>%
  rename(Response_rate = quadrant_prop) %>%
  ggplot(aes(x = quadrant, y = Response_rate)) + geom_col() +
  theme_Publication() + labs(x = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

quadrant_response_proportions

quadrant_response_proportions %>%
  select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()

trial_quadrant_proportions <- table(
  wes_rna_features_df_pre_f$TMB >= 10,
  wes_rna_features_df_pre_f$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre_f$Trial) %>%
  as.data.frame() %>%
  setNames(c("TMB_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB_group == TRUE, "TMB-H", "TMB-L")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
trial_quadrant_proportions %>%
  select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

quadrant_trial_proportions <- table(
  wes_rna_features_df_pre_f$TMB >= 10,
  wes_rna_features_df_pre_f$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre_f$Trial) %>%
  as.data.frame() %>%
  setNames(c("TMB_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(TMB_group = ifelse(TMB_group == TRUE, "TMB-H", "TMB-L")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(TMB_group, "_", mTIL_group)) %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = Freq / Total)

TMB_TIL_quadrant_trial_proportions_plot <- quadrant_trial_proportions %>%
  ggplot(aes(x = quadrant, y = quadrant_prop, fill = Trial)) + geom_col() +
  scale_fill_manual(values = trial_colors) +
  theme_Publication() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 3)) +
  coord_flip()

graph2pdf(x = TMB_TIL_quadrant_trial_proportions_plot,
          file = here(
            output_dir, "_TMB_TIL_quadrant_trial_proportions_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 6)


table(wes_rna_features_df_pre_f$Response)

wes_rna_features_df_pre_f %>%
  mutate(Response = ifelse(Response == "R",
                           1, ifelse(Response == "NR", 0, NA))) %>%
  pull(Response) %>% sum()

```

# TMB TIL score GLMer
```{r, fig.height = 8, fig.width = 8}
cor.test(wes_rna_features_df_pre_f$TMB,
         wes_rna_features_df_pre_f$TIL_score)

wes_rna_features_df_pre_f %>%
  ggplot(aes(x = TMB_scaled, y = TIL_score_scaled)) +
  geom_point() + theme_Publication() + stat_cor()

wes_rna_features_df_pre_f %>%
  ggplot(aes(x = TMB_scaled, y = TIL_score_scaled)) +
  geom_point() + theme_Publication() +
  stat_cor() + facet_wrap(~Trial, scales = "free")
```

### Supplementary Table 14. GLMM analysis TMB and TIL signature score

```{r}
# Cohort Nivo Ipi Mono
model11 <- glmer(Response_n ~ TMB_scaled * TIL_score_scaled  + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_rna_features_df_pre_f,
                family = binomial)

summary(model11)

model12 <- glmer(Response_n ~ TMB_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_rna_features_df_pre_f,
                family = binomial)
summary(model12)

model13 <- glmer(Response_n ~ TIL_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_rna_features_df_pre_f,
                family = binomial)
summary(model13)

model14 <- glmer(Response_n ~ TMB_scaled + TIL_score_scaled + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_rna_features_df_pre_f,
                family = binomial)
summary(model14)

anova(model11, model12)
anova(model11, model13)
anova(model11, model14)
anova(model11, model13, model14)
anova(model11, model12, model14)

```

## Excluding dMMR

### Supplementary Table X. GLMM analysis TMB and TIL signature score excluding dMMR

```{r}
wes_rna_features_df_pre_excl_dMMR <- wes_rna_features_df_pre_f %>%
  filter(! Trial %in% c("NICHE-MSI"))

# Cohort Nivo Ipi Mono
model11 <- glmer(Response_n ~ TMB_scaled * TIL_score_scaled  + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_rna_features_df_pre_excl_dMMR,
                family = binomial)

summary(model11)

model12 <- glmer(Response_n ~ TMB_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_rna_features_df_pre_excl_dMMR,
                family = binomial)
summary(model12)

model13 <- glmer(Response_n ~ TIL_score_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_rna_features_df_pre_excl_dMMR,
                family = binomial)
summary(model13)

anova(model11, model12)
anova(model11, model13)
```
 
# ----------
# Quadrant plots Proliferation TIL score

```{r}
# Not filtering for TMB paired samples
median_Tumor_proliferation_rate_pre <- wes_rna_features_df_pre %>%
  filter(Treatment == "Pre") %>%
  pull(Tumor_proliferation_rate) %>%
  median()
median_Tumor_proliferation_rate_pre

wes_rna_features_df_pre %>%
  filter(Treatment == "Pre") %>% 
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Trial)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = median_Tumor_proliferation_rate_pre,
             linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  scale_color_manual(values = trial_colors) +
  guides(col = guide_legend(nrow = 3))

prolif_TIL_quadrant_trial_plot <- wes_rna_features_df_pre %>%
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Trial)) +
  geom_point() + theme_Publication() +
  geom_vline(xintercept = median_Tumor_proliferation_rate_pre,
             linetype = "dashed") +
  geom_hline(yintercept = median_til_score_pre, linetype = "dashed") +
  scale_color_manual(values = trial_colors) +
  guides(col = guide_legend(nrow = 3)) + facet_wrap(~Response) +
  scale_color_manual(values = trial_colors)

prolif_TIL_quadrant_trial_plot

trial_quadrant_proportions <- table(
  wes_rna_features_df_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre, 
  wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre$Trial) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
dim(wes_rna_features_df_pre)
table(wes_rna_features_df_pre$Response)

quadrant_response_proportions <- table(
  wes_rna_features_df_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre$Response) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Response", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions

quadrant_response_proportions %>%
  select(quadrant, Response, Freq) %>%
  pivot_wider(names_from = Response, values_from = Freq) %>%
  column_to_rownames("quadrant") %>%
  as.matrix() %>%
  chisq.test()

quadrant_response_proportions %>%
  filter(Response == "R") %>%
  rename(Response_rate = quadrant_prop) %>%
  ggplot(aes(x = quadrant, y = Response_rate)) + geom_col() +
  theme_Publication() + labs(x = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

```{r}
wes_rna_features_df_pre %>%
  filter(TIL_score < median_til_score_pre) %>%
  n_distinct()

wes_rna_features_df_pre %>%
  filter(TIL_score >= median_til_score_pre) %>%
  n_distinct()

table(
  wes_rna_features_df_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  wes_rna_features_df_pre$TIL_score > median_til_score_pre)

trial_quadrant_proportions <- table(
  wes_rna_features_df_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre$Trial) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  group_by(Trial) %>% add_count(Trial, wt = Freq, name = "Total") %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group),
         quadrant_prop = Freq / Total)
  
trial_quadrant_proportions

trial_quadrant_proportions %>% select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

quadrant_trial_proportions <- table(
  wes_rna_features_df_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre$Trial) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Trial", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = Freq / Total)

quadrant_response_proportions <- table(
  wes_rna_features_df_pre$Tumor_proliferation_rate > median_Tumor_proliferation_rate_pre,
  wes_rna_features_df_pre$TIL_score > median_til_score_pre,
  wes_rna_features_df_pre$Response) %>%
  as.data.frame() %>%
  setNames(c("Prolif_group", "mTIL_group", "Response", "Freq")) %>%
  mutate(Prolif_group = ifelse(Prolif_group == TRUE,
                               "Prolif_high", "Prolif_low")) %>%
  mutate(mTIL_group = ifelse(mTIL_group == TRUE, "mTIL_high", "mTIL_low")) %>%
  mutate(quadrant = paste0(Prolif_group, "_", mTIL_group))  %>%
  group_by(quadrant) %>% add_count(quadrant, wt = Freq, name = "Total") %>%
  mutate(quadrant_prop = round(Freq / Total, 2))

quadrant_response_proportions

quadrant_response_proportions %>%
  write_csv2(here(output_dir, "quadrant_response_proportions_prolif_til.csv"))

prolif_TIL_quadrant_trial_proportions_plot <- quadrant_trial_proportions %>%
  ggplot(aes(x = quadrant, y = quadrant_prop, fill = Trial)) + geom_col() +
  scale_fill_manual(values = trial_colors) + theme_Publication() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 3)) +
  coord_flip()
  
graph2pdf(x = prolif_TIL_quadrant_trial_proportions_plot,
          file = here(
            output_dir, "_prolif_TIL_quadrant_trial_proportions_plot.pdf"),
          colormodel = "rgb",
          font = "Arial",
          height = 4,
          width = 6)

trial_quadrant_proportions %>%
  select(quadrant, quadrant_prop, Trial) %>%
  ggplot(aes(x = Trial, y = quadrant_prop, fill = quadrant)) +
  geom_col() +
  theme_Publication() +
  coord_flip() +
  labs(x = "", y = "", col = "") +
  guides(fill = guide_legend(nrow = 2))

wes_rna_features_df_pre %>% 
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Trial)) +
  geom_point() + theme_Publication() + facet_wrap(~Response) +
  scale_color_manual(values = trial_colors)

wes_rna_features_df_pre %>% 
  ggplot(aes(x = Tumor_proliferation_rate, y = TIL_score,
             shape = Response, col = Trial)) +
  geom_point() + theme_Publication() + facet_wrap(~Response) +
  scale_color_manual(values = trial_colors)

wes_rna_features_df_pre %>%
  ggplot(aes(x = TIL_score, y = Ayers_T_cell_inflamed)) +
  geom_point() + theme_Publication()


```

# Prolif TIL score GLMer

### Supplementary Table 15. GLMM analysis TMB and Proliferation signature score

```{r}
# Cohort Nivo Ipi Mono
model11 <- glmer(Response_n ~ Proliferation_scaled * TIL_score_scaled  + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_rna_features_df_pre,
                family = binomial)

summary(model11)

model12 <- glmer(Response_n ~ Proliferation_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_rna_features_df_pre,
                family = binomial)
summary(model12)

model13 <- glmer(Response_n ~ TIL_score_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_rna_features_df_pre,
                family = binomial)
summary(model13)

model14 <- glmer(Response_n ~ Proliferation_scaled + TIL_score_scaled  + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_rna_features_df_pre,
                family = binomial)

summary(model14)

anova(model11, model12)
anova(model11, model13)
anova(model11, model14)

anova(model11, model12, model14)
anova(model11, model13, model14)

```

## Excl dMMR
### Supplementary Table X. GLMM analysis TMB and Proliferation signature score excl dMMR

```{r}
wes_rna_features_df_pre_excl_dMMR <- wes_rna_features_df_pre %>%
  filter(!Trial %in% c("NICHE-MSI"))


# Cohort Nivo Ipi Mono
model11 <- glmer(Response_n ~ Proliferation_scaled * TIL_score_scaled  + Sex + Baseline_N_h + Cohort + (1 | Trial),
                data = wes_rna_features_df_pre_excl_dMMR,
                family = binomial)

summary(model11)

model12 <- glmer(Response_n ~ Proliferation_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_rna_features_df_pre_excl_dMMR,
                family = binomial)
summary(model12)

model13 <- glmer(Response_n ~ TIL_score_scaled + Sex + Baseline_N_h + Cohort  + (1 | Trial),
                data = wes_rna_features_df_pre_excl_dMMR,
                family = binomial)
summary(model13)

anova(model11, model12)
anova(model11, model13)

```

# ----------
# Lint script
```{r}
# lint("~/NKI_pancancer_neoadj_io/src/pancan_glmer_sensitivity.Rmd")
lint(filename = "~/NKI_pancancer_neoadj_io/src/pancan_glmer_sensitivity.Rmd",
  linters = unused_import_linter()
)
```

# ----------
# sessionInfo - Package Versions
```{r sessionInfo}
sessionInfo()
```
