---
title: "neoadjuvant pancancer Aggregate meta-analysis"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  export: TRUE
  output_dir: "meta_aggregated"
  metadata: ""
  exclude_niche_msi: FALSE
  run_meta_analysis: TRUE
  exclude_clinical_response: FALSE
  response_definition: "Original" # MPR pCR
---

Author comment:
File description comment:
The purpose of this script is to parse results from meta-analysis subsets, aggregating to visualize and export results

# Loading Packages
```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# Install if necessary install.packages("tidyverse")
# Install if necessary BiocManager::install("DESeq2")
library(tidyverse)
source("helper_functions.R")
source("oncomatrix.R")
library(VariantAnnotation)
library(ggpubr)
library(ggrepel)
library(lintr)
library(export)
MakeDataFrameFromVCF <- getFromNamespace("MakeDataFrameFromVCF", "ICAMS")

set.seed(42)
select <- dplyr::select
rename <- dplyr::rename
here <- here::here
```

# Setup output folder

```{r Setup output, echo = FALSE, include = FALSE}
export <- params$export
response_definition <- params$response_definition

# overwrite clinical exclusion in case of MPR/pCR response
exclude_clinical_response <- params$exclude_clinical_response
if (params$response_definition %in% c("MPR", "pCR")) {
  exclude_clinical_response <- TRUE
}

# generate output directory

output_dir <- here("data", "processed", params$output_dir)
print(output_dir)

print(exclude_clinical_response)
print(response_definition)

# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)
    } else {
  print("Output folder already exists")
}

```

```{r}
wes_dir <- here("data", "processed", "WES_analysis")
metadata_dir <- here("data", "processed", "metadata")
wes_meta_dir <- here("data", "processed", "WES_analysis_meta") 
wes_MPR_meta_dir <- here("data", "processed", "WES_analysis_MPR_meta") 
wes_pCR_meta_dir <- here("data", "processed", "WES_analysis_pCR_meta") 
```

# ----------
# Load metadata

```{r Load metadata}
dna_metadata <- read_csv2(here(metadata_dir, "combined_dna_metadata.csv"))

patient_metadata_combined <- read_csv2(
  here("data", "processed", "metadata", "patient_metadata_combined.csv"))

histology_df <- data.frame(
  Trial = c("BELLINI", "IMCISION", "NABUCCO", "NICHE-MSS", "NICHE-MSI",
            "OpACIN_neo", "PRADO", "MATISSE"),
  Histology = c("TBNC", "HNSCC", "Urothelial", "Colon", "Colon",
                "LN_melanoma", "LN_melanoma", "CSCC"))

```

# Annotation color
```{r}

# Define colours for each group
Color_NR <- "#F8766D"
Color_R <- "#00BA38" # "#619CFF"
Color_NA <- "#CD9600"
Color_Pre <- "#619CFF"
Color_Post <- "#F564E3"
Color_HNSCC <- "#009292"
Color_CSCC <- "#CD9600"
Color_Urothelial <- "#B66DFF"
Color_LN_melanoma <- "#00BFC4"
Color_Colon <- "#F8766D"
Color_Colon_MSI <- "#B73A3A"
Color_Colon_MSS <- "#23C26F"
Color_TNBC <- "#FEB5DA"
Color_BELLINI <- "#FEB5DA"
Color_IMCISION <- "#009292"
Color_MATISSE <- "#CD9600"
Color_NABUCCO <- "#B66DFF"
Color_BELLINI <- "#FEB5DA"
Color_NICHE_MSS <- "#23C26F"
Color_NICHE_MSI <- "#B73A3A"
Color_OpACIN_neo <- "#00BFC4"
Color_PRADO <- "#619CFF"

annotation_color_list = list(
  Response = c(NR = Color_NR, R = Color_R),
  Treatment = c(Pre = Color_Pre, Post = Color_Post),
  Histology = c(LN_melanoma = Color_LN_melanoma, Urothelial = Color_Urothelial,
                Colon = Color_Colon, HNSCC = Color_HNSCC, TBNC = Color_TNBC,
                CSCC = Color_CSCC),
  Trial = c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO))

trial_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_colors_combined <- c(Combined = "#999999", BELLINI = Color_BELLINI,
                           IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_Colon_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

histology_msi_mss_colors <- c(`Colon-MSI`= Color_Colon_MSI,
                              `Colon-MSS`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              LN_melanoma = Color_LN_melanoma,
                              TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)

```

# ----------

# Load WES meta-analysis results 
```{r}
list.files(here("data", "processed"))
list.files(here(wes_meta_dir))

wes_meta_studies_results_random_effects <- read_csv2(
  here(wes_meta_dir, "meta_studies_results_random_effects.csv"))

wes_meta_random_effects <- read_csv2(
    here(wes_meta_dir, "meta_combined_effects_random_effects.csv"))

wes_excl_mono_random_effects <- read_csv2(
    here("data", "processed", "WES_analysis_meta_excl_mono",
         "meta_combined_effects_random_effects.csv"))

wes_excl_clin_random_effects <- read_csv2(
    here("data", "processed", "WES_analysis_meta_excl_clin_response",
         "meta_combined_effects_random_effects.csv"))

wes_MPR_meta_random_effects <- read_csv2(
    here(wes_MPR_meta_dir, "meta_combined_effects_random_effects.csv"))

wes_pCR_meta_random_effects <- read_csv2(
    here(wes_pCR_meta_dir, "meta_combined_effects_random_effects.csv"))

```

## Aggregate meta results
```{r}
wes_meta_aggregated_smd_all_statistics <- wes_meta_random_effects %>%
  mutate(Analysis_subset = "Original") %>%
  rbind(
    wes_MPR_meta_random_effects %>%
      mutate(Analysis_subset = "MPR")) %>%
  rbind(
    wes_pCR_meta_random_effects %>%
      mutate(Analysis_subset = "pCR"))  %>%
  rbind(
    wes_excl_mono_random_effects %>%
      mutate(Analysis_subset = "Excl_mono."))  %>%
  rbind(
    wes_excl_clin_random_effects %>%
      mutate(Analysis_subset = "Excl_clin.")) %>%
  relocate(Analysis_subset, .before = "Variable")

wes_meta_aggregated_smd_all_statistics %>%
  write_csv2(here("data", "processed", "meta_aggregated",
                  "wes_meta_aggregated_smd_all_statistics.csv")
)

```


```{r}
variable_order <- factor(c("TMB", "pvactools_filtered_neoantigen_load",
                    "pyclone_clonal_TMB", "pyclone_subclonal_TMB", 
                    "APOBEC_abs", "UV_abs", "HRD_abs", "dMMR_abs",
                    "TMB_frameshift", "wgII_0.6_bases", 
                             "Aneuploidy_score", "Ploidy", "scna_loss",
                             "Purity"))

wes_meta_aggregated %>%
  mutate(Variable = factor(Variable, levels = fct_rev(variable_order))) %>%
  mutate(label = cut(
    pval.random,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(
      Analysis_subset,
      levels = c("Original", "Excl_mono.", "Excl_clin.", "MPR", "pCR")), 
    y = fct_rev(Variable),
    fill = TE.random,
    label = label)
    ) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "", title = "Meta-analysis sensitivity")

```

# ----------
# Load WES Log OR meta results
```{r}

meta_studies_results_or_all <-  read_csv2(
  here(wes_meta_dir, "meta_studies_results_or_all.csv"))

wes_meta_or <- read_csv2(
  here(wes_meta_dir, "meta_log_or_combined_effects_random_effects.csv"))

wes_excl_mono_or <- read_csv2(
    here("data", "processed", "WES_analysis_meta_excl_mono",
         "meta_log_or_combined_effects_random_effects.csv"))

wes_excl_clin_or <- read_csv2(
    here("data", "processed", "WES_analysis_meta_excl_clin_response",
         "meta_log_or_combined_effects_random_effects.csv"))

wes_MPR_meta_or <- read_csv2(
    here(wes_MPR_meta_dir, "meta_log_or_combined_effects_random_effects.csv"))

wes_pCR_meta_or <- read_csv2(
    here(wes_pCR_meta_dir, "meta_log_or_combined_effects_random_effects.csv"))

```

## Aggregate meta Log OR

```{r}
wes_or_meta_aggregated_all_statistics  <- wes_meta_or %>%
  mutate(Analysis_subset = "Original") %>%
  rbind(
    wes_MPR_meta_or %>%
      mutate(Analysis_subset = "MPR")) %>%
  rbind(
    wes_pCR_meta_or %>%
      mutate(Analysis_subset = "pCR"))  %>%
  rbind(
    wes_excl_mono_or %>%
      mutate(Analysis_subset = "Excl_mono."))  %>%
  rbind(
    wes_excl_clin_or %>%
      mutate(Analysis_subset = "Excl_clin.")) %>%
  relocate(Analysis_subset, .before = "feature") %>%
  rename(Variable = "feature")

wes_or_meta_aggregated_all_statistics %>%
  write_csv2(here(output_dir, "wes_or_meta_aggregated_all_statistics.csv"))
```

```{r}
log_or_variable_order <- c("GIE_AP", "TP53", "BRCA1_2", "SIGPW_Wnt_B_Cat",
                           "SIGPW_PI3K_signaling", "SIGPW_Cell_cycle",
                           "SIGPW_TGFB_signaling", "GIE_IFNy", "RAS_RAF",
                           "HLA_alteration")

wes_or_meta_aggregated_all_statistics %>%
  mutate(feature = factor(feature, levels = fct_rev(log_or_variable_order))) %>%
  mutate(label = cut(
    pval.random,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(Analysis_subset,
               levels = c(
                 "Original", "Excl_mono.", "Excl_clin.", "MPR", "pCR")), 
    y = factor(
      feature, levels = fct_rev(log_or_variable_order)),
    fill = TE.random,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "", title = "Meta-analysis OR sensitivity")

```

## Aggregate WES SMD and OR meta results
```{r}
wes_meta_aggregated_all_statistics <- wes_meta_aggregated_smd_all_statistics %>%
  rbind(wes_or_meta_aggregated_all_statistics) 

wes_meta_aggregated_all_statistics <- wes_meta_aggregated_all_statistics %>%
  group_by(Analysis_subset) %>%
  mutate(padj = p.adjust(pval.random, method = "BH")) %>%
  ungroup()

wes_meta_aggregated_all_statistics %>%
  select(Analysis_subset, Variable, pval.random, padj)

wes_meta_aggregated_all_statistics %>%
  select(Analysis_subset, Variable, pval.random, padj) %>%
  filter(pval.random < 0.1)

wes_meta_aggregated_all_statistics <- wes_meta_aggregated_all_statistics %>%
  relocate(padj, .after = pval.random) %>%
  mutate(Analysis_subset = factor(
    Analysis_subset,
    levels = c("Original", "Excl_clin.", "Excl_mono.", "MPR", "pCR"))) %>%
  arrange(Analysis_subset)

wes_meta_aggregated_all_statistics %>%
  write_csv2(here("data", "processed", "meta_aggregated",
                  "wes_meta_aggregated_all_statistics.csv")
)

colnames(wes_meta_aggregated_all_statistics)
colnames(wes_or_meta_aggregated_all_statistics)
colnames(wes_meta_aggregated_all_statistics)

```

#----------
# Load RNA meta-analysis results
```{r}
output_folders <- list.files(here("data", "processed"))
output_folders[str_detect(output_folders, "meta")]

list.files(here("data", "processed", "gene_signatures_meta"))

rna_meta_random_effects <- read_csv2(
    here("data", "processed", "gene_signatures_meta",
         "meta_combined_effects_random_effects.csv"))

rna_excl_mono_random_effects <- read_csv2(
    here("data", "processed", "gene_signatures_meta_excl_mono",
         "meta_combined_effects_random_effects.csv"))

rna_excl_clin_random_effects <- read_csv2(
    here("data", "processed", "gene_signatures_meta_excl_clin_response",
         "meta_combined_effects_random_effects.csv"))

rna_MPR_meta_random_effects <- read_csv2(
    here("data", "processed", "gene_signatures_meta_MPR",
         "meta_combined_effects_random_effects.csv"))

rna_pCR_meta_random_effects <- read_csv2(
    here("data", "processed", "gene_signatures_meta_pCR",
         "meta_combined_effects_random_effects.csv"))

```

## Aggregate RNA meta results

```{r, fig.width  = 8, fig.height = 11}

rna_meta_aggregated_all_statistics  <- rna_meta_random_effects %>%
  mutate(Analysis_subset = "Original",
         padj = p.adjust(pval.random, method = "BH")) %>%
  rbind(
    rna_MPR_meta_random_effects %>%
      mutate(Analysis_subset = "MPR",
             padj = p.adjust(pval.random, method = "BH"))) %>%
  rbind(
    rna_pCR_meta_random_effects %>%
      mutate(Analysis_subset = "pCR",
             padj = p.adjust(pval.random, method = "BH"))) %>%
  rbind(
    rna_excl_mono_random_effects %>%
      mutate(Analysis_subset = "Excl_mono.",
             padj = p.adjust(pval.random, method = "BH")))  %>%
  rbind(
    rna_excl_clin_random_effects %>%
      mutate(Analysis_subset = "Excl_clin.",
             padj = p.adjust(pval.random, method = "BH"))) %>%
  mutate(Variable = ifelse(Variable == "Gamma_delta_TCR_Yu",
                           "Gamma_delta_TCRs", Variable))%>%
  relocate(Analysis_subset, .before = "Variable")


rna_meta_aggregated_all_statistics <- rna_meta_aggregated_all_statistics %>%
  relocate(padj, .after = pval.random) %>%
  mutate(Analysis_subset = factor(
    Analysis_subset,
    levels = c("Original", "Excl_clin.", "Excl_mono.", "MPR", "pCR"))) %>%
  arrange(Analysis_subset)


rna_meta_aggregated_all_statistics %>%
  write_csv2(here(output_dir, "rna_meta_aggregated_all_statistics.csv"))

# rna_meta_aggregated_all_statistics <- rna_meta_aggregated_all_statistics %>%
#   group_by(Analysis_subset) %>%
#   mutate(padj = p.adjust(pval.random, method = "BH")) %>%
#   ungroup()

colnames(rna_meta_aggregated_all_statistics)
colnames(rna_excl_mono_random_effects)
colnames(rna_excl_clin_random_effects)


rna_meta_studies_results <- read_csv2(
    here("data", "processed", "gene_signatures_meta",
         "meta_studies_results.csv"))

# Filter out meta-analysis variables, used combined analysis subset metrics
rna_meta_studies_results_per_cohort <- rna_meta_studies_results %>%
  select(c("n.e", "mean.e", "sd.e", "n.c", "mean.c", "sd.c", "studlab",
           "TE", "seTE", "statistic", "pval", "lower", "upper", "w.common",
           "w.random", "zval", "w.fixed", "Signature", "Contrast")) %>%
  mutate(Analysis_subset = "Original") %>%
  relocate(c(Analysis_subset, studlab, Signature, Contrast), .before = "n.e") 

rna_meta_studies_results_per_cohort %>%
  write_csv2(here(output_dir, "rna_meta_studies_results_per_cohort.csv"))

wes_meta_studies_results_per_cohort <- wes_meta_studies_results_random_effects %>%
  select(c("n.e", "mean.e", "sd.e", "n.c", "mean.c", "sd.c", "studlab",
           "TE", "seTE", "statistic", "pval", "lower", "upper", "w.common",
           "w.random", "zval", "w.fixed", "Variable", "Contrast")) %>%
  mutate(Analysis_subset = "Original") %>%
  relocate(c(Analysis_subset, studlab, Variable, Contrast), .before = "n.e") %>%
  full_join(meta_studies_results_or_all %>%
  select(c("studlab",
           "TE", "seTE", "statistic", "pval", "lower", "upper", "w.common",
           "w.random", "zval", "w.fixed", "Variable", "Contrast")) %>%
  mutate(Analysis_subset = "Original") %>%
  relocate(c(Analysis_subset, studlab, Variable, Contrast), .before = "TE") )

wes_meta_studies_results_per_cohort %>%
  write_csv2(here(output_dir, "wes_meta_studies_results_per_cohort.csv"))

```


```{r, fig.width  = 8, fig.height = 11}
original_order <- rna_meta_random_effects %>%
  mutate(Variable = ifelse(Variable == "Gamma_delta_TCR_Yu",
                           "Gamma_delta_TCRs", Variable)) %>%
  arrange(TE.random) %>%
  pull("Variable")

rna_meta_aggregated_all_statistics %>%
  mutate(label = cut(
    pval.random,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(Analysis_subset,
               levels = c(
                 "Original", "Excl_mono.", "Excl_clin.", "MPR", "pCR")), 
    y = factor(
      Variable, levels = original_order),
    fill = TE.random,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "", title = "Meta-analysis sensitivity")

```

# ----------
# Heterogeneity statistics

## wes smd meta
```{r}
wes_meta_aggregated_smd_all_statistics %>%
  ggplot(aes(x = TE.random, y = TE.fixed, col = Analysis_subset)) +
  geom_point() +
  theme_Publication() + geom_abline(slope = 1, intercept = 0)

wes_meta_aggregated_smd_all_statistics %>%
  ggplot(aes(x = Q, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = Q, y = -log10(pval.Q), label = Variable),
                  data = wes_meta_aggregated_smd_all_statistics)

wes_meta_aggregated_smd_all_statistics %>%
  ggplot(aes(x = tau2, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = tau2, y = -log10(pval.Q), label = Variable),
                  data = wes_meta_aggregated_smd_all_statistics)

wes_meta_aggregated_smd_all_statistics %>%
  ggplot(aes(x = I2, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = I2, y = -log10(pval.Q), label = Variable),
                  data = wes_meta_aggregated_smd_all_statistics)

wes_meta_aggregated_smd_all_statistics %>%
  filter(pval.random < 0.05) %>%
  ggplot(aes(x = I2, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = I2, y = -log10(pval.Q), label = Variable),
                  data = wes_meta_aggregated_smd_all_statistics %>%
  filter(pval.random < 0.05))

wes_meta_aggregated_smd_all_statistics %>%
  ggplot(aes(x = Q, y = I2, col = Analysis_subset)) +
  geom_point() + theme_Publication() +
  geom_text_repel(aes(x = Q, y = I2, label = Variable),
                  data = wes_meta_aggregated_smd_all_statistics)

wes_meta_aggregated_smd_all_statistics %>%
  ggplot(aes(x = tau2, y = I2, col = Analysis_subset)) +
  geom_point() + theme_Publication() +
  geom_text_repel(aes(x = tau2, y = I2, label = Variable),
                  data = wes_meta_aggregated_smd_all_statistics)

wes_meta_aggregated_smd_all_statistics %>%
  mutate(label = cut(
    pval.Q,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(Analysis_subset), 
    y = factor(
      Variable, levels = rev(order_variables)),
    fill = I2,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

ggsave(here(output_dir, "meta_combined_effects_random_effects_het.png"),
    height = 9, width = 4.5)

heterogeneity_plot <- wes_meta_aggregated_smd_all_statistics %>%
  mutate(label = cut(
    pval.Q,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(Analysis_subset), 
    y = factor(
      Variable, levels = rev(order_variables)),
    fill = I2,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

heterogeneity_plot
graph2pdf(heterogeneity_plot,
          here(output_dir, "meta_combined_effects_random_effects_het.pdf"),
    height = 9, width = 4.5)
```

```{r}
wes_meta_random_effects_all <- wes_meta_random_effects %>%
  select(Variable, TE.random, pval.random) %>%
  mutate(Feature_set = "WES_SMD") %>%
  rbind(
    meta_studies_results_or_all %>%
      select(feature, TE.random, pval.random) %>%
      rename(Variable = feature)%>%
      mutate(Feature_set = "WES_OR")
  )  %>%
  rbind(
    rna_meta_random_effects %>%
      select(Variable, TE.random, pval.random) %>%
      mutate(Feature_set = "RNA_SMD")
    )  %>%
  mutate(meta_FDR = p.adjust(pval.random, method = "BH")) %>%
  arrange(TE.random)

wes_meta_random_effects_all %>%
  ggplot(aes(x = pval.random, y = meta_FDR)) +
  geom_point() + theme_Publication()

wes_meta_random_effects_all %>%
  filter(meta_FDR < 0.05) %>%
  ggplot(aes(x = pval.random, y = meta_FDR)) +
  geom_point() + theme_Publication()

wes_meta_random_effects_all %>%
  arrange(-TE.random) %>% 
  select(Variable, pval.random, meta_FDR, Feature_set) %>%
  mutate(pval.random = round(pval.random, 2)) %>%
  mutate(meta_FDR = round(meta_FDR, 2)) %>%
  arrange(meta_FDR)
```


## OR meta
```{r}

order_variables <- wes_meta_or %>%
  arrange(TE.random) %>%
  pull(feature) 

wes_or_meta_aggregated_all_statistics %>%
  ggplot(aes(x = TE.random, y = TE.fixed, col = Analysis_subset)) +
  geom_point() +
  theme_Publication() + geom_abline(slope = 1, intercept = 0)

wes_or_meta_aggregated_all_statistics %>%
  ggplot(aes(x = Q, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = Q, y = -log10(pval.Q), label = Variable),
                  data = wes_or_meta_aggregated_all_statistics)

wes_or_meta_aggregated_all_statistics %>%
  ggplot(aes(x = tau2, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = tau2, y = -log10(pval.Q), label = Variable),
                  data = wes_or_meta_aggregated_all_statistics)

wes_or_meta_aggregated_all_statistics %>%
  ggplot(aes(x = I2, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = I2, y = -log10(pval.Q), label = Variable),
                  data = wes_or_meta_aggregated_all_statistics)

wes_or_meta_aggregated_all_statistics %>%
  filter(pval.random < 0.05) %>%
  ggplot(aes(x = I2, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = I2, y = -log10(pval.Q), label = Variable),
                  data = wes_or_meta_aggregated_all_statistics %>%
  filter(pval.random < 0.05))

wes_or_meta_aggregated_all_statistics %>%
  ggplot(aes(x = Q, y = I2, col = Analysis_subset)) + geom_point() +
  theme_Publication() +
  geom_text_repel(aes(x = Q, y = I2, label = Variable),
                  data = wes_or_meta_aggregated_all_statistics)

wes_or_meta_aggregated_all_statistics %>%
  ggplot(aes(x = tau2, y = I2, col = Analysis_subset)) + geom_point() +
  theme_Publication() +
  geom_text_repel(aes(x = tau2, y = I2, label = Variable),
                  data = wes_or_meta_aggregated_all_statistics)

wes_or_meta_aggregated_all_statistics %>%
  mutate(label = cut(
    pval.Q,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(Analysis_subset), 
    y = factor(
      Variable, levels = rev(order_variables)),
    fill = I2,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

ggsave(here(output_dir, "meta_combined_effects_random_effects_OR_het.png"),
    height = 9, width = 4.5)

heterogeneity_plot <- wes_or_meta_aggregated_all_statistics %>%
  mutate(label = cut(
    pval.Q,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(Analysis_subset), 
    y = factor(
      Variable, levels = rev(order_variables)),
    fill = I2,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

heterogeneity_plot
graph2pdf(heterogeneity_plot,
          here(output_dir, "meta_combined_effects_random_effects_OR_het.pdf"),
    height = 9, width = 4.5)
```

## RNA SMD meta

```{r, fig.height = 10, fig.width = 8}
order_variables <- rna_meta_random_effects %>%
  mutate(Variable = ifelse(
    Variable == "Gamma_delta_TCR_Yu", "Gamma_delta_TCRs", Variable)) %>%
  arrange(TE.random) %>%
  pull(Variable) %>%
  rev()

rna_meta_aggregated_all_statistics %>%
  ggplot(aes(x = TE.random, y = TE.fixed, col = Analysis_subset)) +
  geom_point() +
  theme_Publication() + geom_abline(slope = 1, intercept = 0)

rna_meta_aggregated_all_statistics %>%
  ggplot(aes(x = Q, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = Q, y = -log10(pval.Q), label = Variable),
                  data = rna_meta_aggregated_all_statistics)

rna_meta_aggregated_all_statistics %>%
  ggplot(aes(x = tau2, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = tau2, y = -log10(pval.Q), label = Variable),
                  data = rna_meta_aggregated_all_statistics)

rna_meta_aggregated_all_statistics %>%
  ggplot(aes(x = I2, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = I2, y = -log10(pval.Q), label = Variable),
                  data = rna_meta_aggregated_all_statistics)

rna_meta_aggregated_all_statistics %>%
  filter(pval.random < 0.05) %>%
  ggplot(aes(x = I2, y = -log10(pval.Q), col = Analysis_subset)) +
  geom_point() +
  theme_Publication() +
  geom_hline(yintercept = -log10(0.05)) +
  geom_text_repel(aes(x = I2, y = -log10(pval.Q), label = Variable),
                  data = rna_meta_aggregated_all_statistics %>%
  filter(pval.random < 0.05))

rna_meta_aggregated_all_statistics %>%
  ggplot(aes(x = Q, y = I2, col = Analysis_subset)) + geom_point() +
  theme_Publication() +
  geom_text_repel(aes(x = Q, y = I2, label = Variable),
                  data = rna_meta_aggregated_all_statistics)

rna_meta_aggregated_all_statistics %>%
  ggplot(aes(x = tau2, y = I2, col = Analysis_subset)) + geom_point() +
  theme_Publication() +
  geom_text_repel(aes(x = tau2, y = I2, label = Variable),
                  data = rna_meta_aggregated_all_statistics)

rna_meta_aggregated_all_statistics %>%
  mutate(label = cut(
    pval.Q,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(Analysis_subset), 
    y = factor(
      Variable, levels = rev(order_variables)),
    fill = I2,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

ggsave(here(output_dir, "meta_combined_effects_random_effects_het_rna.png"),
    height = 9, width = 4.5)

rna_heterogeneity_plot <- rna_meta_aggregated_all_statistics %>%
  mutate(label = cut(
    pval.Q,
    breaks = c(0, 0.001, 0.01, 0.05, 1),
    labels = c("***", "**", "*", "")
  ))  %>%
  ggplot(aes(
    x = factor(Analysis_subset), 
    y = factor(
      Variable, levels = rev(order_variables)),
    fill = I2,
    label = label)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradient2(low = "#075AFF",
                       mid = "white",
                       high = "#FF0000") +
  theme_Publication() +
  labs(y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  theme(legend.position = "right", legend.direction = "vertical") +
  theme(legend.key.size = unit(0.75, 'cm')) +
  labs(x = "")

rna_heterogeneity_plot
graph2pdf(rna_heterogeneity_plot,
          here(output_dir, "meta_combined_effects_random_effects_het_rna.pdf"),
    height = 9, width = 4.5)
```

# ----------
# Lint script
```{r}
# lint("~/NKI_pancancer_neoadj_io/src/pancan_aggregate_meta.Rmd")
lint(filename = "~/NKI_pancancer_neoadj_io/src/pancan_aggregate_meta.Rmd",
  linters = unused_import_linter()
)
```

# ----------

# sessionInfo - Package Versions
```{r sessionInfo}
sessionInfo()
```
