---
title: "neoadjuvant pancancer WES analysis"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  export: TRUE
  output_dir: "Tissue_PDL1_TILs"
  metadata: ""
  filter_and_parse_vcfs: FALSE
  exclude_niche_msi: FALSE
  exclude_clinical_response: FALSE
  response_definition: "Original" # MPR pCR
---

Author comment: 
File description comment:
The purpose of this script is to parse available PDL1 scores and CD8/CD8-PD1 staining data, and perform analysis for nICB response

# Loading Packages

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# Install if necessary install.packages("tidyverse")
# Install if necessary BiocManager::install("DESeq2")

# library("devtools")
# BiocManager::install("VariantAnnotation")
library(tidyverse)
library(readxl)
source("helper_functions.R")
library(ggpubr)
library(meta)
library(forestplot)

set.seed(42)
select <- dplyr::select
rename <- dplyr::rename
here <- here::here

```



# Setup output folder
```{r, echo = FALSE, include = FALSE}
export <- params$export
analyze_specific_trial <- params$analyze_specific_trial
run_meta_analysis <- params$run_meta_analysis   
generate_heatmaps <- params$generate_heatmaps

# generate output directory
output_dir <- here("data", "processed", params$output_dir)
print(output_dir)
output_dir_plots <- here(output_dir, "plots")

data_dir <- here("data", "processed")
combined_dir <- here(data_dir, "gene_signatures")

# if output_dir does not exist, create it
if (!dir.exists(output_dir)) {
    dir.create(output_dir)
    dir.create(output_dir_plots)
} else {
  print("Output folder already exists")
}

```

# ----------
# Annotation color
```{r}

# annotation_color_list = list(
#   Response = c(NR = "#F8766D", R = "#619CFF"),
#   Treatment = c(Pre = "#00BA38", Post = "#F564E3"),
#   Celecoxib = c(Yes = "#00BFC4", No = "#CD9600"),
#   KRAS_12X = c(G12 = "#333333", not_G12 = "#CCCCCC"),
#   B2M = c(Mut = "#333333", WT = "#CCCCCC"),
#   KRAS = c(Mut = "#333333", WT = "#CCCCCC"),
#   TP53 = c(Mut = "#333333", WT = "#CCCCCC")
# )

# Define colours for each group
Color_NR <- "#F8766D"
Color_R <- "#00BA38" # "#619CFF"
Color_NA <- "#CD9600"
Color_Pre <- "#619CFF"
Color_Post <- "#F564E3"
Color_HNSCC <- "#009292"
Color_CSCC <- "#CD9600"
Color_Urothelial <- "#B66DFF"
Color_LN_melanoma <- "#00BFC4"
Color_Colon <- "#F8766D"
Color_Colon_MSI <- "#B73A3A"
Color_Colon_MSS <- "#23C26F"
Color_TNBC <- "#FEB5DA"
Color_BELLINI <- "#FEB5DA"
Color_IMCISION <- "#009292"
Color_MATISSE <- "#CD9600"
Color_NABUCCO <- "#B66DFF"
Color_BELLINI <- "#FEB5DA"
Color_NICHE_MSS <- "#23C26F"
Color_NICHE_MSI <- "#B73A3A"
Color_OpACIN_neo <- "#00BFC4"
Color_PRADO <- "#619CFF"

annotation_color_list = list(
  Response = c(NR = Color_NR, R = Color_R),
  Treatment = c(Pre = Color_Pre, Post = Color_Post),
  Histology = c(LN_melanoma = Color_LN_melanoma, Urothelial = Color_Urothelial,
                Colon = Color_Colon, HNSCC = Color_HNSCC, TBNC = Color_TNBC,
                CSCC = Color_CSCC),
  Trial = c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO))

trial_colors <- c(BELLINI = Color_BELLINI, IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_NICHE_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

trial_colors_combined <- c(Combined = "#999999", BELLINI = Color_BELLINI,
                           IMCISION = Color_IMCISION,
            MATISSE = Color_MATISSE, NABUCCO = Color_NABUCCO,
            `NICHE-MSS` = Color_Colon_MSS, `NICHE-MSI` = Color_NICHE_MSI,
            OpACIN_neo = Color_OpACIN_neo, PRADO = Color_PRADO)

histology_msi_mss_colors <- c(`Colon-MSI`= Color_Colon_MSI,
                              `Colon-MSS`= Color_Colon_MSS,
                              CSCC = Color_CSCC, HNSCC = Color_HNSCC, 
                              LN_melanoma = Color_LN_melanoma,
                              TBNC = Color_TNBC,
                              Urothelial = Color_Urothelial)

```

```{r}
patient_metadata_combined <- read_csv2(
  here("data", "processed", "metadata", "patient_metadata_combined.csv"))
```


# ----------
# BELLINI
```{r}
BELLINI_tissue_raw <- read_excel(
  here("data", "external", "PDL1_CD8_PD1_data",
       "M18BEL_for_Pedro_28042026.xlsx"))

BELLINI_tissue_pre <- BELLINI_tissue_raw %>%
  filter(Timepoint == "Pre") %>% mutate(Patient = paste0("BEL", Patient_ID))

table(BELLINI_tissue_pre$Patient %in% patient_metadata_combined$Patient)
BELLINI_tissue_pre$Patient[
  !BELLINI_tissue_pre$Patient %in% patient_metadata_combined$Patient]

# Patients without RNA/DNA Pre treatment
BELLINI_tissue_pre_f <- BELLINI_tissue_pre %>%
  rename(Cohort_ID_ABC = Cohort) %>%
  filter(Patient %in% patient_metadata_combined$Patient) %>%
  filter(!Patient %in% c("BEL53")) %>%
  left_join(patient_metadata_combined, by = "Patient") %>%
  mutate(PDL1_CPS = parse_number(CPS_pre),
         CD8_per_um2 = parse_number(CD8_area),
         CD8_PD1_per_um2 = parse_number(CD8PD1_area)) %>% 
  mutate(CD8_per_mm2 = CD8_per_um2 * 1e6,
         CD8_PD1_per_mm2 = CD8_PD1_per_um2 * 1e6) # Convert area to mm2

BELLINI_tissue_pre_f %>% factor_summary()
```

```{r}
BELLINI_tissue_pre_f %>%
  ggplot(aes(x = CD8_per_mm2, y = CD8_PD1_per_mm2)) +
  geom_point() + theme_Publication()

BELLINI_tissue_pre_f %>%
  ggplot(aes(x = Response, y = CD8_per_mm2, fill = Response)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "BELLINI") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

BELLINI_tissue_pre_f %>%
  ggplot(aes(x = Response, y = CD8_PD1_per_mm2, fill = Response)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "BELLINI") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

BELLINI_tissue_pre_f %>%
  ggplot(aes(x = Response, y = PDL1_CPS, fill = Response)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15) + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "BELLINI") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

BELLINI_tissue_pre_f %>%
  ggplot(aes(x = Response, y = CD8_per_mm2, fill = Response)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  facet_wrap(~Cohort_id) +
  labs(title = "BELLINI") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

BELLINI_tissue_pre_f %>%
  ggplot(aes(x = Response, y = CD8_PD1_per_mm2, fill = Response)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  facet_wrap(~Cohort_id) +
  labs(title = "BELLINI") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

BELLINI_tissue_pre_f %>%
  ggplot(aes(x = Response, y = PDL1_CPS, fill = Response)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15) + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  facet_wrap(~Cohort_id) +
  labs(title = "BELLINI") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)
```

# ----------
# NICHE
```{r}
NICHE_CD8_PD1_raw <- read_csv2(
  here("data", "external","PDL1_CD8_PD1_data", "NICHE1_pMMR_CD8_PD1.csv"))

table(NICHE_CD8_PD1_raw$Patient %in% patient_metadata_combined$Patient)

NICHE_CD8_PD1 <- NICHE_CD8_PD1_raw %>%
  left_join(patient_metadata_combined, by = c(
    "Patient", "Response", "Response_detail"))

colnames(NICHE_CD8_PD1)

NICHE_CD8_PD1 %>%
  ggplot(aes(x = Response, y = CD8_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "NICHE-1 pMMR") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

NICHE_CD8_PD1 %>%
  ggplot(aes(x = Response, y = CD8_PD1_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "NICHE-1 pMMR") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

```

# ----------
# IMCISION
```{r, fig.height = 8, fig.width = 8}
IMCISION_PDL1_CD3_CD8a <- read_excel(
  here("data", "external", "paper_supplementaries",
       "Vos_IMCISION_41467_2021_26472_MOESM4_ESM.xlsx"),
  sheet = "Supp. Fig. 8a") %>%
  filter(!is.na(ID))

colnames(IMCISION_PDL1_CD3_CD8a)

IMCISION_PDL1_CD3_CD8b <- read_excel(
  here("data", "external", "paper_supplementaries",
       "Vos_IMCISION_41467_2021_26472_MOESM4_ESM.xlsx"),
  sheet = "Supp. Fig. 8b") %>% filter(!is.na(ID))
colnames(IMCISION_PDL1_CD3_CD8b)

IMCISION_PDL1_CD3_CD8 <- IMCISION_PDL1_CD3_CD8a %>%
  full_join(IMCISION_PDL1_CD3_CD8b) %>%
  setNames(c("ID", "Response_orig", "PDL1_CPS", "CD3_CD8_per_mm2",
             "Tumor_mutational_load", "AID_APOBEC_Signature_2")) %>%
  mutate(Patient = paste0("IMC_", ID)) %>%
  left_join(patient_metadata_combined, by = "Patient") %>%
  mutate(PDL1_CPS = extract_numeric(PDL1_CPS),
         CD3_CD8_per_mm2 = extract_numeric(CD3_CD8_per_mm2)) %>%
  mutate(CD8_per_mm2 = CD3_CD8_per_mm2) # Assume all CD8s are CD3 positive

IMCISION_PDL1_CD3_CD8 %>% factor_summary()

rm(IMCISION_PDL1_CD3_CD8a, IMCISION_PDL1_CD3_CD8b)
```

```{r}
table(IMCISION_PDL1_CD3_CD8$Response_orig,
      IMCISION_PDL1_CD3_CD8$Response)

IMCISION_PDL1_CD3_CD8 %>%
  ggplot(aes(x = Response_orig,
             y = CD3_CD8_per_mm2)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  labs(title = "IMCISION")

IMCISION_PDL1_CD3_CD8 %>%
  ggplot(aes(x = Response_orig,
             y = Tumor_mutational_load)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  labs(title = "IMCISION")

IMCISION_PDL1_CD3_CD8 %>%
  ggplot(aes(x = Response_orig,
             y = PDL1_CPS)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  labs(title = "IMCISION")

IMCISION_PDL1_CD3_CD8 %>%
  ggplot(aes(x = Response,
             y = CD3_CD8_per_mm2, fill = Response)) +
  geom_boxplot(outlier.shape = NA) + geom_point() + theme_Publication() +
  stat_compare_means(method = "wilcox") +
  labs(title = "IMCISION")+
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

IMCISION_PDL1_CD3_CD8 %>%
  ggplot(aes(x = Response,
             y = CD3_CD8_per_mm2, fill = Response)) +
  geom_boxplot(outlier.shape = NA) + geom_point() + theme_Publication() +
  facet_wrap(~Response_type) +
  labs(title = "IMCISION") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

IMCISION_PDL1_CD3_CD8 %>%
  ggplot(aes(x = Response, y = Tumor_mutational_load,
             fill = Response)) + geom_boxplot() +
  geom_point(outlier.shape = NA) + theme_Publication() +
  labs(title = "IMCISION") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

IMCISION_PDL1_CD3_CD8 %>%
  ggplot(aes(x = Response, y = PDL1_CPS,
             fill = Response)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() + theme_Publication() +
  labs(title = "IMCISION") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

IMCISION_PDL1_CD3_CD8 %>%
  ggplot(aes(x = Response, y = PDL1_CPS,
             fill = Response)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() + theme_Publication() + facet_wrap(~Response_type) +
  labs(title = "IMCISION") +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

```

# ----------
# NABUCCO
## PDL1
```{r}
NABUCCO_PDL1_raw <- read_excel(here("data", "external", "paper_supplementaries",
                                "Stockem et al 2025 Table S3.xlsx")) %>%
  filter(`Study ID` !=  "Abbreviations: NA = not available")

NABUCCO_PDL1 <- NABUCCO_PDL1_raw
colnames(NABUCCO_PDL1)
colnames(NABUCCO_PDL1) <- c("Study_ID", "PDL1_CPS_discrete",
                            "PDL1_IC_discrete", "PDL1_TPS_discrete",
                            "Response_orig")

NABUCCO_PDL1 <- NABUCCO_PDL1 %>%
  mutate(PDL1_CPS_discrete = ifelse(
    is.na(PDL1_CPS_discrete), "NA", PDL1_CPS_discrete)) %>%
  mutate(PDL1_TPS_discrete = ifelse(
    is.na(PDL1_TPS_discrete), "NA", PDL1_TPS_discrete)) %>%
  mutate(PDL1_IC_discrete = ifelse(
    is.na(PDL1_IC_discrete), "NA", PDL1_IC_discrete)) %>%
  filter(! Study_ID %in% 34)%>%
  filter(! Study_ID %in% c(33, 52))  # NA response

NABUCCO_PDL1 <- NABUCCO_PDL1 %>%
  mutate(PDL1_CPS_discrete = ifelse(
    PDL1_CPS_discrete == "1", "CPS >= 10", PDL1_CPS_discrete)) %>%
  mutate(PDL1_CPS_discrete = ifelse(
    PDL1_CPS_discrete == "0", "CPS < 10", PDL1_CPS_discrete)) %>%
  mutate(PDL1_TPS_discrete = ifelse(
    PDL1_TPS_discrete == "1", "TPS >= 1", PDL1_TPS_discrete)) %>%
  mutate(PDL1_TPS_discrete = ifelse(
    PDL1_TPS_discrete == "0", "TPS < 1", PDL1_TPS_discrete)) %>%
  mutate(PDL1_IC_discrete = ifelse(
    PDL1_IC_discrete == "0", "IC < 1", PDL1_IC_discrete)) %>%
  mutate(PDL1_IC_discrete = ifelse(
    PDL1_IC_discrete == "1", "IC 1-5", PDL1_IC_discrete))%>%
  mutate(PDL1_IC_discrete = ifelse(
    PDL1_IC_discrete == "2", "IC >5", PDL1_IC_discrete))

NABUCCO_PDL1 <- NABUCCO_PDL1 %>%
  mutate(Patient = paste0("NAB_S", Study_ID)) %>%
  left_join(patient_metadata_combined)
table(NABUCCO_PDL1$Patient %in% patient_metadata_combined$Patient)

table(NABUCCO_PDL1$PDL1_CPS_discrete, useNA = "ifany")
table(NABUCCO_PDL1$PDL1_TPS_discrete, useNA = "ifany")
table(NABUCCO_PDL1$PDL1_CPS_discrete,
      NABUCCO_PDL1$PDL1_IC_discrete, useNA = "ifany")
table(NABUCCO_PDL1$PDL1_CPS_discrete,
      NABUCCO_PDL1$PDL1_TPS_discrete, useNA = "ifany")

NABUCCO_PDL1_detailed <- read_excel(
  here("data", "external", "PDL1_CD8_PD1_data",
       "NABUCCO Response per PD-l1_Pedro_16-4-2026.xlsx")) 

NABUCCO_PDL1 %>% factor_summary()
```

## CD8 PD1

```{r}
NABUCCO_CD8_PD1_raw <- read_excel(
  here("data", "external", "PDL1_CD8_PD1_data",
       "cd8pd1_summarized_NABUCCO_1_and_2.xlsx"))

table(NABUCCO_CD8_PD1_raw$`Study ID`, NABUCCO_CD8_PD1_raw$tur_cys) %>%
  as.data.frame() %>%
  filter(Freq > 0) %>%
  pull(Var1) %>%
  table() %>%
  as.data.frame() %>%
  filter(Freq > 1)

# Exclude patients S34 and S33
# S33 non evaluable response
# S34 without RNA or DNA data
NABUCCO_CD8_PD1_raw %>%
  select(`Study ID`, tur_cys, tnumber) %>%
  unique() %>%
  pivot_wider(names_from = tur_cys, values_from = tnumber)

NABUCCO_CD8_PD1_pre <- NABUCCO_CD8_PD1_raw %>%
  filter(tur_cys == "TUR") %>%
  rename(Study_ID = "Study ID") %>%
  mutate(Patient = paste0("NAB_S", Study_ID))

table(NABUCCO_CD8_PD1_pre$Patient %in% patient_metadata_combined$Patient)
NABUCCO_CD8_PD1_pre$Patient[
  !NABUCCO_CD8_PD1_pre$Patient %in% patient_metadata_combined$Patient] %>%
  unique()

NABUCCO_CD8_PD1_pre_f <- NABUCCO_CD8_PD1_pre %>%
  filter(Patient %in% patient_metadata_combined$Patient) %>%
  left_join(patient_metadata_combined, by = "Patient") 

NABUCCO_CD8_PD1_pre %>% factor_summary()

```

```{r}
NABUCCO_CD8_PD1_pre_check <- NABUCCO_CD8_PD1_pre_f %>%
  select(c(tnumber, classification, n, total_n, n_frac, `Tissue Area (μm²)`,
           density, Cohort_id)) %>%
  mutate(n_frac2 = n / total_n,
         density2 = n / `Tissue Area (μm²)`)

# Calculation check
NABUCCO_CD8_PD1_pre_check %>%
  ggplot(aes(x = n_frac, y = n_frac2, color = classification)) +
  geom_point() + theme_Publication() +
  facet_wrap(~classification, scales = "free") +
  labs(title = "NABUCCO")

NABUCCO_CD8_PD1_pre_check %>%
  ggplot(aes(x = density, y = density2 * 1e6)) +
  geom_point() + theme_Publication() +
  labs(title = "NABUCCO") # 1e6 factor

# Distribution check
NABUCCO_CD8_PD1_pre_check %>%
  ggplot(aes(x = classification, y = n_frac, color = classification)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() + geom_boxplot() + theme_Publication() +
  labs(title = "NABUCCO")

NABUCCO_CD8_PD1_pre_check %>%
  ggplot(aes(x = classification, y = density, color = classification)) +
  geom_boxplot(outlier.shape = NA)  +
  geom_point()+ theme_Publication() +
  labs(title = "NABUCCO")

NABUCCO_CD8_PD1_pre_f %>%
  filter(classification %in% c("CD8+PD1+")) %>%
  select(Patient, density) %>% rename(CD8_PD1_per_mm2 = density)

NABUCCO_CD8_PD1_pre_f %>%
  filter(classification %in% c("CD8+PD1-", "CD8+PD1+")) %>%
  group_by(Patient) %>% summarize(CD8_per_mm2 = sum(density)) %>%
  ungroup()

NABUCCO_CD8_PD1_pre_f %>%
  filter(classification %in% c("CD8+PD1-", "CD8+PD1+")) %>%
  group_by(Patient) %>%
  summarize(n = sum(n), `Tissue Area (μm²)` = `Tissue Area (μm²)`) %>% 
  ungroup() %>%
  unique() %>%
  mutate(density = n / `Tissue Area (μm²)` * 1e6)

NABUCCO_CD8_PD1_pre_p <- NABUCCO_CD8_PD1_pre_f %>%
  filter(classification %in% c("CD8+PD1-", "CD8+PD1+")) %>%
  group_by(Patient) %>%
  summarize(CD8_per_mm2 = sum(density)) %>%
  ungroup() %>%
  full_join(NABUCCO_CD8_PD1_pre_f %>%
              filter(classification %in% c("CD8+PD1+")) %>%
              select(Patient, density) %>%
              rename(CD8_PD1_per_mm2 = density)
  ) %>%
  mutate(CD8_PD1_per_mm2 = ifelse(
    is.na(CD8_PD1_per_mm2), 0, CD8_PD1_per_mm2)) %>%
  left_join(patient_metadata_combined)
```


```{r, fig.height = 6, fig.width = 6}

NABUCCO_CD8_PD1_pre_check %>%
  ggplot(aes(x = Cohort_id, y = density, color = classification)) +
  geom_boxplot(outlier.shape = NA)  +
  geom_point() + theme_Publication() +
  facet_wrap(~classification) +
  labs(title = "NABUCCO")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

NABUCCO_CD8_PD1_pre_check %>%
  ggplot(aes(x = Cohort_id, y = density, color = classification)) +
  geom_boxplot(outlier.shape = NA)  +
  geom_point() + theme_Publication() +
  facet_wrap(~classification, scales = "free") +
  expand_limits(y = 0) +
  labs(title = "NABUCCO") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

NABUCCO_CD8_PD1_pre_p %>%
  ggplot(aes(x = Cohort_id, y = CD8_per_mm2)) +
  geom_boxplot(outlier.shape = NA)  +
  geom_point() + theme_Publication() +
  expand_limits(y = 0) +
  labs(title = "NABUCCO") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

table(NABUCCO_CD8_PD1_pre_p$Cohort_id)

NABUCCO_CD8_PD1_pre_p %>%
  ggplot(aes(x = Cohort_id, y = CD8_PD1_per_mm2)) +
  geom_boxplot(outlier.shape = NA)  +
  geom_point() + theme_Publication() +
  expand_limits(y = 0) +
  labs(title = "NABUCCO") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

# ----------
# Opacin Neo
## PDL1 TPS
```{r}
Opacin_neo_PDL1_raw <- read_excel(
  here("data", "external", "PDL1_CD8_PD1_data",
       "PD-L1_OpACINneo_04052026.xlsx"))

table(Opacin_neo_PDL1_raw$`PD-L1`)

Opacin_neo_PDL1 <- Opacin_neo_PDL1_raw %>%
  mutate(Patient = paste0("OpACIN_neo", Patient_ID)) %>%
  select(Patient, `PD-L1`) %>%
  setNames(c("Patient", "PDL1_TPS")) %>%
  filter(!is.na(PDL1_TPS))  %>%
  filter(PDL1_TPS != "NA") %>%
  mutate(PDL1_TPS_cat = ifelse(PDL1_TPS %in% c("1-50%", ">50%"),
                               "TPS >= 1", NA))%>%
  mutate(PDL1_TPS_cat = ifelse(PDL1_TPS %in% c("<1%"),
                               "TPS < 1", PDL1_TPS_cat)) %>%
  filter(Patient %in% patient_metadata_combined$Patient) %>%
  left_join(patient_metadata_combined) 

table(Opacin_neo_PDL1$PDL1_TPS, Opacin_neo_PDL1$Response)
table(Opacin_neo_PDL1$PDL1_TPS_cat, Opacin_neo_PDL1$Response)

```

# ----------
# PRADO
## PDL1 TPS
```{r}
PRADO_PDL1_raw <- read_excel(
  here("data", "external", "PDL1_CD8_PD1_data",
       "PRADO_annotation_DNA_RNA_complete_20221124_copy.xlsx"))
list.files( here("data", "external", "PDL1_CD8_PD1_data"))

table(PRADO_PDL1_raw$PDL1_CAT)

PRADO_PDL1 <- PRADO_PDL1_raw  %>%
  mutate(Patient = paste0("PRADO_", studyID))%>%
  select(Patient, PDL1, PDL1_CAT) %>%
  setNames(c("Patient", "PDL1_TPS", "PDL1_TPS_cat")) %>%
  filter(Patient %in% patient_metadata_combined$Patient) %>%
  left_join(patient_metadata_combined) %>%
  filter(!PDL1_TPS %in% c("unknown")) %>%
  mutate(PDL1_TPS_cat = ifelse(PDL1_TPS_cat %in% c("≥ 1%"),
                               "TPS >= 1", PDL1_TPS_cat))%>%
  mutate(PDL1_TPS_cat = ifelse(PDL1_TPS_cat %in% c("˂ 1%"),
                               "TPS < 1", PDL1_TPS_cat))

table(PRADO_PDL1$PDL1_TPS, PRADO_PDL1$Response)
table(PRADO_PDL1$PDL1_TPS_cat, PRADO_PDL1$Response)

```

# ----------

# Combine CD8 PD1
```{r}
CD8_combined <- NICHE_CD8_PD1 %>%
  select(Patient, Trial, Response, CD8_per_mm2) %>%
  rbind(IMCISION_PDL1_CD3_CD8 %>%
              select(Patient, Trial, Response, CD3_CD8_per_mm2) %>%
              rename(CD8_per_mm2 = "CD3_CD8_per_mm2")) %>%
  rbind(BELLINI_tissue_pre_f %>%
              select(Patient, Trial, Response, CD8_per_mm2)) %>%
  rbind(NABUCCO_CD8_PD1_pre_p %>%
              select(Patient, Trial, Response, CD8_per_mm2)) %>%
  left_join(patient_metadata_combined)

CD8_PD1_combined <- NICHE_CD8_PD1 %>%
  select(Patient, Trial, Response, CD8_PD1_per_mm2) %>%
  rbind(BELLINI_tissue_pre_f %>%
              select(Patient, Trial, Response, CD8_PD1_per_mm2)) %>%
  rbind(NABUCCO_CD8_PD1_pre_p %>%
              select(Patient, Trial, Response, CD8_PD1_per_mm2))  %>%
  left_join(patient_metadata_combined)

CD8_combined %>% write_csv2(here(output_dir, "CD8_combined.csv"))
CD8_PD1_combined %>% write_csv2(here(output_dir, "CD8_PD1_combined.csv"))

table(CD8_combined$Trial)
round(43/44, 2)
round(38/51, 2)
round(29/31, 2)

table(CD8_PD1_combined$Trial)
round(43/44, 2)
round(38/51, 2)
round(29/31, 2)
```


```{r}
CD8_combined %>%
  ggplot(aes(x = Trial, y = CD8_per_mm2, fill = Trial)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = trial_colors) +
  guides(fill = FALSE)

CD8_PD1_combined %>%
  ggplot(aes(x = Trial, y = CD8_PD1_per_mm2, fill = Trial)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = trial_colors) +
  guides(fill = FALSE)

```

```{r}
CD8_combined %>%
  ggplot(aes(x = Response, y = CD8_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE) +
  labs(title = "Combined")
```

```{r, fig.height = 7, fig.width = 10}
table(CD8_combined$Trial, CD8_combined$Response)
table(CD8_PD1_combined$Trial, CD8_PD1_combined$Response,
      is.na(CD8_PD1_combined$CD8_PD1_per_mm2))

CD8_combined %>%
  ggplot(aes(x = Response, y = CD8_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() + facet_wrap(~Trial, scales = "free") +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE) +
  labs(title = "Combined")
```

```{r}
CD8_PD1_combined %>%
  ggplot(aes(x = Response, y = CD8_PD1_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)
```

```{r, fig.height = 5, fig.width = 10}

table(CD8_PD1_combined$Trial, CD8_PD1_combined$Response,
      is.na(CD8_PD1_combined$CD8_PD1_per_mm2))

CD8_PD1_combined %>%
  ggplot(aes(x = Response, y = CD8_PD1_per_mm2, fill = Response)) +
  geom_boxplot() + geom_point() + facet_wrap(~Trial) +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)

```

# ----------

# Combine PDL1

```{r}

PDL1_CPS_combined <- IMCISION_PDL1_CD3_CD8 %>%
  select(Patient, PDL1_CPS) %>%
  rbind(BELLINI_tissue_pre_f %>%
  select(Patient, PDL1_CPS)) %>%
    mutate(PDL1_CPS_discrete = ifelse(PDL1_CPS >= 10,
                                      "CPS >= 10", "CPS < 10")) %>%
  full_join(NABUCCO_PDL1 %>%
              select(Patient, PDL1_CPS_discrete)) %>%
  left_join(patient_metadata_combined)

PDL1_CPS_combined %>% write_csv2(here(output_dir, "PDL1_CPS_combined.csv"))

```

```{r}
PDL1_CPS_combined %>%
  ggplot(aes(x = PDL1_CPS_discrete, y = PDL1_CPS, fill = Trial)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = trial_colors) +
  guides(fill = FALSE) + facet_wrap(~Trial) +
  geom_hline(yintercept = 10, linetype = "dashed")

PDL1_CPS_combined %>%
  ggplot(aes(x = Trial, y = PDL1_CPS, fill = Trial)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = trial_colors) +
  guides(fill = FALSE)

PDL1_CPS_combined %>%
  ggplot(aes(x = Trial, y = PDL1_CPS, fill = Trial)) +
  geom_boxplot() + geom_point() + theme_Publication() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = trial_colors) +
  guides(fill = FALSE)

PDL1_CPS_combined %>%
  ggplot(aes(x = Response, y = PDL1_CPS, fill = Response)) +
  geom_boxplot() + geom_point() + facet_wrap(~Trial) +
  stat_compare_means(method = "wilcox") + theme_Publication() +
  scale_fill_manual(values = c(Color_NR, Color_R)) +
  guides(fill = FALSE)
```

## Discrete
```{r}
table(PDL1_CPS_combined$Trial,
      PDL1_CPS_combined$PDL1_CPS_discrete) %>%
  as.data.frame() %>%
  setNames(c("Trial","PDL1_CPS_discrete", "Freq")) %>%
  ggplot(aes(x = Trial, y = Freq, fill = PDL1_CPS_discrete)) +
  geom_col() + theme_Publication()

table(PDL1_CPS_combined$PDL1_CPS_discrete,
      PDL1_CPS_combined$Response,
      PDL1_CPS_combined$Trial)

table(PDL1_CPS_combined$Trial,
      PDL1_CPS_combined$PDL1_CPS_discrete) %>%
  as.data.frame() %>%
  setNames(c("Trial","PDL1_CPS_discrete", "Freq")) %>%
  ggplot(aes(x = Trial, y = Freq, fill = PDL1_CPS_discrete)) +
  geom_col() + theme_Publication()


table(PDL1_CPS_combined$PDL1_CPS_discrete,
      PDL1_CPS_combined$Response,
      PDL1_CPS_combined$Trial) %>%
  as.data.frame() %>%
  setNames(c("PDL1_CPS_discrete", "Response", "Trial", "Freq"))

table(PDL1_CPS_combined$PDL1_CPS_discrete,
      PDL1_CPS_combined$Response,
      PDL1_CPS_combined$Trial) %>%
  as.data.frame() %>%
  setNames(c("PDL1_CPS_discrete", "Response", "Trial", "Freq")) %>%
  pivot_wider(values_from = Freq, names_from = Response) %>%
  mutate(Total = NR + R, Response_rate = R / Total)

table(PDL1_CPS_combined$PDL1_CPS_discrete,
      PDL1_CPS_combined$Response,
      PDL1_CPS_combined$Trial) %>%
  as.data.frame() %>%
  setNames(c("PDL1_CPS_discrete", "Response", "Trial", "Freq")) %>%
  pivot_wider(values_from = Freq, names_from = Response) %>%
  mutate(Total = NR + R, Response_rate = R / Total) %>%
  ggplot(aes(PDL1_CPS_discrete, y = Response_rate)) + geom_col() +
  theme_Publication() + facet_wrap(~Trial)

PDL1_cat_combined <- PDL1_CPS_combined %>%
  select(Patient, PDL1_CPS_discrete) %>%
  setNames(c("Patient", "PDL1_group")) %>%
  rbind(PRADO_PDL1 %>%
  select(Patient, PDL1_TPS_cat) %>%
  setNames(c("Patient", "PDL1_group"))) %>%
  rbind(Opacin_neo_PDL1 %>%
  select(Patient, PDL1_TPS_cat) %>%
  setNames(c("Patient", "PDL1_group")))  %>%
  mutate(PDL1_discrete_group = ifelse(
    PDL1_group %in% c("TPS >= 1", "CPS >= 10"),
    "PDL1_high", NA)) %>%
  mutate(PDL1_discrete_group = ifelse(
    PDL1_group %in% c("TPS < 1", "CPS < 10"),
    "PDL1_low", PDL1_discrete_group)) %>%
  left_join(patient_metadata_combined)
  
table(PDL1_cat_combined$PDL1_group,
      PDL1_cat_combined$Response,
      PDL1_cat_combined$Trial) %>%
  as.data.frame() %>%
  setNames(c("PDL1_group", "Response", "Trial", "Freq")) %>%
  filter(! (
    Trial %in% c("BELLINI", "IMCISION", "NABUCCO") &
      PDL1_group %in% c("TPS < 1", "TPS >= 1"))) %>%
  filter(! (
    Trial %in% c("PRADO", "OpACIN_neo") &
      PDL1_group %in% c("CPS < 10", "CPS >= 10")))  %>%
  pivot_wider(values_from = Freq, names_from = Response) %>%
  mutate(Total = NR + R, Response_rate = R / Total) %>%
  ggplot(aes(PDL1_group, y = Response_rate)) + geom_col() +
  theme_Publication() + facet_wrap(~Trial, scales = "free_x")

table(PDL1_cat_combined$PDL1_group,
      PDL1_cat_combined$Response,
      PDL1_cat_combined$Trial) %>%
  as.data.frame() %>%
  setNames(c("PDL1_group", "Response", "Trial", "Freq")) %>%
  filter(! (
    Trial %in% c("BELLINI", "IMCISION", "NABUCCO") &
      PDL1_group %in% c("TPS < 1", "TPS >= 1"))) %>%
  filter(! (
    Trial %in% c("PRADO", "OpACIN_neo") &
      PDL1_group %in% c("CPS < 10", "CPS >= 10")))  %>%
  pivot_wider(values_from = Freq, names_from = Response) %>%
  mutate(Total = NR + R, Response_rate = R / Total) %>%
  ggplot(aes(PDL1_group, y = Response_rate)) + geom_col() +
  theme_Publication() + facet_wrap(~Trial, scales = "free")

table(PDL1_cat_combined$PDL1_discrete_group,
      PDL1_cat_combined$Response,
      PDL1_cat_combined$Trial) %>%
  as.data.frame() %>%
  setNames(c("PDL1_discrete_group", "Response", "Trial", "Freq"))  %>%
  pivot_wider(values_from = Freq, names_from = Response) %>%
  mutate(Total = NR + R, Response_rate = R / Total) %>%
  ggplot(aes(x = fct_rev(PDL1_discrete_group), y = Response_rate)) +
  geom_col() + theme_Publication() +
  facet_wrap(~Trial, scales = "free_x") + labs(x = "")

table(PDL1_CPS_combined$Trial)
table(PDL1_cat_combined$Trial)

round(43/44, 2)
round(56/64, 2)
round(63/80, 2)

PDL1_cat_combined %>% write_csv2(here(output_dir, "PDL1_cat_combined.csv"))
```

## Odds ratio

```{r}
PDL1_cat_data <- PDL1_cat_combined %>%
    mutate(PDL1_discrete_group_n = ifelse(
      PDL1_discrete_group %in% c("PDL1_high"), TRUE, FALSE))

table(PDL1_cat_data$PDL1_discrete_group_n, PDL1_cat_data$Trial)
table(PDL1_cat_data$PDL1_discrete_group_n, PDL1_cat_data$Response,
      PDL1_cat_data$Trial)

# Sourced from helper function
trial_odds_ratio_df <- get_trial_or_estimates(
  data = PDL1_cat_combined %>%
    mutate(PDL1_discrete_group_n = ifelse(
      PDL1_discrete_group %in% c("PDL1_high"), TRUE, FALSE)),
  trial_list = c("Combined", "BELLINI", "NABUCCO", "IMCISION",
                 "PRADO", "OpACIN_neo"),
  variable_list = "PDL1_discrete_group_n")

trial_odds_ratio_df

trial_odds_ratio_df %>%
  filter(Trial == "Combined") %>%
  mutate(estimate = round(estimate, 2), p.value = round(p.value, 3)) %>%
  arrange(-estimate) %>%
  forestplot(labeltext = c(feature, n_feature_pos, n_feature_neg,
                           estimate, p.value),
             mean = "estimate", lower = "conf.low", upper = "conf.high",
             clip = c(0.1, 20),
             zero = 1,
             boxsize = 0.375,
             xlog = TRUE,
             xlab = "log(OR) (95% CI)") %>%
  fp_set_style(box = "royalblue",
               line = "darkblue",
               summary = "royalblue",
               txt_gp = fpTxtGp(ticks = gpar(cex = 1),
                                xlab  = gpar(cex = 1))
               ) %>% 
  fp_set_zebra_style("#EFEFEF") %>%
  fp_add_header(feature = c("Feature"),
                estimate = c("OR"),
                n_feature_pos = c("n pos"),
                n_feature_neg = c("n neg"),
                p.value = c("p.value")) %>%
  fp_add_lines()

trial_odds_ratio_df %>%
  mutate(estimate = round(estimate, 2), p.value = round(p.value, 4)) %>%
  arrange(-estimate) %>%
  forestplot(labeltext = c(Trial, n_feature_pos, n_feature_neg,
                           estimate, p.value),
             mean = "estimate", lower = "conf.low", upper = "conf.high",
             clip = c(0.1, 20),
             zero = 1,
             boxsize = 0.375,
             xlog = TRUE,
             xlab = "log(OR) (95% CI)") %>%
  fp_set_style(box = "royalblue",
               line = "darkblue",
               summary = "royalblue",
               txt_gp = fpTxtGp(ticks = gpar(cex = 1),
                                xlab  = gpar(cex = 1))
               ) %>%
  fp_set_zebra_style("#EFEFEF") %>%
  fp_add_header(Trial = c("Trial"),
                estimate = c("OR"),
                n_feature_pos = c("n pos"),
                n_feature_neg = c("n neg"),
                p.value = c("p.value")) %>%
  fp_add_lines()
```

## OR meta
```{r}

feature_odds_ratio_list <- "PDL1_discrete_group_n"

# Sourced from helper function
run_or_meta_analysis(meta_analysis_df = trial_odds_ratio_df %>%
                               filter(!Trial %in% c("Combined",
                                                    "Combined_excl_dMMR")),
                     variable_list = feature_odds_ratio_list,
                     output_suffix = "random_effects")

```

# ----------
# Lint script
```{r}

library(lintr)
# lint("~/NKI_pancancer_neoadj_io/src/pancan_aggregate_PDL1_TILs.Rmd")
lint(filename = "~/NKI_pancancer_neoadj_io/src/pancan_aggregate_PDL1_TILs.Rmd",
  linters = unused_import_linter()
)
```

# ----------

# sessionInfo - Package Versions

```{r sessionInfo}
sessionInfo()
```
